### 2020-01-22 paper.tex
initial commit

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\begin{document}

%% Title information
\title{Elaboration with First-Class Implicit Function Types}

%% Author with single affiliation.
\author{Andr{\'a}s Kov{\'a}cs}
\orcid{0000-0002-6375-9781}
\affiliation{
  \department{Department of Programming Languages and Compilers}
  \institution{E{\"o}tv{\"o}s Lor{\'a}nd University}
  \city{Budapest}
  \country{Hungary}
}
\email{kovacsandras@inf.elte.hu}

\begin{abstract}
Implicit functions are dependently typed functions, such that arguments are
provided (by default) by inference machinery instead of programmers of the
surface language. Implicit functions in Agda are an archetypal example. In the
Haskell language as implemented by the Glasgow Haskell Compiler (GHC),
polymorphic types are another example. Implicit function types are
\emph{first-class} if they are treated as any other type in the surface
language. This is true in Agda and partially true in GHC. Inference and
elaboration in the presence of first-class implicit functions poses a challenge;
in the context of GHC and ML-like languages, this has been dubbed
``impredicative instantiation'' or ``impredicative inference''. We propose a new
framework for elaborating first-class implicit functions, which is applicable
for full dependent type theories and compares favorably to prior solutions in
terms of power, generality and conceptual simplicity. We build atop Norell's
bidirectional elaboration algorithm for Agda, and note that key issue is
incomplete information about insertions of implicit abstractions and
applications. We make it possible to track and refine information related to
such insertions, by adding a new function type to a core Martin-L\"of type
theory, which supports strict (definitional) currying. This allows us to
represent undetermined domain arities of implicit function types, and we can
decide at any point during elaboration whether implicit abstractions should be
inserted.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{impredicative polymorphism, type theory, elaboration, type inference}
\maketitle

\section{Introduction}

\begin{itemize}
\item Elaboration.

\item Hole filling.

\item Insertion of implicits.

\item First-class implicit functions vs Coq.

\item Unique vs non-unique solutions.

\end{itemize}

\section{Bidirectional Elaboration}

\begin{itemize}
\item

\item Limitations in Agda

\item Insertion of implicits.

\item First-class implicit functions vs Coq.

\item Unique vs non-unique solutions.

\item On the uniqueness of implicit insertion.
\end{itemize}

\begin{acks}
  This work was supported by the European Union, co-financed by the
  European Social Fund (EFOP-3.6.3-VEKOP-16-2017-00002) and COST Action
  EUTypes CA15123.
\end{acks}

%% Bibliography
\bibliography{references}

\end{document}

### 2020-01-22 implicit-fun-elab/Elaboration.hs

@@ -1,0 +1,687 @@
+
+{-# options_ghc -Wno-type-defaults #-}
+
+module Elaboration where
+
+import Control.Monad
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Monad.State.Strict
+import Data.Foldable
+import Lens.Micro.Platform
+
+import qualified Data.IntMap.Strict as M
+import qualified Data.IntSet        as IS
+import qualified Data.Set           as S
+
+import Types
+import Evaluation
+
+
+-- Freshness
+--------------------------------------------------------------------------------
+
+data Fresh = Fresh
+  (forall a. Name -> (Name, UnifyM a -> UnifyM a))
+  (forall a. Name -> Val -> (Name, UnifyM a -> UnifyM a))
+
+bindU :: Name -> UnifyCxt -> UnifyCxt
+bindU x (UnifyCxt vs is p) =
+  UnifyCxt ((x, Nothing):vs) ((x, Bound BVSrc):is) p
+
+telBindU :: Name -> VTy -> UnifyCxt -> UnifyCxt
+telBindU x ~a (UnifyCxt vs is p) =
+  UnifyCxt ((x, Nothing):vs) ((x, Bound (BVTel a)):is) p
+
+localFresh :: UnifyM Fresh
+localFresh = do
+  vs <- view vals
+  pure $ Fresh
+    (\x -> let x' = fresh vs x in (x', local (bindU x')))
+    (\x a -> let x' = fresh vs x in (x', local (telBindU x' a)))
+
+-- Telscope constancy
+--------------------------------------------------------------------------------
+
+data Occurs = Rigid | Flex IS.IntSet | None deriving (Eq, Show)
+
+instance Semigroup Occurs where
+  Flex ms <> Flex ms' = Flex (ms <> ms')
+  Rigid   <> _        = Rigid
+  _       <> Rigid    = Rigid
+  None    <> r        = r
+  l       <> None     = l
+
+occurrence :: IS.IntSet -> Occurs
+occurrence ms | IS.null  ms = Rigid
+              | otherwise   = Flex ms
+
+instance Monoid Occurs where
+  mempty = None
+
+occurs :: Name -> Val -> UnifyM Occurs
+occurs topX = go mempty where
+
+  goSp :: IS.IntSet -> Spine -> UnifyM Occurs
+  goSp ms ~sp = forceSpM sp >>= \case
+    SNil -> pure mempty
[…]

### 2020-01-22 implicit-fun-elab/Evaluation.hs

@@ -1,0 +1,299 @@
+
+module Evaluation (
+    apply
+  , applyM
+  , close
+  , closeM
+  , eval
+  , evalM
+  , force
+  , forceM
+  , forceSp
+  , forceSpM
+  , fresh
+  , hlamM
+  , nf
+  , nfM
+  , quote
+  , quoteM
+  , runEval
+  , vApp
+  , vAppM
+  , zonk
+  , zonkM
+  ) where
+
+import Control.Monad.Reader
+import Lens.Micro.Platform
+import qualified Data.IntMap.Strict as M
+
+import Types
+
+--------------------------------------------------------------------------------
+
+runEval :: EvalM a -> MCxt -> Vals -> a
+runEval ma ms vs = runReader ma (EvalEnv ms vs)
+
+embedEvalM :: HasVals cxt Vals => EvalM a -> M cxt a
+embedEvalM ma = runEval ma <$> use mcxt <*> view vals
+
+vProj1 :: Val -> EvalM Val
+vProj1 (VTcons t u)    = pure t
+vProj1 (VNe h sp)      = pure $ VNe h (SProj1 sp)
+vProj1 (VLamTel x a t) = vProj1 =<< vLamTel pure x a t
+vProj1 _               = error "impossible"
+
+vProj2 :: Val -> EvalM Val
+vProj2 (VTcons t u)    = pure u
+vProj2 (VNe h sp)      = pure $ VNe h (SProj2 sp)
+vProj2 (VLamTel x a t) = vProj2 =<< vLamTel pure x a t
+vProj2 _               = error "impossible"
+
+vVar :: Name -> EvalM Val
+vVar x = do
+  vs <- view vals
+  case (lookup x vs) of
+    Nothing -> pure (VVar x) -- error $ "impossible: " ++ x ++ " not in scope"
+    Just mv -> pure $ maybe (VVar x) id mv
+
+lookupMeta :: MId -> EvalM MetaEntry
+lookupMeta m = do
+  ms <- view mcxt
+  case M.lookup m ms of
+    Just e  -> pure e
+    Nothing -> error "impossible"
+
+vMeta :: MId -> EvalM Val
+vMeta m = lookupMeta m >>= \case
+  Solved v -> pure v
+  _        -> pure (VMeta m)
+
+defining :: Name -> Val -> EvalM a -> EvalM a
+defining x ~v ma = local (vals %~ ((x, Just v):)) ma
+
[…]

### 2020-01-22 implicit-fun-elab/Main.hs

@@ -1,0 +1,279 @@
+
+module Main where
+
+import Text.Printf
+import Control.Monad.State.Strict
+import Control.Monad.Reader
+import System.Exit
+import System.Environment
+
+import Types
+import Evaluation
+import Elaboration
+import Parser
+
+helpMsg = unlines [
+  "usage: holes [--help|nf|type]",
+  "  --help : display this message",
+  "  elab   : read & elaborate expression from stdin",
+  "  nf     : read & elaborate expression from stdin, print its normal form",
+  "  type   : read & elaborate expression from stdin, print its type"]
+
+displayError :: String -> Err -> IO ()
+displayError file (Err msg (SourcePos path (unPos -> linum) (unPos -> colnum))) = do
+  let lnum = show linum
+      lpad = map (const ' ') lnum
+  printf "%s:%d:%d:\n" path linum colnum
+  printf "%s |\n"    lpad
+  printf "%s | %s\n" lnum (lines file !! (linum - 1))
+  printf "%s | %s\n" lpad (replicate (colnum - 1) ' ' ++ "^")
+  printf "%s\n\n" (show msg)
+
+elabCxt0 :: ElabCxt
+elabCxt0 = ElabCxt [] [] [] (initialPos "(stdin)") True
+
+st0 :: St
+st0 = St 0 mempty
+
+runElab0 :: ElabM a -> Either Err a
+runElab0 ma = evalStateT (runReaderT ma elabCxt0) st0
+
+mainWith :: IO [String] -> IO (Raw, String) -> IO ()
+mainWith getOpt getTm = do
+
+  let elab :: IO (Tm, Tm, Tm)
+      elab = do
+        (t, src) <- getTm
+        let res = runElab0 $ do
+              (t, a) <- infer MIYes t
+              t      <- zonkM t
+              -- traceShowM "inferred"
+              nt     <- nfM t
+              na     <- quoteM a
+              pure (t, nt, na)
+        case res of
+          Left err -> displayError src err >> exitSuccess
+          Right x  -> pure x
+
+  getOpt >>= \case
+    ["--help"] -> putStrLn helpMsg
+    ["nf"] -> do
+      (t, nt, na) <- elab
+      print nt
+    ["type"] -> do
[…]

### 2020-01-22 implicit-fun-elab/Parser.hs

@@ -1,0 +1,143 @@
+
+module Parser (
+    parseString
+  , parseStdin
+  ) where
+
+import Control.Monad
+import Data.Char
+import Data.Void
+import System.Exit
+import Text.Megaparsec
+
+import qualified Text.Megaparsec.Char       as C
+import qualified Text.Megaparsec.Char.Lexer as L
+
+import Types
+
+--------------------------------------------------------------------------------
+
+type Parser = Parsec Void String
+
+ws :: Parser ()
+ws = L.space C.space1 (L.skipLineComment "--") (L.skipBlockComment "{-" "-}")
+
+withPos :: Parser Raw -> Parser Raw
+withPos p = RSrcPos <$> getSourcePos <*> p
+
+lexeme     = L.lexeme ws
+symbol s   = lexeme (C.string s)
+char c     = lexeme (C.char c)
+parens p   = char '(' *> p <* char ')'
+braces p   = char '{' *> p <* char '}'
+pArrow     = symbol "→" <|> symbol "->"
+pBind      = pIdent <|> symbol "_"
+
+keyword :: String -> Bool
+keyword x =
+  x == "let" || x == "in" || x == "λ" || x == "U" || x == "assume"
+
+pIdent :: Parser Name
+pIdent = try $ do
+  x <- takeWhile1P Nothing isAlphaNum
+  guard (not (keyword x))
+  x <$ ws
+
+pAtom :: Parser Raw
+pAtom  =
+      withPos (    (RVar <$> pIdent)
+               <|> (RU <$ char 'U')
+               <|> (RHole <$ char '_'))
+  <|> parens pTm
+
+pArg :: Parser (Maybe (Either Name Icit, Raw))
+pArg = (Nothing <$ char '!')
+  <|>  (Just <$> (
+           (try $ braces $ do {x <- pIdent; char '='; t <- pTm; pure (Left x, t)})
+       <|> ((Right Impl,) <$> (char '{' *> pTm <* char '}'))
+       <|> ((Right Expl,) <$> pAtom)))
+
+pSpine :: Parser Raw
+pSpine = do
+  h <- pAtom
+  args <- many pArg
+  pure $ foldl
+    (\t -> \case Nothing     -> RStopInsertion t
+                 Just (i, u) -> RApp t u i)
+    h args
+
[…]

### 2020-01-22 implicit-fun-elab/Types.hs

@@ -1,0 +1,357 @@
+
+module Types (
+  module Types,
+  module Text.Megaparsec
+  ) where
+
+import Control.Monad.Except
+import Control.Monad.Reader
+import Control.Monad.State.Strict
+import Data.String
+import Data.List
+import Lens.Micro.Platform
+import Text.Megaparsec (SourcePos(..), unPos, initialPos)
+import Text.Printf
+
+import qualified Data.IntMap.Strict as M
+import qualified Data.IntSet        as IS
+
+import Debug.Trace
+
+--------------------------------------------------------------------------------
+
+type Name = String
+data Icit = Impl | Expl deriving (Eq, Show)
+
+icit :: Icit -> a -> a -> a
+icit Impl i e = i
+icit Expl i e = e
+
+data Raw
+  = RVar Name
+  | RLam Name (Maybe Raw) (Either Name Icit) Raw
+  | RApp Raw Raw (Either Name Icit)
+  | RU
+  | RPi Name Icit Raw Raw
+  | RLet Name Raw Raw Raw
+  | RAssume Name Raw Raw
+  | RHole
+  | RStopInsertion Raw
+  | RSrcPos SourcePos Raw
+
+-- deriving instance Show Raw
+instance Show Raw where
+  show = show . go where
+    go :: Raw -> Tm
+    go = \case
+      RVar x               -> Var x
+      RLam x _ (Left y) t  -> Lam x Impl (go t)
+      RLam x _ (Right i) t -> Lam x i (go t)
+      RApp t u (Left x)    -> App (go t) (go u) Impl
+      RApp t u (Right i)   -> App (go t) (go u) i
+      RU                   -> U
+      RPi x i a b          -> Pi x i (go a) (go b)
+      RLet x a t u         -> Let x (go a) (go t) (go u)
+      RAssume x a t        -> Assume x (go a) (go t)
+      RHole                -> Var "_"
+      RStopInsertion t     -> App (go t) (Var "!") Expl
+      RSrcPos _ t          -> go t
+
+--------------------------------------------------------------------------------
+
+-- | Elaboration problem identifier.
+type MId = Int
+
+-- | List of blocked problems.
+type Blocking = IS.IntSet
[…]

### 2020-01-23 paper.tex
stuff

@@
+%% build: latexmk -pdf -pvc -xelatex paper.tex
+
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
@@
 
+%% ------------------------------------------------------------
+
+%% \usepackage{fontspec}
+%% \fontspec[
+%%   Path = fonts/,
+%%   BoldFont = *-Bold,
+%%   ItalicFont = *-Oblique,
+%%   BoldItalicFont = *-BoldOblique]
+%%   {DejaVuSansMono}
+\setmonofont[Scale=0.8]{DejaVuSansMono}
+
+%% ------------------------------------------------------------
+
 \begin{document}
@@
 terms of power, generality and conceptual simplicity. We build atop Norell's
-bidirectional elaboration algorithm for Agda, and note that key issue is
+bidirectional elaboration algorithm for Agda, and note that the key issue is
 incomplete information about insertions of implicit abstractions and
@@
 
+Programmers and users of proof assistants do not like to write out obvious
+things. Type inference and elaboration in general serve the purpose of filling
+in tedious details, translating terse surface-level languages to explicit core
+languages. Modern compilers such as Agda have gotten quite adept at this
+task. However, in practice the programmer still has to tell the compiler when
+and where to try filling in details on its own.
+
+\textbf{Implicit function types} are a common mechanism for conveying to the
+compiler that particular arguments should be inferred by default. In Agda and
+Coq, one can use bracketed function domains for this purpose:
+\begin{verbatim}
+          -- Agda                       (* Coq *)
+          id : {A : Set} → A → A        Definition id {A:Type}(x : A) := x.
+          id x = x
+\end{verbatim}
+In GHC, one can use \texttt{forall} to define implicit function
+types\footnote{This notation requires language extensions
+  \texttt{KindSignatures} and \texttt{RankNTypes}; one could also write the type
+  \texttt{a -> a} and GHC would silently insert the quantification.}
+\begin{verbatim}
+                         id :: forall (a :: *). a -> a
+                         id x = x
+\end{verbatim}
+
+In all of the above cases, if we apply \texttt{id} to an argument, the implicit
+type argument is provided by elaboration. For example, in Agda \texttt{id true}
+is elaborated to \texttt{id \{Bool\} true}, and analogously in GHC and Coq. In
+all three systems, there is also a way to explicitly specify implicit arguments:
+in Agda we may put arguments in brackets as we have seen, in Coq we can prefix a
+name with \texttt{@} to make every implicit argument explicit, as in \texttt{@id
+  bool true}, and in GHC we can enable the language extension
+\texttt{TypeApplications} and write \texttt{id @Bool true}.
+
+Implicit functions are \textbf{first class} if they can be manipulated like any
+other type. In Agda, this is the case. For example, we have a list type where
+the type of elements is an implicit function, as in \texttt{List (\{A : Set\} →
+  A → A)}. That said, Agda's elaboration has limitations when it comes to
+handling such types, and in many cases programmers must write out
+implicit lambda abstractions.
+
+In Coq, the core language does not have an actual implicit function
+type, instead, implicitness is tied to particular \emph{names}, and while we can
+write \texttt{List (forall \{A : Set\}, A -> A)}, the brackets here are simply
+ignored by Coq. For illustration, Coq accepts the following definition:
+\begin{verbatim}
+      Definition poly : forall (f : forall {A : Type}, A -> A), bool * nat :=
+        fun f => (f bool true, f nat 0).
+\end{verbatim}
+This is a higher-rank polymorphic function which returns a pair. Note that
+\texttt{f} is applied to two arguments, because the implicitness in
+\texttt{forall \{A : Type\}, A -> A} is silently dropped.
+
+In GHC Haskell, \texttt{forall} types are somewhere between Agda and Coq. We can certainly write the following, with \texttt{RankNTypes} enabled:
+\begin{verbatim}
+                    poly :: (forall a. a -> a) -> (Bool, Int)
+                    poly f = (f True, f 0)
+\end{verbatim}
+However, polymorphic types are only supported in function domains and as fields
+of algebraic data constructors. We cannot instantiate an arbitrary type
+parameter to a \texttt{forall}, as in \texttt{[forall a. a -> a]} for the list
+type with polymorphic elements. While this type is technically allowed by the
+\texttt{ImpredicativeTypes} language extension, as of GHC 8.8 this extension is
+deprecated and is not particularly usable in practice.
+
+There is significant literature on type inference in the presence of first-class
+polymorphic types, mainly in relation to GHC and ML-like languages
+\cite{TODO}. However, none of the proposed algorithms have landed so far in
+production compilers, for reasons of complexity, fragility, interaction with
+other language features, or necessity to modify surface or core languages.
+
+\subsection{Contributions}
 \begin{itemize}
-\item Elaboration.
+  \item We propose an elaboration algorithm which translates from a small
+    Agda-like surface language to a small Martin-L\"of type theory extended with
+    implicit function types, telescopes and \emph{strictly curried} function
+    types with telescope domain. The extensions to the target theory are modest
+    and are derivable from W-types.
+  \item Our algorithm is a conservative extension of Norell's bidirectional
+    elaborator for Agda \cite[Chapter~3.6]{norell07thesis}; it accepts strictly more
+    programs, and does not require new constructs in the surface language.
+  \item Our target language serves as a general platform for elaborating
+    implicit functions. The concrete elaborator presented in this paper is a
+    relatively simple one, and there is plenty of room to develop more advanced
+    elaboration and unification. However, our simple algorithm is already
+    comparable or superior to previous solutions for impredicative inference.
+  \item We provide a standalone implementation of the elaborator described in
+    this paper.
+\end{itemize}
 
-\item Hole filling.
+\subsection{Note on Terminology}
 
-\item Insertion of implicits.
-
-\item First-class implicit functions vs Coq.
-
-\item Unique vs non-unique solutions.
+We prefer to avoid the term ``impredicative inference'' in order to avoid
+confusion with impredicativity in type theory, which is an orthogonal notion. In
+type theory, impredicativity is a property of a universe, i.e.\ closure of a
+universe under arbitrary products. In the type inference literature,
+impredicativity means the ability to instantiate type variables and
+metavariables to polymorphic types. In particular, we have that
+\begin{itemize}
+  \item Agda is type-theory-predicative, but implements type-inference-impredicative elaboration.
+  \item Coq has type-theory-impredicative $\mathsf{Prop}$ universe, but
+    implements type-inference-predicative elaboration.
+  \item GHC is type-theory-impredicative with \texttt{RankNTypes} but without
+    \texttt{ImpredicativeTypes}, as we have \texttt{(forall a. a -> a) :: *}.
 
@@
 \begin{itemize}
-\item
-
 \item Limitations in Agda

### 2020-01-23 implicit-fun-elab/Types.hs

@@ -34,7 +34,6 @@ data Raw
   | RU
   | RPi Name Icit Raw Raw
   | RLet Name Raw Raw Raw
-  | RAssume Name Raw Raw
   | RHole
   | RStopInsertion Raw
   | RSrcPos SourcePos Raw
@@ -52,7 +51,6 @@ instance Show Raw where
       RU                   -> U
       RPi x i a b          -> Pi x i (go a) (go b)
       RLet x a t u         -> Let x (go a) (go t) (go u)
-      RAssume x a t        -> Assume x (go a) (go t)
       RHole                -> Var "_"
       RStopInsertion t     -> App (go t) (Var "!") Expl
       RSrcPos _ t          -> go t
@@ -83,7 +81,7 @@ type Info  = Sub VarInfo
 type MCxt  = M.IntMap MetaEntry
 
 data BoundVar = BVSrc | BVTel ~VTy | BVRenamed
-data VarInfo = Bound BoundVar | Defined | Assumed
+data VarInfo = Bound BoundVar | Defined
 
 data ElabCxt = ElabCxt {
   elabCxtVals  :: Vals,
@@ -327,12 +325,10 @@ instance Show ElabError where
     IcitMismatch i i' -> printf (
       "Function icitness mismatch: expected %s, got %s.")
       (show i) (show i')
-    AssumptionNotTopLevel -> "Assumption only allowed in the top scope."
     NonLinearSolution m sp rhs x -> printf
       ("Nonlinear variable %s in meta spine in equation\n\n" ++
        "  %s %s =? %s")
       (show x) (show m) ('[':intercalate ", " sp++"]") (show rhs)
-    UnsolvedAfterAssumption -> "Unsolved meta remains after assumption"
 
 
 report :: HasPos cxt SourcePos => ElabError -> M cxt a

### 2020-01-23 paper.tex
stuff

@@
 \setmonofont[Scale=0.8]{DejaVuSansMono}
+\usepackage{fancyvrb}
+\usepackage{xcolor}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
+\newcommand{\red}[1]{\textcolor{red}{#1}}
 
@@
 Coq, one can use bracketed function domains for this purpose:
-\begin{verbatim}
-          -- Agda                       (* Coq *)
-          id : {A : Set} → A → A        Definition id {A:Type}(x : A) := x.
+\begin{Verbatim}[commandchars=\\\{\}]
+          \red{-- Agda}                     \red{(* Coq *)}
+          \blue{id} : {A : \blue{Set}} → A → A        \blue{Definition} id \{A:Type\}(x : A) := x.
           id x = x
-\end{verbatim}
+\end{Verbatim}
 In GHC, one can use \texttt{forall} to define implicit function
@@
 
-In Coq, the core language does not have an actual implicit function
-type, instead, implicitness is tied to particular \emph{names}, and while we can
-write \texttt{List (forall \{A : Set\}, A -> A)}, the brackets here are simply
-ignored by Coq. For illustration, Coq accepts the following definition:
+In Coq, the core language does not have an actual implicit function type,
+instead, implicitness is tied to particular \emph{names}, and while we can write
+\texttt{List (forall \{A : Set\}, A -> A)}, the brackets here are simply ignored
+by Coq and we get a plain function a type. For illustration, Coq accepts the
+following definition:
 \begin{verbatim}
@@
     \texttt{ImpredicativeTypes}, as we have \texttt{(forall a. a -> a) :: *}.
-
 \end{itemize}

### 2020-01-25 paper.tex
stuff

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review,anonymous,prologue,dvipsnames]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 %%   {DejaVuSansMono}
-\setmonofont[Scale=0.8]{DejaVuSansMono}
+\setmonofont[Colour=blue,Scale=0.8]{DejaVuSansMono}
 \usepackage{fancyvrb}
 \usepackage{xcolor}
-\newcommand{\blue}[1]{\textcolor{blue}{#1}}
-\newcommand{\red}[1]{\textcolor{red}{#1}}
+\usepackage{mathpartir}
+\usepackage{todonotes}
+\presetkeys{todonotes}{inline}{}
+\usepackage{scalerel}
+
+%% \newcommand{\blue}[1]{\textcolor{blue}{#1}}
+%% \newcommand{\red}[1]{\textcolor{red}{#1}}
+%% \newcommand{\green}[1]{\textcolor{ForestGreen}{#1}}
+%% \newcommand{\purple}[1]{\textcolor{purple}{#1}}
+%% \newcommand{\ASet}{\blue{Set}}
+%% \newcommand{\CDefinition}{\blue{Definition}}
+%% \newcommand{\CType}{\blue{Type}}
+%% \newcommand{\Hforall}{\blue{forall}}
+%% \newcommand{\Cforall}{\blue{forall}}
+%% \newcommand{\HStar}{\blue{*}}
+
+\newcommand{\slet}{\mathsf{let}}
+\renewcommand{\sin}{\mathsf{in}}
+\renewcommand{\U}{\mathsf{U}}
+\newcommand{\emptycon}{\scaleobj{.75}\bullet}
 
@@
 Programmers and users of proof assistants do not like to write out obvious
-things. Type inference and elaboration in general serve the purpose of filling
-in tedious details, translating terse surface-level languages to explicit core
+things. Type inference and elaboration serve the purpose of filling in tedious
+details, translating terse surface-level languages to explicit core
 languages. Modern compilers such as Agda have gotten quite adept at this
-task. However, in practice the programmer still has to tell the compiler when
-and where to try filling in details on its own.
+task. However, in practice, programmers still have to tell the compiler when and
+where to try filling in details on its own.
 
 \textbf{Implicit function types} are a common mechanism for conveying to the
-compiler that particular arguments should be inferred by default. In Agda and
-Coq, one can use bracketed function domains for this purpose:
+compiler that particular function arguments should be inferred by default. In
+Agda and Coq, one can use bracketed function domains for this purpose:
 \begin{Verbatim}[commandchars=\\\{\}]
-          \red{-- Agda}                     \red{(* Coq *)}
-          \blue{id} : {A : \blue{Set}} → A → A        \blue{Definition} id \{A:Type\}(x : A) := x.
+          -- Agda                     (* Coq *)
+          id : \{A : Set\} → A → A      Definition id \{A : Type\}(x : A) := x.
           id x = x
@@
   \texttt{a -> a} and GHC would silently insert the quantification.}
-\begin{verbatim}
+\begin{Verbatim}[commandchars=\\\{\}]
                          id :: forall (a :: *). a -> a
                          id x = x
-\end{verbatim}
+\end{Verbatim}
 
 In all of the above cases, if we apply \texttt{id} to an argument, the implicit
-type argument is provided by elaboration. For example, in Agda \texttt{id true}
-is elaborated to \texttt{id \{Bool\} true}, and analogously in GHC and Coq. In
-all three systems, there is also a way to explicitly specify implicit arguments:
-in Agda we may put arguments in brackets as we have seen, in Coq we can prefix a
+type argument is provided by elaboration. For example, in Agda, \texttt{id true}
+is elaborated to \texttt{id \{Bool\} true}, and analogously in GHC and Coq. In all
+three systems, there is also a way to explicitly specify implicit arguments: in
+Agda we may put arguments in brackets as we have seen, in Coq we can prefix a
 name with \texttt{@} to make every implicit argument explicit, as in \texttt{@id
   bool true}, and in GHC we can enable the language extension
-\texttt{TypeApplications} and write \texttt{id @Bool true}.
+\texttt{TypeApplications} and write \texttt{id @Bool True}.
 
-Implicit functions are \textbf{first class} if they can be manipulated like any
-other type. In Agda, this is the case. For example, we have a list type where
-the type of elements is an implicit function, as in \texttt{List (\{A : Set\} →
-  A → A)}. That said, Agda's elaboration has limitations when it comes to
-handling such types, and in many cases programmers must write out
-implicit lambda abstractions.
-
+Implicit functions are \textbf{first-class} if they can be manipulated like any
+other type. Coq is an example for a system where this is \emph{not} the case.
 In Coq, the core language does not have an actual implicit function type,
 instead, implicitness is tied to particular \emph{names}, and while we can write
-\texttt{List (forall \{A : Set\}, A -> A)}, the brackets here are simply ignored
-by Coq and we get a plain function a type. For illustration, Coq accepts the
+\texttt{List (forall \{A : Type\}, A -> A)}, the brackets here are simply
+ignored by Coq and we get a plain function type. For example, Coq accepts the
 following definition:
-\begin{verbatim}
-      Definition poly : forall (f : forall {A : Type}, A -> A), bool * nat :=
+\begin{Verbatim}[commandchars=\\\{\}]
+      Definition poly : forall (f : forall \{A : Type\}, A -> A), bool * nat :=
         fun f => (f bool true, f nat 0).
-\end{verbatim}
+\end{Verbatim}
 This is a higher-rank polymorphic function which returns a pair. Note that
@@
 
-In GHC Haskell, \texttt{forall} types are somewhere between Agda and Coq. We can certainly write the following, with \texttt{RankNTypes} enabled:
-\begin{verbatim}
+In GHC Haskell, \texttt{forall} types are more flexible. We can write
+the following, with \texttt{RankNTypes} enabled:
+\begin{Verbatim}[commandchars=\\\{\}]
                      poly :: (forall a. a -> a) -> (Bool, Int)
                      poly f = (f True, f 0)
-\end{verbatim}
+\end{Verbatim}
 However, polymorphic types are only supported in function domains and as fields
 of algebraic data constructors. We cannot instantiate an arbitrary type
-parameter to a \texttt{forall}, as in \texttt{[forall a. a -> a]} for the list
+parameter to a \texttt{forall}, as in \texttt{[forall a. a -> a]} for a list
 type with polymorphic elements. While this type is technically allowed by the
@@
 
+In Agda, implicit functions are truly a first-class notion, and we may have
+\texttt{List (\{A : Set\} → A → A)} without issue. However, Agda's elaboration
+still has limitations when it comes to handling implicit functions. Assume that
+we have \texttt{[]} for the empty list and \texttt{\_∷\_} for list extension,
+and consider the following code:
+\begin{Verbatim}[commandchars=\\\{\}]
+                     polyList : List (\{A : Set\} → A → A)
+                     polyList = (λ x → x) ∷ []
+\end{Verbatim}
+Agda 2.6.0.1 does not accept this. However, it does accept \texttt{polyList = (λ
+  \{A\} x → x) ∷ []}. The issue is the following. Agda first \emph{infers} a
+type for \texttt{(λ x → x) ∷ []}, then tries to unify the inferred type with the
+given \texttt{List (\{A : Set\} → A → A)} annotation. However, when Agda
+elaborates \texttt{λ x → x}, it does not yet know anything about the element
+type of the list; it is an undetermined unification variable. Hence, Agda does
+not know whether it should insert an extra \texttt{λ\{A\}} or not. If the
+element type is later found to be an implicit function, then it should, otherwise it
+should not. To solve this conundrum, Agda simply assumes that any unknown type
+is \emph{not} an implicit function type, and elects to not insert a lambda. This
+assumption is often correct, but sometimes --- as in the current case --- it is not.
+
 There is significant literature on type inference in the presence of first-class
-polymorphic types, mainly in relation to GHC and ML-like languages
-\cite{TODO}. However, none of the proposed algorithms have landed so far in
-production compilers, for reasons of complexity, fragility, interaction with
-other language features, or necessity to modify surface or core languages.
+polymorphic types, mainly in relation to GHC and ML-like languages \cite{TODO}.
+The above issue in Agda is a specific instance of the challenges described in
+the mentioned works. However, none of the proposed algorithms have landed so far
+in production compilers, for reasons of complexity, fragility and interaction
+with other language features.
+
+The solution presented in this paper is to gradually accummulate information
+about implicit insertions, and to have a setup where insertions can be refined
+and performed at any time after a particular expression is elaborated. In the
+current example, our algorithm wraps \texttt{λ x → x} in an implicit lambda with
+unknown arity, and the domain type is later refined to be \texttt{A : Set} when
+the inferred type is unified with the annotation.
 
@@
     types with telescope domain. The extensions to the target theory are modest
-    and are derivable from W-types.
+    and are derivable from W-types. We use these extensions to accummulate
+    information about implicit insertions.
   \item Our algorithm is a conservative extension of Norell's bidirectional
-    elaborator for Agda \cite[Chapter~3.6]{norell07thesis}; it accepts strictly more
+    elaborator for Agda \cite[Chapter~3]{norell07thesis}; it accepts strictly more
     programs, and does not require new constructs in the surface language.
@@
     comparable or superior to previous solutions for impredicative inference.
-  \item We provide a standalone implementation of the elaborator described in
+  \item We provide an executable implementation of the elaborator described in
     this paper.
@@
 
-\subsection{Note on Terminology}
+\subsubsection{Note on Terminology}
 
 We prefer to avoid the term ``impredicative inference'' in order to avoid
-confusion with impredicativity in type theory, which is an orthogonal notion. In
-type theory, impredicativity is a property of a universe, i.e.\ closure of a
-universe under arbitrary products. In the type inference literature,
-impredicativity means the ability to instantiate type variables and
-metavariables to polymorphic types. In particular, we have that
+confusion with impredicativity in type theory. The two notions historically
+coincided, but has since diverged to the point of being orthogonal. In type
+theory, impredicativity is a property of a universe, i.e.\ closure of a universe
+under arbitrary products. In the type inference literature, impredicativity
+means the ability to instantiate type variables and metavariables to polymorphic
+types. In particular, we have that
 \begin{itemize}
-  \item Agda is type-theory-predicative, but implements type-inference-impredicative elaboration.
-  \item Coq has type-theory-impredicative $\mathsf{Prop}$ universe, but
-    implements type-inference-predicative elaboration.
-  \item GHC is type-theory-impredicative with \texttt{RankNTypes} but without
-    \texttt{ImpredicativeTypes}, as we have \texttt{(forall a. a -> a) :: *}.
+  \item Agda has type-theory-predicative universes, but implements
+    type-inference-impredicative elaboration with first-class implicit function
+    types.
+  \item Coq has type-theory-impredicative \texttt{Prop} universe (and
+    optionally also \texttt{Set}), but implements
+    type-inference-predicative elaboration, because of the lack of implicit
+    function types.
+  \item GHC is type-theory-impredicative with \texttt{RankNTypes} enabled
+    and \texttt{ImpredicativeTypes} disabled, as we have
+    \texttt{(forall (a :: *). a -> a) :: *}.
 \end{itemize}
 
-\section{Bidirectional Elaboration}
+\section{Basic Bidirectional Elaboration}
+\label{sec:basic_bidirectional_elaboration}
 
+First, we present a variant of Norell's bidirectional elaborator
+\cite[Chapter~3]{norell07thesis}. Compared to ibid.\ we make some extensions and
+simplifications; what we end up with can be viewed as a toy version of the
+actual Agda elaborator. In this section, we use it to illustrate the key issues,
+and we extend it in Section TODO with additional rules.
+\begin{figure}[h]
+\begin{alignat*}{4}
+  t,\,u,\,v,\,A,\,B,\,C\, :&:=\quad  && x\hspace{8em}              & \text{variable}                 &  \\
+                           & |       && (x : A)\to B               & \text{function type}            &  \\
+                           & |       && \{x : A\}\to B             & \text{implicit function type}   &  \\
+                           & |       && t\,u                       & \text{application}              &  \\
+                           & |       && t\,\{u\}                   & \text{implicit application}     &  \\
+                           & |       && \lambda\,x.\, t            & \text{lambda abstraction}       &  \\
+                           & |       && \lambda\,\{x\}.\,t         & \text{implicit abstraction}     &  \\
+                           & |       && \U                         & \text{universe}                 &  \\
+                           & |       && \slet\,x : A = t\,\sin\, u & \text{let-definition}           &  \\
+                           & |       && \_                         & \text{hole for inferred term}   &
+\end{alignat*}
+\caption{Syntax of the surface language.}
+\label{fig:surface}
+\end{figure}
+
+\begin{figure}[h]
+
+\begin{alignat*}{2}
+  & \boxed{\Theta\vdash}\hspace{6em}           && \text{\emph{metacontext formation}}\\
+  & \boxed{\Theta|\Gamma\vdash}                && \text{\emph{context formation}}\\
+  & \boxed{\Theta|\Gamma\vdash t : A}          && \text{\emph{typing}}\\
+  & \boxed{\Theta|\Gamma\vdash t \equiv u : A} && \text{\emph{term equality}}
+\end{alignat*}
+
+\begin{mathpar}
+  \inferrule*[lab=metacon/empty]
+             {\\}
+             {\emptycon \vdash}
+
+  \inferrule*[lab=metacon/bind]
+             {\Theta \vdash \\ \Theta \vdash A : \U}
+             {\Theta, \alpha : A \vdash}
+
+  \inferrule*[lab=con/empty]
+             {\Theta \vdash}
+             {\Theta|\emptycon \vdash}
+
+  \inferrule*[lab=con/bind]
+             {\Theta|\Gamma \vdash \\ \Theta|\Gamma \vdash A : \U}
+             {\Theta|\Gamma,\,x : A \vdash}
+
+  \inferrule*[lab=con/define]
+             {\Theta|\Gamma \vdash \\ \Theta|\Gamma \vdash t : A}
+             {\Theta|\Gamma,\,x : A = t \vdash}
+
+  \inferrule*[lab=tm/metavar]
+             {\\}
+             {\Theta_0,\,\alpha : A,\,\Theta_1|\Gamma \vdash \alpha : A}
+
+  \inferrule*[lab=tm/bound-var]
+             {\\}
+             {\Theta|\Gamma,\,x : A,\,\Delta \vdash x : A}
+
+  \inferrule*[lab=tm/defined-var]
+             {\\}
+             {\Theta|\Gamma,\,x : A = t,\,\Delta \vdash x : A}
+
+  \inferrule*[lab=ty/u]
+             {\\}
+             {\Theta|\Gamma \vdash \U : \U}
+
+  \inferrule*[lab=let]
+             {\Theta|\Gamma\vdash t : A \\ \Theta|\Gamma,\,x : A = t \vdash u : B}
+             {\Theta|\Gamma\vdash \slet\,x : A = t\,\sin\,u : B[x\mapsto t]}
+
+  \inferrule*[lab=ty/fun]
+             {\Theta|\Gamma \vdash A : \U \\ \Theta|\Gamma,\,x : A \vdash B : \U}
+             {\Theta|\Gamma \vdash (x : A) \to B : \U}
+
+  \inferrule*[lab=ty/implicit-fun]
+             {\Theta|\Gamma \vdash A : \U \\ \Theta|\Gamma,\,x : A \vdash B : \U}
+             {\Theta|\Gamma \vdash \{x : A\} \to B : \U}
+
+  \inferrule*[lab=tm/app]
+             {\Theta|\Gamma \vdash t : (x : A) \to B \\ \Theta|\Gamma\vdash u : A}
+             {\Theta|\Gamma \vdash t\,u : B[x \mapsto u]}
+
+  \inferrule*[lab=tm/implicit-app]
+             {\Theta|\Gamma \vdash t : \{x : A\} \to B \\ \Theta|\Gamma\vdash u : A}
+             {\Theta|\Gamma \vdash t\,\{u\} : B[x \mapsto u]}
+
+  \inferrule*[lab=tm/lam]
+             {\Theta|\Gamma,\,x : A \vdash t : B}
+             {\Theta|\Gamma \vdash \lambda\,x.\,t : (x : A) \to B}
+
+  \inferrule*[lab=tm/implicit-lam]
+             {\Theta|\Gamma,\,x : A \vdash t : B}
+             {\Theta|\Gamma \vdash \lambda\,\{x\}.\,t : \{x : A\} \to B}
+
+  \inferrule*[lab=fun-$\beta$]
+             {\Theta|\Gamma,\,x : A \vdash t : B \\\Theta|\Gamma\vdash u : A }
+             {\Theta|\Gamma\vdash (\lambda\,x.\,t)\,u \equiv t[x\mapsto u] : B[x\mapsto u]}
+
+  \inferrule*[lab=implicit-fun-$\beta$]
+             {\Theta|\Gamma,\,x : A \vdash t : B \\\Theta|\Gamma\vdash u : A }
+             {\Theta|\Gamma\vdash (\lambda\,\{x\}.\,t)\,\{u\} \equiv t[x\mapsto u] : B[x\mapsto u]}
+
+  \inferrule*[lab=fun-$\eta$]
+             {\Theta|\Gamma\vdash t : (x : A)\to B}
+             {\Theta|\Gamma\vdash (\lambda\,x.\,t\,x) \equiv t : (x : A)\to B}
+
+  \inferrule*[lab=implicit-fun-$\eta$]
+             {\Theta|\Gamma\vdash t : \{x : A\}\to B}
+             {\Theta|\Gamma\vdash (\lambda\,\{x\}.\,t\,\{x\}) \equiv t : \{x : A\}\to B}
+
+  \inferrule*[lab=definition]
+             {\\}
+             {\Theta|\Gamma,\,x : A = t,\,\Delta \vdash x \equiv t : A}
+\end{mathpar}
+\caption{Selected rules of the core language.}
+\label{fig:plaincore}
+\end{figure}
+
+\subsection{Surface syntax}
+Figure \ref{fig:surface} shows the the possible constructs in the surface
+language. We only have terms, as we have Russell-style universe in the core, and
+we can conflate types and terms for convenience. The surface syntax does not
+have semantics or any well-formedness relations attached; its sole purpose is to
+serve as input to elaboration. Hence, the surface syntax can be also viewed as
+a minimal untyped tactic language, which is interpreted by the elaborator.
+
+The syntactic constructs are the almost the same in the surface language as in
+the core syntax. The difference is that $\_$ holes only appear in surface
+syntax. The $\_$ can be used to request a term to be inferred by elaboration,
+the same way as in Agda.  This can be used to give let-definitions without type
+annotation, as in $\slet\,x : \_ = \U\,\sin\,x$.
+
+\subsection{Core syntax}
+
+Figure \ref{fig:plaincore} lists selected rules of the core language. We avoid a
+fully formal presentation in this paper. Some notes on what is elided:
 \begin{itemize}
-\item Limitations in Agda
+  \item We use nameful notation and implicit weakening, i.e.\ whenever a term is
+    well-formed in some context, it is assumed to be well-formed (as it is) in
+    extended contexts. Formally, we would use de Bruijn indices for variables,
+    and define variable renaming and parallel substitution by recursion on
+    presyntax, e.g.\ as in \cite{schafer2015autosubst}. Also, typing is stable
+    under substitution.
+  \item
+    Definitional equality is understood to be a congruence and an equivalence relation,
+    which is respected by substitution and typing.
+  \item
+    We elide a number of well-formedness assumptions in rules. For instance, whenever
+    a context appears in a rule, it is assumed to be well-formed. Likewise, whenever
+    we have $\Theta|\Gamma\vdash t : A$, we assume that $\Theta|\Gamma\vdash A : \U$.
+\end{itemize}
 
-\item Insertion of implicits.
+From now on, we will only consider well-formed core syntax, and only
+constructions which respect definitional equality. In other words, we quotient
+well-formed syntax by definitional equality.
 
-\item First-class implicit functions vs Coq.
+Alternatively, one could present the syntax as a generalized algebraic theory
+\cite{sterling2019algebraic} or a quotient inductive-inductive type
+\cite{ttintt}, in which case we would get congruences and quotienting for
+definitional equality for free. However, it seems that there are a number of
+possible choices for giving an algebraic presentation of metacontexts, and
+existing works on algebraic presentations of dependent modal contexts
+(e.g.\ \cite{birkedal2018modal}, TODO) do not precisely cover our current use
+case. We leave to future work this and the investigation of elaboration from an
+algebraic perspective.
 
-\item Unique vs non-unique solutions.
+\todo{how to refer to mathpartir rules from text?}
 
-\item On the uniqueness of implicit insertion.
-\end{itemize}
+\emph{Metacontexts} are used to record metavariables which are created during
+elaboration. In our case, metacontexts are simply a context prefix, and we have
+variables pointing into it, but we do not support contextual modality e.g.\ as
+in \cite{nanevski2008contextual}. The non-meta typing context additionally
+supports \emph{defined variables}, which is used in the typing rule for
+\emph{let-definitions}, and we have that any defined variable is equal to its
+definition. We mainly support this as a convenience feature in the
+implementation of our elaborator.
+
+\emph{The universe} $\U$ is Russell-style, and we have the type-in-type rule.
+This causes our core syntax to be non-total, and our elaboration algorithm
+to be possibly non-terminating. We use type-in-type for simplicity of
+presentation, since consistent universe setups are orthogonal to the focus of
+this work.
+
+\emph{Function types} only differ from each other in notation: implicit
+functions have the same rules as ``explicit'' functions. The primary purpose of
+implicit function types is to \emph{guide elaboration}: the elaborator will at
+times compute a type and branch on whether it is an implicit function.
+
+\subsubsection{Example programs}
+
+The core syntax is quite expressive as a programming language, thanks to
+let-definitions and the type-in-type rule which allows Church-encodings of a
+large variety of inductive types. For example, the following term computes a
+list of Booleans:
+
+%% \begin{alignat*}{3}
+%%   & \slet
+
+%% \end{alignat*}
 

### 2020-01-25 Notes.hs

@@ -1,0 +1,12 @@
+{-# LANGUAGE KindSignatures #-}
+{-# LANGUAGE RankNTypes #-}
+{-# LANGUAGE TypeApplications #-}
+
+import Prelude hiding (id)
+
+id :: forall (a :: *). a -> a
+id x = x
+
+majom = 100022
+
+foo = id @Bool True

### 2020-01-27 paper.tex
stuff

@@
 
-\newcommand{\slet}{\mathsf{let}}
-\renewcommand{\sin}{\mathsf{in}}
+\newcommand{\slet}{\boldsymbol{\mathsf{let}}}
+\renewcommand{\sin}{\boldsymbol{\mathsf{in}}}
 \renewcommand{\U}{\mathsf{U}}
 \newcommand{\emptycon}{\scaleobj{.75}\bullet}
+\newcommand{\To}{\Rightarrow}
+\newcommand{\p}{\mathsf{p}}
+\newcommand{\id}{\mathsf{id}}
+\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
+\newcommand{\fresh}{\mathsf{fresh}}
+\newcommand{\unify}{\mathsf{unify}}
+\newcommand{\fail}{\mathsf{fail}}
+
+\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,#2\,#3\,#4}
+\newcommand{\einfer}[3]{\llbracket#1\rrbracket\!\Uparrow\,#2\,#3}
+\newcommand{\edo}{\boldsymbol{\mathsf{do}}}
+\newcommand{\ereturn}{\boldsymbol{\mathsf{return}}}
+
+\theoremstyle{remark}
+\newtheorem{notation}{Notation}
 
@@
 
-\begin{figure}[h]
+\begin{figure}
 
@@
 \emph{The universe} $\U$ is Russell-style, and we have the type-in-type rule.
-This causes our core syntax to be non-total, and our elaboration algorithm
-to be possibly non-terminating. We use type-in-type for simplicity of
-presentation, since consistent universe setups are orthogonal to the focus of
-this work.
+This causes our core syntax to be non-total, and our elaboration algorithm to be
+possibly non-terminating. We use type-in-type to simplify presentation, since
+consistent universe setups are orthogonal to the focus of this work.
 
@@
 
-\subsubsection{Example programs}
+\begin{notation}
+Both in the surface and core syntax, we use Agda-like notation:
+\begin{itemize}
+  \item We use $A \to B$ to refer to non-dependent functions.
+  \item We group domain types together in functions, and omit function arrows,
+    as in $\{A\,B : \U\}(x : A) \to B \to A$.
+  \item We group multiple $\lambda$-s, as in $\lambda\,\{A\}\,\{B\}\,x\,y.\,x$.
+\end{itemize}
 
+\end{notation}
+
+\begin{example}
 The core syntax is quite expressive as a programming language, thanks to
@@
 large variety of inductive types. For example, the following term computes a
-list of Booleans:
+list of types, by mapping over the list $cons\,\U\,(cons\,\U\,nil)$.
+\begin{alignat*}{3}
+  & \slet\,List : \U\to\U\\
+  & \hspace{1em}= \lambda\,A.\,(L : \U)\to(A\to L\to L)\to L\to L\,\,\sin\\
+  & \slet\,map : \{A\,B : \U\}\to (A \to B) \to List\,A \to List\,B\\
+  & \hspace{1em}=
+  \lambda\,\{A\}\,\{B\}\,f\,as\,L\,cons\,nil.\,as\,L\,(\lambda\,a.\,cons\,(f\,a))\,nil\,\,\sin\\
+  & map\,\{\U\}\,\{\U\}\,(\lambda\,A.\, A \to A)\,(\lambda\,L\,cons\,nil.\,cons\,\U\,(cons\,\U\,nil))
+\end{alignat*}
+In the core syntax, all applications must be explicit, so using implicit
+function type for $map$ above does not make much difference in optics.
+\end{example}
 
-%% \begin{alignat*}{3}
-%%   & \slet
+\subsection{Metasubstitutions}
 
-%% \end{alignat*}
+Before we can move on to the description of the elaborator, we need to specify
+metasubstitutions. These are essentially just parallel substitutions of
+metacontexts, and their purpose is to keep track of meta-operations (e.g.\ fresh
+meta creation or solution of a meta).
+
+\begin{itemize}
+  \item
+    A metasubstitution $\boxed{\theta : \Theta_0 \To \Theta_1}$ assigns to each variable
+    in $\Theta_1$ a term in $\Theta_0$ , hence it is represented as a list of
+    terms $(\alpha_1 \mapsto t_1,\,...\,\,\alpha_i \mapsto t_i)$.
+  \item
+    We define the action of a metasubstitution on contexts and terms by
+    recursion; we notate action on contexts as $\Gamma[\theta]$ and action on
+    terms as $t[\theta]$. We remark that there is no abstraction for
+    metavariables in the core syntax, so we do not have to handle variable
+    capture (or index shifting).
+\end{itemize}
+The following are admissible:
+\begin{mathpar}
+  \inferrule*[lab=metasub/empty]
+             {\Theta_0\vdash}
+             {() : \Theta_0\To \emptycon}
+
+  \inferrule*[lab=metasub/extended]
+             {\theta : \Theta_0\To\Theta_1 \\ \Theta_0|\emptycon\vdash t : A[\theta]}
+             {(\theta,\, \alpha\mapsto t) : \Theta_0\To(\Theta_1,\,\alpha : A)}
+
+  \inferrule*[lab=metasub/con-action]
+             {\Theta_1|\Gamma\vdash \\ \theta : \Theta_0\To\Theta_1}
+             {\Theta_0|\Gamma[\theta]\vdash}
+
+  \inferrule*[lab=metasub/tm-action]
+             {\Theta_1|\Gamma\vdash t : A \\ \theta : \Theta_0\To\Theta_1}
+             {\Theta_0|\Gamma[\theta] \vdash t[\theta] : A[\theta]}
+
+  \inferrule*[lab=metasub/identity]
+             {\\}
+             {\id : \Theta\To\Theta}
+
+  \inferrule*[lab=metasub/composition]
+             {\theta_{0} : \Theta_1 \To \Theta_2 \\ \theta_1 : \Theta_0 \To \Theta_1}
+             {\theta_0 \circ \theta_1 : \Theta_0 \To \Theta_2}
+
+  \inferrule*[lab=metasub/weakening]
+             {\\}
+             {\p : (\Theta,\,x : A) \To \Theta}
+\end{mathpar}
+
+The identity substitution $\id$ maps each variable to itself. Composition
+is given by pointwise term substitution, and $\id$ and $\blank\circ\blank$
+yields a category. The weakening substitution $\p$ (the naming comes from
+categories-with-families terminology \cite{dybjer1995internal}) can be defined
+as dropping the last entry from $\id : (\Theta,\,x : A) \To (\Theta,\,x : A)$.
+
+\subsection{Fresh Metavariables}
+
+Using \emph{contextual metavariables} is a standard practice in the
+implementation of dependently typed languages. This means that every ``hole'' in
+the surface language is represented as an unknown function which abstracts over
+all bound variables in the scope of a hole. Unlike \cite{nanevski2008contextual}
+and similarly to \cite{gundry2013type}, we do not have a first-class notion of
+contextual types, and instead reuse the standard dependent function type to
+abstract over enclosing contexts.
+
+\begin{definition}[Closing type] For each $\Theta|\Gamma \vdash A : \U$, we define
+$\Gamma \To A$ by recursion on $\Gamma$ such that $\Theta|\emptycon\vdash \Gamma
+\To A : \U$.
+  \begin{alignat*}{3}
+    &((\Gamma,\, x : A) \To B)     && :\equiv (\Gamma \To ((x : A) \to B))\\
+    &((\Gamma,\, x : A = t) \To B) && :\equiv (\Gamma \To B[x\mapsto t])\\
+    &(\emptycon \To B)             && :\equiv B
+  \end{alignat*}
+\end{definition}
+
+\begin{definition}[Contextualization] For each $\Theta|\Gamma \vdash t : \Gamma\To A$,
+we define $t\,\overline{\id_{\Gamma}}$ by recursion on $\Gamma$ such that
+$\Theta|\Gamma\vdash t\,\overline{\id_{\Gamma}} : A$. Informally, this is $t$
+applied to all bound variables in $\Gamma$.
+  \begin{alignat*}{3}
+    &(t\,\overline{\id_{\Gamma,\,x:A}})   && :\equiv (t\,\overline{\id_{\Gamma}})\,x \\
+    &(t\,\overline{\id_{\Gamma,\,x:A=t}}) && :\equiv (t\,\overline{\id_{\Gamma}}) \\
+    &(t\,\overline{\id_{\emptycon}})     && :\equiv t
+  \end{alignat*}
+\end{definition}
+
+\begin{example} If we have $\Gamma \equiv (\emptycon,\,A : \U,\,B : A \to \U)$, then
+$(\Gamma \To \U) \equiv ((A : \U)(B : A \to \U) \to \U)$ and
+$t\,\overline{\id_{\Gamma}} \equiv t\,A\,B$.
+\end{example}
+
+\begin{definition}[Fresh meta creation] We specify $\fresh_A$ as follows:
+  \begin{mathpar}
+  \inferrule*{\Theta|\Gamma\vdash A : \U}
+             {\fresh_A \in \{(\Theta',\,\theta,\,t)\,|\,(\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t : A[\theta])\}}
+  \end{mathpar}
+The definition $\fresh_A :\equiv ((\Theta,\,\alpha:\Gamma\To
+A),\,\p,\,\alpha\,\overline{\id_{\Gamma}})$, where $\alpha$ is fresh in
+$\Theta$, satisfies this specification. We extend $\Theta$ with a fresh meta,
+which has the closing type $\Gamma \To A$. The $\p$ weakening relates the new
+metacontext to the old one, by ``dropping'' the new entry. Lastly,
+$\alpha\,\overline{\id_{\Gamma}}$ is the new meta
+applied to all bound variables.
+\end{definition}
+
+\subsection{Unification}
+
+We assume that there is a partial unification procedure, which returns a unifying
+metasubstitution on success. We only have \emph{homogeneous} unification, i.e.\ the
+two terms to be unified must have the same type. The specification is as follows:
+\begin{mathpar}
+\inferrule*{\Theta|\Gamma\vdash t : A \\ \Theta|\Gamma\vdash u : A}
+           {\unify\,t\,u \in \{(\Theta',\,\theta)\,|\,(\theta : \Theta' \To \Theta)
+             \,\land\,(\Theta'|\Gamma[\theta]\vdash
+             t[\theta] \equiv u[\theta] : A[\theta])\}\,\cup\,\{\fail\}}
+\end{mathpar}
+
+Here, we do not require that unification returns most general unifiers, nor do
+we go into the details of how unification is implemented. Gundry describes
+unification in detail in \cite[Chapter~4]{gundry2013type} for a similar syntax,
+with a similar (though more featureful) setup for metacontexts. Note that our
+unification algorithm does not support \emph{constraint postponing}, since we
+have not specified constraints. In our concrete implementation, unification supports
+basic higher-order pattern unification and metavariable pruning \cite{TODO}.
+
+\subsection{Elaboration}
+
+Elaboration consists of two (partial) functions, checking and inferring, which
+are defined by mutual induction on surface syntax. They are specified as
+follows:
+
+\begin{mathpar}
+\inferrule*[lab=checking]
+           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
+           {\echeck{t}{\Theta}{\Gamma}{A} \in
+             \{(\Theta',\,\theta,\,t')\,|\,(\theta : \Theta' \To \Theta)\,\land\,
+               (\Theta'|\Gamma[\theta]\vdash t' : A[\theta])\}\cup\{\fail\}
+           }
+
+\inferrule*[lab=inferring]
+           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
+           {\einfer{t}{\Theta}{\Gamma} \in
+             \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
+               (\Theta'|\Gamma[\theta]\vdash t' : A)\}\cup\{\fail\}
+           }
+\end{mathpar}
+In the following, we use a monadic pseudocode notation, where the side effect is
+failure.
+
+\begin{alignat*}{3}
+  &\echeck{\lambda\,x.\,t}{\Theta}{\Gamma}{((x : A)\to B)} :\equiv \edo \\
+  &\hspace{2em}(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\hspace{2em}\ereturn\,(\Theta',\,\theta,\,\lambda\,x.\,t')\\
+  &\echeck{\lambda\,x.\,t}{\Theta}{\Gamma}{((x : A)\to B)} :\equiv \edo \\
+\end{alignat*}
 

### 2020-02-02 paper.tex
stuff

@@
 \newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
-\newcommand{\fresh}{\mathsf{fresh}}
+\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,#1\,#2\,#3}
 \newcommand{\unify}{\mathsf{unify}}
@@
 
+\newcommand{\echeckblank}{\llbracket\blank\rrbracket\!\Downarrow}
+\newcommand{\einferblank}{\llbracket\blank\rrbracket\!\Uparrow}
 \newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,#2\,#3\,#4}
-\newcommand{\einfer}[3]{\llbracket#1\rrbracket\!\Uparrow\,#2\,#3}
+\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,#2\,#3\,#4}
+\newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
+\newcommand{\einfert}[2]{\llbracket#1\rrbracket\!\Uparrow\,#2}
 \newcommand{\edo}{\boldsymbol{\mathsf{do}}}
 \newcommand{\ereturn}{\boldsymbol{\mathsf{return}}}
+\newcommand{\eif}{\boldsymbol{\mathsf{if}}}
+\newcommand{\ethen}{\boldsymbol{\mathsf{then}}}
+\newcommand{\eelse}{\boldsymbol{\mathsf{else}}}
+\newcommand{\ecase}{\boldsymbol{\mathsf{case}}}
+\newcommand{\eof}{\boldsymbol{\mathsf{of}}}
+\newcommand{\true}{\mathsf{true}}
+\newcommand{\false}{\mathsf{false}}
+\newcommand{\einsert}[5]{\mathsf{insert}\,#1\,#2\,#3\,#4\,#5}
+
+\newcommand{\Nat}{\mathsf{Nat}}
+\newcommand{\zero}{\mathsf{zero}}
+\newcommand{\suc}{\mathsf{suc}}
 
@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
     well-formed in some context, it is assumed to be well-formed (as it is) in
-    extended contexts. Formally, we would use de Bruijn indices for variables,
+    extended contexts. We also assume that any specifically mentioned name is
+    fresh, e.g.\ when we write $\Theta,\,\alpha : A$, we assume that $\alpha$ is
+    fresh in $\Theta$.  Formally, we would use de Bruijn indices for variables,
     and define variable renaming and parallel substitution by recursion on
-    presyntax, e.g.\ as in \cite{schafer2015autosubst}. Also, typing is stable
+    presyntax, e.g.\ as in \cite{schafer2015autosubst}.
+  \item Fixing any $\Theta$ metacontext, parallel substitutions of bound and
+    defined variables form morphisms of a category, where the identity
+    substitution $\id$ maps each variable to itself and composition
+    $\blank\circ\blank$ is given by pointwise substitution. The action of
+    parallel substitution on terms is functorial, i.e.\ $t[\sigma][\delta]
+    \equiv t[\sigma\circ\delta]$ and $t[\id] \equiv t$, and typing is stable
     under substitution.
@@
 \cite{sterling2019algebraic} or a quotient inductive-inductive type
-\cite{ttintt}, in which case we would get congruences and quotienting for
-definitional equality for free. However, it seems that there are a number of
-possible choices for giving an algebraic presentation of metacontexts, and
-existing works on algebraic presentations of dependent modal contexts
-(e.g.\ \cite{birkedal2018modal}, TODO) do not precisely cover our current use
-case. We leave to future work this and the investigation of elaboration from an
-algebraic perspective.
+\cite{ttintt}, in which case we would get congruences and quotienting for free,
+and we would also get a rich model theory for our syntax. However, it seems that
+there are a number of possible choices for giving an algebraic presentation of
+metacontexts, and existing works on algebraic presentations of dependent modal
+contexts (e.g.\ \cite{birkedal2018modal}, TODO) do not precisely cover the
+current use case. We leave this to future work, along with the investigation of
+elaboration from an algebraic perspective.
 
@@
 \end{itemize}
+\end{notation}
 
+\begin{notation}[Spines]\label{not:spines}
+  We sometimes use a spine notation for neutral terms. A spine is a list of
+  terms, noted as $\overline{t}$, where terms may be wrapped in brackets to
+  signal implicit application. For example, if $\overline{u} \equiv
+  (\{A\},\,\{B\},\,x)$, then $t\,\overline{u}$ denotes $t\,\{A\}\,\{B\}\,x$.  In
+  $t\,\overline{u}$, we call $t$ the \emph{head} of the neutral term. In
+  particular, if $t$ is a metavariable, the neutral term is \emph{meta-headed}.
 \end{notation}
@@
 let-definitions and the type-in-type rule which allows Church-encodings of a
-large variety of inductive types. For example, the following term computes a
-list of types, by mapping over the list $cons\,\U\,(cons\,\U\,nil)$.
+large class of inductive types. For example, the following term computes a list
+of types, by mapping over the list $cons\,\U\,(cons\,\U\,nil)$.
 \begin{alignat*}{3}
@@
 
-The identity substitution $\id$ maps each variable to itself. Composition
-is given by pointwise term substitution, and $\id$ and $\blank\circ\blank$
-yields a category. The weakening substitution $\p$ (the naming comes from
+The identity substitution $\id$ maps each variable to itself. Composition is
+given by pointwise term substitution, $\id$ and $\blank\circ\blank$ yields a
+category, and the action of metasubstitution on contexts and terms is
+functorial. The weakening substitution $\p$ (the naming comes from
 categories-with-families terminology \cite{dybjer1995internal}) can be defined
@@
 \begin{definition}[Closing type] For each $\Theta|\Gamma \vdash A : \U$, we define
-$\Gamma \To A$ by recursion on $\Gamma$ such that $\Theta|\emptycon\vdash \Gamma
-\To A : \U$.
+$\Gamma \To A$ by recursion on $\Gamma$, such that $\Theta|\emptycon\vdash
+  \Gamma \To A : \U$.
   \begin{alignat*}{3}
@@
 \begin{definition}[Contextualization] For each $\Theta|\Gamma \vdash t : \Gamma\To A$,
-we define $t\,\overline{\id_{\Gamma}}$ by recursion on $\Gamma$ such that
-$\Theta|\Gamma\vdash t\,\overline{\id_{\Gamma}} : A$. Informally, this is $t$
-applied to all bound variables in $\Gamma$.
+we define the spine $\overline{\mathsf{vars}_{\Gamma}}$ such that that
+$\Theta|\Gamma\vdash t\,\overline{\mathsf{vars}_{\Gamma}} : A$. Informally, this
+is $t$ applied to all bound variables in $\Gamma$.
   \begin{alignat*}{3}
-    &(t\,\overline{\id_{\Gamma,\,x:A}})   && :\equiv (t\,\overline{\id_{\Gamma}})\,x \\
-    &(t\,\overline{\id_{\Gamma,\,x:A=t}}) && :\equiv (t\,\overline{\id_{\Gamma}}) \\
-    &(t\,\overline{\id_{\emptycon}})     && :\equiv t
+    &(t\,\overline{\mathsf{vars}_{\Gamma,\,x:A}})   && :\equiv (t\,\overline{\mathsf{vars}_{\Gamma}})\,x \\
+    &(t\,\overline{\mathsf{vars}_{\Gamma,\,x:A=t}}) && :\equiv (t\,\overline{\mathsf{vars}_{\Gamma}}) \\
+    &(t\,\overline{\mathsf{vars}_{\emptycon}})     && :\equiv t
   \end{alignat*}
@@
 $(\Gamma \To \U) \equiv ((A : \U)(B : A \to \U) \to \U)$ and
-$t\,\overline{\id_{\Gamma}} \equiv t\,A\,B$.
+$t\,\overline{\mathsf{vars}_{\Gamma}} \equiv t\,A\,B$.
 \end{example}
 
-\begin{definition}[Fresh meta creation] We specify $\fresh_A$ as follows:
+\begin{definition}[Fresh meta creation] We specify $\freshMeta{\Theta}{\Gamma}{A}$ as follows:
   \begin{mathpar}
   \inferrule*{\Theta|\Gamma\vdash A : \U}
-             {\fresh_A \in \{(\Theta',\,\theta,\,t)\,|\,(\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t : A[\theta])\}}
+             {\freshMeta{\Theta}{\Gamma}{A} \in \{(\Theta',\,\theta,\,t)\,|\,(\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t : A[\theta])\}}
   \end{mathpar}
-The definition $\fresh_A :\equiv ((\Theta,\,\alpha:\Gamma\To
-A),\,\p,\,\alpha\,\overline{\id_{\Gamma}})$, where $\alpha$ is fresh in
+The definition $\freshMeta{\Theta}{\Gamma}{A} :\equiv ((\Theta,\,\alpha:\Gamma\To
+A),\,\p,\,\alpha\,\overline{\mathsf{vars}_{\Gamma}})$, where $\alpha$ is fresh in
 $\Theta$, satisfies this specification. We extend $\Theta$ with a fresh meta,
@@
 metacontext to the old one, by ``dropping'' the new entry. Lastly,
-$\alpha\,\overline{\id_{\Gamma}}$ is the new meta
+$\alpha\,\overline{\mathsf{vars}_{\Gamma}}$ is the new meta
 applied to all bound variables.
 \end{definition}
+
+\subsection{Implicit Argument Insertion}
+
+We define a helper function which inserts implicit applications around a core
+term. For example, if we have a defined name $id$ with type $\{A : \U\}\to A \to
+A$ occurring in the surface program, we usually want to expand $id$ to
+$id\,\{\alpha\}$, where $\alpha$ is a fresh metavariable.
+\begin{mathpar}
+\inferrule*{ins \in \{\true,\,\false\} \\ \Theta|\Gamma\vdash t : A}
+           {\einsert{ins}{\Theta}{\Gamma}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
+             (\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t' : A')\}}
+\end{mathpar}
+We have an additional $ins \in \{\true,\,\false\}$ parameter, which simply
+toggles whether any insertion is to be performed. We will use this in the
+definition of elaboration. Insertion is defined by recursion on the $A$ core
+type, as follows.
+\begin{alignat*}{3}
+  &\einsert{\false}{\Theta}{\Gamma}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
+  &\einsert{\true\,}{\Theta}{\Gamma}{t}{(\{x : A\}\to B)} && :\equiv&&
+       \slet\,(\Theta',\,\theta,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
+  & && &&\sin\,\einsert{\true}{\Theta'}{(\Gamma[\theta])}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
+  &\einsert{\true\,}{\Theta}{\Gamma}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
+\end{alignat*}
 
@@
 unification in detail in \cite[Chapter~4]{gundry2013type} for a similar syntax,
-with a similar (though more featureful) setup for metacontexts. Note that our
-unification algorithm does not support \emph{constraint postponing}, since we
-have not specified constraints. In our concrete implementation, unification supports
-basic higher-order pattern unification and metavariable pruning \cite{TODO}.
+with a similar (though more featureful) setup for metacontexts.
+
+\todo{example for meta solution, more references}
+
+Note that our unification algorithm does not support \emph{constraint
+  postponing}, since we have not specified constraints. In our concrete
+implementation, unification supports basic higher-order pattern unification and
+metavariable pruning \cite{TODO}.
 
@@
 
+In this section we define the elaboration algorithm. First, we explain the used notations.
+
+\begin{itemize}
+  \item We use a Haskell-like monadic pseudocode notation, where the side effect is
+    failure via $\fail$.
+  \item We use pattern matching notation on core terms; e.g.\ we may match on
+    whether a type is a function type. This requires an evaluation/normalization
+    procedure on core terms, but we already assume this in unification.
+\end{itemize}
+
 Elaboration consists of two (partial) functions, checking and inferring, which
 are defined by mutual induction on surface syntax. They are specified as
-follows:
+follows. We also have the additional $ins \in \{\true,\,\false\}$ argument for
+inference, which toggles whether the output has inserted implicit arguments.  We
+explain the cases of the definition in order.
 
@@
 \inferrule*[lab=inferring]
-           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
-           {\einfer{t}{\Theta}{\Gamma} \in
+           {ins \in \{\true,\,\false\} \\
+                \text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
+           {\einfer{t}{ins}{\Theta}{\Gamma} \in
              \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
@@
 \end{mathpar}
-In the following, we use a monadic pseudocode notation, where the side effect is
-failure.
 
+\begingroup
+\allowdisplaybreaks
 \begin{alignat*}{3}
   &\echeck{\lambda\,x.\,t}{\Theta}{\Gamma}{((x : A)\to B)} :\equiv \edo \\
-  &\hspace{2em}(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
-  &\hspace{2em}\ereturn\,(\Theta',\,\theta,\,\lambda\,x.\,t')\\
-  &\echeck{\lambda\,x.\,t}{\Theta}{\Gamma}{((x : A)\to B)} :\equiv \edo \\
+  &\quad(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\quad\ereturn\,(\Theta',\,\theta,\,\lambda\,x.\,t')\\
+  &\echeck{\lambda\,\{x\}.\,t}{\Theta}{\Gamma}{(\{x : A\}\to B)} :\equiv \edo \\
+  &\quad(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\quad\ereturn\,(\Theta',\,\theta,\,\lambda\,\{x\}.\,t')\\
+  &\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)} : \equiv \edo \\
+  &\quad (\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\quad (\Theta',\,\theta,\,\lambda\,\{x\}.\,t')\\
+  &\echeck{\slet\,x:A=t\,\sin\,u}{\Theta_0}{\Gamma}{B} :\equiv \edo \\
+  &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_0}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,t') \leftarrow \echeck{t}
+                {\Theta_1}{(\Gamma[\theta_1])}{A'}\\
+  &\quad(\Theta_3,\,\theta_3,\,u') \leftarrow
+                \echeck{u}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(B[\theta_1\circ\theta_2])}\\
+  &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u')\\
+  &\echeck{\_}{\Theta}{\Gamma}{A} :\equiv \edo \\
+  &\quad\ereturn\,(\freshMeta{\Theta}{\Gamma}{A})\\
+  &\echeck{t}{\Theta_0}{\Gamma}{A} :\equiv \edo \\
+  &\quad(\Theta_1,\,\theta_1,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
+  &\quad(\Theta_2,\,\theta_2) \leftarrow \unify\,(A[\theta_1])\,B\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,t'[\theta_2])\\\\
+  &\einfer{x}{ins}{\Theta}{\Gamma} :\equiv \edo\\
+  &\quad\eif\,(x : A \in \Gamma)\,\lor\,(x : A = t \in \Gamma) \\
+  &\qquad   \ethen\, \ereturn\,(\einsert{ins}{\Theta}{\Gamma}{x}{A})\\
+  &\qquad   \eelse\, \fail\\
+  &\einfer{\U}{ins}{\Theta}{\Gamma} :\equiv \edo\\
+  &\quad\ereturn\,(\Theta,\,\id,\,\U,\,\U)\\
+  &\einfer{(x : A)\to B}{ins}{\Theta_0}{\Gamma} :\equiv \edo\\
+  &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_1}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,B') \leftarrow
+                 \echeck{B}{\Theta_2}{(\Gamma[\theta_1],\,x : A')}{\U}\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,
+                 \,((x : A'[\theta_2])\to B'),\,\U)\\
+  &\einfer{\{x : A\}\to B}{ins}{\Theta_0}{\Gamma} :\equiv \edo\\
+  &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_1}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,B') \leftarrow
+                 \echeck{B}{\Theta_2}{(\Gamma[\theta_1],\,x : A')}{\U}\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,
+                 \,(\{x : A'[\theta_2]\}\to B'),\,\U)\\
+  &\einfer{\lambda\,x.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{(\Gamma[\theta_1],\,x:A)}\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\
+  &\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{(\Gamma[\theta_1],\,x:A)}\\
+  &\quad\ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(\lambda\,\{x\}.\,t')}{(\{x : A[\theta_2]\}\to B)})\\
+  &\einfer{t\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
+  &\quad \ecase\,\,A\,\,\eof\\
+  &\qquad   ((x : A)\to B) \To \edo\\
+  &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
+                    \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
+  &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,u')}{(B[\theta_2][x\mapsto u'])})\\
+  &\qquad A \To \fail\\
+  &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\false}{\Theta_0}{\Gamma}\\
+  &\quad \ecase\,\,A\,\,\eof\\
+  &\qquad   (\{x : A\}\to B) \To \edo\\
+  &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
+                    \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
+  &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,\{u'\})}{(B[\theta_2][x\mapsto u'])})\\
+  &\qquad A \To \fail\\
+  &\einfer{\slet\,x:A=t\,\sin\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_0}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,t') \leftarrow \echeck{t}
+                {\Theta_1}{(\Gamma[\theta_1])}{A'}\\
+  &\quad(\Theta_3,\,\theta_3,\,u',\,B) \leftarrow
+                \einfer{u}{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}\\
+  &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,(\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u'),\,B)\\
+  &\einfer{\_}{ins}{\Theta}{\Gamma} :\equiv \edo \\
+  &\quad \slet\,(\Theta',\,\theta,\,A) = \freshMeta{\Theta}{\Gamma}{\U}\\
+  &\quad \ereturn\,(\freshMeta{\Theta'}{(\Gamma[\theta])}{A})\\
 \end{alignat*}
+\endgroup
+
+\subsubsection{Checking}
+The first two clauses are checking $\lambda$-s, where the expected type exactly
+matches the $\lambda$ expressions. Hence, we simply check under binders with
+$\echeck{t}{\Theta}{(\Gamma,\,x:A)} B$, and wrap the resulting term in the
+appropriate $\lambda$.
+
+The third clause for $\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$ is more
+interesting. Here, we are checking a surface term which is \emph{not} a
+$\lambda$ (this follows from our top-down pattern matching notation), with an
+implicit function expected type. Here, we check $t$ in the extended $\Gamma,\,x
+: A$ context, and we insert a new implicit $\lambda$ in the elaboration
+output. This is the only point where implicit $\lambda$-s are introduced by
+elaboration. Practically, this rule is commonly useful whenever we have a
+higher-order function where some arguments have implicit function type. For
+example, in the surface syntax, assume natural numbers, and an induction
+principle for them:
+\[
+NatInd : \{P : \Nat \to \U\}\to \,P\,\zero \to (\{n : \Nat\}\to P\,n \to
+         P\,(\suc\,n)) \to (n : \Nat) \to P\,n
+\]
+Then, define addition using induction:
+\begin{alignat*}{1}
+& \slet\, NatPlus : \Nat \to \Nat \to \Nat\\
+& \qquad = NatInd\,(\lambda\,n.\,\Nat\to\Nat)\,(\lambda\,m.\,m)\,
+                        (\lambda\,f\,m.\,\suc\,(f\,m))\,\,\sin\,\,...
+\end{alignat*}
+When the above is elaborated, the $\lambda\,f\,m.\,\suc\,(f\,m)$ function is
+checked with the expected type $\{n : \Nat\}\to (\Nat \to \Nat) \to (\Nat \to
+\Nat)$, and the elaboration output is $\lambda\,\{n\}\,f\,m.\,\suc\,(f\,m)$.
+Hence, in this case we do not have to write implicit $\lambda$ in the surface
+syntax.
+
+For $\echeck{\slet\,x:A=t\,\sin\,u}{\Theta_0}{\Gamma}{B}$, we simply let
+checking fall through. The definition here is a bit noisy, because we need to
+thread metasubstitutions through, and we always have to ``update'' core terms
+and contexts with the current metasubstitution.
+
+For $\echeck{\_}{\Theta}{\Gamma}{A}$, we return a fresh metavariable with the
+expected type. In any other $\echeck{t}{\Theta_0}{\Gamma}{A}$ case, we have a
+\emph{change of direction}: we infer a type for $t$, then unify the expected and
+inferred types.
+
+\subsubsection{Inferring}
+
+For $\einfer{x}{ins}{\Theta}{\Gamma}$, we look up the type of $x$ in $\Gamma$,
+and insert implicit applications if needed. In the case of $\U$, we always
+succeed and infer $\U$ as type. In the cases for function types, we check that
+the domains and codomains have type $\U$.
+
+For $\lambda$-s, we create a fresh meta for the domain type (since our surface
+$\lambda$-s are not annotated), and infer the bodies. In the case of
+$\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma}$, we additionally perform
+$\einsert$ on the output. \todo{explain this weird case}
+
+The $\einfer{t\,u}{ins}{\Theta_0}{\Gamma}$ and
+$\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma}$ cases are again interesting. Here, we
+first infer a type for the term which is being applied, and only proceed if the
+type is an appropriate function. Note a difference between the explicit and
+implicit case. In the former case, we use $\einfer{t}{\true}{\Theta_0}{\Gamma}$,
+which inserts implicit applications. In the latter case we do not insert
+implicit applications. This, together with the inference definition for
+variables, ensures that implicit applications in the surface syntax behave
+similarly as in Agda. For example, given $id : \{A : \U\}\to A\to A$ in scope,
+we elaborate $id\,\U$ as follows:
+\begin{enumerate}
+  \item The expression is an explicit application, so we infer $\id$ and insert implicit
+        arguments, returning $id\,\{\alpha\}$, where $\alpha$ is a fresh meta.
+  \item We check that $\U$ has type $\alpha$. Here we immediately change direction,
+        inferring $\U$ as type for $\U$ and unifying $\alpha$ with $\U$.
+  \item Hence, the resulting output is $id\,\{\U\}\,\U$.
+\end{enumerate}
+On the other hand, we elaborate $id\,\{\U\}$ as follows:
+\begin{enumerate}
+  \item This is an implicit application, so we infer a type for $id$ without inserting
+    implicit arguments. This yields the inferred type $\{A : \U\}\to A\to A$, and
+    we check that $\U$ has type $\U$.
+  \item This changes direction again, and we infer $\U$ type for $\U$, and
+    successfully unify $\U$ with $\U$.
+\end{enumerate}
+\todo{explain that we fail if $t$ has meta-headed type or handle it in elab}
+
+In the case of $\slet$, inference again just falls through, and we infer a type
+for the $\slet$ body. For $\einfer{\_}{ins}{\Theta}{\Gamma}$, we create a fresh meta
+for the type of the hole, and another fresh meta for the hole itself.
+
+\subsubsection{On soundness, completeness, and other properties}
+We briefly discuss what we can and cannot say about the currently presented
+elaborator. First, elaboration is \emph{sound} in the sense that it never
+produces malformed core syntax.
+
+\begin{theorem}[Soundness of elaboration] The definitions of $\echeckblank$ and
+$\einferblank$ conform to the specification \ref{TODO}.
+This follows by induction on surface syntax and analyzing the definitions, while also relying on properties of substitution, metasubstitution, $\unify$ and $\freshMeta$.
+\qed
+\end{theorem}
+
+We remark that this notion of soundness is only a ``sanity'' or well-typing
+statement for elaboration. In fact, we could define elaboration as a constantly
+failing partial function, and it would also conform to the specification. The
+right way to view this, is that $\echeckblank$ and $\einferblank$ together
+with their specification constitute the semantics of surface syntax. We do not
+give any other semantics to the surface syntax, nor does it support any other
+operation.
+
+We do not present any \emph{completeness} result for elaboration in this
+paper. For an example of what this would entail, in \cite{dunfield2013complete},
+completeness means that whenever there is a way to fill in missing details in
+the surface syntax, algorithmic typechecking \emph{always} finds it. In
+ibid.\ this means figuring out domain types for $\lambda$-s and inserting all
+implicit applications. However, our elaborator targets a far stronger theory,
+and it is beyond our reach to succinctly characterize which annotations are
+inferable in the surface language.
+
+We can still say something about what kind of work our elaborator does. For
+this, we consider a translation from core terms to surface terms, the obvious
+``forgetful'' translation, which maps terms to surface counterparts. Now, this
+is an ``evil'' construction on core terms, since it does not preserve
+definitional equality. We shall only use this evil notion in the following
+statement.
+
+\begin{theorem}[Conservativity]
+Elaboration is conservative over the surface syntax, in the sense that for any
+surface term $t$, if checking or inference outputs $t'$, then the forgetful
+translation of $t'$ differs from $t$ only by
+  \begin{itemize}
+    \item Having all $\_$ holes filled in by expressions.
+    \item Having extra implicit $\lambda$-s and implicit applications inserted.
+  \end{itemize}
+This follows by straightforward induction on surface syntax. \qed
+\end{theorem}
+
+Also, note that we do not support \emph{let-generalization}. This is an open
+research topic in settings with dependent types, and we make no attempt at
+tackling it. See \cite{eisenberg2016dependent} for a treatment in a proposed
+dependent version of Haskell.
+
+\section{Issues with First-Class Implicit Functions}
+
+We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
+assume the following:
+\begin{alignat*}{3}
+  & List && : \U \to \U \\
+  & nil  && : \{A : \U\} \to List\,A\\
+  & cons && : \{A : \U\} \to A \to List\,A
+\end{alignat*}
+In the following, we present a trace of checking $cons\,(\lambda\,x.\,x)\,nil$
+at type $List\,(\{A : \U\}\to A \to A)$. We omit context and metacontext
+parameters everywhere, and notate recursive calls by indentation. We also omit
+some checking, inference, implicit insertion and unification calls which are not
+essential for illustration.
+\begin{alignat*}{3}
+  & \scriptstyle{0 }\qquad\qquad\quad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}\\
+  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil}{\true} \\
+  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}{\true}\\
+  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}{\true}\\
+  & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
+  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
+  & \scriptstyle{6 }  && \qquad\quad = \lambda\,x.\,x\\
+  & \scriptstyle{7 }  && \qquad = cons\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x) : List\,(\alpha_1\to\alpha_1) \to List\,(\alpha_1\to\alpha_1)\\
+  & \scriptstyle{8 }  && \qquad \echeckt{nil}{(List\,(\alpha_1\to\alpha_1))}\\
+  & \scriptstyle{9 }  && \qquad = nil\,\{\alpha_1\to\alpha_1\}\\
+  & \scriptstyle{10} && \quad = cons\,\{\alpha_1 \to \alpha_1\}\,(\lambda\,x.\,x)\,(nil\,\{\alpha_1\to\alpha_1\}): List\,(\alpha_1\to\alpha_1)\\
+  & \scriptstyle{11} && \quad \unify\,(List\,(\{A : \U\}\to A \to A))\,(List\,(\alpha_1\to\alpha_1))\\
+  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\alpha_1\to\alpha_1)\\
+  & \scriptstyle{13} && \qquad = \fail
+\end{alignat*}
+
+Above, we first infer $cons\,(\lambda\,x.\,x)\,nil$, which inserts implicit
+applications to fresh metas in $cons$ and $nil$, and returns $cons\,\{\alpha_1
+\to \alpha_1\}\,(\lambda\,x.\,x)\,(nil\,\{\alpha_1\to\alpha_1\}) :
+List\,(\alpha_1\to\alpha_1)$. Here, the $\alpha_0$ meta is refined to $\alpha_1
+\to \alpha_1$ when we check $\lambda\,x.\,x$. In the end, we need to unify the
+expected and inferred types, which fails, since we have an implicit function
+type on one side and an explicit function on the other side.
+
+Why does this fail? The culprit is line $\scriptstyle{5}$, where we call
+$\echeckt{\lambda\,x.\,x}{\alpha_0}$. At this point, the checking type is not an
+implicit function type (it is a meta), so we do not insert an implicit
+$\lambda$. At the heart of the issue is that elaboration makes insertion choices
+based on core types.
+\begin{enumerate}
+\item $\echeck{t}{\Theta}{\Gamma}{A}$ can insert a $\lambda$ only if $A$ is an implicit function type.
+\item $\einsert{\true}{\Theta}{\Gamma}{A}$ inserts an application only if $A$ is an implicit function type.
+\end{enumerate}
+In both of these cases, if $A$ is of the form $\alpha\,\overline{u}$
+(i.e.\ meta-headed), then it is possible that $\alpha$ is later refined to an
+implicit function, but at that point we have already missed our shot at implicit
+insertion.
+
+At least for $\lambda$-insertion, there is a potential solution: just
+\emph{postpone} checking a term until the shape of the checking type is known
+for sure. This was included as part of a proposed solution for smarter
+$\lambda$-insertions in \cite{johansson2015eliminating}. This means that
+checking with a meta-headed type returns a ``guarded constant''
+\cite[Chapter~3]{norell07thesis}, an opaque stand-in which only computes to an
+actual core term when the checking type becomes known. In practice, this
+solution has a painful drawback: \emph{we get no information at all from checked
+  terms before the guard is unblocked}. For an example for unexpected behavior
+with this solution, let us assume $Bool : \U$ and $true : Bool$, and try to
+infer type for the following surface term:
+\begin{alignat*}{2}
+  \slet\,x : \_ = true\,\sin\, x
+\end{alignat*}
+We first insert a fresh meta $\alpha$ for the hole, and then check $true$ with
+$\alpha$. We postpone this checking, returning a guarded constant, and then
+infer a type for $x$, which is $\alpha$. Hence, this small example yields an
+unsolved meta and a guarded constant in the output.
+
+Now, this particular example can be repaired by special-casing the elaboration
+of a $\slet$-definition without an explicit type annotation. However, the
+current author's experience from playing with an implementation of this
+solution, is that we are missing too much information by postponing, and this
+cascades in an unfortunate way: postponing yields more unsolved metas, which
+cause more postponing.
+
+\section{Telescopes and Strictly Curried Functions}
 

### 2020-02-05 implicit-fun-elab/ElabState.hs
start revamping Haskell implementation

@@ -1,0 +1,34 @@
+
+module ElabState where
+
+import Data.IORef
+import System.IO.Unsafe
+import qualified Data.IntMap.Strict as IM
+
+import Types
+
+mcxt :: IORef MCxt
+mcxt = unsafeDupablePerformIO (newIORef mempty)
+{-# noinline mcxt #-}
+
+nextMId :: IORef Int
+nextMId = unsafeDupablePerformIO (newIORef 0)
+{-# noinline nextMId #-}
+
+lookupMetaIO :: MId -> IO MetaEntry
+lookupMetaIO m = do
+  ms <- readIORef mcxt
+  case IM.lookup m ms of
+    Just e -> pure e
+    _      -> error "impossible"
+
+lookupMeta :: MId -> MetaEntry
+lookupMeta m = unsafeDupablePerformIO (lookupMetaIO m)
+
+writeMetaIO :: MId -> MetaEntry -> IO ()
+writeMetaIO m e = modifyIORef' mcxt (IM.insert m e)
+
+reset :: IO ()
+reset = do
+  writeIORef mcxt mempty
+  writeIORef nextMId 0

### 2020-02-05 implicit-fun-elab/Elaboration.hs

@@ -1,7 +1,11 @@
-
 {-# options_ghc -Wno-type-defaults #-}
 
 module Elaboration where
+
+
+
+{-
+
 
 import Control.Monad
 import Control.Monad.Except
@@ -685,3 +689,4 @@ infer ins t = case t of
     nty <- quoteM (snd res)
     debugM ("inferred", fst res, nty)
     pure res
+-}

### 2020-02-05 implicit-fun-elab/Evaluation.hs

@@ -1,299 +1,185 @@
 
-module Evaluation (
-    apply
-  , applyM
-  , close
-  , closeM
-  , eval
-  , evalM
-  , force
-  , forceM
-  , forceSp
-  , forceSpM
-  , fresh
-  , hlamM
-  , nf
-  , nfM
-  , quote
-  , quoteM
-  , runEval
-  , vApp
-  , vAppM
-  , zonk
-  , zonkM
-  ) where
-
-import Control.Monad.Reader
-import Lens.Micro.Platform
-import qualified Data.IntMap.Strict as M
+module Evaluation where
 
 import Types
+import ElabState
 
---------------------------------------------------------------------------------
-
-runEval :: EvalM a -> MCxt -> Vals -> a
-runEval ma ms vs = runReader ma (EvalEnv ms vs)
-
-embedEvalM :: HasVals cxt Vals => EvalM a -> M cxt a
-embedEvalM ma = runEval ma <$> use mcxt <*> view vals
-
-vProj1 :: Val -> EvalM Val
-vProj1 (VTcons t u)    = pure t
-vProj1 (VNe h sp)      = pure $ VNe h (SProj1 sp)
-vProj1 (VLamTel x a t) = vProj1 =<< vLamTel pure x a t
+vProj1 :: Val -> Val
+vProj1 (VTcons t u)    = t
+vProj1 (VNe h sp)      = VNe h (SProj1 sp)
+vProj1 (VLamTel x a t) = vProj1 (vLamTel id x a t)
 vProj1 _               = error "impossible"
 
-vProj2 :: Val -> EvalM Val
-vProj2 (VTcons t u)    = pure u
-vProj2 (VNe h sp)      = pure $ VNe h (SProj2 sp)
-vProj2 (VLamTel x a t) = vProj2 =<< vLamTel pure x a t
+vProj2 :: Val -> Val
+vProj2 (VTcons t u)    = u
+vProj2 (VNe h sp)      = VNe h (SProj2 sp)
+vProj2 (VLamTel x a t) = vProj2 (vLamTel id x a t)
 vProj2 _               = error "impossible"
 
-vVar :: Name -> EvalM Val
-vVar x = do
-  vs <- view vals
-  case (lookup x vs) of
-    Nothing -> pure (VVar x) -- error $ "impossible: " ++ x ++ " not in scope"
-    Just mv -> pure $ maybe (VVar x) id mv
+vVar :: Ix -> Vals -> Val
+vVar 0 (VDef vs v) = v
[…]

### 2020-02-05 implicit-fun-elab/Parser.hs

@@ -50,21 +50,17 @@ pAtom  =
                <|> (RHole <$ char '_'))
   <|> parens pTm
 
-pArg :: Parser (Maybe (Either Name Icit, Raw))
-pArg = (Nothing <$ char '!')
-  <|>  (Just <$> (
-           (try $ braces $ do {x <- pIdent; char '='; t <- pTm; pure (Left x, t)})
-       <|> ((Right Impl,) <$> (char '{' *> pTm <* char '}'))
-       <|> ((Right Expl,) <$> pAtom)))
+pArg :: Parser (Either Name Icit, Raw)
+pArg =
+      (try $ braces $ do {x <- pIdent; char '='; t <- pTm; pure (Left x, t)})
+  <|> ((Right Impl,) <$> (char '{' *> pTm <* char '}'))
+  <|> ((Right Expl,) <$> pAtom)
 
 pSpine :: Parser Raw
 pSpine = do
   h <- pAtom
   args <- many pArg
-  pure $ foldl
-    (\t -> \case Nothing     -> RStopInsertion t
-                 Just (i, u) -> RApp t u i)
-    h args
+  pure $ foldl (\t (i, u) -> RApp t u i) h args
 
 pLamBinder :: Parser (Name, Maybe Raw, Either Name Icit)
 pLamBinder =
@@ -112,17 +108,8 @@ pLet = do
   u <- pTm
   pure $ RLet x (maybe RHole id ann) t u
 
-pAssume :: Parser Raw
-pAssume = do
-  symbol "assume"
-  x <- pIdent
-  ann <- char ':' *> pTm
-  symbol "in"
-  u <- pTm
-  pure $ RAssume x ann u
-
 pTm :: Parser Raw
-pTm = withPos (pLam <|> pLet <|> pAssume <|> try pPi <|> pFunOrSpine)
+pTm = withPos (pLam <|> pLet <|> try pPi <|> pFunOrSpine)
 
 pSrc :: Parser Raw
 pSrc = ws *> pTm <* eof

### 2020-02-05 implicit-fun-elab/Pretty.hs

@@ -1,0 +1,139 @@
+
+
+module Pretty where
+
+import Data.List (intercalate)
+import Text.Printf
+
+import Types
+
+-- | Assumption: the `[Name]` input does not have shadowing.
+prettyTm :: Int -> [Name] -> Tm -> ShowS
+prettyTm prec = go (prec /= 0) where
+
+  fresh :: [Name] -> Name -> Name
+  fresh ns n | elem n ns = fresh ns (n++"'")
+             | otherwise = n
+
+  goVar :: [Name] -> Ix -> ShowS
+  goVar ns topX = go ns topX where
+    go []     _ = (show topX++)
+    go (n:ns) 0 = (n++)
+    go (n:ns) x = go ns (x - 1)
+
+  goArg :: [Name] -> Tm -> Icit -> ShowS
+  goArg ns t i = icit i (bracket (go False ns t)) (go True ns t)
+
+  goLamBind :: Name -> Icit -> ShowS
+  goLamBind x i = icit i bracket id ((if null x then "_" else x) ++)
+
+  bracket :: ShowS -> ShowS
+  bracket s = ('{':).s.('}':)
+
+  goLam :: [Name] -> Tm -> ShowS
+  goLam ns (Lam (fresh ns -> x) i a t)  = (' ':) . goLamBind x i . goLam (x:ns) t
+  goLam ns (LamTel(fresh ns -> x) a t) =
+    (' ':) . bracket ((x++) . (" : "++) . go False ns a) . goLam (x:ns) t
+  goLam ns t = (". "++) . go False ns t
+
+  goPiBind :: [Name] -> Name -> Icit -> Tm -> ShowS
+  goPiBind ns x i a =
+    icit i bracket (showParen True) ((x++) . (" : "++) . go False ns a)
+
+  goPi :: [Name] -> Bool -> Tm -> ShowS
+  goPi ns p (Pi (fresh ns -> x) i a b)
+    | x /= "_" = goPiBind ns x i a . goPi ns True b
+    | otherwise =
+       (if p then (" → "++) else id) .
+       go (case a of App{} -> False; AppTel{} -> False; _ -> True) ns a .
+       (" → "++) . go False (x:ns) b
+
+  goPi ns p (PiTel (fresh ns -> x) a b)
+    | x /= "_" = goPiBind ns x Impl a . goPi ns True b
+    | otherwise =
+       (if p then (" → "++) else id) .
[…]

### 2020-02-05 implicit-fun-elab/Types.hs

@@ -4,20 +4,15 @@ module Types (
   module Text.Megaparsec
   ) where
 
-import Control.Monad.Except
 import Control.Monad.Reader
-import Control.Monad.State.Strict
-import Data.String
-import Data.List
 import Lens.Micro.Platform
 import Text.Megaparsec (SourcePos(..), unPos, initialPos)
-import Text.Printf
 
-import qualified Data.IntMap.Strict as M
+import qualified Data.IntMap.Strict as IM
 import qualified Data.IntSet        as IS
+import qualified Data.Map.Strict    as M
 
-import Debug.Trace
-
+-- Raw syntax
 --------------------------------------------------------------------------------
 
 type Name = String
@@ -35,82 +30,58 @@ data Raw
   | RPi Name Icit Raw Raw
   | RLet Name Raw Raw Raw
   | RHole
-  | RStopInsertion Raw
   | RSrcPos SourcePos Raw
 
--- deriving instance Show Raw
-instance Show Raw where
-  show = show . go where
-    go :: Raw -> Tm
-    go = \case
-      RVar x               -> Var x
-      RLam x _ (Left y) t  -> Lam x Impl (go t)
-      RLam x _ (Right i) t -> Lam x i (go t)
-      RApp t u (Left x)    -> App (go t) (go u) Impl
-      RApp t u (Right i)   -> App (go t) (go u) i
-      RU                   -> U
-      RPi x i a b          -> Pi x i (go a) (go b)
-      RLet x a t u         -> Let x (go a) (go t) (go u)
-      RHole                -> Var "_"
-      RStopInsertion t     -> App (go t) (Var "!") Expl
-      RSrcPos _ t          -> go t
+deriving instance Show Raw
 
+-- Types
 --------------------------------------------------------------------------------
 
 -- | Elaboration problem identifier.
 type MId = Int
 
--- | List of blocked problems.
-type Blocking = IS.IntSet
+-- | Blocked problems.
+type Blocking  = IS.IntSet
 type BlockedBy = IS.IntSet
 
 data MetaEntry
-  = Unsolved Blocking
+  = Unsolved Blocking ~VTy
   | Solved Val
 
[…]

### 2020-02-05 paper.tex
tidy code examples

@@
 
-%% \usepackage{fontspec}
-%% \fontspec[
-%%   Path = fonts/,
-%%   BoldFont = *-Bold,
-%%   ItalicFont = *-Oblique,
-%%   BoldItalicFont = *-BoldOblique]
-%%   {DejaVuSansMono}
-\setmonofont[Colour=blue,Scale=0.8]{DejaVuSansMono}
-\usepackage{fancyvrb}
 \usepackage{xcolor}
@@
 \usepackage{scalerel}
-
-%% \newcommand{\blue}[1]{\textcolor{blue}{#1}}
-%% \newcommand{\red}[1]{\textcolor{red}{#1}}
-%% \newcommand{\green}[1]{\textcolor{ForestGreen}{#1}}
-%% \newcommand{\purple}[1]{\textcolor{purple}{#1}}
-%% \newcommand{\ASet}{\blue{Set}}
-%% \newcommand{\CDefinition}{\blue{Definition}}
-%% \newcommand{\CType}{\blue{Type}}
-%% \newcommand{\Hforall}{\blue{forall}}
-%% \newcommand{\Cforall}{\blue{forall}}
-%% \newcommand{\HStar}{\blue{*}}
+\usepackage{cleveref}
 
@@
 \newcommand{\fail}{\mathsf{fail}}
+\newcommand{\kw}[1]{{\mathsf{#1}}}
 
@@
 \newcommand{\einsert}[5]{\mathsf{insert}\,#1\,#2\,#3\,#4\,#5}
+\newcommand{\Set}{\mathsf{Set}}
 
@@
 \emph{first-class} if they are treated as any other type in the surface
-language. This is true in Agda and partially true in GHC. Inference and
+language. This holds in Agda and partially holds in GHC. Inference and
 elaboration in the presence of first-class implicit functions poses a challenge;
@@
 Agda and Coq, one can use bracketed function domains for this purpose:
-\begin{Verbatim}[commandchars=\\\{\}]
-          -- Agda                     (* Coq *)
-          id : \{A : Set\} → A → A      Definition id \{A : Type\}(x : A) := x.
-          id x = x
-\end{Verbatim}
-In GHC, one can use \texttt{forall} to define implicit function
+\begin{alignat*}{3}
+  & id : \{A : \kw{Set}\}\to A \to A \hspace{5em} \kw{Definition}\,id\,\{A : \kw{Type}\}(x : A) := x.\\
+  & id\,x = x
+\end{alignat*}
+In GHC, one can use $\kw{forall}$ to define implicit function
 types\footnote{This notation requires language extensions
-  \texttt{KindSignatures} and \texttt{RankNTypes}; one could also write the type
-  \texttt{a -> a} and GHC would silently insert the quantification.}
-\begin{Verbatim}[commandchars=\\\{\}]
-                         id :: forall (a :: *). a -> a
-                         id x = x
-\end{Verbatim}
-
-In all of the above cases, if we apply \texttt{id} to an argument, the implicit
-type argument is provided by elaboration. For example, in Agda, \texttt{id true}
-is elaborated to \texttt{id \{Bool\} true}, and analogously in GHC and Coq. In all
-three systems, there is also a way to explicitly specify implicit arguments: in
-Agda we may put arguments in brackets as we have seen, in Coq we can prefix a
-name with \texttt{@} to make every implicit argument explicit, as in \texttt{@id
-  bool true}, and in GHC we can enable the language extension
-\texttt{TypeApplications} and write \texttt{id @Bool True}.
+  $\mathsf{KindSignatures}$ and $\mathsf{RankNTypes}$; one could also write the type
+  $a \to a$ and GHC would silently insert the quantification.}
+\begin{alignat*}{3}
+  & id :: \kw{forall}\,(a :: \kw{*}).\,a \to a\\
+  & id\,x = x
+\end{alignat*}
+In all of the above cases, if we apply $id$ to an argument, the implicit type
+argument is provided by elaboration. For example, in Agda, $id\,true$ is
+elaborated to $id\,Bool\,true$, and analogously in GHC and Coq. In all three
+systems, there is also a way to explicitly specify implicit arguments: in Agda
+we may put arguments in brackets as we have seen, in Coq we can prefix a name
+with $\kw{@}$ to make every implicit argument explicit, as in
+$\kw{@}id\,bool\,true$, and in GHC we can enable the language extension
+$\mathsf{TypeApplications}$ and write $id\,\kw{@}Bool\,True$.
 
@@
 instead, implicitness is tied to particular \emph{names}, and while we can write
-\texttt{List (forall \{A : Type\}, A -> A)}, the brackets here are simply
-ignored by Coq and we get a plain function type. For example, Coq accepts the
-following definition:
-\begin{Verbatim}[commandchars=\\\{\}]
-      Definition poly : forall (f : forall \{A : Type\}, A -> A), bool * nat :=
-        fun f => (f bool true, f nat 0).
-\end{Verbatim}
-This is a higher-rank polymorphic function which returns a pair. Note that
-\texttt{f} is applied to two arguments, because the implicitness in
-\texttt{forall \{A : Type\}, A -> A} is silently dropped.
-
-In GHC Haskell, \texttt{forall} types are more flexible. We can write
-the following, with \texttt{RankNTypes} enabled:
-\begin{Verbatim}[commandchars=\\\{\}]
-                     poly :: (forall a. a -> a) -> (Bool, Int)
-                     poly f = (f True, f 0)
-\end{Verbatim}
+$list\,(\kw{forall}\,\{A : \kw{Type}\},\,A\to A)$ for a list type of polymorphic
+functions, the brackets here are simply ignored by Coq and we get a plain
+function type. For example, Coq accepts the following definition:
+\begin{alignat*}{3}
+  & \kw{Definition}\,poly\,: \kw{forall}\,(f : \kw{forall}\,\{A : \kw{Type}\},\,A\to A),\,bool\,*\,nat :=\\
+  & \quad \kw{fun}\,f \Rightarrow (f\,bool\,true,\,f\,nat\,0)
+\end{alignat*}
+This is a higher-rank polymorphic function which returns a pair. Note that $f$
+is applied to two arguments, because the implicitness in $\kw{forall}\,\{A :
+\kw{Type}\},\,A\to A$ is silently dropped.
+
+In GHC Haskell, $\kw{forall}$ types are more flexible. We can write the
+following, with $\kw{RankNTypes}$ enabled:
+\begin{alignat*}{3}
+  & poly :: (\kw{forall}\,a.\,a \to a) \to (Bool,\,Int) \\
+  & poly\,f = (f\,True,\,f\,0)
+\end{alignat*}
 However, polymorphic types are only supported in function domains and as fields
 of algebraic data constructors. We cannot instantiate an arbitrary type
-parameter to a \texttt{forall}, as in \texttt{[forall a. a -> a]} for a list
-type with polymorphic elements. While this type is technically allowed by the
-\texttt{ImpredicativeTypes} language extension, as of GHC 8.8 this extension is
+parameter to a $\kw{forall}$, as in $[\kw{forall}\,a.\,a\to a]$ for a list type
+with polymorphic elements. While this type is technically allowed by the
+$\kw{ImpredicativeTypes}$ language extension, as of GHC 8.8 this extension is
 deprecated and is not particularly usable in practice.
@@
 In Agda, implicit functions are truly a first-class notion, and we may have
-\texttt{List (\{A : Set\} → A → A)} without issue. However, Agda's elaboration
+$List\,(\{A : \kw{Set}\}\to A \to A)$ without issue. However, Agda's elaboration
 still has limitations when it comes to handling implicit functions. Assume that
-we have \texttt{[]} for the empty list and \texttt{\_∷\_} for list extension,
+we have $[]$ for the empty list and $\blank::\blank$ for list extension,
 and consider the following code:
-\begin{Verbatim}[commandchars=\\\{\}]
-                     polyList : List (\{A : Set\} → A → A)
-                     polyList = (λ x → x) ∷ []
-\end{Verbatim}
-Agda 2.6.0.1 does not accept this. However, it does accept \texttt{polyList = (λ
-  \{A\} x → x) ∷ []}. The issue is the following. Agda first \emph{infers} a
-type for \texttt{(λ x → x) ∷ []}, then tries to unify the inferred type with the
-given \texttt{List (\{A : Set\} → A → A)} annotation. However, when Agda
-elaborates \texttt{λ x → x}, it does not yet know anything about the element
-type of the list; it is an undetermined unification variable. Hence, Agda does
-not know whether it should insert an extra \texttt{λ\{A\}} or not. If the
-element type is later found to be an implicit function, then it should, otherwise it
-should not. To solve this conundrum, Agda simply assumes that any unknown type
-is \emph{not} an implicit function type, and elects to not insert a lambda. This
-assumption is often correct, but sometimes --- as in the current case --- it is not.
+\begin{alignat*}{3}
+  & polyList : List\,(\{A : \kw{Set}\}\to A \to A)\\
+  & polyList = (\lambda\,x \to x)\,::\,[]
+\end{alignat*}
+Agda 2.6.0.1 does not accept this. However, it does accept $polyList =
+(\lambda\,\{A\}\,x \to x)\,::\,[]$. The issue is the following. Agda first
+\emph{infers} a type for $(\lambda\,x \to x)\,::\,[]$, then tries to unify the
+inferred type with the given $List\,(\{A : \kw{Set}\}\to A \to A)$
+annotation. However, when Agda elaborates $\lambda\,x \to x$, it does not yet
+know anything about the element type of the list; it is an undetermined
+unification variable. Hence, Agda does not know whether it should insert an
+extra $\lambda\,\{A\}$ or not. If the element type is later found to be an
+implicit function, then it should, otherwise it should not. To solve this
+conundrum, Agda simply assumes that any unknown type is \emph{not} an implicit
+function type, and elects to not insert a lambda. This assumption is often
+correct, but sometimes --- as in the current case --- it is not.
 
@@
 and performed at any time after a particular expression is elaborated. In the
-current example, our algorithm wraps \texttt{λ x → x} in an implicit lambda with
-unknown arity, and the domain type is later refined to be \texttt{A : Set} when
-the inferred type is unified with the annotation.
+current example, our algorithm wraps $\lambda\,x\to x$ in an implicit lambda
+with unknown arity, whose domain is later refined to be $A : \kw{Set}$ when the
+inferred type is unified with the annotation.
 
@@
 
-\subsubsection{Note on Terminology}
+\subsubsection{Note on terminology}
 
@@
     types.
-  \item Coq has type-theory-impredicative \texttt{Prop} universe (and
-    optionally also \texttt{Set}), but implements
-    type-inference-predicative elaboration, because of the lack of implicit
-    function types.
-  \item GHC is type-theory-impredicative with \texttt{RankNTypes} enabled
-    and \texttt{ImpredicativeTypes} disabled, as we have
-    \texttt{(forall (a :: *). a -> a) :: *}.
+  \item Coq has type-theory-impredicative $\kw{Prop}$ universe (and optionally
+    also $\kw{Set}$), but implements type-inference-predicative elaboration,
+    because of the lack of implicit function types.
+  \item GHC is type-theory-impredicative with $\kw{RankNTypes}$ enabled and
+    $\kw{ImpredicativeTypes}$ disabled, as we have $(\kw{forall}\,(a :: *).\, a
+    \to a) :: *$.
 \end{itemize}
@@
 In the core syntax, all applications must be explicit, so using implicit
-function type for $map$ above does not make much difference in optics.
+function type for $map$ above does not make much difference in notation.
 \end{example}
@@
 toggles whether any insertion is to be performed. We will use this in the
-definition of elaboration. Insertion is defined by recursion on the $A$ core
-type, as follows.
+definition of elaboration. Insertion is defined by recursion on $ins$ and the
+$A$ type, as follows.
 \begin{alignat*}{3}
@@
 
-We assume that there is a partial unification procedure, which returns a unifying
-metasubstitution on success. We only have \emph{homogeneous} unification, i.e.\ the
-two terms to be unified must have the same type. The specification is as follows:
+We assume that there is a unification procedure, which returns a unifying
+metasubstitution on success. We only have \emph{homogeneous} unification,
+i.e.\ the two terms to be unified must have the same type. The specification is
+as follows:
 \begin{mathpar}
@@
 
-\todo{example for meta solution, more references}
+\todo{example for meta solution, more references, describe at least pattern unif and pruning briefly}
 
@@
 \subsubsection{On soundness, completeness, and other properties}
-We briefly discuss what we can and cannot say about the currently presented
+We briefly discuss what we can and cannot say about the presented
 elaborator. First, elaboration is \emph{sound} in the sense that it never
@@
 
-We can still say something about what kind of work our elaborator does. For
-this, we consider a translation from core terms to surface terms, the obvious
-``forgetful'' translation, which maps terms to surface counterparts. Now, this
-is an ``evil'' construction on core terms, since it does not preserve
-definitional equality. We shall only use this evil notion in the following
-statement.
+We can still say something about the behavior of our elaborator. For this, we
+consider a translation from core terms to surface terms, the obvious forgetful
+translation, which maps core terms to surface counterparts. Now, this is an
+``evil'' construction on core terms, since it does not preserve definitional
+equality. We shall only use this evil notion in the following statement.
 
@@
 \end{theorem}
+
+\todo{Say something about termination of elab?}
 

### 2020-02-05 implicit-fun-elab/Elaboration.hs

@@ -2,7 +2,7 @@
 
 module Elaboration where
 
-
+-- TODODODOO
 
 {-
 

### 2020-02-07 paper.tex
stuff

@@
 We prefer to avoid the term ``impredicative inference'' in order to avoid
-confusion with impredicativity in type theory. The two notions historically
-coincided, but has since diverged to the point of being orthogonal. In type
+confusion with impredicativity in type theory. The two notions sometimes
+coincided historically, but currently they are largely orthogonal. In type
 theory, impredicativity is a property of a universe, i.e.\ closure of a universe
@@
   \item GHC is type-theory-impredicative with $\kw{RankNTypes}$ enabled and
-    $\kw{ImpredicativeTypes}$ disabled, as we have $(\kw{forall}\,(a :: *).\, a
+    $\kw{ImpredicativeTypes}$ \emph{disabled}, as we have $(\kw{forall}\,(a :: *).\, a
     \to a) :: *$.
@@
 
-\emph{Metacontexts} are used to record metavariables which are created during
+Metacontexts are used to record metavariables which are created during
 elaboration. In our case, metacontexts are simply a context prefix, and we have
-variables pointing into it, but we do not support contextual modality e.g.\ as
-in \cite{nanevski2008contextual}. The non-meta typing context additionally
-supports \emph{defined variables}, which is used in the typing rule for
-\emph{let-definitions}, and we have that any defined variable is equal to its
-definition. We mainly support this as a convenience feature in the
+variables pointing into it, but we do not support any contextual modality
+e.g.\ as in \cite{nanevski2008contextual}. The non-meta typing context
+additionally supports \emph{defined variables}, which is used in the typing rule
+for \emph{let-definitions}, and we have that any defined variable is equal to
+its definition. We mainly support this as a convenience feature in the
 implementation of our elaborator.
 
-\emph{The universe} $\U$ is Russell-style, and we have the type-in-type rule.
+The universe $\U$ is Russell-style, and we have the type-in-type rule.
 This causes our core syntax to be non-total, and our elaboration algorithm to be
@@
 
-\emph{Function types} only differ from each other in notation: implicit
+Function types only differ from each other in notation: implicit
 functions have the same rules as ``explicit'' functions. The primary purpose of
@@
 
-\begin{notation}[Spines]\label{not:spines}
+\begin{notation}\label{not:spines}
   We sometimes use a spine notation for neutral terms. A spine is a list of
@@
 \end{alignat*}
-In the core syntax, all applications must be explicit, so using implicit
-function type for $map$ above does not make much difference in notation.
+In the core syntax, all applications must be explicit. In the surface syntax we
+will be able to omit implicit arguments to $map$.
+
 \end{example}

### 2020-02-07 implicit-fun-elab/Elaboration.hs

@@ -1,51 +1,21 @@
 {-# options_ghc -Wno-type-defaults #-}
 
+-- 1. implement constancy, pattern unify, but no pruning
+-- 2. implement pruning
+-- 3. see if we can merge stuff
+
 module Elaboration where
 
--- TODODODOO
-
-{-
-
-
-import Control.Monad
-import Control.Monad.Except
-import Control.Monad.Reader
-import Control.Monad.State.Strict
-import Data.Foldable
-import Lens.Micro.Platform
-
-import qualified Data.IntMap.Strict as M
-import qualified Data.IntSet        as IS
-import qualified Data.Set           as S
+import qualified Data.IntSet as IS
+import qualified Data.IntMap as IM
+import Data.IORef
 
 import Types
 import Evaluation
+import ElabState
 
-
--- Freshness
---------------------------------------------------------------------------------
-
-data Fresh = Fresh
-  (forall a. Name -> (Name, UnifyM a -> UnifyM a))
-  (forall a. Name -> Val -> (Name, UnifyM a -> UnifyM a))
-
-bindU :: Name -> UnifyCxt -> UnifyCxt
-bindU x (UnifyCxt vs is p) =
-  UnifyCxt ((x, Nothing):vs) ((x, Bound BVSrc):is) p
-
-telBindU :: Name -> VTy -> UnifyCxt -> UnifyCxt
-telBindU x ~a (UnifyCxt vs is p) =
-  UnifyCxt ((x, Nothing):vs) ((x, Bound (BVTel a)):is) p
-
-localFresh :: UnifyM Fresh
-localFresh = do
-  vs <- view vals
-  pure $ Fresh
-    (\x -> let x' = fresh vs x in (x', local (bindU x')))
-    (\x a -> let x' = fresh vs x in (x', local (telBindU x' a)))
-
--- Telscope constancy
---------------------------------------------------------------------------------
+-- occurrence check for the purpose of constancy
+------------------------------------------------------------
 
 data Occurs = Rigid | Flex IS.IntSet | None deriving (Eq, Show)
 
@@ -63,630 +33,168 @@ occurrence ms | IS.null  ms = Rigid
 instance Monoid Occurs where
   mempty = None
 
-occurs :: Name -> Val -> UnifyM Occurs
-occurs topX = go mempty where
+occurs :: Lvl -> Lvl -> Val -> Occurs
[…]

### 2020-02-07 implicit-fun-elab/Evaluation.hs

@@ -93,49 +93,51 @@ vAppSp h = go where
   go (SProj2 sp)      = vProj2 (go sp)
 
 eval :: Vals -> Tm -> Val
-eval vs = \case
-  Var x        -> vVar x vs
-  Let x _ t u  -> eval (VDef vs (eval vs t)) u
-  U            -> VU
-  Meta m       -> vMeta m
-  Pi x i a b   -> VPi x i (eval vs a) $ \x -> eval (VDef vs x) b
-  Lam x i a t  -> VLam x i (eval vs a) $ \x -> eval (VDef vs x) t
-  App t u i    -> vApp (eval vs t) (eval vs u) i
-  Tel          -> VTel
-  TEmpty       -> VTEmpty
-  TCons x a b  -> VTCons x (eval vs a) $ \x -> eval (VDef vs x) b
-  Rec a        -> VRec (eval vs a)
-  Tempty       -> VTempty
-  Tcons t u    -> VTcons (eval vs t) (eval vs u)
-  Proj1 t      -> vProj1 (eval vs t)
-  Proj2 t      -> vProj2 (eval vs t)
-  PiTel x a b  -> vPiTel id x (eval vs a) $ \x -> eval (VDef vs x) b
-  AppTel a t u -> vAppTel (eval vs a) (eval vs t) (eval vs u)
-  LamTel x a t -> vLamTel id x (eval vs a) $ \x -> eval (VDef vs x) t
+eval vs t = case (eval vs, \t x -> eval (VDef vs x) t) of
+  (eval, evalBind) -> case t of
+    Var x        -> vVar x vs
+    Let x _ t u  -> evalBind u (eval t)
+    U            -> VU
+    Meta m       -> vMeta m
+    Pi x i a b   -> VPi x i (eval a) (evalBind b)
+    Lam x i a t  -> VLam x i (eval a) (evalBind t)
+    App t u i    -> vApp (eval t) (eval u) i
+    Tel          -> VTel
+    TEmpty       -> VTEmpty
+    TCons x a b  -> VTCons x (eval a) (evalBind b)
+    Rec a        -> VRec (eval a)
+    Tempty       -> VTempty
+    Tcons t u    -> VTcons (eval t) (eval u)
+    Proj1 t      -> vProj1 (eval t)
+    Proj2 t      -> vProj2 (eval t)
+    PiTel x a b  -> vPiTel id x (eval a) (evalBind b)
[…]

### 2020-02-07 implicit-fun-elab/Pretty.hs

@@ -1,139 +1,0 @@
-
-
-module Pretty where
-
-import Data.List (intercalate)
-import Text.Printf
-
-import Types
-
--- | Assumption: the `[Name]` input does not have shadowing.
-prettyTm :: Int -> [Name] -> Tm -> ShowS
-prettyTm prec = go (prec /= 0) where
-
-  fresh :: [Name] -> Name -> Name
-  fresh ns n | elem n ns = fresh ns (n++"'")
-             | otherwise = n
-
-  goVar :: [Name] -> Ix -> ShowS
-  goVar ns topX = go ns topX where
-    go []     _ = (show topX++)
-    go (n:ns) 0 = (n++)
-    go (n:ns) x = go ns (x - 1)
-
-  goArg :: [Name] -> Tm -> Icit -> ShowS
-  goArg ns t i = icit i (bracket (go False ns t)) (go True ns t)
-
-  goLamBind :: Name -> Icit -> ShowS
-  goLamBind x i = icit i bracket id ((if null x then "_" else x) ++)
-
-  bracket :: ShowS -> ShowS
-  bracket s = ('{':).s.('}':)
-
-  goLam :: [Name] -> Tm -> ShowS
-  goLam ns (Lam (fresh ns -> x) i a t)  = (' ':) . goLamBind x i . goLam (x:ns) t
-  goLam ns (LamTel(fresh ns -> x) a t) =
-    (' ':) . bracket ((x++) . (" : "++) . go False ns a) . goLam (x:ns) t
-  goLam ns t = (". "++) . go False ns t
-
-  goPiBind :: [Name] -> Name -> Icit -> Tm -> ShowS
-  goPiBind ns x i a =
-    icit i bracket (showParen True) ((x++) . (" : "++) . go False ns a)
-
-  goPi :: [Name] -> Bool -> Tm -> ShowS
-  goPi ns p (Pi (fresh ns -> x) i a b)
-    | x /= "_" = goPiBind ns x i a . goPi ns True b
-    | otherwise =
-       (if p then (" → "++) else id) .
-       go (case a of App{} -> False; AppTel{} -> False; _ -> True) ns a .
-       (" → "++) . go False (x:ns) b
-
-  goPi ns p (PiTel (fresh ns -> x) a b)
-    | x /= "_" = goPiBind ns x Impl a . goPi ns True b
-    | otherwise =
-       (if p then (" → "++) else id) .
[…]

### 2020-02-07 implicit-fun-elab/Types.hs

@@ -4,13 +4,14 @@ module Types (
   module Text.Megaparsec
   ) where
 
-import Control.Monad.Reader
-import Lens.Micro.Platform
+import Control.Exception
 import Text.Megaparsec (SourcePos(..), unPos, initialPos)
+import Text.Printf
 
 import qualified Data.IntMap.Strict as IM
 import qualified Data.IntSet        as IS
 import qualified Data.Map.Strict    as M
+
 
 -- Raw syntax
 --------------------------------------------------------------------------------
@@ -63,18 +64,40 @@ type MCxt  = IM.IntMap MetaEntry
 data NameOrigin = NOSource | NOInserted
 type NameTable  = M.Map Name (NameOrigin, Lvl)
 
+lvlName :: [Name] -> Lvl -> Name
+lvlName ns x = ns !! (length ns - x - 1)
+
+typesLen :: Types -> Int
+typesLen = go 0 where
+  go acc TNil           = acc
+  go acc (TDef tys _)   = go (acc + 1) tys
+  go acc (TBound tys _) = go (acc + 1) tys
+
+ixType :: Types -> Ix -> VTy
+ixType TNil           _ = error "impossible"
+ixType (TDef   tys a) 0 = a
+ixType (TBound tys a) 0 = a
+ixType (TDef   tys a) x = ixType tys (x - 1)
+ixType (TBound tys a) x = ixType tys (x - 1)
+
+lvlType :: Types -> Lvl -> VTy
+lvlType tys x = ixType tys (typesLen tys - x - 1)
+
 data ElabCxt = ElabCxt {
   elabCxtVals      :: Vals,
   elabCxtTypes     :: Types,
+  elabCxtNames     :: [Name],
   elabCxtNameTable :: NameTable}
 
 data UnifyCxt  = UnifyCxt {
   unifyCxtVals  :: Vals,
-  unifyCxtTypes :: Types}
+  unifyCxtTypes :: Types,
+  unifyCxtNames :: [Name]}
 
-data Err    = Err {errErr :: ElabError, errPos :: Maybe SourcePos}
-type ElabM  = ReaderT ElabCxt IO
-type UnifyM = ReaderT UnifyCxt IO
+data Err = Err {
+  errNames :: [Name],
+  errErr   :: ElabError,
+  errPos   :: Maybe SourcePos}
 
 data Tm
   = Var Ix
[…]

### 2020-02-07 implicit-fun-elab/Elaboration.hs
stuff

@@ -1,8 +1,9 @@
 {-# options_ghc -Wno-type-defaults #-}
 
--- 1. implement constancy, pattern unify, but no pruning
--- 2. implement pruning
--- 3. see if we can merge stuff
+-- TODO: - implement simple elab without telescope/pruning
+--       - add pruning
+--       - add telescopes
+
 
 module Elaboration where
 
@@ -65,7 +66,9 @@ occurs d topX = occurs' d mempty where
       VPiTel x a b  -> go a <> goBind b
       VLamTel x a t -> go a <> goBind t
 
-------------------------------------------------------------
+
+-- unification
+--------------------------------------------------------------------------------
 
 -- If (Just l)   : rename to l
 --    Nothing    : nonlinear var (error)
@@ -198,3 +201,13 @@ unify ns tys d topT topT' = go topT topT' where
     (SApp sp u i, SApp sp' u' i') | i == i' -> goSp sp sp' >> go u u'
     (SAppTel a sp u, SAppTel a' sp' u')     -> go a a' >> goSp sp sp' >> go u u'
     _                                       -> report ns $ UnifyError ntopT ntopT'
+
+-- Elaboration
+--------------------------------------------------------------------------------
+
+check :: Cxt -> Raw -> VTy -> IO Tm
+check cxt t a = undefined
+
+
+infer :: Cxt -> Raw -> IO (Tm, VTy)
+infer cxt t = undefined

### 2020-02-07 implicit-fun-elab/Types.hs

@@ -7,6 +7,7 @@ module Types (
 import Control.Exception
 import Text.Megaparsec (SourcePos(..), unPos, initialPos)
 import Text.Printf
+import Lens.Micro.Platform
 
 import qualified Data.IntMap.Strict as IM
 import qualified Data.IntSet        as IS
@@ -83,16 +84,11 @@ ixType (TBound tys a) x = ixType tys (x - 1)
 lvlType :: Types -> Lvl -> VTy
 lvlType tys x = ixType tys (typesLen tys - x - 1)
 
-data ElabCxt = ElabCxt {
-  elabCxtVals      :: Vals,
-  elabCxtTypes     :: Types,
-  elabCxtNames     :: [Name],
-  elabCxtNameTable :: NameTable}
-
-data UnifyCxt  = UnifyCxt {
-  unifyCxtVals  :: Vals,
-  unifyCxtTypes :: Types,
-  unifyCxtNames :: [Name]}
+data Cxt = Cxt {
+  cxtVals      :: Vals,
+  cxtTypes     :: Types,
+  cxtNames     :: [Name],
+  cxtNameTable :: NameTable}
 
 data Err = Err {
   errNames :: [Name],
@@ -331,3 +327,9 @@ showError ns = \case
      "  %s =? %s")
     (show x)
     (showTm ns lhs) (showTm ns rhs)
+
+-- Lenses
+--------------------------------------------------------------------------------
+
+makeFields ''Cxt
+makeFields ''Err

### 2020-02-07 implicit-fun-elab/Elaboration.hs
elab cleanup

@@ -1,15 +1,18 @@
 {-# options_ghc -Wno-type-defaults #-}
 
--- TODO: - implement simple elab without telescope/pruning
+-- TODO: - Only use elabcxt and unifycxt
+--       - try to decorate errors with info when info is available (catch/rethrow)
+--       - implement simple elab without telescope/pruning
 --       - add pruning
 --       - add telescopes
-
 
 module Elaboration where
 
 import qualified Data.IntSet as IS
 import qualified Data.IntMap as IM
+import qualified Data.Map.Strict as M
 import Data.IORef
+import Lens.Micro.Platform
 
 import Types
 import Evaluation
@@ -76,24 +79,24 @@ occurs d topX = occurs' d mempty where
 type SpineRenaming = IM.IntMap (Maybe Lvl)
 
 -- | Expects a forced spine. Returns a partial renaming and the length of the spine.
-checkSp :: ([Name], Tm, Tm) -> Spine -> (SpineRenaming, Lvl)
-checkSp (report -> report, lhs, rhs) = go where
+checkSp :: UnifyCxt -> (Tm, Tm) -> Spine -> (SpineRenaming, Lvl)
+checkSp cxt (lhs, rhs) = go where
   go :: Spine -> (SpineRenaming, Lvl)
   go = \case
     SNil        -> (mempty, 0)
     SApp sp u i -> case go sp of
       (!r, !d) -> case force u of
         VVar x -> (IM.insert x (if IM.member x r then Nothing else Just d) r, d + 1)
-        _      -> report $ SpineNonVar lhs rhs
+        _      -> report (cxt^.names) $ SpineNonVar lhs rhs
     SAppTel a sp u -> case go sp of
       (!r, !d) -> case force u of
         VVar x -> (IM.insert x (if IM.member x r then Nothing else Just d) r, d + 1)
-        _    -> report $ SpineNonVar lhs rhs
-    SProj1 _ -> report SpineProjection
-    SProj2 _ -> report SpineProjection
+        _    -> report (cxt^.names) $ SpineNonVar lhs rhs
+    SProj1 _ -> report (cxt^.names) SpineProjection
+    SProj2 _ -> report (cxt^.names) SpineProjection
 
-quoteRhs :: ([Name], Tm, Tm) -> MId -> (SpineRenaming, Lvl) -> Val -> Tm
[…]

### 2020-02-07 implicit-fun-elab/Evaluation.hs

@@ -139,8 +139,8 @@ quote d v = case (quote d, \t -> quote (d + 1) (t (VVar d))) of
     VPiTel x a b  -> PiTel x (quote a) (quoteBind b)
     VLamTel x a t -> LamTel x (quote a) (quoteBind t)
 
-nf :: Vals -> Tm -> Tm
-nf vs t = quote (valsLen vs) (eval vs t)
+nf :: Vals -> Lvl -> Tm -> Tm
+nf vs d t = quote d (eval vs t)
 
 -- zonking
 --------------------------------------------------------------------------------

### 2020-02-07 implicit-fun-elab/Parser.hs

@@ -25,13 +25,13 @@ ws = L.space C.space1 (L.skipLineComment "--") (L.skipBlockComment "{-" "-}")
 withPos :: Parser Raw -> Parser Raw
 withPos p = RSrcPos <$> getSourcePos <*> p
 
-lexeme     = L.lexeme ws
-symbol s   = lexeme (C.string s)
-char c     = lexeme (C.char c)
-parens p   = char '(' *> p <* char ')'
-braces p   = char '{' *> p <* char '}'
-pArrow     = symbol "→" <|> symbol "->"
-pBind      = pIdent <|> symbol "_"
+lexeme   = L.lexeme ws
+symbol s = lexeme (C.string s)
+char c   = lexeme (C.char c)
+parens p = char '(' *> p <* char ')'
+braces p = char '{' *> p <* char '}'
+pArrow   = symbol "→" <|> symbol "->"
+pBind    = pIdent <|> symbol "_"
 
 keyword :: String -> Bool
 keyword x =
@@ -50,11 +50,10 @@ pAtom  =
                <|> (RHole <$ char '_'))
   <|> parens pTm
 
-pArg :: Parser (Either Name Icit, Raw)
+pArg :: Parser (Icit, Raw)
 pArg =
-      (try $ braces $ do {x <- pIdent; char '='; t <- pTm; pure (Left x, t)})
-  <|> ((Right Impl,) <$> (char '{' *> pTm <* char '}'))
-  <|> ((Right Expl,) <$> pAtom)
+      ((Impl,) <$> (char '{' *> pTm <* char '}'))
+  <|> ((Expl,) <$> pAtom)
 
 pSpine :: Parser Raw
 pSpine = do
@@ -62,12 +61,11 @@ pSpine = do
   args <- many pArg
   pure $ foldl (\t (i, u) -> RApp t u i) h args
 
-pLamBinder :: Parser (Name, Maybe Raw, Either Name Icit)
+pLamBinder :: Parser (Name, Maybe Raw, Icit)
 pLamBinder =
-      ((,Nothing,Right Expl) <$> pBind)
-  <|> parens ((,,Right Expl) <$> pBind <*> optional (char ':' *> pTm))
-  <|> try (braces ((,,Right Impl) <$> pBind <*> optional (char ':' *> pTm)))
-  <|> braces (do {x <- pIdent; char '='; y <- pBind; pure (y, Nothing, Left x)})
[…]

### 2020-02-07 implicit-fun-elab/Types.hs

@@ -26,8 +26,8 @@ icit Expl i e = e
 
 data Raw
   = RVar Name
-  | RLam Name (Maybe Raw) (Either Name Icit) Raw
-  | RApp Raw Raw (Either Name Icit)
+  | RLam Name (Maybe Raw) Icit Raw
+  | RApp Raw Raw Icit
   | RU
   | RPi Name Icit Raw Raw
   | RLet Name Raw Raw Raw
@@ -68,11 +68,12 @@ type NameTable  = M.Map Name (NameOrigin, Lvl)
 lvlName :: [Name] -> Lvl -> Name
 lvlName ns x = ns !! (length ns - x - 1)
 
-typesLen :: Types -> Int
-typesLen = go 0 where
-  go acc TNil           = acc
-  go acc (TDef tys _)   = go (acc + 1) tys
-  go acc (TBound tys _) = go (acc + 1) tys
+-- -- clean these up
+-- typesLen :: Types -> Int
+-- typesLen = go 0 where
+--   go acc TNil           = acc
+--   go acc (TDef tys _)   = go (acc + 1) tys
+--   go acc (TBound tys _) = go (acc + 1) tys
 
 ixType :: Types -> Ix -> VTy
 ixType TNil           _ = error "impossible"
@@ -81,14 +82,20 @@ ixType (TBound tys a) 0 = a
 ixType (TDef   tys a) x = ixType tys (x - 1)
 ixType (TBound tys a) x = ixType tys (x - 1)
 
-lvlType :: Types -> Lvl -> VTy
-lvlType tys x = ixType tys (typesLen tys - x - 1)
+-- lvlType :: Types -> Lvl -> VTy
+-- lvlType tys x = ixType tys (typesLen tys - x - 1)
 
 data Cxt = Cxt {
   cxtVals      :: Vals,
   cxtTypes     :: Types,
   cxtNames     :: [Name],
-  cxtNameTable :: NameTable}
+  cxtNameTable :: NameTable,
+  cxtLen       :: Int}
+
+data UnifyCxt = UCxt {
+  unifyCxtTypes :: Types,
+  unifyCxtNames :: [Name],
+  unifyCxtLen   :: Int }
 
 data Err = Err {
   errNames :: [Name],
@@ -134,13 +141,13 @@ valsLen = go 0 where
   go acc (VDef vs _) = go (acc + 1) vs
   go acc (VSkip vs)  = go (acc + 1) vs
 
-spLen :: Spine -> Int
-spLen = go 0 where
-  go n SNil             = n
[…]

### 2020-02-14 implicit-fun-elab/Elaboration.hs
refactor strengthening, errors

@@ -84,113 +84,103 @@ type Renaming = IM.IntMap Lvl
 -- unification
 --------------------------------------------------------------------------------
 
--- | Expects a forced spine. Returns a partial renaming and the length of the spine.
---   Nonlinear spine is always an error.
-checkSp :: UnifyCxt -> (Tm, Tm) -> Spine -> (Renaming, Lvl)
-checkSp cxt (lhs, rhs) = go where
+-- | Returns a partial renaming and the length of the spine.
+--   Nonlinear spine is always an error. Throws SpineError.
+checkSp :: Spine -> (Renaming, Lvl)
+checkSp = go . forceSp where
   go :: Spine -> (Renaming, Lvl)
   go = \case
     SNil        -> (mempty, 0)
     SApp sp u i -> case go sp of
       (!r, !d) -> case force u of
-        VVar x | IM.member x r -> report (cxt^.names) $ NonLinearSpine lhs rhs x
+        VVar x | IM.member x r -> throw $ NonLinearSpine x
                | otherwise     -> (IM.insert x d r, d + 1)
-        _      -> report (cxt^.names) $ SpineNonVar lhs rhs
+        _      -> throw SpineNonVar
     SAppTel a sp u -> case go sp of
       (!r, !d) -> case force u of
-        VVar x | IM.member x r -> report (cxt^.names) $ NonLinearSpine lhs rhs x
+        VVar x | IM.member x r -> throw $ NonLinearSpine x
                | otherwise     -> (IM.insert x d r, d + 1)
-        _    -> report (cxt^.names) $ SpineNonVar lhs rhs
-    SProj1 _ -> report (cxt^.names) SpineProjection
-    SProj2 _ -> report (cxt^.names) SpineProjection
+        _    -> throw SpineNonVar
+    SProj1 _ -> throw SpineProjection
+    SProj2 _ -> throw SpineProjection
 
 
-quoteRhs :: UnifyCxt -> (Tm, Tm) -> MId -> (Renaming, Lvl) -> Val -> Tm
-quoteRhs cxt (topLhs, topRhs) topMeta (r, splen) = go (cxt^.len) splen r where
+data Str = Str {
+  _strDom :: Lvl,           -- ^ size of renaming domain
+  _strCod :: Lvl,           -- ^ size of renaming codomain
+  _strRen :: IM.IntMap Lvl, -- ^ partial renaming
[…]

### 2020-02-14 implicit-fun-elab/Evaluation.hs

@@ -93,94 +93,103 @@ vAppSp h = go where
   go (SProj2 sp)      = vProj2 (go sp)
 
 eval :: Vals -> Tm -> Val
-eval vs t = case (eval vs, \t x -> eval (VDef vs x) t) of
-  (eval, evalBind) -> case t of
+eval vs = go where
+  go = \case
     Var x        -> vVar x vs
-    Let x _ t u  -> evalBind u (eval t)
+    Let x _ t u  -> goBind u (go t)
     U            -> VU
     Meta m       -> vMeta m
-    Pi x i a b   -> VPi x i (eval a) (evalBind b)
-    Lam x i a t  -> VLam x i (eval a) (evalBind t)
-    App t u i    -> vApp (eval t) (eval u) i
+    Pi x i a b   -> VPi x i (go a) (goBind b)
+    Lam x i a t  -> VLam x i (go a) (goBind t)
+    App t u i    -> vApp (go t) (go u) i
     Tel          -> VTel
     TEmpty       -> VTEmpty
-    TCons x a b  -> VTCons x (eval a) (evalBind b)
-    Rec a        -> VRec (eval a)
+    TCons x a b  -> VTCons x (go a) (goBind b)
+    Rec a        -> VRec (go a)
     Tempty       -> VTempty
-    Tcons t u    -> VTcons (eval t) (eval u)
-    Proj1 t      -> vProj1 (eval t)
-    Proj2 t      -> vProj2 (eval t)
-    PiTel x a b  -> vPiTel id x (eval a) (evalBind b)
-    AppTel a t u -> vAppTel (eval a) (eval t) (eval u)
-    LamTel x a t -> vLamTel id x (eval a) (evalBind t)
+    Tcons t u    -> VTcons (go t) (go u)
+    Proj1 t      -> vProj1 (go t)
+    Proj2 t      -> vProj2 (go t)
+    PiTel x a b  -> vPiTel id x (go a) (goBind b)
+    AppTel a t u -> vAppTel (go a) (go t) (go u)
+    LamTel x a t -> vLamTel id x (go a) (goBind t)
+
+  goBind t x = eval (VDef vs x) t
+
 
 quote :: Lvl -> Val -> Tm
-quote d v = case (quote d, \t -> quote (d + 1) (t (VVar d))) of
-  (quote, quoteBind) -> case force v of
+quote d = go where
+
+  go v = case force v of
     VNe h sp ->
-      let go SNil = case h of
[…]

### 2020-02-14 implicit-fun-elab/Main.hs

@@ -77,12 +77,10 @@ test = main' "elab" $ unlines [
   "let zero : Nat = λ _ s z. z in",
   "let inc : Nat → Nat = λ n _ s z. s (n _ s z) in",
   "let mul : Nat -> Nat -> Nat",
-  "    = \\a b N s z. a _ (b _ s) z in",
+  "    = \\a b N s z. a _ (b N s) z in",
   "let ten : Nat",
   "    = \\N s z. s (s (s (s (s (s (s (s (s (s z))))))))) in",
   "let hundred = mul ten ten in",
-
-
 
   "let List : U -> U",
   "    = \\A. (L : U) -> (A -> L -> L) -> L -> L in",

### 2020-02-14 implicit-fun-elab/Types.hs

@@ -107,11 +107,6 @@ data UnifyCxt = UCxt {
   unifyCxtNames :: [Name],
   unifyCxtLen   :: Int }
 
-data Err = Err {
-  errNames :: [Name],
-  errErr   :: ElabError,
-  errPos   :: Maybe SPos}
-
 data Tm
   = Var Ix
   | Let Name Ty Tm Tm
@@ -189,17 +184,33 @@ pattern VVar x = VNe (HVar x) SNil
 pattern VMeta :: MId -> Val
 pattern VMeta m = VNe (HMeta m) SNil
 
-data ElabError
-  = SpineNonVar Tm Tm                    -- ^ lhs, rhs
+data SpineError
+  = SpineNonVar
   | SpineProjection
-  | ScopeError Tm Tm Lvl                 -- ^ Lhs, rhs, offending variable
-  | OccursCheck Tm Tm                    -- ^ Lhs, rhs
-  | UnifyError Tm Tm                     -- ^ Lhs, rhs
-  | UnifyErrorWhile Tm Tm Tm Tm
+  | NonLinearSpine Lvl
+  deriving (Show, Exception)
+
+data StrengtheningError
+  = ScopeError Lvl
+  | OccursCheck
+  deriving (Show, Exception)
+
+data UnifyError
+  = UnifyError [Name] Tm Tm
+  deriving (Show, Exception)
+
+data ElabError
+  = StrengtheningError Tm Tm StrengtheningError
+  | SpineError Tm Tm SpineError
+  | UnifyErrorWhile Tm Tm UnifyError
   | NameNotInScope Name
-  | ExpectedFunction Tm                  -- ^ Inferred type.
+  | ExpectedFunction Tm
   | IcitMismatch Icit Icit
-  | NonLinearSpine Tm Tm Lvl          -- ^ Lhs, rhs, offending variable
+
+data Err = Err {
+  errNames :: [Name],
+  errErr   :: ElabError,
+  errPos   :: Maybe SPos}
 
 instance Show Err where
   show _ = "Error"
@@ -314,26 +325,28 @@ instance Show Tm where show = showTm []
 
 showError :: [Name] -> ElabError -> String
 showError ns = \case
-  SpineNonVar lhs rhs -> printf (
-    "Non-bound-variable value in meta spine in equation:\n\n" ++
-    "  %s =? %s")
-    (showTm ns lhs) (showTm ns rhs)
-  SpineProjection -> "Projection in meta spine"
-  ScopeError lhs rhs x -> printf (
[…]

### 2020-02-14 implicit-fun-elab/Elaboration.hs
errors

@@ -190,7 +190,7 @@ solveMeta cxt m sp rhs = do
       ~rhst = quote (cxt^.len) rhs
   (r, d) <- pure $ checkSp sp
   rhs <- pure (strengthen (Str d (cxt^.len) r (Just m)) rhs)
-         `catch` \e -> report (cxt^.names) $ StrengtheningError lhst rhst e
+         `catch` \e -> throw $ StrengtheningError (cxt^.names) lhst rhst e
 
   let mty = case lookupMeta m of
         Unsolved _ ty -> ty
@@ -231,7 +231,7 @@ freshMeta cxt a = do
 unify :: UnifyCxt -> Val -> Val -> IO ()
 unify cxt topT topT' = go topT topT' where
 
-  ~report =
+  ~unifyError =
     throw $ UnifyError (cxt^.names) (quote (cxt^.len) topT) (quote (cxt^.len) topT')
 
   go t t' = case (force t, force t') of
@@ -251,7 +251,7 @@ unify cxt topT topT' = go topT topT' where
     (VNe h sp, VNe h' sp') | h == h'          -> goSp (forceSp sp) (forceSp sp')
     (VNe (HMeta m) sp, t')                    -> solveMeta cxt m sp t'
     (t, VNe (HMeta m') sp')                   -> solveMeta cxt m' sp' t
-    _                                         -> report
+    _                                         -> unifyError
 
   goBind x a t t' =
     let UCxt tys ns d = cxt
@@ -261,7 +261,7 @@ unify cxt topT topT' = go topT topT' where
     (SNil, SNil)                            -> pure ()
     (SApp sp u i, SApp sp' u' i') | i == i' -> goSp sp sp' >> go u u'
     (SAppTel a sp u, SAppTel a' sp' u')     -> go a a' >> goSp sp sp' >> go u u'
-    _                                       -> report
+    _                                       -> unifyError
 
 -- Elaboration
 --------------------------------------------------------------------------------

### 2020-02-14 implicit-fun-elab/Main.hs

@@ -127,7 +127,7 @@ test = main' "elab" $ unlines [
 
   -- "let t5 = λ xs. poly (head xs) in", -- pruning
 
-  "not"
+  "hundred"
   ]
 
 -- example for when a telescope cannot depend on local A% B% vars.

### 2020-02-14 implicit-fun-elab/Types.hs

@@ -197,11 +197,11 @@ data StrengtheningError
 
 data UnifyError
   = UnifyError [Name] Tm Tm
+  | StrengtheningError [Name] Tm Tm StrengtheningError
   deriving (Show, Exception)
 
 data ElabError
-  = StrengtheningError Tm Tm StrengtheningError
-  | SpineError Tm Tm SpineError
+  = SpineError Tm Tm SpineError
   | UnifyErrorWhile Tm Tm UnifyError
   | NameNotInScope Name
   | ExpectedFunction Tm
@@ -325,6 +325,7 @@ instance Show Tm where show = showTm []
 
 showError :: [Name] -> ElabError -> String
 showError ns = \case
+
   SpineError lhs rhs e -> case e of
     SpineNonVar -> printf (
       "Non-bound-variable value in meta spine in equation:\n\n" ++
@@ -337,25 +338,30 @@ showError ns = \case
        "  %s =? %s")
       (lvlName ns x)
       (showTm ns lhs) (showTm ns rhs)
-  StrengtheningError lhs rhs e -> case e of
-    ScopeError x -> printf (
-      "Variable %s is out of scope in equation\n\n" ++
-      "  %s =? %s")
-      (lvlName ns x) (showTm ns lhs) (showTm ns rhs)
-    OccursCheck -> printf (
-      "Meta occurs cyclically in its solution candidate in equation:\n\n" ++
-      "  %s =? %s")
-      (showTm ns lhs) (showTm ns rhs)
-  UnifyErrorWhile lhs rhs (UnifyError ns' lhs' rhs') -> printf
-    ("Cannot unify\n\n" ++
-     "  %s\n\n" ++
-     "with\n\n" ++
-     "  %s\n\n" ++
-     "while trying to unify\n\n" ++
-     "  %s\n\n" ++
-     "with\n\n" ++
-     "  %s")
-    (showTm ns' lhs') (showTm ns' rhs') (showTm ns lhs) (showTm ns rhs)
+
+  UnifyErrorWhile lhs rhs e ->
+    let err1 = case e of
+          UnifyError ns' lhs' rhs' -> printf
+            ("Cannot unify\n\n" ++
+             "  %s\n\n" ++
+             "with\n\n" ++
+             "  %s\n\n")
+            (showTm ns' lhs') (showTm ns' rhs')
+          StrengtheningError ns lhs rhs e -> case e of
+            ScopeError x -> printf (
+              "Variable %s is out of scope in equation\n\n" ++
[…]

### 2020-02-17 paper.tex
pruning

@@
 elaboration. In our case, metacontexts are simply a context prefix, and we have
-variables pointing into it, but we do not support any contextual modality
-e.g.\ as in \cite{nanevski2008contextual}. The non-meta typing context
+variables pointing into it. This corresponds to a particularly simple variant of
+\emph{crisp type theory} \cite{licata2018internal}, where we do not have modal
+type operators or functions with crisp (``meta'') domain. The non-meta typing context
 additionally supports \emph{defined variables}, which is used in the typing rule

### 2020-02-17 implicit-fun-elab/Elaboration.hs

@@ -86,20 +86,22 @@ type Renaming = IM.IntMap Lvl
 
 -- | Returns a partial renaming and the length of the spine.
 --   Nonlinear spine is always an error. Throws SpineError.
-checkSp :: Spine -> (Renaming, Lvl)
+checkSp :: Spine -> IO (Renaming, Lvl)
 checkSp = go . forceSp where
-  go :: Spine -> (Renaming, Lvl)
+  go :: Spine -> IO (Renaming, Lvl)
   go = \case
-    SNil        -> (mempty, 0)
-    SApp sp u i -> case go sp of
-      (!r, !d) -> case force u of
+    SNil        -> pure (mempty, 0)
+    SApp sp u i -> do
+      (!r, !d) <- go sp
+      case force u of
         VVar x | IM.member x r -> throw $ NonLinearSpine x
-               | otherwise     -> (IM.insert x d r, d + 1)
+               | otherwise     -> pure (IM.insert x d r, d + 1)
         _      -> throw SpineNonVar
-    SAppTel a sp u -> case go sp of
-      (!r, !d) -> case force u of
+    SAppTel a sp u -> do
+      (!r, !d) <- go sp
+      case force u of
         VVar x | IM.member x r -> throw $ NonLinearSpine x
-               | otherwise     -> (IM.insert x d r, d + 1)
+               | otherwise     -> pure (IM.insert x d r, d + 1)
         _    -> throw SpineNonVar
     SProj1 _ -> throw SpineProjection
     SProj2 _ -> throw SpineProjection
@@ -111,76 +113,132 @@ data Str = Str {
   _strRen :: IM.IntMap Lvl, -- ^ partial renaming
   _strOcc :: Maybe MId      -- ^ meta for occurs strengthening
   }
-
 makeFields ''Str
 
-strengthen :: Str -> Val -> Tm
-strengthen (Str d c r occurs) = go where
+liftStr :: Str -> Str
+liftStr (Str c d r occ) = Str (c + 1) (d + 1) (IM.insert d c r) occ
 
-  prune m sp = Nothing
+skipStr :: Str -> Str
+skipStr (Str c d r occ) = Str c (d + 1) r occ
+
+
+-- | A left-to-right representation for pruned spines (reversed compared to
+--   `Spine`).  This is convenient for pruning types, where we process function
+--   arguments one-by-one from the left.
+data PrunedSpine
+  = PSNil
[…]

### 2020-02-17 implicit-fun-elab/Evaluation.hs

@@ -113,6 +113,7 @@ eval vs = go where
     PiTel x a b  -> vPiTel id x (go a) (goBind b)
     AppTel a t u -> vAppTel (go a) (go t) (go u)
     LamTel x a t -> vLamTel id x (go a) (goBind t)
+    Skip t       -> eval (VSkip vs) t
 
   goBind t x = eval (VDef vs x) t
 
@@ -193,3 +194,4 @@ zonk vs t = go t where
                       Left t  -> quote (valsLen vs) (vAppTel (eval vs a) t (eval vs u))
                       Right t -> AppTel (go a) t (go u)
     LamTel x a b -> LamTel x (go a) (goBind b)
+    Skip t       -> Skip (goBind t)

### 2020-02-17 implicit-fun-elab/Main.hs

@@ -65,6 +65,19 @@ main = mainWith getArgs parseStdin
 main' :: String -> String -> IO ()
 main' mode src = mainWith (pure [mode]) ((,src) <$> parseString src)
 
+pruneTest = main' "elab" $ unlines [
+  "λ (Bool : U)(not : Bool → Bool → Bool)(true : Bool).",
+  "λ f (x : Bool). not (f x)"
+  ]
+
+test0 = main' "elab" $ unlines [
+  "let Nat : U",
+  "    = (N : U) -> (N -> N) -> N -> N in",
+  "let zero : Nat = λ _ s z. z in",
+  "let id : {A} → A → A = λ x. x in",
+  "let id2 : {A} → A → A = id in",
+  "id zero"
+  ]
 
 test = main' "elab" $ unlines [
   "let the : (A : U) → A → A = \\A x.x in",
@@ -75,7 +88,8 @@ test = main' "elab" $ unlines [
   "let Nat : U",
   "    = (N : U) -> (N -> N) -> N -> N in",
   "let zero : Nat = λ _ s z. z in",
-  "let inc : Nat → Nat = λ n _ s z. s (n _ s z) in",
+  "let inc : Nat → Nat = λ n N s z. s (n _ s z) in",
+
   "let mul : Nat -> Nat -> Nat",
   "    = \\a b N s z. a _ (b N s) z in",
   "let ten : Nat",
@@ -118,9 +132,9 @@ test = main' "elab" $ unlines [
   "let false : Bool",
   "    = \\B t f. f in",
   "let not : Bool → Bool",
-  "   = λ b B t f. b B f t in",
+  "   = λ b B t f. b _ f t in",
 
-  -- "let poly : ({A} → A → A) → Pair Nat Bool = λ f. pair (f zero) (f true) in",
+  "let poly : ({A} → A → A) → Pair Nat Bool = λ f. pair (f zero) (f true) in",
   "let auto : ({A} → A → A) → ({A} → A → A) = id in",
   "let app  : {A B} → (A → B) → A → B = λ f a. f a in",
   "let revapp : {A B} → A → (A → B) → B = λ a f. f a in",
@@ -138,19 +152,19 @@ outOfScopeTelRefinement = main' "elab" $ unlines [
   "the ({A B} → A → B → A) f"
   ]
 
-issue = main' "elab" $ unlines [
- "assume Nat : U in",
- "assume Dec : U → U in",
- "assume meh : {A} → Dec A in",
[…]

### 2020-02-17 implicit-fun-elab/Types.hs

@@ -133,6 +133,8 @@ data Tm
   | U
   | Meta MId
 
+  | Skip Tm  -- explicit weakening (convenience feature for fresh meta gen)
+
 data Spine
   = SNil
   | SApp Spine ~Val Icit
@@ -144,6 +146,12 @@ valsLen :: Vals -> Int
 valsLen = go 0 where
   go acc VNil        = acc
   go acc (VDef vs _) = go (acc + 1) vs
+  go acc (VSkip vs)  = go (acc + 1) vs
+
+boundVars :: Vals -> Int
+boundVars = go 0 where
+  go acc VNil        = acc
+  go acc (VDef vs _) = go acc vs
   go acc (VSkip vs)  = go (acc + 1) vs
 
 -- spLen :: Spine -> Int
@@ -197,12 +205,12 @@ data StrengtheningError
 
 data UnifyError
   = UnifyError [Name] Tm Tm
+  | SpineError Tm Tm SpineError
   | StrengtheningError [Name] Tm Tm StrengtheningError
   deriving (Show, Exception)
 
 data ElabError
-  = SpineError Tm Tm SpineError
-  | UnifyErrorWhile Tm Tm UnifyError
+  = UnifyErrorWhile Tm Tm UnifyError
   | NameNotInScope Name
   | ExpectedFunction Tm
   | IcitMismatch Icit Icit
@@ -235,7 +243,8 @@ prettyTm prec = go (prec /= 0) where
   goVar :: [Name] -> Ix -> ShowS
   -- goVar ns topX = (show topX++)
   goVar ns topX = go ns topX where
-    go []     _ = error "impossible"
+    -- go []     _ = error "impossible"
+    go []     _ = (show topX++)
     go (n:ns) 0 = (n++)
     go (n:ns) x = go ns (x - 1)
 
@@ -316,6 +325,8 @@ prettyTm prec = go (prec /= 0) where
     LamTel x a t -> showParen p (("λ"++)
                    . bracket ((x++) . (" : "++) . go False ns a) . goLam ns t)
 
+    Skip t -> go p ("_":ns) t
+
 showTm :: [Name] -> Tm -> String
 showTm ns t = prettyTm 0 ns t []
 -- showTm ns t = show t
@@ -326,35 +337,34 @@ instance Show Tm where show = showTm []
 showError :: [Name] -> ElabError -> String
 showError ns = \case
 
-  SpineError lhs rhs e -> case e of
-    SpineNonVar -> printf (
[…]

### 2020-02-17 implicit-fun-elab/Elaboration.hs
cleaning up

@@ -16,6 +16,7 @@ import qualified Data.IntSet as IS
 import qualified Data.IntMap as IM
 import Data.IORef
 import Lens.Micro.Platform
+import Data.Maybe
 
 import Types
 import Evaluation
@@ -74,18 +75,17 @@ occurs d topX = occurs' d mempty where
       VPiTel x a b  -> go a <> goBind b
       VLamTel x a t -> go a <> goBind t
 
--- pruning
+
+-- unification
 --------------------------------------------------------------------------------
 
 -- | A partial mapping from levels to levels. Undefined domain reresents
 --   out-of-scope or "illegal" variables.
 type Renaming = IM.IntMap Lvl
 
--- unification
---------------------------------------------------------------------------------
-
--- | Returns a partial renaming and the length of the spine.
---   Nonlinear spine is always an error. Throws SpineError.
+-- | Checks that a spine consists only of distinct bound vars.
+--   Returns a partial variable renaming on success, alongside the size
+--   of the spine.
 checkSp :: Spine -> IO (Renaming, Lvl)
 checkSp = go . forceSp where
   go :: Spine -> IO (Renaming, Lvl)
@@ -106,32 +106,43 @@ checkSp = go . forceSp where
     SProj1 _ -> throw SpineProjection
     SProj2 _ -> throw SpineProjection
 
-
+-- | A strengthening. We use this for pruning and checking meta solution
+--   candidates.
 data Str = Str {
-  _strDom :: Lvl,           -- ^ size of renaming domain
-  _strCod :: Lvl,           -- ^ size of renaming codomain
-  _strRen :: IM.IntMap Lvl, -- ^ partial renaming
-  _strOcc :: Maybe MId      -- ^ meta for occurs strengthening
+  _strDom :: Lvl,        -- ^ size of renaming domain
+  _strCod :: Lvl,        -- ^ size of renaming codomain
+  _strRen :: Renaming,   -- ^ partial renaming
+  _strOcc :: Maybe MId   -- ^ meta for occurs strengthening
   }
 makeFields ''Str
 
+-- | Lift over a bound variable.
 liftStr :: Str -> Str
 liftStr (Str c d r occ) = Str (c + 1) (d + 1) (IM.insert d c r) occ
 
[…]

### 2020-02-17 implicit-fun-elab/Types.hs

@@ -76,13 +76,6 @@ pattern TSnoc as a <- ((\case TBound as a -> Just (as, a)
 lvlName :: [Name] -> Lvl -> Name
 lvlName ns x = ns !! (length ns - x - 1)
 
--- -- clean these up
--- typesLen :: Types -> Int
--- typesLen = go 0 where
---   go acc TNil           = acc
---   go acc (TDef tys _)   = go (acc + 1) tys
---   go acc (TBound tys _) = go (acc + 1) tys
-
 ixType :: Types -> Ix -> VTy
 ixType TNil           _ = error "impossible"
 ixType (TDef   tys a) 0 = a
@@ -90,10 +83,9 @@ ixType (TBound tys a) 0 = a
 ixType (TDef   tys a) x = ixType tys (x - 1)
 ixType (TBound tys a) x = ixType tys (x - 1)
 
--- lvlType :: Types -> Lvl -> VTy
--- lvlType tys x = ixType tys (typesLen tys - x - 1)
-
 data NameOrigin = NOSource | NOInserted
+
+type MetaInsertion = Bool
 
 data Cxt = Cxt {
   cxtVals       :: Vals,
@@ -148,20 +140,6 @@ valsLen = go 0 where
   go acc (VDef vs _) = go (acc + 1) vs
   go acc (VSkip vs)  = go (acc + 1) vs
 
-boundVars :: Vals -> Int
-boundVars = go 0 where
-  go acc VNil        = acc
-  go acc (VDef vs _) = go acc vs
-  go acc (VSkip vs)  = go (acc + 1) vs
-
--- spLen :: Spine -> Int
--- spLen = go 0 where
---   go n SNil             = n
---   go n (SApp sp _ _)    = go (n + 1) sp
---   go n (SAppTel _ sp _) = go (n + 1) sp
---   go n (SProj1 sp)      = go (n + 1) sp
---   go n (SProj2 sp)      = go (n + 1) sp
-
 data Head
   = HVar Lvl
   | HMeta MId
@@ -184,13 +162,14 @@ data Val
   | VPiTel Name ~Val (Val -> Val)
   | VLamTel Name ~Val (Val -> Val)
 
-type MetaInsertion = Bool
-
 pattern VVar :: Lvl -> Val
 pattern VVar x = VNe (HVar x) SNil
 
 pattern VMeta :: MId -> Val
 pattern VMeta m = VNe (HMeta m) SNil
+
+-- Errors
[…]

### 2020-02-17 implicit-fun-elab/ElabState.hs
more cleanup, start telescope unify

@@ -18,18 +18,33 @@ nextMId :: IORef Int
 nextMId = runIO (newIORef 0)
 {-# noinline nextMId #-}
 
-lookupMetaIO :: MId -> IO MetaEntry
-lookupMetaIO m = do
+lookupMeta :: MId -> IO MetaEntry
+lookupMeta m = do
   ms <- readIORef mcxt
   case IM.lookup m ms of
     Just e -> pure e
     _      -> error "impossible"
 
-lookupMeta :: MId -> MetaEntry
-lookupMeta m = runIO (lookupMetaIO m)
+runLookupMeta :: MId -> MetaEntry
+runLookupMeta m = runIO (lookupMeta m)
 
-writeMetaIO :: MId -> MetaEntry -> IO ()
-writeMetaIO m e = modifyIORef' mcxt (IM.insert m e)
+writeMeta :: MId -> MetaEntry -> IO ()
+writeMeta m e = modifyIORef' mcxt (IM.alter go m) where
+  go Just{} = Just e
+  go _      = error "impossible"
+
+newMetaEntry :: MetaEntry -> IO MId
+newMetaEntry e = do
+  m <- readIORef nextMId
+  writeIORef nextMId $! (m + 1)
+  let f = maybe (Just e) (\_ -> error "impossible")
+  modifyIORef' mcxt (IM.alter f m)
+  pure m
+
+modifyMeta :: MId -> (MetaEntry -> MetaEntry) -> IO ()
+modifyMeta m f = modifyIORef' mcxt (IM.alter go m) where
+  go Nothing  = error "impossible"
+  go (Just e) = Just (f e)
 
 reset :: IO ()
 reset = do

### 2020-02-17 implicit-fun-elab/Elaboration.hs

@@ -9,14 +9,14 @@
 
 module Elaboration where
 
-import Data.Foldable
 import Control.Exception
 import Control.Monad
-import qualified Data.IntSet as IS
-import qualified Data.IntMap as IM
+import Data.Foldable
 import Data.IORef
-import Lens.Micro.Platform
 import Data.Maybe
+import Lens.Micro.Platform
+import qualified Data.IntMap as IM
+import qualified Data.IntSet as IS
 
 import Types
 import Evaluation
@@ -24,7 +24,8 @@ import ElabState
 
 import Debug.Trace
 
--- occurrence check for constancy
+
+-- Telescope constancy
 --------------------------------------------------------------------------------
 
 data Occurs = Rigid | Flex IS.IntSet | None deriving (Eq, Show)
@@ -76,12 +77,62 @@ occurs d topX = occurs' d mempty where
       VLamTel x a t -> go a <> goBind t
 
 
--- unification
---------------------------------------------------------------------------------
+-- newConstancy :: MId -> Spine -> Closure -> IO ()
+-- newConstancy telMeta sp cl = do
+--   x <- fresh <$> view vals <*> pure "_"
+--   ~codomain <- applyM cl (VVar x)
+--   constM <- newMetaEntry (Constancy telMeta sp x codomain mempty)
+--   tryConstancy constM
 
--- | A partial mapping from levels to levels. Undefined domain reresents
---   out-of-scope or "illegal" variables.
-type Renaming = IM.IntMap Lvl
+-- | Attempt to solve a constancy constraint.
+tryConstancy :: MId -> IO ()
+tryConstancy constM = lookupMeta constM >>= \case
+  Constancy cxt telMeta sp codomain blockers -> do
+
+    -- clear blockers
+    forM_ (IS.toList blockers) $ \m -> do
+      modifyMeta m $ \case
+        Unsolved ms a -> Unsolved (IS.delete constM ms) a
+        Solved t      -> Solved t
+        Constancy{}   -> error "impossible"
+
+    writeMeta constM $ Constancy cxt telMeta sp codomain mempty
+
+    lookupMeta telMeta >>= \case
+      Constancy{} -> error "impossible"
+      Solved{}    -> pure ()
+      Unsolved{}  ->
+        case occurs (cxt^.len + 1) (cxt^.len) codomain of
[…]

### 2020-02-17 implicit-fun-elab/Evaluation.hs

@@ -24,7 +24,7 @@ vVar x (VSkip vs)  = vVar (x - 1) vs
 vVar _ _           = error "impossible"
 
 vMeta :: MId -> Val
-vMeta m = case lookupMeta m of
+vMeta m = case runLookupMeta m of
   Unsolved{} -> VMeta m
   Solved v   -> v
   _          -> error "impossible"
@@ -33,7 +33,7 @@ vMeta m = case lookupMeta m of
 --   of a neutral value.
 force :: Val -> Val
 force = \case
-  v@(VNe (HMeta m) sp) -> case lookupMeta m of
+  v@(VNe (HMeta m) sp) -> case runLookupMeta m of
     Unsolved{} -> v
     Solved v   -> force (vAppSp v sp)
     _          -> error "impossible"
@@ -155,7 +155,7 @@ zonk vs t = go t where
 
   goSp :: Tm -> Either Val Tm
   goSp = \case
-    Meta m       -> case lookupMeta m of
+    Meta m       -> case runLookupMeta m of
                       Solved v -> Left v
                       _        -> Right (Meta m)
     App t u ni   -> case goSp t of
@@ -170,7 +170,7 @@ zonk vs t = go t where
 
   go = \case
     Var x        -> Var x
-    Meta m       -> case lookupMeta m of
+    Meta m       -> case runLookupMeta m of
                       Solved v   -> quote (valsLen vs) v
                       Unsolved{} -> Meta m
                       _          -> error "impossible"

### 2020-02-17 implicit-fun-elab/Types.hs

@@ -58,7 +58,31 @@ data MetaEntry
 
   -- | Telescope constancy constraint. When the closure becomes constant,
   --   we unify the telescope with the empty telescope.
-  | Constancy MId Spine Name Val BlockedBy
+  --   Constancy context (telescope meta) (meta spine) codomain blockers
+  | Constancy UnifyCxt MId Spine Val BlockedBy
+
+
+-- | A partial mapping from levels to levels. Undefined domain reresents
+--   out-of-scope or "illegal" variables.
+type Renaming = IM.IntMap Lvl
+
+-- | A strengthening. We use this for pruning and checking meta solution
+--   candidates.
+data Str = Str {
+  _strDom :: Lvl,        -- ^ size of renaming domain
+  _strCod :: Lvl,        -- ^ size of renaming codomain
+  _strRen :: Renaming,   -- ^ partial renaming
+  _strOcc :: Maybe MId   -- ^ meta for occurs strengthening
+  }
+
+-- | Lift over a bound variable.
+liftStr :: Str -> Str
+liftStr (Str c d r occ) = Str (c + 1) (d + 1) (IM.insert d c r) occ
+
+-- | Skip a bound variable.
+skipStr :: Str -> Str
+skipStr (Str c d r occ) = Str c (d + 1) r occ
+
 
 data Vals  = VNil | VDef Vals ~Val | VSkip Vals
 data Types = TNil | TDef Types ~VTy | TBound Types ~VTy
@@ -366,12 +390,13 @@ showError ns = \case
 makeFields ''Cxt
 makeFields ''UnifyCxt
 makeFields ''Err
+makeFields ''Str
 
 ucxt :: Lens' Cxt UnifyCxt
 ucxt f (Cxt vs tys ns no d) =
   (\(UCxt tys ns d) -> Cxt vs tys ns no d) <$> f (UCxt tys ns d)
 
-instance HasNames  [Name]  [Name]  where names = id
-instance HasVals   Vals    Vals    where vals  = id
-instance HasTypes  Types   Types   where types = id
-instance HasLen    Int     Int     where len   = id
+-- instance HasNames  [Name]  [Name]  where names = id
+-- instance HasVals   Vals    Vals    where vals  = id
+-- instance HasTypes  Types   Types   where types = id
+-- instance HasLen    Int     Int     where len   = id

### 2020-02-17 implicit-fun-elab/Elaboration.hs
telescope progress

@@ -1,11 +1,4 @@
 
--- TODO: - Only use elabcxt and unifycxt
---       - try to decorate errors with info when info is available (catch/rethrow)
---       - implement simple elab without telescope/pruning
---       - add pruning
---       - add telescopes
-
--- ISSUE : don't have any nonlinear solutions!
 
 module Elaboration where
 
@@ -77,13 +70,6 @@ occurs d topX = occurs' d mempty where
       VLamTel x a t -> go a <> goBind t
 
 
--- newConstancy :: MId -> Spine -> Closure -> IO ()
--- newConstancy telMeta sp cl = do
---   x <- fresh <$> view vals <*> pure "_"
---   ~codomain <- applyM cl (VVar x)
---   constM <- newMetaEntry (Constancy telMeta sp x codomain mempty)
---   tryConstancy constM
-
 -- | Attempt to solve a constancy constraint.
 tryConstancy :: MId -> IO ()
 tryConstancy constM = lookupMeta constM >>= \case
@@ -120,15 +106,9 @@ tryConstancy constM = lookupMeta constM >>= \case
 
 newConstancy :: UnifyCxt -> MId -> Spine -> (Val -> Val) -> IO ()
 newConstancy cxt telMeta sp b = do
-  let ~codomain = b (VVar (cxt^.len))
-  undefined
-
--- newConstancy :: MId -> Spine -> Closure -> UnifyM ()
--- newConstancy telMeta sp cl = do
---   x <- fresh <$> view vals <*> pure "_"
---   ~codomain <- applyM cl (VVar x)
---   constM <- newMetaEntry (Constancy telMeta sp x codomain mempty)
---   tryConstancy constM
+  let codomain = b (VVar (cxt^.len))
+  constM <- newMetaEntry $ Constancy cxt telMeta sp codomain mempty
+  tryConstancy constM
 
 
 -- Unification
@@ -291,18 +271,23 @@ solveMeta cxt m sp rhs = do
   -- try solving unblocked constraints
   forM_ (IS.toList blocked) tryConstancy
 
-
-freshMeta :: UnifyCxt -> VTy -> IO Tm
+freshMeta :: UnifyCxt -> VTy -> IO (MId, Spine)
[…]

### 2020-02-17 implicit-fun-elab/Evaluation.hs

@@ -64,7 +64,7 @@ vPiTel k x a b = case force a of
                         x2 = x ++ "2"
                     in VPi x1 Impl a $ \ ~x1 ->
                        vPiTel id x2 (as x1) $ \ ~x2 -> b (VTcons x1 x2)
-  a -> VPiTel x a b
+  a              -> VPiTel x a b
 
 vLamTel :: (Val -> Val) -> Name -> VTy -> (Val -> Val) -> Val
 vLamTel k x a t = case force a of

### 2020-02-17 implicit-fun-elab/Main.hs

@@ -74,9 +74,8 @@ test0 = main' "elab" $ unlines [
   "let Nat : U",
   "    = (N : U) -> (N -> N) -> N -> N in",
   "let zero : Nat = λ _ s z. z in",
-  "let id : {A} → A → A = λ x. x in",
-  "let id2 : {A} → A → A = id in",
-  "id zero"
+  "let foo = zero in",
+  "foo"
   ]
 
 test = main' "elab" $ unlines [
@@ -138,6 +137,16 @@ test = main' "elab" $ unlines [
   "let auto : ({A} → A → A) → ({A} → A → A) = id in",
   "let app  : {A B} → (A → B) → A → B = λ f a. f a in",
   "let revapp : {A B} → A → (A → B) → B = λ a f. f a in",
+
+  -- dependent function composition
+  "let comp : {A}{B : A -> U}{C : {a} -> B a -> U}",
+  "           (f : {a}(b : B a) -> C b)",
+  "           (g : (a : A) -> B a)",
+  "           (a : A)",
+  "           -> C (g a)",
+  "    = \\f g a. f (g a) in",
+
+  "let compExample = comp (cons true) (cons false) nil in",
 
   -- "let t5 = λ xs. poly (head xs) in", -- pruning
 

### 2020-02-18 paper.tex
trying some notation changes

@@
 \newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
-\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,#1\,#2\,#3}
+\newcommand{\freshMeta}[3]{\mathsf{freshMeta}_{#1|#2}\,#3}
 \newcommand{\unify}{\mathsf{unify}}
@@
 \newcommand{\einferblank}{\llbracket\blank\rrbracket\!\Uparrow}
-\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,#2\,#3\,#4}
-\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,#2\,#3\,#4}
+\newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,_{#2|#3}\,#4}
+\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,#2\,_{#3|#4}}
 \newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
@@
 \newcommand{\false}{\mathsf{false}}
-\newcommand{\einsert}[5]{\mathsf{insert}\,#1\,#2\,#3\,#4\,#5}
+\newcommand{\einsert}[3]{\mathsf{insert}\,#1\,#2\,#3}
+\newcommand{\einsertpar}[5]{\mathsf{insert}_{#1|#2}\,#3\,#4\,#5}
 \newcommand{\Set}{\mathsf{Set}}
@@
 elaboration in the presence of first-class implicit functions poses a challenge;
-in the context of GHC and ML-like languages, this has been dubbed
+in the context of Haskell and ML-like languages, this has been dubbed
 ``impredicative instantiation'' or ``impredicative inference''. We propose a new
@@
 instead, implicitness is tied to particular \emph{names}, and while we can write
-$list\,(\kw{forall}\,\{A : \kw{Type}\},\,A\to A)$ for a list type of polymorphic
-functions, the brackets here are simply ignored by Coq and we get a plain
-function type. For example, Coq accepts the following definition:
+$list\,(\kw{forall}\,\{A : \kw{Type}\},\,A\to A)$ for a list type with
+polymorphic elements, the brackets here are simply ignored by Coq. For example,
+Coq accepts the following definition:
 \begin{alignat*}{3}
@@
 There is significant literature on type inference in the presence of first-class
-polymorphic types, mainly in relation to GHC and ML-like languages \cite{TODO}.
+polymorphic types, mainly in relation to GHC and ML-like languages \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy}.
 The above issue in Agda is a specific instance of the challenges described in
@@
     types with telescope domain. The extensions to the target theory are modest
-    and are derivable from W-types. We use these extensions to accummulate
-    information about implicit insertions.
+    and are derivable from W-types and $\Sigma$-types. We use these extensions
+    to accummulate information about implicit insertions.
   \item Our algorithm is a conservative extension of Norell's bidirectional
@@
 have semantics or any well-formedness relations attached; its sole purpose is to
-serve as input to elaboration. Hence, the surface syntax can be also viewed as
-a minimal untyped tactic language, which is interpreted by the elaborator.
+serve as input to elaboration. Hence, the surface syntax can be also viewed as a
+small untyped tactic language which is interpreted by the elaborator.
 
@@
 metacontexts, and existing works on algebraic presentations of dependent modal
-contexts (e.g.\ \cite{birkedal2018modal}, TODO) do not precisely cover the
-current use case. We leave this to future work, along with the investigation of
-elaboration from an algebraic perspective.
+contexts (e.g.\ \cite{birkedal2018modal}) do not precisely cover the current use
+case. We leave this to future work, along with the investigation of elaboration
+from an algebraic perspective.
 
@@
 \emph{crisp type theory} \cite{licata2018internal}, where we do not have modal
-type operators or functions with crisp (``meta'') domain. The non-meta typing context
-additionally supports \emph{defined variables}, which is used in the typing rule
-for \emph{let-definitions}, and we have that any defined variable is equal to
-its definition. We mainly support this as a convenience feature in the
-implementation of our elaborator.
+type operators or functions with crisp (``meta'') domain. The non-meta typing
+context additionally supports \emph{defined variables}, which is used in the
+typing rule for $\slet$-definitions, and we have that any defined variable is
+equal to its definition. We mainly support this as a convenience feature in the
+surface language.
 
@@
 The core syntax is quite expressive as a programming language, thanks to
-let-definitions and the type-in-type rule which allows Church-encodings of a
-large class of inductive types. For example, the following term computes a list
-of types, by mapping over the list $cons\,\U\,(cons\,\U\,nil)$.
+$\slet$-definitions\footnote{In dependent type theories, the $\slet$ rule is not
+  derivable from function application, unlike in simple type theories.} and the
+type-in-type rule which allows Church-encodings of a large class of inductive
+types. For example, the following term computes a list of types by mapping:
 \begin{alignat*}{3}
@@
 \inferrule*{ins \in \{\true,\,\false\} \\ \Theta|\Gamma\vdash t : A}
-           {\einsert{ins}{\Theta}{\Gamma}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
+           {\einsert{ins}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
              (\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t' : A')\}}
@@
 definition of elaboration. Insertion is defined by recursion on $ins$ and the
-$A$ type, as follows.
-\begin{alignat*}{3}
-  &\einsert{\false}{\Theta}{\Gamma}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
-  &\einsert{\true\,}{\Theta}{\Gamma}{t}{(\{x : A\}\to B)} && :\equiv&&
+$A$ type, as follows. Here, we notate $\Theta$ and $\Gamma$ as subscripts but
+later we will leave them implicit.
+\begin{alignat*}{3}
+  &\einsertpar{\Theta}{\Gamma}{\false}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
+  &\einsertpar{\Theta}{\Gamma}{\true\,}{t}{(\{x : A\}\to B)} && :\equiv&&
        \slet\,(\Theta',\,\theta,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
-  & && &&\sin\,\einsert{\true}{\Theta'}{(\Gamma[\theta])}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
-  &\einsert{\true\,}{\Theta}{\Gamma}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
+  & && &&\sin\,\einsertpar{\Theta'}{\Gamma[\theta']}{\true}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
+  &\einsertpar{\Theta}{\Gamma}{\true\,}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
 \end{alignat*}
@@
 unification in detail in \cite[Chapter~4]{gundry2013type} for a similar syntax,
-with a similar (though more featureful) setup for metacontexts.
-
-\todo{example for meta solution, more references, describe at least pattern unif and pruning briefly}
-
-Note that our unification algorithm does not support \emph{constraint
-  postponing}, since we have not specified constraints. In our concrete
-implementation, unification supports basic higher-order pattern unification and
-metavariable pruning \cite{TODO}.
+with a similar (though more featureful) setup for metacontexts. See also
+\cite{abel2011higher} for a reference on unification. Note that our unification
+algorithm does not support \emph{constraint postponing}, as we have not talked
+about constraints at all. In our concrete implementation, unification supports
+basic pattern unification and metavariable pruning.
+
+We describe a simple example. We have $\Theta :\equiv (\emptycon,\,\alpha :
+\U,\, \beta : \U)$, and we compute $\unify\,\alpha\,(\beta\to\beta)$. This
+yields $\Theta' :\equiv (\emptycon,\,\beta : \U)$ and the substitution $\theta
+:\equiv (\alpha \mapsto (\beta\to\beta),\,\beta\mapsto \beta)$, where $\theta :
+\Theta' \To \Theta$.
 
@@
     whether a type is a function type. This requires an evaluation/normalization
-    procedure on core terms, but we already assume this in unification.
+    procedure on core terms, but note that we already assume this in implicit
+    argument insertion and unification.
 \end{itemize}
@@
   &\quad\eif\,(x : A \in \Gamma)\,\lor\,(x : A = t \in \Gamma) \\
-  &\qquad   \ethen\, \ereturn\,(\einsert{ins}{\Theta}{\Gamma}{x}{A})\\
+  &\qquad   \ethen\, \ereturn\,(\einsert{ins}{x}{A})\\
   &\qquad   \eelse\, \fail\\
@@
   &\einfer{\lambda\,x.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta}{\Gamma}{\U}\\
+  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
   &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{(\Gamma[\theta_1],\,x:A)}\\
@@
   &\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta}{\Gamma}{\U}\\
+  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
   &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{(\Gamma[\theta_1],\,x:A)}\\
-  &\quad\ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(\lambda\,\{x\}.\,t')}{(\{x : A[\theta_2]\}\to B)})\\
+  &\quad\ereturn\,(\einsert{ins}{(\lambda\,\{x\}.\,t')}{(\{x : A[\theta_2]\}\to B)})\\
   &\einfer{t\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
   &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
-  &\quad \ecase\,\,A\,\,\eof\\
-  &\qquad   ((x : A)\to B) \To \edo\\
-  &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
-                    \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
-  &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,u')}{(B[\theta_2][x\mapsto u'])})\\
-  &\qquad A \To \fail\\
+  &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
+  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,x:A_0}{\U}\\
+  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,((x : A_0[\theta_3])\to A_1)\\
+  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{(\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4])}{(A_0[\theta_3\circ\theta_4])}\\
+  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,u')}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\
   &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
   &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\false}{\Theta_0}{\Gamma}\\
-  &\quad \ecase\,\,A\,\,\eof\\
-  &\qquad   (\{x : A\}\to B) \To \edo\\
-  &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
-                    \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
-  &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,\{u'\})}{(B[\theta_2][x\mapsto u'])})\\
-  &\qquad A \To \fail\\
+  &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
+  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,x:A_0}{\U}\\
+  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,(\{x : A_0[\theta_3]\}\to A_1)\\
+  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{(\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4])}{(A_0[\theta_3\circ\theta_4])}\\
+  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,\{u'\})}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\
   &\einfer{\slet\,x:A=t\,\sin\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
@@
   &\quad \slet\,(\Theta',\,\theta,\,A) = \freshMeta{\Theta}{\Gamma}{\U}\\
-  &\quad \ereturn\,(\freshMeta{\Theta'}{(\Gamma[\theta])}{A})\\
+  &\quad \ereturn\,(\freshMeta{\Theta'}{\Gamma[\theta]}{A})\\
 \end{alignat*}
 \endgroup
+
+%% %% alternative application infer
+  %% &\einfer{t\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  %% &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
+  %% &\quad \ecase\,\,A\,\,\eof\\
+  %% &\qquad   ((x : A)\to B) \To \edo\\
+  %% &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
+  %%                   \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
+  %% &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,u')}{(B[\theta_2][x\mapsto u'])})\\
+  %% &\qquad A \To \fail\\
+  %% &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  %% &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\false}{\Theta_0}{\Gamma}\\
+  %% &\quad \ecase\,\,A\,\,\eof\\
+  %% &\qquad   (\{x : A\}\to B) \To \edo\\
+  %% &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
+  %%                   \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
+  %% &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,\{u'\})}{(B[\theta_2][x\mapsto u'])})\\
+  %% &\qquad A \To \fail\\
 
@@
 $\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma}$, we additionally perform
-$\einsert$ on the output. \todo{explain this weird case}
+$\einsert$ on the output.
 
@@
 $\einferblank$ conform to the specification \ref{TODO}.
-This follows by induction on surface syntax and analyzing the definitions, while also relying on properties of substitution, metasubstitution, $\unify$ and $\freshMeta$.
+This follows by induction on surface syntax and analyzing the definitions, while also relying on properties of substitution, metasubstitution, $\unify$ and $\mathsf{freshMeta}$.
 \qed

### 2020-02-18 implicit-fun-elab/Elaboration.hs

@@ -139,10 +139,12 @@ checkSp = go . forceSp where
 
 -- | Close a type in a cxt by wrapping it in Pi types and explicit weakenings.
 closingTy :: UnifyCxt -> Ty -> Ty
-closingTy (UCxt TNil           []     d) b = b
-closingTy (UCxt (TDef tys a)   (x:ns) d) b = closingTy (UCxt tys ns (d-1)) (Skip b)
-closingTy (UCxt (TBound tys a) (x:ns) d) b = closingTy (UCxt tys ns (d-1)) (Pi x Expl (quote (d-1) a) b)
-closingTy _                              _ = error "impossible"
+closingTy = go where
+  go (UCxt TNil                  []     d) b = b
+  go (UCxt (TDef tys a)          (x:ns) d) b = go (UCxt tys ns (d-1)) (Skip b)
+  go (UCxt (TBound tys (VRec a)) (x:ns) d) b = go (UCxt tys ns (d-1)) (PiTel x (quote (d-1) a) b)
+  go (UCxt (TBound tys a)        (x:ns) d) b = go (UCxt tys ns (d-1)) (Pi x Expl (quote (d-1) a) b)
+  go _                                     _ = error "impossible"
 
 -- | Close a term by wrapping it in `Int` number of lambdas, while taking the domain
 --   types from the `VTy`.
@@ -275,11 +277,13 @@ freshMeta :: UnifyCxt -> VTy -> IO (MId, Spine)
 freshMeta cxt (quote (cxt^.len) -> a) = do
   let metaTy = closingTy cxt a
   m <- newMetaEntry $ Unsolved mempty (eval VNil metaTy)
+  -- traceShowM ("freshMeta", m, metaTy)
 
   let vars :: Types -> (Spine, Lvl)
-      vars TNil                          = (SNil, 0)
-      vars (TDef (vars -> (sp, !d)) _)   = (sp, d + 1)
-      vars (TBound (vars -> (sp, !d)) _) = (SApp sp (VVar d) Expl, d + 1)
+      vars TNil                                 = (SNil, 0)
+      vars (TDef (vars -> (sp, !d)) _)          = (sp, d + 1)
+      vars (TBound (vars -> (sp, !d)) (VRec a)) = (SAppTel a sp (VVar d), d + 1)
+      vars (TBound (vars -> (sp, !d)) _)        = (SApp sp (VVar d) Expl, d + 1)
 
   let sp = fst $ vars (cxt^.types)
   pure (m, sp)
[…]

### 2020-02-18 implicit-fun-elab/Main.hs

@@ -71,12 +71,12 @@ pruneTest = main' "elab" $ unlines [
   ]
 
 test0 = main' "elab" $ unlines [
-  "let Nat : U",
-  "    = (N : U) -> (N -> N) -> N -> N in",
-  "let zero : Nat = λ _ s z. z in",
-  "let foo = zero in",
-  "foo"
+  "let IdTy = {A} → A → A in",
+  "let f = λ (g:IdTy). g g in",
+  -- "let id : IdTy = f in",
+  "U"
   ]
+
 
 test = main' "elab" $ unlines [
   "let the : (A : U) → A → A = \\A x.x in",
@@ -175,11 +175,22 @@ outOfScopeTelRefinement = main' "elab" $ unlines [
 --  "U"
 --   ]
 
+-- ex1 = main' "elab" $ unlines [
+--   "λ (List : U → U)(nil : {A} → List A)(cons : {A} → A → List A → List A)(append : {A} → List A → List A → List A).",
+
+
+--   "let l1 = nil in",
+--   "let l2 : List ({A} → List A) = cons l1 nil in",
+
+--   "U"
+--   ]
+
 
 ex1 = main' "elab" $ unlines [
   "let the : (A : U) → A → A = \\A x.x in",
   "let id : {A} → A → A = λ x. x in",
   "let choose : {A} → A → A → A = λ x y. x in",
+  "λ (Bool : U)(true : Bool)(false : Bool).",
 
   "let Nat : U",
   "    = (N : U) -> (N -> N) -> N -> N in",
@@ -221,15 +232,6 @@ ex1 = main' "elab" $ unlines [
   "let fst : {A B} → Pair A B → A = λ p. p _ (λ x y. x) in",
   "let snd : {A B} → Pair A B → B = λ p. p _ (λ x y. y) in",
 
-  "let Bool : U",
-  "    = {B} -> B -> B -> B in",
-  "let true : Bool",
-  "    = \\t f. t in",
-  "let false : Bool",
-  "    = \\t f. f in",
-  "let not : Bool → Bool",
-  "   = λ b t f. b f t in",
-
   "let poly : ({A} → A → A) → Pair Nat Bool = λ f. pair (f zero) (f true) in",
   "let auto : ({A} → A → A) → ({A} → A → A) = id in",
   "let app  : {A B} → (A → B) → A → B = λ f a. f a in",
@@ -251,23 +253,24 @@ ex1 = main' "elab" $ unlines [
   "let l = cons const2 (nil {{A B} → A → B → A}) in",
 
[…]

### 2020-02-18 implicit-fun-elab/Types.hs

@@ -326,14 +326,13 @@ prettyTm prec = go (prec /= 0) where
       showParen p (go True ns t . (' ':)
                    . bracket (go False ns u  . (" : "++) . go False ns a))
     LamTel x a t -> showParen p (("λ"++)
-                   . bracket ((x++) . (" : "++) . go False ns a) . goLam ns t)
+                   . bracket ((x++) . (" : "++) . go False ns a) . goLam (x:ns) t)
 
     Skip t -> go p ("_":ns) t
 
 showTm :: [Name] -> Tm -> String
 showTm ns t = prettyTm 0 ns t []
 -- showTm ns t = show t
-
 -- deriving instance Show Tm
 instance Show Tm where show = showTm []
 

### 2020-02-18 paper.tex
stuff

@@
 \subsubsection{On soundness, completeness, and other properties}
-We briefly discuss what we can and cannot say about the presented
-elaborator. First, elaboration is \emph{sound} in the sense that it never
-produces malformed core syntax.
+We briefly discuss what we can say about the presented elaborator. First,
+elaboration is \emph{sound} in the sense that it never produces malformed core
+syntax.
 
 \begin{theorem}[Soundness of elaboration] The definitions of $\echeckblank$ and
-$\einferblank$ conform to the specification \ref{TODO}.
-This follows by induction on surface syntax and analyzing the definitions, while also relying on properties of substitution, metasubstitution, $\unify$ and $\mathsf{freshMeta}$.
-\qed
+$\einferblank$ conform to the specification \ref{TODO}. This follows by
+  induction on surface syntax, while also relying on the properties of
+  substitution, metasubstitution, $\mathsf{insert}$, $\unify$ and
+  $\mathsf{freshMeta}$. \qed
 \end{theorem}
@@
 
-\todo{Say something about termination of elab?}
+\todo{polymorphic subtyping, starting and ending elaboration}
 

### 2020-02-19 paper.tex
stuff

@@
 \usepackage{scalerel}
-\usepackage{cleveref}
+\usepackage{amssymb}
+%% \usepackage{cleveref}
+%% \usepackage{hyperref}
+%% \usepackage{nameref}
 
@@
 \newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}}
-\newcommand{\freshMeta}[3]{\mathsf{freshMeta}_{#1|#2}\,#3}
+\newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,_{#1|#2}\,#3}
 \newcommand{\unify}{\mathsf{unify}}
@@
 \newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,_{#2|#3}\,#4}
-\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,#2\,_{#3|#4}}
+\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,_{#3|#4}\,#2}
 \newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
@@
 \newcommand{\suc}{\mathsf{suc}}
+\newcommand{\Tel}{\mathsf{Tel}}
+\newcommand{\TCons}{\triangleright}
+\newcommand{\Rec}{\mathsf{Rec}}
+\newcommand{\ol}[1]{\overline{#1}}
 
@@
 There is significant literature on type inference in the presence of first-class
-polymorphic types, mainly in relation to GHC and ML-like languages \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy}.
+polymorphic types, mainly in relation to GHC and ML-like languages \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy,serrano2020a}.
 The above issue in Agda is a specific instance of the challenges described in
@@
     implicit function types, telescopes and \emph{strictly curried} function
-    types with telescope domain. The extensions to the target theory are modest
-    and are derivable from W-types and $\Sigma$-types. We use these extensions
-    to accummulate information about implicit insertions.
+    types with telescope domain. We use these extensions to accummulate
+    information about implicit insertions.
   \item Our algorithm is a conservative extension of Norell's bidirectional
@@
 
-  \inferrule*[lab=tm/metavar]
+  \inferrule*[lab=metavar]
              {\\}
@@
 
-  \inferrule*[lab=tm/bound-var]
+  \inferrule*[lab=bound-var]
              {\\}
@@
 
-  \inferrule*[lab=tm/defined-var]
+  \inferrule*[lab=defined-var]
              {\\}
@@
 
-  \inferrule*[lab=ty/u]
+  \inferrule*[lab=universe]
              {\\}
@@
 
-  \inferrule*[lab=ty/fun]
+  \inferrule*[lab=fun]
              {\Theta|\Gamma \vdash A : \U \\ \Theta|\Gamma,\,x : A \vdash B : \U}
@@
 
-  \inferrule*[lab=ty/implicit-fun]
+  \inferrule*[lab=implicit-fun]
              {\Theta|\Gamma \vdash A : \U \\ \Theta|\Gamma,\,x : A \vdash B : \U}
@@
 
-  \inferrule*[lab=tm/app]
+  \inferrule*[lab=app]
              {\Theta|\Gamma \vdash t : (x : A) \to B \\ \Theta|\Gamma\vdash u : A}
@@
 
-  \inferrule*[lab=tm/implicit-app]
+  \inferrule*[lab=implicit-app]
              {\Theta|\Gamma \vdash t : \{x : A\} \to B \\ \Theta|\Gamma\vdash u : A}
@@
 
-  \inferrule*[lab=tm/lam]
+  \inferrule*[lab=lam]
              {\Theta|\Gamma,\,x : A \vdash t : B}
@@
 
-  \inferrule*[lab=tm/implicit-lam]
+  \inferrule*[lab=implicit-lam]
              {\Theta|\Gamma,\,x : A \vdash t : B}
@@
 
-\todo{how to refer to mathpartir rules from text?}
-
 Metacontexts are used to record metavariables which are created during
@@
 \end{mathpar}
+For a simple example, assuming $\Theta :\equiv (\emptycon,\,\alpha : \U,\, \beta
+: \U)$, $\unify\,\alpha\,(\beta\to\beta)$ yields $\Theta'
+:\equiv (\emptycon,\,\beta : \U)$ and the substitution $\theta :\equiv (\alpha
+\mapsto (\beta\to\beta),\,\beta\mapsto \beta)$, where $\theta : \Theta' \To
+\Theta$.
 
@@
 algorithm does not support \emph{constraint postponing}, as we have not talked
-about constraints at all. In our concrete implementation, unification supports
-basic pattern unification and metavariable pruning.
-
-We describe a simple example. We have $\Theta :\equiv (\emptycon,\,\alpha :
-\U,\, \beta : \U)$, and we compute $\unify\,\alpha\,(\beta\to\beta)$. This
-yields $\Theta' :\equiv (\emptycon,\,\beta : \U)$ and the substitution $\theta
-:\equiv (\alpha \mapsto (\beta\to\beta),\,\beta\mapsto \beta)$, where $\theta :
-\Theta' \To \Theta$.
+about constraints at all. In our concrete prototype implementation, unification
+supports basic pattern unification and metavariable pruning.
 
@@
   \item We use pattern matching notation on core terms; e.g.\ we may match on
-    whether a type is a function type. This requires an evaluation/normalization
-    procedure on core terms, but note that we already assume this in implicit
+    whether a type is a function type. This assumes an evaluation/normalization
+    procedure on core terms; but note that we already assume this in implicit
     argument insertion and unification.
@@
 \begin{mathpar}
-\inferrule*[lab=checking]
+\inferrule*[lab=check]
            {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash A : \U}
@@
 
-\inferrule*[lab=inferring]
+\inferrule*[lab=infer]
            {ins \in \{\true,\,\false\} \\
@@
   &\echeck{\lambda\,x.\,t}{\Theta}{\Gamma}{((x : A)\to B)} :\equiv \edo \\
-  &\quad(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\quad(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{\Gamma,\,x:A} B\\
   &\quad\ereturn\,(\Theta',\,\theta,\,\lambda\,x.\,t')\\
   &\echeck{\lambda\,\{x\}.\,t}{\Theta}{\Gamma}{(\{x : A\}\to B)} :\equiv \edo \\
-  &\quad(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\quad(\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{\Gamma,\,x:A} B\\
   &\quad\ereturn\,(\Theta',\,\theta,\,\lambda\,\{x\}.\,t')\\
   &\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)} : \equiv \edo \\
-  &\quad (\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{(\Gamma,\,x:A)} B\\
+  &\quad (\Theta',\,\theta,\,t') \leftarrow \echeck{t}{\Theta}{\Gamma,\,x:A} B\\
   &\quad (\Theta',\,\theta,\,\lambda\,\{x\}.\,t')\\
@@
   &\quad(\Theta_2,\,\theta_2,\,t') \leftarrow \echeck{t}
-                {\Theta_1}{(\Gamma[\theta_1])}{A'}\\
+                {\Theta_1}{\Gamma[\theta_1]}{A'}\\
   &\quad(\Theta_3,\,\theta_3,\,u') \leftarrow
-                \echeck{u}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(B[\theta_1\circ\theta_2])}\\
+                \echeck{u}{\Theta_2}{\Gamma[\theta_1\circ\theta_2]}{(B[\theta_1\circ\theta_2])}\\
   &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u')\\
@@
   &\einfer{x}{ins}{\Theta}{\Gamma} :\equiv \edo\\
-  &\quad\eif\,(x : A \in \Gamma)\,\lor\,(x : A = t \in \Gamma) \\
+  &\quad\eif\,((x : A) \in \Gamma)\,\lor\,((x : A = t) \in \Gamma) \\
   &\qquad   \ethen\, \ereturn\,(\einsert{ins}{x}{A})\\
@@
   &\quad(\Theta_2,\,\theta_2,\,B') \leftarrow
-                 \echeck{B}{\Theta_2}{(\Gamma[\theta_1],\,x : A')}{\U}\\
+                 \echeck{B}{\Theta_2}{\Gamma[\theta_1],\,x : A'}{\U}\\
   &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,
@@
   &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
-  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{(\Gamma[\theta_1],\,x:A)}\\
+  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
   &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\
@@
   &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
-  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{(\Gamma[\theta_1],\,x:A)}\\
+  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
   &\quad\ereturn\,(\einsert{ins}{(\lambda\,\{x\}.\,t')}{(\{x : A[\theta_2]\}\to B)})\\
@@
   &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,((x : A_0[\theta_3])\to A_1)\\
-  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{(\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4])}{(A_0[\theta_3\circ\theta_4])}\\
+  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4]}{(A_0[\theta_3\circ\theta_4])}\\
   &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,u')}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\
@@
   &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,(\{x : A_0[\theta_3]\}\to A_1)\\
-  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{(\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4])}{(A_0[\theta_3\circ\theta_4])}\\
+  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4]}{(A_0[\theta_3\circ\theta_4])}\\
   &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,\{u'\})}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\
@@
   &\quad(\Theta_2,\,\theta_2,\,t') \leftarrow \echeck{t}
-                {\Theta_1}{(\Gamma[\theta_1])}{A'}\\
+                {\Theta_1}{\Gamma[\theta_1]}{A'}\\
   &\quad(\Theta_3,\,\theta_3,\,u',\,B) \leftarrow
-                \einfer{u}{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}\\
+                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_1\circ\theta_2]}\\
   &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,(\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u'),\,B)\\
@@
 $\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma}$ cases are again interesting. Here, we
-first infer a type for the term which is being applied, and only proceed if the
-type is an appropriate function. Note a difference between the explicit and
-implicit case. In the former case, we use $\einfer{t}{\true}{\Theta_0}{\Gamma}$,
-which inserts implicit applications. In the latter case we do not insert
-implicit applications. This, together with the inference definition for
-variables, ensures that implicit applications in the surface syntax behave
-similarly as in Agda. For example, given $id : \{A : \U\}\to A\to A$ in scope,
-we elaborate $id\,\U$ as follows:
+first infer a type for the term which is being applied, then refine the type to
+a function type, and lastly check the argument with the domain type.  Note the
+difference between the explicit and implicit case. In the former case, we use
+$\einfer{t}{\true}{\Theta_0}{\Gamma}$, which inserts implicit applications. In
+the latter case we do no insertion.  This, together with the inference
+definition for variables, ensures that implicit applications in the surface
+syntax behave similarly as in Agda. For example, given $id : \{A : \U\}\to A\to
+A$ in scope, we elaborate $id\,\U$ as follows:
 \begin{enumerate}
@@
 \end{enumerate}
-\todo{explain that we fail if $t$ has meta-headed type or handle it in elab}
 
@@
 
-\subsubsection{On soundness, completeness, and other properties}
-We briefly discuss what we can say about the presented elaborator. First,
-elaboration is \emph{sound} in the sense that it never produces malformed core
-syntax.
-
-\begin{theorem}[Soundness of elaboration] The definitions of $\echeckblank$ and
-$\einferblank$ conform to the specification \ref{TODO}. This follows by
-  induction on surface syntax, while also relying on the properties of
-  substitution, metasubstitution, $\mathsf{insert}$, $\unify$ and
-  $\mathsf{freshMeta}$. \qed
+\subsubsection{Starting and finishing elaboration} Given a surface term $t$,
+we do elaboration by computing $\einfer{t}{\true}{\emptycon}{\emptycon}$. If this
+succeeds, we get a $(\Theta,\,\theta,t')$ result. We say that elaboration succeeds
+if $\Theta = \emptycon$ and $\theta = \id$, i.e.\ no unsolved metas remain.
+
+\subsubsection{Properties of elaboration}
+First, elaboration is \emph{sound} in the sense that it only produces well-formed
+output.
+
+\begin{theorem}[Soundness] The definitions of $\echeckblank$ and
+$\einferblank$ conform to the \LabTirName{check} and \LabTirName{infer}
+  specifications. This follows by induction on surface syntax, while also
+  relying on the properties of substitution, metasubstitution,
+  $\mathsf{insert}$, $\unify$ and $\mathsf{freshMeta}$. \qed
 \end{theorem}
@@
 
-\todo{polymorphic subtyping, starting and ending elaboration}
-
-Also, note that we do not support \emph{let-generalization}. This is an open
+\subsubsection{Omitted features}
+
+\begin{itemize}
+  \item \emph{Let-generalization}. This is an open
     research topic in settings with dependent types, and we make no attempt at
@@
     dependent version of Haskell.
+  \item \emph{Polymorphic subtyping}. In some prior works, e.g.\ in
+    \cite{dunfield2013complete, vytiniotis2008fph}, there is a subtyping
+    relation arising along instantiations of polymorphic types. In GHC 8
+    polymorphic subtyping is implemented for function types only. Polymorphic
+    subtyping complicates type inference, and to our knowledge it has not been
+    researched or implemented in dependently typed settings. Hence we omit it as
+    well. GHC also seems to abandon polymorphic subtyping in recent developments
+    of impredicative inference \cite{serrano2020a}.
+
+\end{itemize}
 
@@
 \item $\echeck{t}{\Theta}{\Gamma}{A}$ can insert a $\lambda$ only if $A$ is an implicit function type.
-\item $\einsert{\true}{\Theta}{\Gamma}{A}$ inserts an application only if $A$ is an implicit function type.
+\item $\einsert{\true}{A}$ inserts an application only if $A$ is an implicit function type.
 \end{enumerate}
@@
 
+As part of the proposed solution, we extend the core theory with telescopes
+and strictly curried functions.
+
+\begin{figure}[h]
+\begin{mathpar}
+  \inferrule*[lab=tel]
+             {\\}
+             {\Theta|\Gamma \vdash \Tel : \U}
+
+  \inferrule*[lab=empty-tel]
+             {\\}
+             {\Theta|\Gamma \vdash \epsilon : \Tel}
+
+  \inferrule*[lab=nonempty-tel]
+             {\Theta|\Gamma \vdash A : \U \\ \Theta|\Gamma,\,x : A \vdash B : \Tel}
+             {\Theta|\Gamma \vdash (x : A) \TCons B : \Tel}
+
+  \inferrule*[lab=record-type]
+             {\Theta|\Gamma \vdash A : \Tel}
+             {\Theta|\Gamma \vdash \Rec\,A : \U}
+
+  \inferrule*[lab=empty-rec]
+             {\\}
+             {\Theta|\Gamma \vdash [] : \Rec\,\epsilon}
+
+  \inferrule*[lab=nonempty-rec]
+             {\Theta|\Gamma \vdash t : A \\ \Theta|\Gamma \vdash u : \Rec\,(B[x\mapsto t])}
+             {\Theta|\Gamma \vdash t :: u : \Rec\,((x : A) \TCons B)}
+
+  \inferrule*[lab=rec-projection]
+             {\Theta|\Gamma \vdash t : \Rec\,((x : A) \TCons B)}
+             {\Theta|\Gamma \vdash \pi_1\,t : A \\
+               \Theta|\Gamma \vdash \pi_2\,t : \Rec\,(B[x\mapsto \pi_1\,t])}
+\end{mathpar}
+
+\begin{alignat*}{3}
+  & \text{\LabTirName{$\pi_1$-$\beta$}}\hspace{2em}  && \pi_1\,(t :: u) && \equiv t\\
+  & \text{\LabTirName{$\pi_2$-$\beta$}}  && \pi_2\,(t :: u) && \equiv u\\
+  & \text{\LabTirName{$::$-$\eta$}}  && (\pi_1\,t :: \pi_2\,t) && \equiv t\\
+  & \text{\LabTirName{${[}{]}$-$\eta$}}  && t && \equiv []
+\end{alignat*}
+
+  %% \inferrule*[lab=rec-projection]
+  %%            {\Theta|\Gamma \vdash t : \Rec\,((x : A) \TCons B)}
+  %%            {\Theta|\Gamma \vdash \pi_1\,t : A \\ \Theta|\Gamma \vdash \pi_2\,t : \Rec\,(B[x\mapsto \pi_1\,t])}
+
+  %% \inferrule*[lab=projection-$\beta$]
+  %%            {\\}
+  %%            {\pi_1\,(t :: u) \equiv t \\ \pi_2\,(t :: u) \equiv u}
+
+  %% \inferrule*[lab=$::$-$\eta$]
+  %%            {\\}
+  %%            {(\pi_1\,t :: \pi_2\,t) \equiv t}
+
+  %% \inferrule*[lab=${[}{]}$-$\eta$]
+  %%            {\Theta|\Gamma \vdash t : \Rec\,\epsilon}
+  %%            {\Theta|\Gamma \vdash t \equiv []}
+
+\caption{Rules for telescopes.}
+\label{fig:telescopes}
+\end{figure}
+
+\begin{figure}[h]
+\begin{mathpar}
+  \inferrule*[lab=curried-fun]
+             {\Theta|\Gamma \vdash A : \Tel \\ \Theta|\Gamma,\,x : \Rec\,A \vdash B : \U}
+             {\Theta|\Gamma \vdash \{x : \ol{A}\} \to B : \U}
+
+  \inferrule*[lab=curried-lam]
+             {\Theta|\Gamma,\,x : \Rec\,A \vdash t : B}
+             {\Theta|\Gamma \vdash \lambda\,\{x : \ol{A}\}.\,t : \{x : \ol{A}\} \to B}
+
+  \inferrule*[lab=curried-app]
+             {\Theta|\Gamma \vdash t : \{x : \ol{A}\} \to B \\ \Theta|\Gamma \vdash u : \Rec\,A}
+             {\Theta|\Gamma \vdash t\,\{u : \ol{A}\} : B[x \mapsto u]}
+\end{mathpar}
+
+\begin{alignat*}{3}
+  & \text{\LabTirName{fun-$\epsilon$}}\hspace{3em} && \{x : \ol{\epsilon}\}\to B && \equiv B[x \mapsto []]\\
+  & \text{\LabTirName{fun-$\TCons$}}         && \{x : \ol{(y : A) \TCons B}\}\to C \quad && \equiv
+                                               \{y : A\}\to(\{b : \ol{B}\}\to B[x \mapsto (y :: b)])\\
+  & \text{\LabTirName{lam-$\epsilon$}}       && \lambda\,\{x : \ol{\epsilon}\}.\,t && \equiv t[x \mapsto []]\\
+  & \text{\LabTirName{lam-$\TCons$}}         && \lambda\,\{x : \ol{(y : A)\TCons B}\}.\,t && \equiv
+                                         \lambda\{y\}.\,\lambda\,\{b : \ol{B}\}.\,t[x \mapsto (y :: b)]\\
+  & \text{\LabTirName{app-$\epsilon$}}       && t\,\{u : \ol{\epsilon}\} && \equiv t\\
+  & \text{\LabTirName{app-$\TCons$}}         && t\,\{u : \ol{(x : A) \TCons B}\} &&\equiv
+                                         t\,\{\pi_1\,u\}\,\{\pi_2\,u : \ol{B[x \mapsto \pi_1\,u]}\}\\
+  & \text{\LabTirName{curried-$\beta$}} && \lambda\,(\{x : \ol{A}\}.\,t)\,\{u:\ol{A}\} &&\equiv
+                                                             t[x \mapsto u]\\
+  & \text{\LabTirName{curried-$\eta$}}  && \lambda\,\{x : \ol{A}\}.\,t\,\{x : \ol{A}\} && \equiv t
+\end{alignat*}
+
+\caption{Rules for strictly curried functions.}
+\label{fig:curriedfunctions}
+\end{figure}
+
 \begin{acks}

### 2020-02-19 implicit-fun-elab/Elaboration.hs

@@ -307,8 +307,8 @@ unify cxt l r =
 
     go t t' = case (force t, force t') of
       (VLam x _ a t, VLam _ _ _ t')            -> goBind x a t t'
-      (VLam x i a t, t')                       -> goBind x a t (\v -> vApp t' v i)
-      (t, VLam x' i' a' t')                    -> goBind x' a' (\v -> vApp t v i') t'
+      (VLam x i a t, t')                       -> goBind x a t (\ ~v -> vApp t' v i)
+      (t, VLam x' i' a' t')                    -> goBind x' a' (\ ~v -> vApp t v i') t'
       (VPi x i a b, VPi x' i' a' b') | i == i' -> go a a' >> goBind x a b b'
       (VU, VU)                                 -> pure ()
       (VTel, VTel)                             -> pure ()
@@ -317,8 +317,10 @@ unify cxt l r =
       (VTCons x a b, VTCons x' a' b')          -> go a a' >> goBind x a b b'
       (VTempty, VTempty)                       -> pure ()
       (VTcons t u, VTcons t' u')               -> go t t' >> go u u'
-      (VPiTel x a b, VPiTel x' a' b')          -> go a a' >> goBind x a b b'
-      (VLamTel x a t, VLamTel x' a' t')        -> goBind x a t t'
+      (VPiTel x a b, VPiTel x' a' b')          -> go a a' >> goBind x (VRec a) b b'
+      (VLamTel x a t, VLamTel x' a' t')        -> goBind x (VRec a) t t'
+      -- (VLamTel x a t, t')                      -> goBind x (VRec a) t (vAppTel a t')
+      -- (t, VLamTel x' a' t')                    -> goBind x' (VRec a') (vAppTel a' t) t'
       (VNe h sp, VNe h' sp') | h == h'         -> goSp (forceSp sp) (forceSp sp')
       (VNe (HMeta m) sp, t')                   -> solveMeta cxt m sp t'
       (t, VNe (HMeta m') sp')                  -> solveMeta cxt m' sp' t

### 2020-02-19 implicit-fun-elab/Evaluation.hs

@@ -80,8 +80,8 @@ vAppTel a ~t ~u = case force a of
   VTEmpty       -> t
   VTCons _ a as -> let u1 = vProj1 u in vAppTel (as u1) (vApp t u1 Impl) (vProj2 u)
   a             -> case t of
-                     VLamTel _ _ t -> t u
                      VNe h sp      -> VNe h (SAppTel a sp u)
+                     VLamTel _ _ t -> t u
                      _             -> error "impossible"
 
 vAppSp :: Val -> Spine -> Val

### 2020-02-19 implicit-fun-elab/Main.hs

@@ -70,6 +70,14 @@ pruneTest = main' "elab" $ unlines [
   "λ f (x : Bool). not (f x)"
   ]
 
+pruneTest2 = main' "elab" $ unlines [
+ "λ (ℕ : U)(zero : ℕ)(suc : ℕ → ℕ)(add : ℕ → ℕ → ℕ).",
+ "let test1 = (P : _) → (n : ℕ) → P n → P n in",
+ "let test2 = λ f x . add (suc (f x)) x in",
+ "U"
+ ]
+
+
 test0 = main' "elab" $ unlines [
   "let IdTy = {A} → A → A in",
   "let f = λ (g:IdTy). g g in",
@@ -148,7 +156,7 @@ test = main' "elab" $ unlines [
 
   "let compExample = comp (cons true) (cons false) nil in",
 
-  -- "let t5 = λ xs. poly (head xs) in", -- pruning
+  -- "let t5 = λ xs. poly (head xs) in",
 
   "hundred"
   ]

### 2020-02-19 paper.tex
stuff

@@
 \newcommand{\Set}{\mathsf{Set}}
+\renewcommand{\tt}{\mathsf{tt}}
 
@@
 \end{enumerate}
+Note that it would be more efficient (and also allow more user-friendly error
+messages) to not immediately refine the inferred type of $t$ to a function type,
+but rather match on the inferred type, and only perform refining when the type
+is meta-headed. We present the unoptimized version here for the sake of brevity.
 
@@
 \subsubsection{Starting and finishing elaboration} Given a surface term $t$,
-we do elaboration by computing $\einfer{t}{\true}{\emptycon}{\emptycon}$. If this
-succeeds, we get a $(\Theta,\,\theta,t')$ result. We say that elaboration succeeds
+we initiate elaboration by computing $\einfer{t}{\true}{\emptycon}{\emptycon}$. If this
+succeeds, we get a $(\Theta,\,\theta,t')$ result. Elaboration is successful overall
 if $\Theta = \emptycon$ and $\theta = \id$, i.e.\ no unsolved metas remain.
@@
     subtyping complicates type inference, and to our knowledge it has not been
-    researched or implemented in dependently typed settings. Hence we omit it as
-    well. GHC also seems to abandon polymorphic subtyping in recent developments
-    of impredicative inference \cite{serrano2020a}.
-
+    implemented in any dependently typed setting. Hence we omit it as well. GHC
+    also seems to abandon polymorphic subtyping in recent developments of
+    impredicative inference \cite{serrano2020a}.
 \end{itemize}
@@
 
-As part of the proposed solution, we extend the core theory with telescopes
-and strictly curried functions.
+As part of the proposed solution, we extend the core theory with telescopes and
+strictly curried functions. Figure \ref{fig:telescopes} lists the typing rules and
+definitional equalities.
+
+\subsection{Telescopes}
+Telescopes can be viewed as a generic implementation of record types. We have
+$\Tel$ as the type of telescopes. Elements of $\Tel$ are right-nested telescopes
+of types, with $\epsilon$ denoting the empty telescope, and $\blank\TCons\blank$
+telescope extension. For example, we can define the signature of natural number
+algebras as follows:
+\begin{alignat*}{3}
+  & \slet\,NatAlgSig : \Tel = (N : \U) \TCons (zero : N) \TCons (suc : N \to N) \TCons \epsilon\,\sin\,...
+\end{alignat*}
+We interpret an $A : \Tel$ as a record type as $\Rec\,A$, which behaves as the
+evident iterated $\Sigma$-type corresponding to the telescope. Hence,
+$\Rec\,\epsilon$ is isomorphic to the unit type, with inhabitant ${[}{]}$, and
+$\Rec\,((x : A) \TCons B)$ behaves as a $\Sigma$-type, with pairing constructor
+$\blank::\blank$ and projections $\pi_1$ and $\pi_2$. We also have the $\beta$
+and $\eta$ rules for record constructors and projections in Figure
+\ref{fig:telescopes}. We present definitional equalities in a compact form, but
+note that they still stand for $\boxed{\Theta|\Gamma\vdash t \equiv u : A}$
+judgments. Hence, the sides of the equations must have the same types, and in
+particular the left side of the \LabTirName{${[}{]}$-$\eta$} rule has type
+$\Rec\,\epsilon$.
+
+Our telescopes and records are derivable from natural numbers, the unit type and
+$\Sigma$-types. We use Agda-like pattern matching notation in the following.
+First, we define length-indexed telescopes.
+\begin{alignat*}{3}
+  & \rlap{$\Tel' : \Nat \to \U$}\\
+  & \Tel'\,\zero      && :\equiv \top\\
+  & \Tel'\,(\suc\,n)  && :\equiv \Sigma(A : \U).\,(A \to \Tel'\,n)
+\end{alignat*}
+Then, we have $\Tel :\equiv \Sigma(n : \Nat).\,(\Tel'\,n)$, and define records:
+\begin{alignat*}{3}
+  & \rlap{$\Rec : \Tel \to \U$}\\
+  & \Rec\,(\zero,\,\_)        &&:\equiv \top\\
+  & \Rec\,(\suc\,n,\,(A,\,B)) &&:\equiv \Sigma(a : A).\,(\Rec\,(n,\,B\,a))
+\end{alignat*}
+From the above, $\epsilon$, $\blank\TCons\blank$, $\blank::\blank$ and ${[}{]}$
+are evident, and all expected equalities hold definitionally. Derivability is
+good news because we inherit nice properties of the type theory which only
+contains the base type formers. For instance, if we have a consistent universe
+setup, we inherit consistency, canonicity and decidability of conversion. We
+currently use native telescopes instead of $\Nat$, $\top$ and $\Sigma$ because
+in unification and elaboration it is convenient that we are able to restrict
+some types to records types.
 
@@
 
-  \inferrule*[lab=empty-rec]
+  \inferrule*[lab=empty-record]
              {\\}
@@
 
-  \inferrule*[lab=nonempty-rec]
+  \inferrule*[lab=nonempty-record]
              {\Theta|\Gamma \vdash t : A \\ \Theta|\Gamma \vdash u : \Rec\,(B[x\mapsto t])}
@@
 
-  \inferrule*[lab=rec-projection]
+  \inferrule*[lab=record-projection]
              {\Theta|\Gamma \vdash t : \Rec\,((x : A) \TCons B)}
@@
                \Theta|\Gamma \vdash \pi_2\,t : \Rec\,(B[x\mapsto \pi_1\,t])}
-\end{mathpar}
-
-\begin{alignat*}{3}
-  & \text{\LabTirName{$\pi_1$-$\beta$}}\hspace{2em}  && \pi_1\,(t :: u) && \equiv t\\
-  & \text{\LabTirName{$\pi_2$-$\beta$}}  && \pi_2\,(t :: u) && \equiv u\\
-  & \text{\LabTirName{$::$-$\eta$}}  && (\pi_1\,t :: \pi_2\,t) && \equiv t\\
-  & \text{\LabTirName{${[}{]}$-$\eta$}}  && t && \equiv []
-\end{alignat*}
-
-  %% \inferrule*[lab=rec-projection]
-  %%            {\Theta|\Gamma \vdash t : \Rec\,((x : A) \TCons B)}
-  %%            {\Theta|\Gamma \vdash \pi_1\,t : A \\ \Theta|\Gamma \vdash \pi_2\,t : \Rec\,(B[x\mapsto \pi_1\,t])}
-
-  %% \inferrule*[lab=projection-$\beta$]
-  %%            {\\}
-  %%            {\pi_1\,(t :: u) \equiv t \\ \pi_2\,(t :: u) \equiv u}
-
-  %% \inferrule*[lab=$::$-$\eta$]
-  %%            {\\}
-  %%            {(\pi_1\,t :: \pi_2\,t) \equiv t}
-
-  %% \inferrule*[lab=${[}{]}$-$\eta$]
-  %%            {\Theta|\Gamma \vdash t : \Rec\,\epsilon}
-  %%            {\Theta|\Gamma \vdash t \equiv []}
-
-\caption{Rules for telescopes.}
-\label{fig:telescopes}
-\end{figure}
-
-\begin{figure}[h]
-\begin{mathpar}
+
   \inferrule*[lab=curried-fun]
@@
 \begin{alignat*}{3}
+  & \text{\LabTirName{$\pi_1$-$\beta$}}\hspace{2em}  && \pi_1\,(t :: u) && \equiv t\\
+  & \text{\LabTirName{$\pi_2$-$\beta$}}  && \pi_2\,(t :: u) && \equiv u\\
+  & \text{\LabTirName{$::$-$\eta$}}  && (\pi_1\,t :: \pi_2\,t) && \equiv t\\
+  & \text{\LabTirName{${[}{]}$-$\eta$}}  && t && \equiv []\\
   & \text{\LabTirName{fun-$\epsilon$}}\hspace{3em} && \{x : \ol{\epsilon}\}\to B && \equiv B[x \mapsto []]\\
   & \text{\LabTirName{fun-$\TCons$}}         && \{x : \ol{(y : A) \TCons B}\}\to C \quad && \equiv
-                                               \{y : A\}\to(\{b : \ol{B}\}\to B[x \mapsto (y :: b)])\\
+                                               \{y : A\}\to(\{b : \ol{B}\}\to C[x \mapsto (y :: b)])\\
   & \text{\LabTirName{lam-$\epsilon$}}       && \lambda\,\{x : \ol{\epsilon}\}.\,t && \equiv t[x \mapsto []]\\
@@
 
-\caption{Rules for strictly curried functions.}
-\label{fig:curriedfunctions}
+\caption{Rules for telescopes and strictly curried functions.}
+\label{fig:telescopes}
 \end{figure}
+
+\subsection{Strictly Curried Functions}
+
+These are function types whose domains are telescopes, and they are immediately
+computed to iterated implicit function types when the domain telescope is known
+to be empty or non-empty. See \LabTirName{fun-$\epsilon$} and
+\LabTirName{fun-$\TCons$}: a curried function with empty domain computes to
+simply the codomain, while a function with a non-empty domain computes to an
+implicit function type. We explicitly notate telescopes in both
+$\lambda$-abstractions and applications for strictly curried functions, since
+they are relevant in the computation rules.
+
+Curried function types tend to be computed away, but they can persist if the
+domain telescope is neutral, and in particular when it is meta-headed. For
+example, assuming a meta $\alpha : \Tel$, the type $\{x : \ol{\alpha}\}\to B$
+cannot be computed further. During elaboration, we will use strictly curried
+function types to represent unknown insertions, but these types are eventually
+computed away if a surface expression can be successfully elaborated. Since the
+surface language remains unchanged, telescopes and curried functions are merely
+an internal implementation detail from the perspective of programmers.
+
+Curried functions are \emph{mostly} derivable from $\Nat$, $\top$ and $\Sigma$.
+\begin{alignat*}{3}
+& \rlap{$\Pi^C : (A : \Tel) \to (\Rec\,A \to \U) \to \U$}\\
+& \Pi^C (\zero,\,\_)\,B && :\equiv B\,\tt\\
+& \Pi^C (\suc\,n,\,(A,\,B))\,C && :\equiv \{a : A\}\to \Pi^C\,(n,\,B\,a) (\lambda\,b.\,C\,(a,\,b))
+\end{alignat*}
+
+%% Π : (Δ : Tel) → (Rec Δ → Set) → Set
+%% Π ε       B = B tt
+%% Π (A ▶ Δ) B = (x : A) → Π (_ , Δ x) λ δ → B (x , δ)
+
+%% app : ∀ {Δ B} → Π Δ B → (δ : Rec Δ) → B δ
+%% app {ε}     f x         = f
+%% app {A ▶ Δ} f (x₀ , x₁) = app (f x₀) x₁
+
+%% lam : ∀ {Δ B} → ((δ : Rec Δ) → B δ) → Π Δ B
+%% lam {ε}     f = f tt
+%% lam {A ▶ Δ} f = λ x → lam λ δ → f (x , δ)
 

### 2020-02-20 paper.tex
stuff

@@
 Curried functions are \emph{mostly} derivable from $\Nat$, $\top$ and $\Sigma$.
+The type former is defined as follows:
 \begin{alignat*}{3}
@@
 \end{alignat*}
+With this, we can also define $\mathsf{app} : \Pi^C\,A\,B \to (a : \Rec\,A) \to
+B\, a$ and $\mathsf{lam} : ((a : \Rec\,A) \to B\, a) \to \Pi^C\,A\,B$, and all
+equations in Figure \ref{fig:telescopes} hold definitionally, except
+\LabTirName{curried-$\beta$} and \LabTirName{curried-$\eta$}. These do not hold
+strictly, because $\Pi^C$, $\mathsf{app}$ and $\mathsf{lam}$ are all defined by
+recursion on the $A$ telescope, but the $\beta\eta$ rules are specified
+generically for an arbitrary (possibly neutral)
+telescope. \LabTirName{curried-$\beta$} is still provable as a propositional
+equality, and assuming function extensionality \LabTirName{curried-$\eta$} is
+provable as well.
 

### 2020-02-20 paper.tex
stuff

@@
     subtyping complicates type inference, and to our knowledge it has not been
-    implemented in any dependently typed setting. Hence we omit it as well. GHC
-    also seems to abandon polymorphic subtyping in recent developments of
-    impredicative inference \cite{serrano2020a}.
+    implemented in any dependently typed setting. We also believe that it is an
+    anti-feature in dependent settings, because the elaboration of subtyping
+    must insert coercions which significantly change the intensional character
+    of programs. For example, if we have covariant list types, then coercing $t
+    : List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
+    mapping over $t$ and inserting implicit applications to $Bool$ for each list
+    element. In System F, this is harmless, since types are computationally
+    irrelevant, but in our core syntax we have implicit functions with arbitrary
+    (relevant) domains. In GHC, subtyping coercions for functions change
+    operational semantics; this is a reason for abandoning subtyping in recent
+    developments of impredicative inference for GHC \cite{serrano2020a}.
 \end{itemize}
@@
 contains the base type formers. For instance, if we have a consistent universe
-setup, we inherit consistency, canonicity and decidability of conversion. We
+setup\footnote{Recall that we currently assume type-in-type, which causes consistency and normalization to fail.}, we inherit consistency, canonicity and decidability of conversion. We
 currently use native telescopes instead of $\Nat$, $\top$ and $\Sigma$ because
@@
 equality, and assuming function extensionality \LabTirName{curried-$\eta$} is
-provable as well.
+provable as well. For details, see our Agda formalization of these definitions,
+which is included alongside the prototype implementation.
+
+Hence, we can derive a somewhat weaker version of curried functions, with
+propositional $\beta$ and $\eta$. From this, we still get consistency almost for
+free. That is because whenever we have a model of the base theory with $\Nat$,
+$\top$ and $\Sigma$, such that the model also validates equality reflection,
+then every propositionally provable equation is also validated as a definitional
+equation. And fortunately for us, standard models which prove consistency
+usually validate equality reflection, e.g.\ set-theoretical models or standard
+models in extensional type theory.
+
+In contrast, showing canonicity, normalization and decidability of conversion
+would require some extra work. We leave this to future work, but we expect that
+it is straightforward to extend previous proofs to cover strict $\beta$ and
+$\eta$ for curried functions.
 

### 2020-02-20 implicit-fun-elab/Main.hs

@@ -265,6 +265,7 @@ ex1 = main' "elab" $ unlines [
   "let t4 = id poly (λ x. x) in",
 
   "let foo = head ids true in",
+  "let kek = λ x. cons x ids in",
   -- "let t5 = λ xs. poly (head xs) in", -- pruning
   -- "let t6 = length ids in",
   -- "let t7 = append (single inc) (single id) in",

### 2020-02-20 paper.tex
stuff

@@
 
-\begin{definition}[Closing type] For each $\Theta|\Gamma \vdash A : \U$, we define
+\begin{definition}[Closing type]\label{def:closingtype} For each $\Theta|\Gamma \vdash A : \U$, we define
 $\Gamma \To A$ by recursion on $\Gamma$, such that $\Theta|\emptycon\vdash
@@
   &\einfer{x}{ins}{\Theta}{\Gamma} :\equiv \edo\\
-  &\quad\eif\,((x : A) \in \Gamma)\,\lor\,((x : A = t) \in \Gamma) \\
+  &\quad\eif\,(\Gamma = (\Gamma_0,\,x : A,\,\Gamma_1))\,\lor\,(\Gamma = (\Gamma_0,\,x : A = t,\,\Gamma_1)) \\
   &\qquad   \ethen\, \ereturn\,(\einsert{ins}{x}{A})\\
@@
 
-\begin{theorem}[Conservativity]
+\begin{theorem}[Conservativity]\label{thm:conservativity}
 Elaboration is conservative over the surface syntax, in the sense that for any
@@
 \begin{itemize}
-  \item \emph{Let-generalization}. This is an open
-    research topic in settings with dependent types, and we make no attempt at
-    tackling it. See \cite{eisenberg2016dependent} for a treatment in a proposed
-    dependent version of Haskell.
+  \item \emph{Let-generalization}. This is an open research topic in settings
+    with dependent types, and we make no attempt at covering it. See
+    \cite{eisenberg2016dependent} for a treatment in a proposed dependent
+    version of Haskell.
   \item \emph{Polymorphic subtyping}. In some prior works, e.g.\ in
@@
     implemented in any dependently typed setting. We also believe that it is an
-    anti-feature in dependent settings, because the elaboration of subtyping
-    must insert coercions which significantly change the intensional character
-    of programs. For example, if we have covariant list types, then coercing $t
-    : List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
+    anti-feature in dependent settings, because elaboration of subtyping must
+    insert coercions which significantly change the intensional character of
+    programs. For example, if we have covariant list types, then coercing $t :
+    List\,(\{A : \U\}\to A \to A)$ to $t : List\,(Bool \to Bool)$ requires
     mapping over $t$ and inserting implicit applications to $Bool$ for each list
-    element. In System F, this is harmless, since types are computationally
-    irrelevant, but in our core syntax we have implicit functions with arbitrary
-    (relevant) domains. In GHC, subtyping coercions for functions change
-    operational semantics; this is a reason for abandoning subtyping in recent
-    developments of impredicative inference for GHC \cite{serrano2020a}.
+    element. In System F, all such coercions are erasible, since types are
+    computationally irrelevant, but in our core syntax we have implicit
+    functions with arbitrary (relevant) domains. In GHC, subtyping coercions for
+    functions change operational semantics; this is a reason for abandoning
+    subtyping in recent developments of impredicative inference for GHC
+    \cite{serrano2020a}.
 \end{itemize}
@@
 \section{Issues with First-Class Implicit Functions}
+\label{sec:issues}
 
@@
 
-%% Π : (Δ : Tel) → (Rec Δ → Set) → Set
-%% Π ε       B = B tt
-%% Π (A ▶ Δ) B = (x : A) → Π (_ , Δ x) λ δ → B (x , δ)
-
-%% app : ∀ {Δ B} → Π Δ B → (δ : Rec Δ) → B δ
-%% app {ε}     f x         = f
-%% app {A ▶ Δ} f (x₀ , x₁) = app (f x₀) x₁
-
-%% lam : ∀ {Δ B} → ((δ : Rec Δ) → B δ) → Π Δ B
-%% lam {ε}     f = f tt
-%% lam {A ▶ Δ} f = λ x → lam λ δ → f (x , δ)
+\section{Extending Elaboration}
+
+We shall utilize the extended core theory to implement smarter
+elaboration. Recall from Section \ref{sec:issues} that the old elaborator makes
+two kinds of unforced insertion choices:
+\begin{enumerate}
+\item $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
+\item $\einsert{\true}{A}$ does not insert an implicit application if $A$ is meta-headed.
+\end{enumerate}
+In the following, we shall only handle the case for $\lambda$ insertions. This
+allows a simple implementation which only requires minimal changes to
+unification, and which is already remarkably powerful. It seems that handling
+implicit application insertions requires extending unification; we will talk
+about this in Section \ref{TODO}. But before delving into elaboration, we need to
+rectify a curious issue.
+
+\subsection{Superfluous Implicit Functions}
+
+Assuming $Bool : \U$, $true : Bool$ and $false : Bool$, consider the following surface expression:
+\[
+  \slet\,x : \_ = true\,\sin\,x
+\]
+What should this expression elaborate to? We would expect the result to be simply
+\[
+  \slet\,x : Bool = true\,\sin\,x
+\]
+However, there are infinitely many core terms which are conservative over the
+surface expression in the sense of Theorem \ref{thm:conservativity}. That is, we
+can wrap definitions with any number of implicit $\lambda$-s, and add implicit
+applications accordingly to usage sites of the defined name. For example, we could have
+\[
+  \slet\,x : \{y : Bool\} \to Bool = \lambda\,\{y\}.\,true\,\sin\,x\,\{true\}
+\]
+This is clearly undesirable. With the type $\{y : Bool\} \to Bool$, the implicit
+argument $y$ is never inferable, because the codomain type does not depend on
+the domain, and the argument is never constrained. Hence, with the above
+definition, we always have to write $x\,\{true\}$ or $x\,\{false\}$ when we want
+to use $x$. To rule out such nonsense, we adopt the following principle:
+
+\begin{quote}
+  \emph{Elaboration should never invent non-dependent implicit function types.}
+\end{quote}
+
+This was a non-issue in the old elaborator, because it was not able to invent
+implicit function types; it was only utilizing the type annotations present in
+the surface input. With $\slet\,x : \_ = true$, the old elaborator checks $true$
+with a fresh meta, and just assumes that the meta does not stand for an
+implicit function type.
+
+To implement our principle, we need \emph{constancy constraints}. These are
+constraints in the sense of GHC's OutsideIn(X) \cite{vytiniotis2011outsidein}
+and also in the sense of the implementation in Agda and Coq. Rules for constancy
+constraints are as follows.
+
+\begin{mathpar}
+  \inferrule*[lab=metacon/constr]
+             {\Theta \vdash \alpha : \Gamma \To \Tel \\ \Theta|\Delta,\,x : \Rec\,(\alpha\,\ol{t}) \vdash B : \U}
+             {\Theta,\,\mathsf{constancy}_{\Delta,\,x : \Rec\,(\alpha\,\ol{t})}\,B\vdash}
+
+  \inferrule*[lab=constr/solve]
+             {\Theta_0 \vdash \alpha : \Gamma_0 \To \Tel \\ x \notin FreeVars(B)}
+             {\Theta_0,\,\mathsf{constancy}_{\Delta,\,x : \Rec\,(\alpha\,\ol{t})}\,B,\,\Theta_1|\Gamma_1
+               \vdash \alpha \equiv \lambda\,\overline{\mathsf{vars}_{\Gamma_0}}.\,\epsilon}
+\end{mathpar}
+
+In the rule \LabTirName{metacon/constr}, we allow extension of the metacontext
+with a constraint.  In the constraint, the metavariable $\alpha : \Gamma \To
+\Tel$ represents an unknown domain for a curried function type, and $B$ is the
+codomain of the function. Here, $\Gamma \To \Tel$ refers to a closing type, which
+was defined in Section \ref{def:closingtype}.
+
+\todo{first, define new closing types and fresh metas!}
+
+The rule \LabTirName{constr/solve} says that if $B$ does not depend on the $x :
+\Rec\,(\alpha\,\ol{t})$ domain variable, then $\alpha$ is equal to the constant
+function which returns the empty telescope. We use the abbreviation
+$\lambda\,\overline{\mathsf{vars}_{\Gamma_0}}$ to refer to an iterated $\lambda$
+which abstracts over all variables in $\Gamma_0$.
 

### 2020-02-22 paper.tex
stuff

@@
 \newcommand{\freshMeta}[3]{\mathsf{freshMeta}\,_{#1|#2}\,#3}
+\newcommand{\constancy}{\mathsf{constancy}}
+\newcommand{\newConstancy}[6]{\mathsf{newConstancy}_{#1|#2,\,#3 : \Rec\,(#4\,#5)}\,#6}
 \newcommand{\unify}{\mathsf{unify}}
@@
 \newcommand{\kw}[1]{{\mathsf{#1}}}
+\newcommand{\FreeVars}{\mathsf{FreeVars}}
 
@@
 terms of power, generality and conceptual simplicity. We build atop Norell's
-bidirectional elaboration algorithm for Agda, and note that the key issue is
+bidirectional elaboration algorithm for Agda, and we note that the key issue is
 incomplete information about insertions of implicit abstractions and
 applications. We make it possible to track and refine information related to
-such insertions, by adding a new function type to a core Martin-L\"of type
-theory, which supports strict (definitional) currying. This allows us to
-represent undetermined domain arities of implicit function types, and we can
-decide at any point during elaboration whether implicit abstractions should be
-inserted.
+such insertions, by adding a function type to a core Martin-L\"of type theory,
+which supports strict (definitional) currying. This allows us to represent
+undetermined domain arities of implicit function types, and we can decide at any
+point during elaboration whether implicit abstractions should be inserted.
 \end{abstract}
@@
 argument is provided by elaboration. For example, in Agda, $id\,true$ is
-elaborated to $id\,Bool\,true$, and analogously in GHC and Coq. In all three
+elaborated to $id\,\{Bool\}\,true$, and analogously in GHC and Coq. In all three
 systems, there is also a way to explicitly specify implicit arguments: in Agda
@@
 (\lambda\,\{A\}\,x \to x)\,::\,[]$. The issue is the following. Agda first
-\emph{infers} a type for $(\lambda\,x \to x)\,::\,[]$, then tries to unify the
+infers a type for $(\lambda\,x \to x)\,::\,[]$, then tries to unify the
 inferred type with the given $List\,(\{A : \kw{Set}\}\to A \to A)$
@@
 There is significant literature on type inference in the presence of first-class
-polymorphic types, mainly in relation to GHC and ML-like languages \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy,serrano2020a}.
+polymorphic types, mainly in relation to GHC and ML-like languages; see
+e.g.\ \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy}.
 The above issue in Agda is a specific instance of the challenges described in
-the mentioned works. However, none of the proposed algorithms have landed so far
-in production compilers, for reasons of complexity, fragility and interaction
-with other language features.
+the mentioned works. So far, none of the above solutions have landed in
+production compilers, for reasons of complexity, fragility and interaction with
+other language features. A recent GHC development \cite{serrano2020a} offers a
+solution which is relatively simple, and which is likely to land in an official
+GHC release. However, none of these solutions support dependent types, and they
+also have other limitations which we would like to address.
 
@@
 actual Agda elaborator. In this section, we use it to illustrate the key issues,
-and we extend it in Section TODO with additional rules.
+and we extend it in Section \ref{sec:extending_elaboration}.
 \begin{figure}[h]
@@
 
-From now on, we will only consider well-formed core syntax, and only
-constructions which respect definitional equality. In other words, we quotient
-well-formed syntax by definitional equality.
+From now on, we will only consider well-formed core syntax, and unless otherwise
+mentioned, constructions on core syntax which respect definitional equality.
 
@@
 
-\subsection{Metasubstitutions}
+\subsection{Metasubstitutions}\label{sec:metasubstitutions}
 
@@
 
-\begin{definition}[Contextualization] For each $\Theta|\Gamma \vdash t : \Gamma\To A$,
+\begin{definition}[Contextualization]\label{def:contextualization} For each $\Theta|\Gamma \vdash t : \Gamma\To A$,
 we define the spine $\overline{\mathsf{vars}_{\Gamma}}$ such that that
@@
 
-\begin{definition}[Fresh meta creation] We specify $\freshMeta{\Theta}{\Gamma}{A}$ as follows:
+\begin{definition}[Fresh meta creation]\label{def:freshmeta} We specify $\freshMeta{\Theta}{\Gamma}{A}$ as follows:
   \begin{mathpar}
@@
 term. For example, if we have a defined name $id$ with type $\{A : \U\}\to A \to
-A$ occurring in the surface program, we usually want to expand $id$ to
-$id\,\{\alpha\}$, where $\alpha$ is a fresh metavariable.
+A$ in a surface program, we usually want to expand $id$ to $id\,\{\alpha\}$,
+where $\alpha$ is a fresh metavariable.
 \begin{mathpar}
@@
 $A$ type, as follows. Here, we notate $\Theta$ and $\Gamma$ as subscripts but
-later we will leave them implicit.
+later we will leave them implicit. The clauses of the following definition are
+matched top-down.
 \begin{alignat*}{3}
@@
 follows. We also have the additional $ins \in \{\true,\,\false\}$ argument for
-inference, which toggles whether the output has inserted implicit arguments.  We
-explain the cases of the definition in order.
+inference, which toggles whether we perform implicit argument insertion on the
+output. We explain the cases of the definition in order.
 
@@
 The first two clauses are checking $\lambda$-s, where the expected type exactly
-matches the $\lambda$ expressions. Hence, we simply check under binders with
+matches the $\lambda$ binders. Hence, we simply check under binders with
 $\echeck{t}{\Theta}{(\Gamma,\,x:A)} B$, and wrap the resulting term in the
-appropriate $\lambda$.
+appropriate (implicit or explicit) $\lambda$.
 
@@
 \[
-NatInd : \{P : \Nat \to \U\}\to \,P\,\zero \to (\{n : \Nat\}\to P\,n \to
-         P\,(\suc\,n)) \to (n : \Nat) \to P\,n
+NatInd : \{P : Nat \to \U\}\to \,P\,\zero \to (\{n : Nat\}\to P\,n \to
+         P\,(\suc\,n)) \to (n : Nat) \to P\,n
 \]
@@
 \begin{alignat*}{1}
-& \slet\, NatPlus : \Nat \to \Nat \to \Nat\\
-& \qquad = NatInd\,(\lambda\,n.\,\Nat\to\Nat)\,(\lambda\,m.\,m)\,
+& \slet\, NatPlus : Nat \to Nat \to Nat\\
+& \qquad = NatInd\,(\lambda\,n.\,\Nat\to Nat)\,(\lambda\,m.\,m)\,
                         (\lambda\,f\,m.\,\suc\,(f\,m))\,\,\sin\,\,...
@@
 When the above is elaborated, the $\lambda\,f\,m.\,\suc\,(f\,m)$ function is
-checked with the expected type $\{n : \Nat\}\to (\Nat \to \Nat) \to (\Nat \to
-\Nat)$, and the elaboration output is $\lambda\,\{n\}\,f\,m.\,\suc\,(f\,m)$.
-Hence, in this case we do not have to write implicit $\lambda$ in the surface
-syntax.
+checked with the expected type $\{n : Nat\}\to (Nat \to Nat) \to (Nat \to Nat)$,
+and the elaboration output is $\lambda\,\{n\}\,f\,m.\,\suc\,(f\,m)$.  Hence, in
+this case we do not have to write implicit $\lambda$ in the surface syntax.
 
@@
 For $\lambda$-s, we create a fresh meta for the domain type (since our surface
-$\lambda$-s are not annotated), and infer the bodies. In the case of
+$\lambda$-s are not annotated), and infer types for the bodies. In the case of
 $\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma}$, we additionally perform
@@
 These are function types whose domains are telescopes, and they are immediately
-computed to iterated implicit function types when the domain telescope is known
-to be empty or non-empty. See \LabTirName{fun-$\epsilon$} and
-\LabTirName{fun-$\TCons$}: a curried function with empty domain computes to
-simply the codomain, while a function with a non-empty domain computes to an
-implicit function type. We explicitly notate telescopes in both
-$\lambda$-abstractions and applications for strictly curried functions, since
-they are relevant in the computation rules.
+computed to iterated implicit function types when the domain telescope is
+canonical.  See \LabTirName{fun-$\epsilon$} and \LabTirName{fun-$\TCons$}: a
+curried function with empty domain computes to simply the codomain, while a
+function with a non-empty domain computes to an implicit function type. We
+explicitly notate telescopes in both $\lambda$-abstractions and applications for
+strictly curried functions, since they are relevant in the computation rules.
 
@@
 \section{Extending Elaboration}
+\label{sec:extending_elaboration}
 
@@
 \end{enumerate}
-In the following, we shall only handle the case for $\lambda$ insertions. This
-allows a simple implementation which only requires minimal changes to
-unification, and which is already remarkably powerful. It seems that handling
-implicit application insertions requires extending unification; we will talk
-about this in Section \ref{TODO}. But before delving into elaboration, we need to
-rectify a curious issue.
-
-\subsection{Superfluous Implicit Functions}
-
-Assuming $Bool : \U$, $true : Bool$ and $false : Bool$, consider the following surface expression:
+In the following, we shall only enhance $\lambda$-insertions. This allows a
+simple implementation which only requires minimal changes to unification, and
+which is already remarkably powerful. It seems that handling implicit
+application insertions requires extending unification; we discuss this in
+Section \ref{TODO}. First, we modify closing types and contextualization to take
+advantage of telescopes.
+
+\begin{definition}[Closing types]
+We use curried function space to close over record types in the scope. We
+prepend the following clause to Definition \ref{def:closingtype}:
+\[
+  ((\Gamma,\, x : \Rec\,A) \To B) :\equiv (\Gamma \To (\{x : \ol{A}\} \to B))
+\]
+\end{definition}
+\begin{definition}[Contextualization]
+We extend our spine notation to applications of curried functions. For example,
+we may have a spine $\ol{t} \equiv (\{x : \ol{A}\}\,\{y : \ol{B}\})$. We
+accordingly revise Definition \ref{def:contextualization} for
+$\overline{\mathsf{vars}_{\Gamma}}$ so that we use curried function application
+for each record type in $\Gamma$.
+\end{definition}
+
+\subsection{Handling Superfluous Implicit Functions}
+
+Before we can move on to unification, checking and inference, we have to address
+a curious issue. Assuming $Bool : \U$, $true : Bool$ and $false : Bool$,
+consider the following surface expression:
 \[
@@
 definition, we always have to write $x\,\{true\}$ or $x\,\{false\}$ when we want
-to use $x$. To rule out such nonsense, we adopt the following principle:
-
-\begin{quote}
-  \emph{Elaboration should never invent non-dependent implicit function types.}
-\end{quote}
+to use $x$. In order to avoid such nonsense, we adopt the following principle:
+\emph{elaboration should never invent non-dependent implicit function types.}
 
@@
 implicit function types; it was only utilizing the type annotations present in
-the surface input. With $\slet\,x : \_ = true$, the old elaborator checks $true$
-with a fresh meta, and just assumes that the meta does not stand for an
-implicit function type.
-
-To implement our principle, we need \emph{constancy constraints}. These are
-constraints in the sense of GHC's OutsideIn(X) \cite{vytiniotis2011outsidein}
-and also in the sense of the implementation in Agda and Coq. Rules for constancy
-constraints are as follows.
-
+the surface input. In the case of $\slet\,x : \_ = true$, the old elaborator
+checks $true$ with a fresh meta, and just assumes that the meta does not stand
+for an implicit function type.
+
+\subsubsection{Constancy constraints}
+We use \emph{constancy constraints} to get rid of curried function types as
+soon as we learn that they are non-dependent. These are constraints in the usual
+sense in unification algorithms (e.g.\ as in \cite{abel2011higher} or
+\cite{vytiniotis2011outsidein}). We formalize them in a compact way, by adding a
+new kind of context extension for metacontexts. The rules are given in Figure
+\ref{fig:constancy}.
+
+\begin{figure}
 \begin{mathpar}
-  \inferrule*[lab=metacon/constr]
-             {\Theta \vdash \alpha : \Gamma \To \Tel \\ \Theta|\Delta,\,x : \Rec\,(\alpha\,\ol{t}) \vdash B : \U}
-             {\Theta,\,\mathsf{constancy}_{\Delta,\,x : \Rec\,(\alpha\,\ol{t})}\,B\vdash}
-
-  \inferrule*[lab=constr/solve]
-             {\Theta_0 \vdash \alpha : \Gamma_0 \To \Tel \\ x \notin FreeVars(B)}
-             {\Theta_0,\,\mathsf{constancy}_{\Delta,\,x : \Rec\,(\alpha\,\ol{t})}\,B,\,\Theta_1|\Gamma_1
-               \vdash \alpha \equiv \lambda\,\overline{\mathsf{vars}_{\Gamma_0}}.\,\epsilon}
+  \inferrule*[lab=metacon/constancy]
+             {\Theta|\emptycon \vdash \alpha : \Delta \To \Tel \\ \Theta|\Gamma,\,x : \Rec\,(\alpha\,\ol{t}) \vdash B : \U}
+             {\Theta,\,\mathsf{constancy}_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B\vdash}
+
+  \inferrule*[lab=constancy-$\equiv$]
+             {\Theta_0|\emptycon \vdash \alpha : \Delta \To \Tel \\ x \notin \FreeVars(B)}
+             {\Theta_0,\,\mathsf{constancy}_{\Gamma_0,\,x : \Rec\,(\alpha\,\ol{t})}\,B,\,\Theta_1|\Gamma_1
+               \vdash \alpha \equiv (\lambda\,\overline{\mathsf{vars}_{\Delta}}.\,\epsilon) : \Delta \To \Tel}
+
+  \inferrule*[lab=metasub/constancy]
+             {\Theta_1|\emptycon \vdash \alpha : \Delta \To \Tel \\ \theta : \Theta_0 \To \Theta_1 \\\\ (x \notin \FreeVars(B[\theta]))\,\,\,\text{implies}\,\,\,(\alpha[\theta] \equiv (\lambda\,\overline{\mathsf{vars}_{\Delta[\theta]}}.\,\epsilon))}
+             {(\theta,\,\mathsf{solve}) : \Theta_0 \To (\Theta_1,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B)}
+
+  \inferrule*[lab=metasub/weaken-constancy]
+             {\\}
+             {\p : (\Theta,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B) \To \Theta}
 \end{mathpar}
-
-In the rule \LabTirName{metacon/constr}, we allow extension of the metacontext
-with a constraint.  In the constraint, the metavariable $\alpha : \Gamma \To
-\Tel$ represents an unknown domain for a curried function type, and $B$ is the
-codomain of the function. Here, $\Gamma \To \Tel$ refers to a closing type, which
-was defined in Section \ref{def:closingtype}.
-
-\todo{first, define new closing types and fresh metas!}
-
-The rule \LabTirName{constr/solve} says that if $B$ does not depend on the $x :
-\Rec\,(\alpha\,\ol{t})$ domain variable, then $\alpha$ is equal to the constant
-function which returns the empty telescope. We use the abbreviation
+\caption{Rules for constancy constraints}
+\label{fig:constancy}
+\end{figure}
+
+In the rule \LabTirName{metacon/constancy} we specify extension of a
+metacontext with a constraint. The metavariable $\alpha : \Gamma \To \Tel$
+represents an unknown domain for a curried function type, and $B$ is the
+codomain of the function.
+
+The \LabTirName{constancy-$\equiv$} rule expresses that, assuming we have a
+constancy constraint for $\alpha$ and $B$ in context, if $B$ does not depend on
+the $x : \Rec\,(\alpha\,\ol{t})$ domain variable, then $\alpha$ is equal to the
+constant function which returns the empty telescope. We use the abbreviation
 $\lambda\,\overline{\mathsf{vars}_{\Gamma_0}}$ to refer to an iterated $\lambda$
 which abstracts over all variables in $\Gamma_0$.
+
+The \LabTirName{metasub/constancy} rule defines metasubstitutions whose
+codomains are extended with constraints. Intuitively, while the
+\LabTirName{metasub/extended} rule from Section \ref{sec:metasubstitutions} can
+be used to solve a metavariable (by mapping it to a term),
+\LabTirName{metasub/constancy} solves a constraint. We can only extend $\theta :
+\Theta_0 \To \Theta_1$ to map into an additional constraint in the codomain if
+$\theta$ forces the constraint to hold. In
+\LabTirName{metasub/weaken-constancy}, we overload $\p$ for the weakening
+substitution which drops a constraint.
+
+\begin{definition}[Creating a new constraint]
+We define this similarly to Definition \ref{def:freshmeta}, by simply
+returning a weakening substitution.
+\[
+\newConstancy{\Theta}{\Gamma}{x}{\alpha}{\ol{t}}{B} : \equiv
+             ((\Theta,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B),\,\p)
+\]
+\end{definition}
+
+\subsubsection{Algorithmic implementation of constraint solving}
+The above specification for constancy constraints is compact but not
+particularly algorithmic: we just magically get new definitional equalities
+whenever we have constraints in contexts. In our prototype implementation, we
+augment the unification procedure with eager removal of solvable constraints.
+
+After solving a meta $\beta$ during unification, which yields a unifying
+$\theta$ substitution, we review all ($\constancy_{\Gamma,\,x :
+  \Rec\,(\alpha\,\ol{t})}\,B$) constraints in the context, where $x$ occurs in
+$B$ inside a $\ol{t}$ spine of some $\beta\,\ol{t}$ term. In other words, we
+review constraints where the new meta solution might make a difference.
+\begin{enumerate}
+  \item If we have $x \in \FreeVars(B[\theta])$, where $x$ occurs rigidly in
+    $B[\theta]$, i.e.\ the occurrence is not in a spine of a meta, then no
+    metasubstitution can possibly remove this occurrence. In this case the
+    constraint holds vacuously, so we can use the rule
+    \LabTirName{metasub/constancy} to return a $\theta'$ substitution which also
+    solves the constraint.
+  \item If we have $x \notin \FreeVars(B[\theta])$, we solve $\alpha$ to be
+    constantly $\epsilon$, thereby getting a $\theta'$ substitution. Since
+    $\theta'$ forces the constrain to hold, we can again use
+    \LabTirName{metasub/constancy} to solve the constraint. Then, since we have
+    just solved $\alpha$, we recursively check for solvable constraints.
+  \item
+    In any other case we simply return $\theta$ and keep the constraint around.
+\end{enumerate}
+
+\emph{Remark.} In the case with $x \in \FreeVars(B[\theta])$, it would be also
+sound to solve the constraint when the occurrence is not necessarily
+rigid. However, this way we could lose potential non-$\epsilon$ solutions of
+$\alpha$.
+
+\subsection{Unification For Strictly Curried Functions}
+Although we omit most details of unification, we shall discuss it for curried
+functions, as it is essential in the extended elaboration algorithm.  The most
+interesting case is when we unify a curried function type with an implicit
+function type. In this case, we learn that the domain of the curried function is
+non-empty, so we refine the domain telescope.
+\begin{alignat*}{3}
+& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x : A'\}\to B') :\equiv \edo\\
+& \quad (\Theta_1,\,\theta_1,\,\alpha\,\ol{t}) \leftarrow \freshMeta{\Theta_0}{\Gamma,\,x : A'}{\Tel}\\
+& \quad (\Theta_2,\,\theta_2) \leftarrow
+  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A'[\theta_1])\TCons \alpha\,\ol{t})\\
+& \quad (\Theta_3,\,\theta_3) \leftarrow \newConstancy{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,y : A'[\theta_1\circ\theta_2]}{z}{(\alpha}{\ol{t})[\theta_2]}{(B[\theta_1\circ\theta_2][x\mapsto (y :: z)])}\\
+& \unify_{\Theta_3|\Gamma[\theta_{123}]}\,(\{z : \ol{\alpha\,\ol{t}}
+\end{alignat*}
 

### 2020-02-22 implicit-fun-elab/Elaboration.hs

@@ -18,7 +18,7 @@ import ElabState
 import Debug.Trace
 
 
--- Telescope constancy
+-- Occurs check for constancy constraints
 --------------------------------------------------------------------------------
 
 data Occurs = Rigid | Flex IS.IntSet | None deriving (Eq, Show)

### 2020-02-22 implicit-fun-elab/Main.hs

@@ -65,293 +65,303 @@ main = mainWith getArgs parseStdin
 main' :: String -> String -> IO ()
 main' mode src = mainWith (pure [mode]) ((,src) <$> parseString src)
 
-pruneTest = main' "elab" $ unlines [
-  "λ (Bool : U)(not : Bool → Bool → Bool)(true : Bool).",
-  "λ f (x : Bool). not (f x)"
-  ]
-
-pruneTest2 = main' "elab" $ unlines [
- "λ (ℕ : U)(zero : ℕ)(suc : ℕ → ℕ)(add : ℕ → ℕ → ℕ).",
- "let test1 = (P : _) → (n : ℕ) → P n → P n in",
- "let test2 = λ f x . add (suc (f x)) x in",
- "U"
- ]
-
-
-test0 = main' "elab" $ unlines [
-  "let IdTy = {A} → A → A in",
-  "let f = λ (g:IdTy). g g in",
-  -- "let id : IdTy = f in",
-  "U"
-  ]
-
-
-test = main' "elab" $ unlines [
-  "let the : (A : U) → A → A = \\A x.x in",
-  "let IdTy = {A} → A → A in",
-  "let id : IdTy = λ x. x in",
-  "let choose : {A} → A → A → A = λ x y. x in",
-
-  "let Nat : U",
-  "    = (N : U) -> (N -> N) -> N -> N in",
-  "let zero : Nat = λ _ s z. z in",
-  "let inc : Nat → Nat = λ n N s z. s (n _ s z) in",
-
-  "let mul : Nat -> Nat -> Nat",
-  "    = \\a b N s z. a _ (b N s) z in",
-  "let ten : Nat",
-  "    = \\N s z. s (s (s (s (s (s (s (s (s (s z))))))))) in",
-  "let hundred = mul ten ten in",
-
-  "let List : U -> U",
-  "    = \\A. (L : U) -> (A -> L -> L) -> L -> L in",
-  "let nil : {A : U} -> List A",
-  "    = \\cons L nil. nil in",
-  "let cons : {A : U} -> A -> List A -> List A",
-  "    = \\{A} x xs L cons nil. cons x (xs _ cons nil) in",
-
-  "let append : {A} → List A → List A → List A",
-  "    = λ xs ys L cons nil. xs _ cons (ys _ cons nil) in",
-  "let map : {A B} → (A → B) → List A → List B = λ f as _ c n. as _ (λ a. c (f a)) n in",
-  "let head : {A} → List A → A = λ as. _ in",
-
[…]

### 2020-02-22 implicit-fun-elab/Types.hs

@@ -100,12 +100,12 @@ pattern TSnoc as a <- ((\case TBound as a -> Just (as, a)
 lvlName :: [Name] -> Lvl -> Name
 lvlName ns x = ns !! (length ns - x - 1)
 
-ixType :: Types -> Ix -> VTy
-ixType TNil           _ = error "impossible"
-ixType (TDef   tys a) 0 = a
-ixType (TBound tys a) 0 = a
-ixType (TDef   tys a) x = ixType tys (x - 1)
-ixType (TBound tys a) x = ixType tys (x - 1)
+-- ixType :: Types -> Ix -> VTy
+-- ixType TNil           _ = error "impossible"
+-- ixType (TDef   tys a) 0 = a
+-- ixType (TBound tys a) 0 = a
+-- ixType (TDef   tys a) x = ixType tys (x - 1)
+-- ixType (TBound tys a) x = ixType tys (x - 1)
 
 data NameOrigin = NOSource | NOInserted
 
@@ -394,8 +394,3 @@ makeFields ''Str
 ucxt :: Lens' Cxt UnifyCxt
 ucxt f (Cxt vs tys ns no d) =
   (\(UCxt tys ns d) -> Cxt vs tys ns no d) <$> f (UCxt tys ns d)
-
--- instance HasNames  [Name]  [Name]  where names = id
--- instance HasVals   Vals    Vals    where vals  = id
--- instance HasTypes  Types   Types   where types = id
--- instance HasLen    Int     Int     where len   = id

### 2020-02-22 implicit-fun-elab/ElabState.hs
streamlined constancy

@@ -28,23 +28,21 @@ lookupMeta m = do
 runLookupMeta :: MId -> MetaEntry
 runLookupMeta m = runIO (lookupMeta m)
 
-writeMeta :: MId -> MetaEntry -> IO ()
-writeMeta m e = modifyIORef' mcxt (IM.alter go m) where
-  go Just{} = Just e
-  go _      = error "impossible"
-
-newMetaEntry :: MetaEntry -> IO MId
-newMetaEntry e = do
-  m <- readIORef nextMId
-  writeIORef nextMId $! (m + 1)
-  let f = maybe (Just e) (\_ -> error "impossible")
-  modifyIORef' mcxt (IM.alter f m)
-  pure m
+alterMeta :: MId -> (Maybe MetaEntry -> Maybe MetaEntry) -> IO ()
+alterMeta m f = modifyIORef' mcxt (IM.alter f m)
 
 modifyMeta :: MId -> (MetaEntry -> MetaEntry) -> IO ()
-modifyMeta m f = modifyIORef' mcxt (IM.alter go m) where
-  go Nothing  = error "impossible"
-  go (Just e) = Just (f e)
+modifyMeta m f = alterMeta m (maybe (error "impossible") (Just . f))
+
+writeMeta :: MId -> MetaEntry -> IO ()
+writeMeta m e = modifyMeta m (const e)
+
+newMeta :: MetaEntry -> IO MId
+newMeta e = do
+  m <- readIORef nextMId
+  writeIORef nextMId $! (m + 1)
+  alterMeta m (maybe (Just e) (\_ -> error "impossible"))
+  pure m
 
 reset :: IO ()
 reset = do

### 2020-02-22 implicit-fun-elab/Elaboration.hs

@@ -17,6 +17,34 @@ import ElabState
 
 import Debug.Trace
 
+-- Context operations
+--------------------------------------------------------------------------------
+
+emptyCxt :: Cxt
+emptyCxt = Cxt VNil TNil [] [] 0
+
+bind :: Name -> NameOrigin -> VTy -> Cxt -> Cxt
+bind x o ~a (Cxt vs tys ns no d) =
+  Cxt (VSkip vs) (TBound tys a) (x:ns) (o:no) (d + 1)
+
+bindSrc :: Name -> VTy -> Cxt -> Cxt
+bindSrc x = bind x NOSource
+
+define :: Name -> VTy -> Val -> Cxt -> Cxt
+define x ~a ~t (Cxt vs tys ns no d) =
+  Cxt (VDef vs t) (TDef tys a) (x:ns) (NOSource:no) (d + 1)
+
+addSrcPos :: SPos -> IO a -> IO a
+addSrcPos p act = act `catch` \(e::Err) ->
+  case e^.pos of
+    Nothing -> throwIO (e & pos .~ Just p)
+    _       -> throwIO e
+
+-- | Lift ("skolemize") a value in an extended context to a function in a
+--   non-extended context.
+liftVal :: Cxt -> Val -> (Val -> Val)
+liftVal cxt t = \ ~x -> eval (VDef (cxt^.vals) x) $ quote (cxt^.len+1) t
+
 
 -- Occurs check for constancy constraints
 --------------------------------------------------------------------------------
@@ -73,7 +101,7 @@ occurs d topX = occurs' d mempty where
 -- | Attempt to solve a constancy constraint.
 tryConstancy :: MId -> IO ()
 tryConstancy constM = lookupMeta constM >>= \case
-  Constancy cxt telMeta sp codomain blockers -> do
+  Constancy cxt dom cod blockers -> do
 
     -- clear blockers
     forM_ (IS.toList blockers) $ \m -> do
@@ -82,34 +110,25 @@ tryConstancy constM = lookupMeta constM >>= \case
         Solved t      -> Solved t
         Constancy{}   -> error "impossible"
 
-    writeMeta constM $ Constancy cxt telMeta sp codomain mempty
+    let dropConstancy = alterMeta constM (const Nothing)
 
-    lookupMeta telMeta >>= \case
-      Constancy{} -> error "impossible"
-      Solved{}    -> pure ()
[…]

### 2020-02-22 implicit-fun-elab/Main.hs

@@ -65,303 +65,301 @@ main = mainWith getArgs parseStdin
 main' :: String -> String -> IO ()
 main' mode src = mainWith (pure [mode]) ((,src) <$> parseString src)
 
--- pruneTest = main' "elab" $ unlines [
---   "λ (Bool : U)(not : Bool → Bool → Bool)(true : Bool).",
---   "λ f (x : Bool). not (f x)"
---   ]
+pruneTest = main' "elab" $ unlines [
+  "λ (Bool : U)(not : Bool → Bool → Bool)(true : Bool).",
+  "λ f (x : Bool). not (f x)"
+  ]
 
--- pruneTest2 = main' "elab" $ unlines [
---  "λ (ℕ : U)(zero : ℕ)(suc : ℕ → ℕ)(add : ℕ → ℕ → ℕ).",
---  "let test1 = (P : _) → (n : ℕ) → P n → P n in",
---  "let test2 = λ f x . add (suc (f x)) x in",
+pruneTest2 = main' "elab" $ unlines [
+ "λ (ℕ : U)(zero : ℕ)(suc : ℕ → ℕ)(add : ℕ → ℕ → ℕ).",
+ "let test1 = (P : _) → (n : ℕ) → P n → P n in",
+ "let test2 = λ f x . add (suc (f x)) x in",
+ "U"
+ ]
+
+test0 = main' "elab" $ unlines [
+  "let the : (A : U) → A → A = λ A x. x in",
+  "let IdTy = {A} → A → A in",
+  "let choose : {A} → A → A → A = λ x y. x in",
+  "let id : IdTy = λ x. x in",
+  "let auto2 : {B} → IdTy → B → B = λ f b. b in",
+  "let r : ({A} → A → {B} → B → B) → U = λ _. U in",
+  "let foo = r (λ x y. y) in",
+  "U"
+  ]
+
+
+test = main' "elab" $ unlines [
+  "let the : (A : U) → A → A = \\A x.x in",
+  "let IdTy = {A} → A → A in",
+  "let id : IdTy = λ x. x in",
+  "let choose : {A} → A → A → A = λ x y. x in",
+
+  "let Nat : U",
+  "    = (N : U) -> (N -> N) -> N -> N in",
+  "let zero : Nat = λ _ s z. z in",
+  "let inc : Nat → Nat = λ n N s z. s (n _ s z) in",
+
+  "let mul : Nat -> Nat -> Nat",
[…]

### 2020-02-22 implicit-fun-elab/Types.hs

@@ -58,8 +58,8 @@ data MetaEntry
 
   -- | Telescope constancy constraint. When the closure becomes constant,
   --   we unify the telescope with the empty telescope.
-  --   Constancy context (telescope meta) (meta spine) codomain blockers
-  | Constancy UnifyCxt MId Spine Val BlockedBy
+  --   Constancy context domain codomain blockers
+  | Constancy Cxt VTy VTy BlockedBy
 
 
 -- | A partial mapping from levels to levels. Undefined domain reresents
@@ -117,11 +117,6 @@ data Cxt = Cxt {
   cxtNames      :: [Name],
   cxtNameOrigin :: [NameOrigin],
   cxtLen        :: Int}
-
-data UnifyCxt = UCxt {
-  unifyCxtTypes :: Types,
-  unifyCxtNames :: [Name],
-  unifyCxtLen   :: Int }
 
 data Tm
   = Var Ix
@@ -382,15 +377,9 @@ showError ns = \case
     "Function icitness mismatch: expected %s, got %s.")
     (show i) (show i')
 
-
 -- Lenses
 --------------------------------------------------------------------------------
 
 makeFields ''Cxt
-makeFields ''UnifyCxt
 makeFields ''Err
 makeFields ''Str
-
-ucxt :: Lens' Cxt UnifyCxt
-ucxt f (Cxt vs tys ns no d) =
-  (\(UCxt tys ns d) -> Cxt vs tys ns no d) <$> f (UCxt tys ns d)

### 2020-02-23 paper.tex
stuff

@@
 \newcommand{\constancy}{\mathsf{constancy}}
-\newcommand{\newConstancy}[6]{\mathsf{newConstancy}_{#1|#2,\,#3 : \Rec\,(#4\,#5)}\,#6}
+\newcommand{\newConstancy}[5]{\mathsf{newConstancy}_{#1|#2,\,#3 : \Rec\,#4}\,#5}
 \newcommand{\unify}{\mathsf{unify}}
@@
 
-\section{Basic Bidirectional Elaboration}
-\label{sec:basic_bidirectional_elaboration}
+\section{Bidirectional Elaboration}
+\label{sec:bidirectional_elaboration}
 
@@
 simplifications; what we end up with can be viewed as a toy version of the
-actual Agda elaborator. In this section, we use it to illustrate the key issues,
-and we extend it in Section \ref{sec:extending_elaboration}.
+actual Agda elaborator. In this section, we use it to build the backbone of our
+algorithm and illustrate the key issues. We extend this elaborator in Section
+\ref{sec:extending_elaboration}.
 \begin{figure}[h]
@@
 \begin{notation}\label{not:spines}
-  We sometimes use a spine notation for neutral terms. A spine is a list of
+  We use a spine notation for neutral terms. A spine is a list of
   terms, noted as $\overline{t}$, where terms may be wrapped in brackets to
@@
 \end{alignat*}
-In the core syntax, all applications must be explicit. In the surface syntax we
-will be able to omit implicit arguments to $map$.
-
 \end{example}
+%% In the core syntax, all abstractions and applications must be notated. Thanks to
+%% elaboration, in the surface syntax we may omit some of these.
 
@@
   \inferrule*[lab=metasub/empty]
-             {\Theta_0\vdash}
-             {() : \Theta_0\To \emptycon}
+             {\Theta\vdash}
+             {() : \Theta\To \emptycon}
 
@@
   \inferrule*[lab=metasub/con-action]
-             {\Theta_1|\Gamma\vdash \\ \theta : \Theta_0\To\Theta_1}
+             {\theta : \Theta_0\To\Theta_1 \\ \Theta_1|\Gamma\vdash}
              {\Theta_0|\Gamma[\theta]\vdash}
@@
   \inferrule*[lab=metasub/tm-action]
-             {\Theta_1|\Gamma\vdash t : A \\ \theta : \Theta_0\To\Theta_1}
+             {\theta : \Theta_0\To\Theta_1 \\ \Theta_1|\Gamma\vdash t : A}
              {\Theta_0|\Gamma[\theta] \vdash t[\theta] : A[\theta]}
@@
     argument insertion and unification.
+  \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$, $\theta_1
+    \circ \theta_2 \circ \theta_3$ as $\theta_{123}$ and analogously in other
+    cases. We do this to reduce the visual noise caused by threading composed
+    metasubstitutions everywhere in the elaboration algorithm.
 \end{itemize}
@@
   &\quad(\Theta_3,\,\theta_3,\,u') \leftarrow
-                \echeck{u}{\Theta_2}{\Gamma[\theta_1\circ\theta_2]}{(B[\theta_1\circ\theta_2])}\\
-  &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u')\\
+                \echeck{u}{\Theta_2}{\Gamma[\theta_{12}]}{(B[\theta_{12}])}\\
+  &\quad\ereturn\,(\Theta_3,\,\theta_{123},\,\slet\,x:A'[\theta_{23}]=t'[\theta_3]\,\sin\,u')\\
   &\echeck{\_}{\Theta}{\Gamma}{A} :\equiv \edo \\
@@
   &\quad(\Theta_2,\,\theta_2) \leftarrow \unify\,(A[\theta_1])\,B\\
-  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,t'[\theta_2])\\\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,t'[\theta_2])\\\\
   &\einfer{x}{ins}{\Theta}{\Gamma} :\equiv \edo\\
@@
                  \echeck{B}{\Theta_2}{\Gamma[\theta_1],\,x : A'}{\U}\\
-  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,
+  &\quad\ereturn\,(\Theta_2,\,\theta_{12},
                  \,((x : A'[\theta_2])\to B'),\,\U)\\
@@
                  \echeck{B}{\Theta_2}{\Gamma[\theta_1],\,x : A'}{\U}\\
-  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,
+  &\quad\ereturn\,(\Theta_2,\,\theta_{12},
                  \,(\{x : A'[\theta_2]\}\to B'),\,\U)\\
@@
   &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
-  &\quad\ereturn\,(\Theta_2,\,\theta_1\circ\theta_2,\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\
   &\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
@@
   &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
-  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,x:A_0}{\U}\\
-  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,((x : A_0[\theta_3])\to A_1)\\
-  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4]}{(A_0[\theta_3\circ\theta_4])}\\
-  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,u')}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\
+  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
+  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_{23}])\,((x : A_0[\theta_3])\to A_1)\\
+  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_{1234}]}{(A_0[\theta_{34}])}\\
+  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,u')}\,{(A_1[\theta_{45}][x\mapsto u'])})\\
   &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
@@
   &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
-  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,x:A_0}{\U}\\
-  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_2\circ\theta_3])\,(\{x : A_0[\theta_3]\}\to A_1)\\
-  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_1\circ\theta_2\circ\theta_3\circ\theta_4]}{(A_0[\theta_3\circ\theta_4])}\\
-  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_2\circ\theta_3\circ\theta_4\circ\theta_5])\,\{u'\})}\,{(A_1[\theta_4\circ\theta_5][x\mapsto u'])})\\
+  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
+  &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_{23}])\,(\{x : A_0[\theta_3]\}\to A_1)\\
+  &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_{1234}]}{(A_0[\theta_{34}])}\\
+  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,\{u'\})}\,{(A_1[\theta_{45}][x\mapsto u'])})\\
   &\einfer{\slet\,x:A=t\,\sin\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
@@
   &\quad(\Theta_3,\,\theta_3,\,u',\,B) \leftarrow
-                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_1\circ\theta_2]}\\
-  &\quad\ereturn\,(\Theta_3,\,\theta_1\circ\theta_2\circ\theta_3,\,(\slet\,x:A'[\theta_2\circ\theta_3]=t'[\theta_3]\,\sin\,u'),\,B)\\
+                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_{12}]}\\
+  &\quad\ereturn\,(\Theta_3,\,\theta_{123},\,(\slet\,x:A'[\theta_{23}]=t'[\theta_3]\,\sin\,u'),\,B)\\
   &\einfer{\_}{ins}{\Theta}{\Gamma} :\equiv \edo \\
@@
     subtyping complicates type inference, and to our knowledge it has not been
-    implemented in any dependently typed setting. We also believe that it is an
-    anti-feature in dependent settings, because elaboration of subtyping must
+    implemented in any dependently typed setting. We also believe that it is
+    undesirable in dependent settings, because elaboration of subtyping must
     insert coercions which significantly change the intensional character of
@@
 \begin{definition}[Closing types]
-We use curried function space to close over record types in the scope. We
-prepend the following clause to Definition \ref{def:closingtype}:
+We use curried function types to close over record types in the scope. If a
+bound variable does not have a record type, then we do as before\footnote{This
+  implies that we close over meta-headed types using plain functions. In theory,
+  this causes a higher-order version of the basic implicit insertion problem: we
+  are uncertain about whether we should be uncertain about implicit
+  insertions. So far, this higher-order insertion problem seems to seems to be
+  irrelevant in practice, in the prototype implementation.}. We prepend the
+following clause to Definition
+\ref{def:closingtype}:
 \[
@@
 \subsubsection{Constancy constraints}
-We use \emph{constancy constraints} to get rid of curried function types as
-soon as we learn that they are non-dependent. These are constraints in the usual
-sense in unification algorithms (e.g.\ as in \cite{abel2011higher} or
+We use these constraints to get rid of curried function types as soon as we
+learn that they are non-dependent. They are constraints in the usual sense in
+unification algorithms (e.g.\ as in \cite{abel2011higher} or
 \cite{vytiniotis2011outsidein}). We formalize them in a compact way, by adding a
@@
   \inferrule*[lab=metacon/constancy]
-             {\Theta|\emptycon \vdash \alpha : \Delta \To \Tel \\ \Theta|\Gamma,\,x : \Rec\,(\alpha\,\ol{t}) \vdash B : \U}
-             {\Theta,\,\mathsf{constancy}_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B\vdash}
+             {\Theta|\Gamma,\,x : \Rec\,A \vdash B : \U}
+             {\Theta,\,\mathsf{constancy}_{\Gamma,\,x : \Rec\,A}\,B\vdash}
 
   \inferrule*[lab=constancy-$\equiv$]
-             {\Theta_0|\emptycon \vdash \alpha : \Delta \To \Tel \\ x \notin \FreeVars(B)}
-             {\Theta_0,\,\mathsf{constancy}_{\Gamma_0,\,x : \Rec\,(\alpha\,\ol{t})}\,B,\,\Theta_1|\Gamma_1
-               \vdash \alpha \equiv (\lambda\,\overline{\mathsf{vars}_{\Delta}}.\,\epsilon) : \Delta \To \Tel}
+             {x \notin \FreeVars(B)}
+             {\Theta_0,\,\mathsf{constancy}_{\Gamma_0,\,x : \Rec\,A}\,B,\,\Theta_1|\Gamma_1 \vdash A \equiv \epsilon : \Tel}
 
   \inferrule*[lab=metasub/constancy]
-             {\Theta_1|\emptycon \vdash \alpha : \Delta \To \Tel \\ \theta : \Theta_0 \To \Theta_1 \\\\ (x \notin \FreeVars(B[\theta]))\,\,\,\text{implies}\,\,\,(\alpha[\theta] \equiv (\lambda\,\overline{\mathsf{vars}_{\Delta[\theta]}}.\,\epsilon))}
-             {(\theta,\,\mathsf{solve}) : \Theta_0 \To (\Theta_1,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B)}
+             {\theta : \Theta_0 \To \Theta_1 \\\\ x \notin \FreeVars(B[\theta])\,\,\,\text{implies}\,\,\,\Theta_0|\Gamma[\theta]\vdash A[\theta] \equiv \epsilon : \Tel}
+             {(\theta,\,\mathsf{solve}_{\Gamma,\,x : \Rec\,A}\,B) : \Theta_0 \To (\Theta_1,\,\constancy_{\Gamma,\,x : \Rec\,A}\,B)}
 
@@
              {\\}
-             {\p : (\Theta,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B) \To \Theta}
+             {\p : (\Theta,\,\constancy_{\Gamma,\,x : \Rec\,A}\,B) \To \Theta}
 \end{mathpar}
@@
 
-In the rule \LabTirName{metacon/constancy} we specify extension of a
-metacontext with a constraint. The metavariable $\alpha : \Gamma \To \Tel$
-represents an unknown domain for a curried function type, and $B$ is the
-codomain of the function.
-
-The \LabTirName{constancy-$\equiv$} rule expresses that, assuming we have a
-constancy constraint for $\alpha$ and $B$ in context, if $B$ does not depend on
-the $x : \Rec\,(\alpha\,\ol{t})$ domain variable, then $\alpha$ is equal to the
-constant function which returns the empty telescope. We use the abbreviation
-$\lambda\,\overline{\mathsf{vars}_{\Gamma_0}}$ to refer to an iterated $\lambda$
-which abstracts over all variables in $\Gamma_0$.
+In the rule \LabTirName{metacon/constancy} we specify extension of a metacontext
+with a constraint. The \LabTirName{constancy-$\equiv$} rule expresses that,
+assuming we have a constancy constraint for $A$ and $B$ in context, if $B$ does
+not depend on the $x : \Rec\,A$ domain variable, then $A$ is equal to the the
+empty telescope $\epsilon$.
 
@@
 \LabTirName{metasub/constancy} solves a constraint. We can only extend $\theta :
-\Theta_0 \To \Theta_1$ to map into an additional constraint in the codomain if
-$\theta$ forces the constraint to hold. In
-\LabTirName{metasub/weaken-constancy}, we overload $\p$ for the weakening
-substitution which drops a constraint.
+\Theta_0 \To \Theta_1$ to map into an additional constraint if $\theta$ forces
+the constraint to hold. In \LabTirName{metasub/weaken-constancy}, we overload
+$\p$ for the weakening substitution which drops a constraint.
 
 \begin{definition}[Creating a new constraint]
-We define this similarly to Definition \ref{def:freshmeta}, by simply
+We do this similarly to Definition \ref{def:freshmeta}, by simply
 returning a weakening substitution.
 \[
-\newConstancy{\Theta}{\Gamma}{x}{\alpha}{\ol{t}}{B} : \equiv
-             ((\Theta,\,\constancy_{\Gamma,\,x : \Rec\,(\alpha\,\ol{t})}\,B),\,\p)
+\newConstancy{\Theta}{\Gamma}{x}{A}{B} : \equiv
+             ((\Theta,\,\constancy_{\Gamma,\,x : \Rec\,A}\,B),\,\p)
 \]
@@
 
-After solving a meta $\beta$ during unification, which yields a unifying
-$\theta$ substitution, we review all ($\constancy_{\Gamma,\,x :
-  \Rec\,(\alpha\,\ol{t})}\,B$) constraints in the context, where $x$ occurs in
-$B$ inside a $\ol{t}$ spine of some $\beta\,\ol{t}$ term. In other words, we
-review constraints where the new meta solution might make a difference.
+After solving a meta $\alpha$ during unification, which yields a unifying
+$\theta$ substitution, we review all ($\constancy_{\Gamma,\,x : \Rec\,A}\,B$)
+constraints in the context, such that $x$ occurs in $B$ inside a $\ol{t}$ spine
+of some $\alpha\,\ol{t}$ term. In other words, we review constraints where the
+new meta solution might make a difference.
 \begin{enumerate}
@@
     solves the constraint.
-  \item If we have $x \notin \FreeVars(B[\theta])$, we solve $\alpha$ to be
-    constantly $\epsilon$, thereby getting a $\theta'$ substitution. Since
-    $\theta'$ forces the constrain to hold, we can again use
-    \LabTirName{metasub/constancy} to solve the constraint. Then, since we have
-    just solved $\alpha$, we recursively check for solvable constraints.
+  \item If we have $x \notin \FreeVars(B[\theta])$, we recursively unify
+    $A[\theta]$ with $\epsilon$. If that succeeds, we get a $\theta'$ which
+    unifies $A$ and $\epsilon$ and thus forces the constraint to hold, so we
+    can again use \LabTirName{metasub/constancy} to solve the constraint.
   \item
@@
 \emph{Remark.} In the case with $x \in \FreeVars(B[\theta])$, it would be also
-sound to solve the constraint when the occurrence is not necessarily
-rigid. However, this way we could lose potential non-$\epsilon$ solutions of
-$\alpha$.
+sound to solve the constraint when the occurrence is not rigid. However, this
+way we could lose potential non-$\epsilon$ solutions of $A$.
 
@@
 function type. In this case, we learn that the domain of the curried function is
-non-empty, so we refine the domain telescope.
-\begin{alignat*}{3}
-& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x : A'\}\to B') :\equiv \edo\\
-& \quad (\Theta_1,\,\theta_1,\,\alpha\,\ol{t}) \leftarrow \freshMeta{\Theta_0}{\Gamma,\,x : A'}{\Tel}\\
+non-empty, so we refine the $A$ domain to an extended $(x_0 : A_0) \TCons A_1$
+telescope. Since we invent a fresh $A_1$ domain for a curried function type, we
+need to add a constancy constraint for it as well.
+\begin{alignat*}{3}
+& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B') :\equiv \edo\\
+& \quad (\Theta_1,\,\theta_1,\,A_1) \leftarrow \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
 & \quad (\Theta_2,\,\theta_2) \leftarrow
-  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A'[\theta_1])\TCons \alpha\,\ol{t})\\
-& \quad (\Theta_3,\,\theta_3) \leftarrow \newConstancy{\Theta_2}{\Gamma[\theta_1\circ\theta_2],\,y : A'[\theta_1\circ\theta_2]}{z}{(\alpha}{\ol{t})[\theta_2]}{(B[\theta_1\circ\theta_2][x\mapsto (y :: z)])}\\
-& \unify_{\Theta_3|\Gamma[\theta_{123}]}\,(\{z : \ol{\alpha\,\ol{t}}
-\end{alignat*}
+  \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
+& \quad (\Theta_3,\,\theta_3) \leftarrow
+    \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
+& \quad\unify_{\Theta_3|\Gamma[\theta_{123}],\,x_0 : A_0[\theta_{123}]}\,
+         (\{x_1 : A_1[\theta_{23}]\}\to B[\theta_{123}][x\mapsto (x_0 :: x_1)])\,B'
+\end{alignat*}
+We have the symmetric $\unify_{\Theta_0|\Gamma}(\{x_0 : A_0\}\to B')\,(\{x :
+\ol{A}\}\to B)$ case the same way as above.
+
+Now, let us assume that $B'$ is not an implicit function type and not
+meta-headed. Then, we have the following case, where we solve a telescope domain
+to be empty.
+\begin{alignat*}{3}
+& \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,B' :\equiv \edo\\
+& \quad (\Theta_1,\,\theta_1) \leftarrow \unify_{\Theta_0|\Gamma}\,A\,\epsilon\\
+& \quad \unify_{\Theta_1|\Gamma[\theta_1]}\,(B[\theta_1][x\mapsto[]])\,(B'[\theta_1])
+\end{alignat*}
+Again, we also have the symmetric case. For $\lambda\,\{x : \ol{A}\}.\,t$ and
+$t\,\{u : \ol{A}\}$, unification is structural, and other cases remain the same
+as in the the basic elaborator of Section \ref{sec:bidirectional_elaboration}.
+
+\subsection{Elaboration}
 

### 2020-02-23 implicit-fun-elab/Elaboration.hs

@@ -4,7 +4,6 @@ module Elaboration where
 
 import Control.Exception
 import Control.Monad
-import Data.Foldable
 import Data.IORef
 import Data.Maybe
 import Lens.Micro.Platform

### 2020-02-23 implicit-fun-elab/Elaboration.hs
fix meta solution binder naming

@@ -134,23 +134,23 @@ newConstancy cxt dom cod =
 
 -- | Checks that a spine consists only of distinct bound vars.
 --   Returns a partial variable renaming on success, alongside the size
---   of the spine.
-checkSp :: Spine -> IO (Renaming, Lvl)
-checkSp = go . forceSp where
-  go :: Spine -> IO (Renaming, Lvl)
+--   of the spine, and the list of variables in the spine.
+checkSp :: Spine -> IO (Renaming, Lvl, [Lvl])
+checkSp = (over _3 reverse <$>) . go . forceSp where
+  go :: Spine -> IO (Renaming, Lvl, [Lvl])
   go = \case
-    SNil        -> pure (mempty, 0)
+    SNil        -> pure (mempty, 0, [])
     SApp sp u i -> do
-      (!r, !d) <- go sp
+      (!r, !d, !xs) <- go sp
       case force u of
         VVar x | IM.member x r -> throwIO $ NonLinearSpine x
-               | otherwise     -> pure (IM.insert x d r, d + 1)
+               | otherwise     -> pure (IM.insert x d r, d + 1, x:xs)
         _      -> throwIO SpineNonVar
     SAppTel a sp u -> do
-      (!r, !d) <- go sp
+      (!r, !d, !xs) <- go sp
       case force u of
         VVar x | IM.member x r -> throwIO $ NonLinearSpine x
-               | otherwise     -> pure (IM.insert x d r, d + 1)
+               | otherwise     -> pure (IM.insert x d r, d + 1, x:xs)
         _    -> throwIO SpineNonVar
     SProj1 _ -> throwIO SpineProjection
     SProj2 _ -> throwIO SpineProjection
@@ -165,13 +165,18 @@ closingTy cxt = go (cxt^.types) (cxt^.names) (cxt^.len) where
   go _                     _      _ _ = error "impossible"
 
 -- | Close a term by wrapping it in `Int` number of lambdas, while taking the domain
---   types from the `VTy`.
-closingTm :: (VTy, Int) -> Tm -> Tm
+--   types from the `VTy`, and the binder names from a list.
+closingTm :: (VTy, Int, [Name]) -> Tm -> Tm
 closingTm = go 0 where
-  go d (a, 0)   rhs = rhs
[…]

### 2020-02-23 implicit-fun-elab/Evaluation.hs

@@ -150,6 +150,8 @@ quote d = go where
 -- zonking
 --------------------------------------------------------------------------------
 
+-- | Unfold all metas and evaluate meta-headed spines, but don't evaluate
+--   anything else.
 zonk :: Vals -> Tm -> Tm
 zonk vs t = go t where
 

### 2020-02-23 implicit-fun-elab/Main.hs

@@ -164,7 +164,7 @@ test = main' "elab" $ unlines [
   "hundred"
   ]
 
--- example for when a telescope cannot depend on local A% B% vars.
+-- example for when a telescope cannot depend on local A B vars.
 -- this would only work with let generalization!
 outOfScopeTelRefinement = main' "elab" $ unlines [
   "let the : (A : U) → A → A = \\A x.x in",
@@ -291,10 +291,9 @@ ex1 = main' "elab" $ unlines [
   ]
 
 ex2 = main' "elab" $ unlines [
-  "let id : {A : U} → A → A = λ x. x in",
+  -- "let id : {A : U} → A → A = λ x. x in",
 
-  "let id2 : {A : U} → A → A = id id in",
-  "id U"
+  -- "let id2 : {A : U} → A → A = id id in",
 
   -- "let const : {A B} -> A -> B -> A",
   -- "    = \\x y. x in",
@@ -302,8 +301,8 @@ ex2 = main' "elab" $ unlines [
   -- -- definition types can be omitted
   -- "let constTy = {A B} → A → B → A in",
 
-  -- -- explicit id function, used for annotation as in Idris
-  -- "let the : (A : _) -> A -> A = \\_ x. x in",
+  -- explicit id function, used for annotation as in Idris
+  -- "let the : (A : _) -> A -> A = \\A x. x in",
 
   -- -- bool
   -- "let Bool : U",
@@ -346,14 +345,16 @@ ex2 = main' "elab" $ unlines [
   -- "    = \\N s z. s (s (s (s (s (s (s (s (s (s z))))))))) in",
   -- "let hundred = mul ten ten in",
 
-  -- -- Leibniz equality
-  -- "let Eq : {A} -> A -> A -> U",
-  -- "    = \\{A} x y. (P : A -> U) -> P x -> P y in",
-  -- "let refl : {A}{x : A} -> Eq x x",
-  -- "    = \\_ px. px in",
+  -- Leibniz equality
+  "let Eq : {A} -> A -> A -> U",
+  "    = \\{A} x y. (P : A -> U) -> P x -> P y in",
+  "let refl : {A}{x : A} -> Eq x x",
+  "    = \\P px. px in",
+  "let foo : Eq {U → U → U}(λ x y. x) (λ x y. x) = refl {U → U → U} in",
 
-  -- "the (Eq (mul ten ten) hundred) refl"
+  "foo"
   ]
+
 
   -- ?4 {A} (?3 {A})* {?6 {A}},A
 

### 2020-02-23 implicit-fun-elab/Types.hs

@@ -100,13 +100,6 @@ pattern TSnoc as a <- ((\case TBound as a -> Just (as, a)
 lvlName :: [Name] -> Lvl -> Name
 lvlName ns x = ns !! (length ns - x - 1)
 
--- ixType :: Types -> Ix -> VTy
--- ixType TNil           _ = error "impossible"
--- ixType (TDef   tys a) 0 = a
--- ixType (TBound tys a) 0 = a
--- ixType (TDef   tys a) x = ixType tys (x - 1)
--- ixType (TBound tys a) x = ixType tys (x - 1)
-
 data NameOrigin = NOSource | NOInserted
 
 type MetaInsertion = Bool
@@ -298,11 +291,11 @@ prettyTm prec = go (prec /= 0) where
     t@Pi{}         -> showParen p (goPi ns False t)
     U              -> ("U"++)
     Tel            -> ("Tel"++)
-    TEmpty         -> ("∙"++)
-    TCons "_" a as -> showParen p (go False ns a . (" ▶ "++). go False ns as)
+    TEmpty         -> ("ε"++)
+    TCons "_" a as -> showParen p (go False ns a . (" ▷ "++). go False ns as)
     TCons (fresh ns -> x) a as ->
               showParen p (showParen True ((x++) . (" : "++) . go False ns a)
-            . (" ▶ "++). go False (x:ns) as)
+            . (" ▷ "++). go False (x:ns) as)
     Tempty         -> ("[]"++)
     Rec a          -> showParen p (("Rec "++) . go True ns a)
     Tcons t u      -> showParen p (go True ns t . (" ∷ "++). go False ns u)

### 2020-02-23 paper.tex
stuff

@@
 
-%% %% alternative application infer
-  %% &\einfer{t\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  %% &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
-  %% &\quad \ecase\,\,A\,\,\eof\\
-  %% &\qquad   ((x : A)\to B) \To \edo\\
-  %% &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
-  %%                   \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
-  %% &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,u')}{(B[\theta_2][x\mapsto u'])})\\
-  %% &\qquad A \To \fail\\
-  %% &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  %% &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\false}{\Theta_0}{\Gamma}\\
-  %% &\quad \ecase\,\,A\,\,\eof\\
-  %% &\qquad   (\{x : A\}\to B) \To \edo\\
-  %% &\qquad\quad  (\Theta_2,\,\theta_2,\,u') \leftarrow
-  %%                   \echeck{u}{\Theta_1}{(\Gamma[\theta_1])}{A}\\
-  %% &\qquad\quad  \ereturn\,(\einsert{ins}{\Theta_2}{(\Gamma[\theta_1\circ\theta_2])}{(t'[\theta_2]\,\{u'\})}{(B[\theta_2][x\mapsto u'])})\\
-  %% &\qquad A \To \fail\\
-
 \subsubsection{Checking}
@@
 \begin{alignat*}{3}
-  & \scriptstyle{0 }\qquad\qquad\quad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}\\
+  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}
+     \hspace{14em}\\
   & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil}{\true} \\
@@
 
-\subsection{Handling Superfluous Implicit Functions}
-
-Before we can move on to unification, checking and inference, we have to address
+\subsection{Handling Superfluous Implicit Functions}\label{sec:superfluous}
+
+Before we can move on to unification and elaboration, we have to address
 a curious issue. Assuming $Bool : \U$, $true : Bool$ and $false : Bool$,
@@
 \subsubsection{Algorithmic implementation of constraint solving}
+\label{sec:algorithmic}
 The above specification for constancy constraints is compact but not
@@
 whenever we have constraints in contexts. In our prototype implementation, we
-augment the unification procedure with eager removal of solvable constraints.
+implement eager removal of solvable constraints.
 
@@
 \end{enumerate}
+Also, when we create a new constancy constraint, we immediately review it
+as described above.
 
@@
 
-Now, let us assume that $B'$ is not an implicit function type and not
-meta-headed. Then, we have the following case, where we solve a telescope domain
-to be empty.
+Now, let us assume that $B'$ is not an implicit function type, curried function
+type or meta-headed. Then, we have the following case, where we solve a
+telescope domain to be empty.
 \begin{alignat*}{3}
@@
 \subsection{Elaboration}
+
+In the definition of checking, we insert a new clause after
+$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
+\begin{alignat*}{3}
+& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{t})} :\equiv \edo\\
+& \quad (\Theta_1,\,\theta_1,\,A) \leftarrow \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
+& \quad (\Theta_2,\,\theta_2,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
+& \quad (\Theta_3,\,\theta_3) \leftarrow \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
+& \quad (\Theta_4,\,\theta_4) \leftarrow
+            \unify\, ((\alpha\,\ol{t})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
+& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
+\end{alignat*}
+Hence, when checking a term with a meta-headed type, we create a fresh meta with
+$\Tel$ type, infer a type for the term, and wrap the result in a strictly
+curried $\lambda$. We again need to create new constancy constraint. This way,
+if $B$ does not depend on $x : \Rec\,A$, the strictly curried $\lambda$ in the
+output immediately computes away, since $A$ is solved to $\epsilon$.
+
+This concludes the definition of the extended elaborator. The new algorithm is
+sound with respect to the extended core syntax, and it also has the
+conservativity property from Theorem \ref{thm:conservativity} if we additionally
+allow elaboration to insert $\lambda$-s for strictly curried functions. We present
+some examples of the algorithm at action.
+
+\begin{example}
+We return to the $polyList$ example from Section \ref{sec:issues}. We trace
+elaboration using the extended algorithm.
+\begin{alignat*}{3}
+  & \scriptstyle{0 }\qquad\qquad && \echeckt{cons\,(\lambda\,x.\,x)\,nil}{(List\,(\{A : \U\}\to A \to A))}
+      \hspace{14em}\\
+  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil}{\true} \\
+  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}{\true}\\
+  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}{\true}\\
+  & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
+  & \scriptstyle{5 }  && \qquad\quad \echeckt{\lambda\,x.\,x}{\alpha_0}\\
+  & \scriptstyle{6 }  && \qquad\quad = \lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x\\
+  & \scriptstyle{7 }  && \qquad = cons\,
+                       \{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,
+                       (\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\\
+  & \scriptstyle{8 }  && \qquad \echeckt{nil}{(List\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))}\\
+  & \scriptstyle{9 }  && \qquad = nil\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\\
+                       & \scriptstyle{10} && \quad = cons\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\}\,(\lambda\,\{y : \ol{\alpha_1}\}.\,\lambda\,x.\,x)\,\\
+                       & && \hspace{4.4em} (nil\,\{\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}\})\\
+  & && \hspace{3em} : List\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
+  & \scriptstyle{11} && \quad \unify\,(List\,(\{A : \U\}\to A \to A))\,(List\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\}))\\
+  & \scriptstyle{12} && \qquad \unify\,(\{A : \U\}\to A \to A)\,(\{y : \ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y : \ol{\alpha_1}\})\\
+  & \scriptstyle{13} && \qquad\quad \unify\,\alpha_1\,((A : \U) \TCons (\alpha_3\,A))\\
+  & \scriptstyle{14} && \qquad \quad \unify\,(A \to A)\,(\{z : \ol{\alpha_3\,A}\} \to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}\to \alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\})\\
+  & \scriptstyle{15} && \qquad \qquad \unify\,(\alpha_3\,A)\,\epsilon\\
+  & \scriptstyle{16} && \qquad \qquad \unify\,(A \to A)\,(\alpha_2\,\{A\} \to \alpha_2\,\{A\})\\
+  & \scriptstyle{17} && \qquad \qquad \quad \unify\,A\,(\alpha_2\,\{A\})\\
+  & \scriptstyle{18} && \qquad \qquad \quad \unify\,A\, A\\
+  & \scriptstyle{19} && = cons\,\{\{A : \U\}\to A \to A\}\,(\lambda\,\{A\}\,x.\,x)\,(nil\,\{\{A : \U\}\to A \to A\})
+\end{alignat*}
+We diverge from the previous attempt at line $\scriptstyle{5}$. Here, we check
+$\lambda\,x.\,x$ with the meta $\alpha_0$, so we create a fresh $\alpha_1 :
+\Tel$ meta and wrap the result as $\lambda\,\{y :
+\ol{\alpha_1}\}.\,\lambda\,x.\,x$. This result has the inferred type $\{y :
+\ol{\alpha_1}\} \to \alpha_2\,\{y : \ol{\alpha_1}\}\to \alpha_2\,\{y :
+\ol{\alpha_1}\}$, and we promptly unify $\alpha_0$ (which stands for the list
+element type) with it. On line $\scriptstyle{10}$, we return the elaborated
+$cons$ expression, where the list element type is made explicit. It only remains
+to unify the inferred and expected types. On line $\scriptstyle{12}$ we have the
+case where a curried function type is matched with an implicit function type, so
+on line $\scriptstyle{13}$ we refine $\alpha_1$ to a non-empty telescope, and
+proceed unifying the domains. Note that on line $\scriptstyle{14}$ we have
+$\alpha_2\,\{A\}\,\{z : \ol{\alpha_3\,A}\}$, which results from applying the
+substitution $y \mapsto (A :: z)$, and computing $\alpha_2\,\{(A :: z) : \ol{(A :
+  \U) \TCons (\alpha_3\,A)}\}$ further using the \LabTirName{app-$\TCons$} rule
+from Figure \ref{fig:telescopes}.
+
+On line $\scriptstyle{14}$ we have the case when a curried function type is
+matched with a type which is not a curried function, implicit function or a
+meta-headed type. We accordingly unify domain telescope $\alpha_3\,A$ with
+$\epsilon$, which causes $\alpha_3$ to be solved as $\lambda\,A.\,\epsilon$
+according to standard pattern unification. Now, $\alpha_2\,\{A\}\,\{z :
+\ol{\alpha_3\,A}\}$ computes to $\alpha_2\,\{A\}\,\{z : \epsilon\}$, which
+further computes to $\alpha_2\,\{A\}$ by \LabTirName{app-$\epsilon$}. From here,
+unification finishes with success. On the last line, the expected output is
+returned: since the $\alpha_1$ telescope meta is now solved, curried function
+types and abstractions are computed away.
+\end{example}
+
+\begin{example}
+We illustrate now the action of constancy constraints, using the example from
+Section \ref{sec:superfluous}. Again, assume $Bool : \U$ and $true : Bool$.
+\begin{alignat*}{3}
+& \scriptstyle{0}\qquad\qquad && \einfert{\slet\,x : \_ = true\,\sin\,x}{\true}
+      \hspace{21em}\\
+& \scriptstyle{1} && \quad \echeckt{\_}{\U}\\
+& \scriptstyle{2} && \quad = \alpha_0\\
+& \scriptstyle{3} && \quad \echeckt{true}{\alpha_0}\\
+& \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel\\
+& \scriptstyle{5} && \qquad = \alpha_1\\
+& \scriptstyle{6} && \qquad \einfert{true}{\true}\\
+& \scriptstyle{7} && \qquad = true : Bool\\
+& \scriptstyle{8} && \qquad \mathsf{newConstancy}_{\emptycon,\,x : \Rec\,\alpha_1}\,Bool\\
+& \scriptstyle{9} && \qquad\quad \unify\,\alpha_1\,\epsilon\\
+& \scriptstyle{10} && \qquad \unify\,\alpha_0\,Bool\\
+& \scriptstyle{11} && \quad = true\\
+& \scriptstyle{12} && \quad \einfert{x}{\true}\\
+& \scriptstyle{13} && \quad = true : Bool\\
+& \scriptstyle{14} && = (\slet\,x : Bool = true\,\sin\,x) : Bool
+\end{alignat*}
+First we create a fresh meta $\alpha_0$ for the hole on line $\scriptstyle{2}$,
+then we check $\true$ with it. Since we are checking with a meta-headed type, we
+create a fresh telescope meta $\alpha_1$, then infer $Bool$ for $true$. On line
+$\scriptstyle{8}$ we create a constancy constraint, and immediately try to solve
+it, in accordance with the algorithmic implementation in Section
+\ref{sec:algorithmic}. Since $Bool$ does not depend on the domain, we solve the
+constraint and thereby solve $\alpha_1$ as $\epsilon$. Hence, we simply return
+$true$ on line $\scriptstyle{11}$, since $\lambda\,\{x : \ol{\epsilon}\}.\,true$
+computes to that, and elaboration succeeds with the expected output.
+\end{example}
 

### 2020-02-23 implicit-fun-elab/Main.hs

@@ -78,14 +78,9 @@ pruneTest2 = main' "elab" $ unlines [
  ]
 
 test0 = main' "elab" $ unlines [
-  "let the : (A : U) → A → A = λ A x. x in",
-  "let IdTy = {A} → A → A in",
-  "let choose : {A} → A → A → A = λ x y. x in",
-  "let id : IdTy = λ x. x in",
-  "let auto2 : {B} → IdTy → B → B = λ f b. b in",
-  "let r : ({A} → A → {B} → B → B) → U = λ _. U in",
-  "let foo = r (λ x y. y) in",
-  "U"
+  "λ (Bool : U)(true : Bool).",
+
+  "let x : _ = true in x"
   ]
 
 
@@ -291,68 +286,67 @@ ex1 = main' "elab" $ unlines [
   ]
 
 ex2 = main' "elab" $ unlines [
-  -- "let id : {A : U} → A → A = λ x. x in",
+  "let id : {A : U} → A → A = λ x. x in",
 
-  -- "let id2 : {A : U} → A → A = id id in",
+  "let id2 : {A : U} → A → A = id id in",
 
-  -- "let const : {A B} -> A -> B -> A",
-  -- "    = \\x y. x in",
+  "let const : {A B} -> A -> B -> A",
+  "    = \\x y. x in",
 
-  -- -- definition types can be omitted
-  -- "let constTy = {A B} → A → B → A in",
+  -- definition types can be omitted
+  "let constTy = {A B} → A → B → A in",
 
   -- explicit id function, used for annotation as in Idris
-  -- "let the : (A : _) -> A -> A = \\A x. x in",
+  "let the : (A : _) -> A -> A = \\A x. x in",
 
-  -- -- bool
-  -- "let Bool : U",
-  -- "    = (B : _) -> B -> B -> B in",
-  -- "let true : Bool",
-  -- "    = \\B t f. t in",
-  -- "let false : Bool",
-  -- "    = \\B t f. f in",
-  -- "let not : Bool -> Bool",
-  -- "    = \\b B t f. b B f t in",
+  -- bool
+  "let Bool : U",
+  "    = (B : _) -> B -> B -> B in",
+  "let true : Bool",
+  "    = \\B t f. t in",
+  "let false : Bool",
+  "    = \\B t f. f in",
+  "let not : Bool -> Bool",
+  "    = \\b B t f. b B f t in",
 
-  -- -- lists
-  -- "let List : U -> U",
[…]

### 2020-02-23 implicit-fun-elab/Elaboration.hs
stuff

@@ -22,22 +22,19 @@ import Debug.Trace
 emptyCxt :: Cxt
 emptyCxt = Cxt VNil TNil [] [] 0
 
+-- | Add a bound variable.
 bind :: Name -> NameOrigin -> VTy -> Cxt -> Cxt
 bind x o ~a (Cxt vs tys ns no d) =
   Cxt (VSkip vs) (TBound tys a) (x:ns) (o:no) (d + 1)
 
+-- | Add a bound variable which comes from surface syntax.
 bindSrc :: Name -> VTy -> Cxt -> Cxt
 bindSrc x = bind x NOSource
 
+-- | Define a new variable.
 define :: Name -> VTy -> Val -> Cxt -> Cxt
 define x ~a ~t (Cxt vs tys ns no d) =
   Cxt (VDef vs t) (TDef tys a) (x:ns) (NOSource:no) (d + 1)
-
-addSrcPos :: SPos -> IO a -> IO a
-addSrcPos p act = act `catch` \(e::Err) ->
-  case e^.pos of
-    Nothing -> throwIO (e & pos .~ Just p)
-    _       -> throwIO e
 
 -- | Lift ("skolemize") a value in an extended context to a function in a
 --   non-extended context.
@@ -45,7 +42,7 @@ liftVal :: Cxt -> Val -> (Val -> Val)
 liftVal cxt t = \ ~x -> eval (VDef (cxt^.vals) x) $ quote (cxt^.len+1) t
 
 
--- Occurs check for constancy constraints
+-- Constancy constraints
 --------------------------------------------------------------------------------
 
 data Occurs = Rigid | Flex IS.IntSet | None deriving (Eq, Show)
@@ -64,6 +61,7 @@ occurrence ms | IS.null  ms = Rigid
 instance Monoid Occurs where
   mempty = None
 
+-- | Occurs check for the purpose of constraint solving.
 occurs :: Lvl -> Lvl -> Val -> Occurs
 occurs d topX = occurs' d mempty where
 
@@ -165,7 +163,8 @@ closingTy cxt = go (cxt^.types) (cxt^.names) (cxt^.len) where
   go _                     _      _ _ = error "impossible"
 
 -- | Close a term by wrapping it in `Int` number of lambdas, while taking the domain
---   types from the `VTy`, and the binder names from a list.
+--   types from the `VTy`, and the binder names from a list. If we run out of provided
[…]

### 2020-02-23 implicit-fun-elab/Types.hs

@@ -15,6 +15,7 @@ import qualified Data.IntSet        as IS
 -- Raw syntax
 --------------------------------------------------------------------------------
 
+-- | We wrap `SourcePos` to avoid printing it in `Show`.
 newtype SPos = SPos SourcePos deriving (Eq, Ord, Read)
 instance Show SPos where show _ = ""
 
@@ -30,14 +31,15 @@ icit Impl i e = i
 icit Expl i e = e
 
 data Raw
-  = RVar Name
-  | RLam Name (Maybe Raw) Icit Raw
-  | RApp Raw Raw Icit
-  | RU
-  | RPi Name Icit Raw Raw
-  | RLet Name Raw Raw Raw
-  | RHole
-  | RSrcPos SPos Raw
+  = RVar Name                        -- ^ x
+  | RLam Name (Maybe Raw) Icit Raw   -- ^ λx.t  or λ{x}.t with optional type annotation
+                                     --   on x
+  | RApp Raw Raw Icit                -- ^ t u  or  t {u}
+  | RU                               -- ^ U
+  | RPi Name Icit Raw Raw            -- ^ (x : A) → B  or  {x : A} → B
+  | RLet Name Raw Raw Raw            -- ^ let x : A = t in u
+  | RHole                            -- ^ _
+  | RSrcPos SPos Raw                 -- ^ source position annotation, added by parsing
 
 deriving instance Show Raw
 
@@ -56,9 +58,8 @@ data MetaEntry
   = Unsolved Blocking ~VTy
   | Solved Val
 
-  -- | Telescope constancy constraint. When the closure becomes constant,
-  --   we unify the telescope with the empty telescope.
-  --   Constancy context domain codomain blockers
+  -- | Constancy (Γ, x : Rec A) B   + a list of blocking metas.
+  --   When B becomes constant, A is solved to ε
   | Constancy Cxt VTy VTy BlockedBy
 
 
@@ -66,7 +67,7 @@ data MetaEntry
 --   out-of-scope or "illegal" variables.
 type Renaming = IM.IntMap Lvl
 
--- | A strengthening. We use this for pruning and checking meta solution
[…]

### 2020-02-23 implicit-fun-elab/Elaboration.hs
stuff, minor flex-flex improvement

@@ -322,6 +322,11 @@ unify cxt l r = go l r where
   unifyError =
     throwIO $ UnifyError (cxt^.names) (quote (cxt^.len) l) (quote (cxt^.len) r)
 
+  flexFlex m sp m' sp' = do
+    try @SpineError (checkSp sp) >>= \case
+      Left{}  -> solveMeta cxt m' sp' (VNe (HMeta m) sp)
+      Right{} -> solveMeta cxt m sp (VNe (HMeta m') sp')
+
   go t t' = case (force t, force t') of
     (VLam x _ a t, VLam _ _ _ t')            -> goBind x a t t'
     (VLam x i a t, t')                       -> goBind x a t (\ ~v -> vApp t' v i)
@@ -338,6 +343,7 @@ unify cxt l r = go l r where
     (VLamTel x a t, VLamTel x' a' t')        -> goBind x (VRec a) t t'
     (VLamTel x a t, t')                      -> goBind x (VRec a) t (vAppTel a t')
     (t, VLamTel x' a' t')                    -> goBind x' (VRec a') (vAppTel a' t) t'
+    (VNe (HMeta m) sp, VNe (HMeta m') sp')   -> flexFlex m sp m' sp'
     (VNe h sp, VNe h' sp') | h == h'         -> goSp (forceSp sp) (forceSp sp')
     (VNe (HMeta m) sp, t')                   -> solveMeta cxt m sp t'
     (t, VNe (HMeta m') sp')                  -> solveMeta cxt m' sp' t

### 2020-02-23 implicit-fun-elab/Types.hs

@@ -235,7 +235,18 @@ addSrcPos p act = act `catch` \case
 --------------------------------------------------------------------------------
 
 prettyTm :: Int -> [Name] -> Tm -> ShowS
-prettyTm prec = go (prec /= 0) where
+prettyTm prec = topLams (prec /= 0) "λ" "" where
+
+  -- special case printing of top lambdas, since they are usually used
+  -- to postulate stuff
+  topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
+  topLams p pre post ns (Lam (fresh ns -> x) i a t) =
+    showParen p (
+      (pre++)
+    . icit i bracket parens (
+           ((if null x then "_" else x)++) . (" : "++) . go False ns a)
+    . topLams False "\n" ".\n\n" (x:ns) t)
+  topLams _ pre post ns t = (post++) . go False ns t
 
   fresh :: [Name] -> Name -> Name
   fresh _ "_" = "_"
@@ -255,8 +266,8 @@ prettyTm prec = go (prec /= 0) where
   goLamBind :: Name -> Icit -> ShowS
   goLamBind x i = icit i bracket id ((if null x then "_" else x) ++)
 
-  bracket :: ShowS -> ShowS
   bracket s = ('{':).s.('}':)
+  parens  s = ('(':).s.(')':)
 
   goLam :: [Name] -> Tm -> ShowS
   goLam ns (Lam (fresh ns -> x) i a t)  = (' ':) . goLamBind x i . goLam (x:ns) t
@@ -309,8 +320,8 @@ prettyTm prec = go (prec /= 0) where
     Tempty         -> ("[]"++)
     Rec a          -> showParen p (("Rec "++) . go True ns a)
     Tcons t u      -> showParen p (go True ns t . (" ∷ "++). go False ns u)
-    Proj1 t        -> showParen p (("₁ "++). go True ns t)
-    Proj2 t        -> showParen p (("₂ "++). go True ns t)
+    Proj1 t        -> showParen p (("π₁ "++). go True ns t)
+    Proj2 t        -> showParen p (("π₂ "++). go True ns t)
     t@PiTel{}      -> showParen p (goPi ns False t)
     AppTel a (App t u i) u'  ->
       showParen p (go False ns t . (' ':) . goArg ns u i . (' ':) .
@@ -336,7 +347,6 @@ instance Show Tm where show = showTm []
 
 showError :: [Name] -> ElabError -> String
[…]

### 2020-02-24 paper.tex
stuff
module reorg

@@
 and it is beyond our reach to succinctly characterize which annotations are
-inferable in the surface language.
+inferable in the surface language, and we do not know of any prior work which
+accomplishes this for a comparably strong elaborator. Our experience in Agda is
+that it is not tractable in general to figure out which arguments are inferable,
+by looking at function types; often we have to run elaboration to see what
+works.
 
@@
 ``evil'' construction on core terms, since it does not preserve definitional
-equality. We shall only use this evil notion in the following statement.
+equality, but we shall only use this evil notion in the following statement.
 
@@
   \begin{itemize}
-    \item Having all $\_$ holes filled in by expressions.
+    \item Having all $\_$ holes filled with core expressions.
     \item Having extra implicit $\lambda$-s and implicit applications inserted.
@@
 \end{theorem}
+
+\emph{Remark.} It is \emph{not} the case that for every term in the core syntax
+there exists a surface term which elaborates to it. The main reason is that
+$\lambda$-s in the surface syntax are not annotated, and it is easy to find core
+$\lambda$ expressions with uninferable domain types. We skip optional surface
+$\lambda$ domain type annotations for the sake of brevity.
 
@@
 
+\section{Implementation}
+\label{sec:impl_and_eval}
+
+We provide a implementation of the elaborator described in Section
+\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
+a surface expression from standard input, and outputs the result of elaboration,
+or optionally the type or the normal form of the result.
+
 \begin{acks}

### 2020-02-24 implicit-fun-elab/Elaboration.hs

@@ -13,6 +13,7 @@ import qualified Data.IntSet as IS
 import Types
 import Evaluation
 import ElabState
+import Errors
 
 import Debug.Trace
 

### 2020-02-24 implicit-fun-elab/Errors.hs

@@ -1,0 +1,98 @@
+
+module Errors where
+
+import Control.Exception
+import Text.Printf
+
+import Types
+import Pretty
+
+--------------------------------------------------------------------------------
+
+data SpineError
+  = SpineNonVar
+  | SpineProjection
+  | NonLinearSpine Lvl
+  deriving (Show, Exception)
+
+data StrengtheningError
+  = ScopeError Lvl
+  | OccursCheck
+  deriving (Show, Exception)
+
+data UnifyError
+  = UnifyError [Name] Tm Tm
+  | SpineError [Name] Tm Tm SpineError
+  | StrengtheningError [Name] Tm Tm StrengtheningError
+  deriving (Show, Exception)
+
+data ElabError
+  = UnifyErrorWhile Tm Tm UnifyError
+  | NameNotInScope Name
+  | ExpectedFunction Tm
+  | IcitMismatch Icit Icit
+
+data Err = Err {
+  errNames :: [Name],
+  errErr   :: ElabError,
+  errPos   :: Maybe SPos}
+
+instance Show Err where
+  show _ = "Error"
+
+instance Exception Err
+
+report :: [Name] -> ElabError -> a
+report ns e = throw (Err ns e Nothing)
+
+-- | Rethrow an `Err` with source position attached.
+addSrcPos :: SPos -> IO a -> IO a
+addSrcPos p act = act `catch` \case
+  Err ns e Nothing -> throwIO (Err ns e (Just p))
+  e                -> throwIO e
+
+
+showError :: [Name] -> ElabError -> String
+showError ns = \case
+  UnifyErrorWhile lhs rhs e ->
+    let err1 = case e of
+          UnifyError ns lhs rhs -> printf
+            ("Cannot unify\n\n" ++
+             "  %s\n\n" ++
+             "with\n\n" ++
+             "  %s\n\n")
+            (showTm ns lhs) (showTm ns rhs)
+          StrengtheningError ns lhs rhs e -> case e of
+            ScopeError x -> printf (
+              "Variable %s is out of scope in equation\n\n" ++
+              "  %s =? %s\n\n")
+              (lvlName ns x) (showTm ns lhs) (showTm ns rhs)
+            OccursCheck -> printf (
+              "Meta occurs cyclically in its solution candidate in equation:\n\n" ++
+              "  %s =? %s\n\n")
+              (showTm ns lhs) (showTm ns rhs)
+          SpineError ns lhs rhs e -> case e of
[…]

### 2020-02-24 implicit-fun-elab/Evaluation.hs

@@ -145,55 +145,3 @@ quote d = go where
     VLamTel x a t -> LamTel x (go a) (goBind t)
 
   goBind t = quote (d + 1) (t (VVar d))
-
-
--- zonking
---------------------------------------------------------------------------------
-
--- | Unfold all metas and evaluate meta-headed spines, but don't evaluate
---   anything else.
-zonk :: Vals -> Tm -> Tm
-zonk vs t = go t where
-
-  goSp :: Tm -> Either Val Tm
-  goSp = \case
-    Meta m       -> case runLookupMeta m of
-                      Solved v -> Left v
-                      _        -> Right (Meta m)
-    App t u ni   -> case goSp t of
-                      Left t  -> Left (vApp t (eval vs u) ni)
-                      Right t -> Right $ App t (go u) ni
-    AppTel a t u -> case goSp t of
-                      Left t  -> Left (vAppTel (eval vs a) t (eval vs u))
-                      Right t -> Right $ AppTel (go a) t (go u)
-    t            -> Right (zonk vs t)
-
-  goBind = zonk (VSkip vs)
-
-  go = \case
-    Var x        -> Var x
-    Meta m       -> case runLookupMeta m of
-                      Solved v   -> quote (valsLen vs) v
-                      Unsolved{} -> Meta m
-                      _          -> error "impossible"
-    U            -> U
-    Pi x i a b   -> Pi x i (go a) (goBind b)
-    App t u ni   -> case goSp t of
-                      Left t  -> quote (valsLen vs) (vApp t (eval vs u) ni)
-                      Right t -> App t (go u) ni
-    Lam x i a t  -> Lam x i (go a) (goBind t)
-    Let x a t u  -> Let x (go a) (go t) (goBind u)
-    Tel          -> Tel
-    TEmpty       -> TEmpty
-    TCons x t u  -> TCons x (go t) (goBind u)
-    Rec a        -> Rec (go a)
-    Tempty       -> Tempty
-    Tcons t u    -> Tcons (go t) (go u)
-    Proj1 t      -> Proj1 (go t)
-    Proj2 t      -> Proj2 (go t)
-    PiTel x a b  -> PiTel x (go a) (goBind b)
-    AppTel a t u -> case goSp t of
-                      Left t  -> quote (valsLen vs) (vAppTel (eval vs a) t (eval vs u))
-                      Right t -> AppTel (go a) t (go u)
-    LamTel x a b -> LamTel x (go a) (goBind b)
-    Skip t       -> Skip (goBind t)

### 2020-02-24 implicit-fun-elab/Main.hs

@@ -7,10 +7,14 @@ import System.Exit
 import System.Environment
 
 import Types
+import Pretty
 import Evaluation
 import Elaboration
+import Errors
 import Parser
 import ElabState
+import Zonk
+
 
 helpMsg = unlines [
   "usage: holes [--help|nf|type]",
@@ -348,13 +352,3 @@ ex2 = main' "elab" $ unlines [
 
   "the (Eq hundred hundred) refl"
   ]
-
-
-  -- ?4 {A} (?3 {A})* {?6 {A}},A
-
-  -- α A {β A : γ A} =? A
-
-  -- γ = λ A. A ▶ ∙
-  -- α = λ _ A. A
-
--- the ? (λ {y : γ} x → x){α:γ}

### 2020-02-24 implicit-fun-elab/Pretty.hs

@@ -1,0 +1,116 @@
+{-# options_ghc -Wno-orphans #-}
+
+module Pretty where
+
+import Types
+
+prettyTm :: Int -> [Name] -> Tm -> ShowS
+prettyTm prec = topLams (prec /= 0) "λ" "" where
+
+  -- special case printing of top lambdas, since they are usually used
+  -- to postulate stuff
+  topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
+  topLams p pre post ns (Lam (fresh ns -> x) i a t) =
+    showParen p (
+      (pre++)
+    . icit i bracket parens (
+           ((if null x then "_" else x)++) . (" : "++) . go False ns a)
+    . topLams False "\n" ".\n\n" (x:ns) t)
+  topLams _ pre post ns t = (post++) . go False ns t
+
+  fresh :: [Name] -> Name -> Name
+  fresh _ "_" = "_"
+  fresh ns n | elem n ns = fresh ns (n++"'")
+             | otherwise = n
+
+  goVar :: [Name] -> Ix -> ShowS
+  goVar ns topX = go ns topX where
+    -- go []     _ = error "impossible"
+    go []     _ = (show topX++)
+    go (n:ns) 0 = (n++)
+    go (n:ns) x = go ns (x - 1)
+
+  goArg :: [Name] -> Tm -> Icit -> ShowS
+  goArg ns t i = icit i (bracket (go False ns t)) (go True ns t)
+
+  goLamBind :: Name -> Icit -> ShowS
+  goLamBind x i = icit i bracket id ((if null x then "_" else x) ++)
+
+  bracket s = ('{':).s.('}':)
+  parens  s = ('(':).s.(')':)
+
+  goLam :: [Name] -> Tm -> ShowS
+  goLam ns (Lam (fresh ns -> x) i a t)  = (' ':) . goLamBind x i . goLam (x:ns) t
+  goLam ns (LamTel(fresh ns -> x) a t) =
+    (' ':) . bracket ((x++) . (" : "++) . go False ns a) . goLam (x:ns) t
+  goLam ns t = (". "++) . go False ns t
+
+  goPiBind :: [Name] -> Name -> Icit -> Tm -> ShowS
+  goPiBind ns x i a =
+    icit i bracket (showParen True) ((x++) . (" : "++) . go False ns a)
+
+  goPi :: [Name] -> Bool -> Tm -> ShowS
[…]

### 2020-02-24 implicit-fun-elab/Types.hs

@@ -185,214 +185,8 @@ pattern VVar x = VNe (HVar x) SNil
 pattern VMeta :: MId -> Val
 pattern VMeta m = VNe (HMeta m) SNil
 
--- Errors
---------------------------------------------------------------------------------
-
-data SpineError
-  = SpineNonVar
-  | SpineProjection
-  | NonLinearSpine Lvl
-  deriving (Show, Exception)
-
-data StrengtheningError
-  = ScopeError Lvl
-  | OccursCheck
-  deriving (Show, Exception)
-
-data UnifyError
-  = UnifyError [Name] Tm Tm
-  | SpineError [Name] Tm Tm SpineError
-  | StrengtheningError [Name] Tm Tm StrengtheningError
-  deriving (Show, Exception)
-
-data ElabError
-  = UnifyErrorWhile Tm Tm UnifyError
-  | NameNotInScope Name
-  | ExpectedFunction Tm
-  | IcitMismatch Icit Icit
-
-data Err = Err {
-  errNames :: [Name],
-  errErr   :: ElabError,
-  errPos   :: Maybe SPos}
-
-instance Show Err where
-  show _ = "Error"
-
-instance Exception Err
-
-report :: [Name] -> ElabError -> a
-report ns e = throw (Err ns e Nothing)
-
--- | Rethrow an `Err` with source position attached.
-addSrcPos :: SPos -> IO a -> IO a
-addSrcPos p act = act `catch` \case
-  Err ns e Nothing -> throwIO (Err ns e (Just p))
-  e                -> throwIO e
-
-
--- Pretty printing
---------------------------------------------------------------------------------
-
-prettyTm :: Int -> [Name] -> Tm -> ShowS
-prettyTm prec = topLams (prec /= 0) "λ" "" where
-
-  -- special case printing of top lambdas, since they are usually used
-  -- to postulate stuff
-  topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
-  topLams p pre post ns (Lam (fresh ns -> x) i a t) =
-    showParen p (
-      (pre++)
-    . icit i bracket parens (
-           ((if null x then "_" else x)++) . (" : "++) . go False ns a)
-    . topLams False "\n" ".\n\n" (x:ns) t)
-  topLams _ pre post ns t = (post++) . go False ns t
-
-  fresh :: [Name] -> Name -> Name
-  fresh _ "_" = "_"
[…]

### 2020-02-24 implicit-fun-elab/Zonk.hs

@@ -1,0 +1,54 @@
+
+module Zonk where
+
+import Types
+import Evaluation
+import ElabState
+
+-- | Unfold all metas and evaluate meta-headed spines, but don't evaluate
+--   anything else.
+zonk :: Vals -> Tm -> Tm
+zonk vs t = go t where
+
+  goSp :: Tm -> Either Val Tm
+  goSp = \case
+    Meta m       -> case runLookupMeta m of
+                      Solved v -> Left v
+                      _        -> Right (Meta m)
+    App t u ni   -> case goSp t of
+                      Left t  -> Left (vApp t (eval vs u) ni)
+                      Right t -> Right $ App t (go u) ni
+    AppTel a t u -> case goSp t of
+                      Left t  -> Left (vAppTel (eval vs a) t (eval vs u))
+                      Right t -> Right $ AppTel (go a) t (go u)
+    t            -> Right (zonk vs t)
+
+  goBind = zonk (VSkip vs)
+
+  go = \case
+    Var x        -> Var x
+    Meta m       -> case runLookupMeta m of
+                      Solved v   -> quote (valsLen vs) v
+                      Unsolved{} -> Meta m
+                      _          -> error "impossible"
+    U            -> U
+    Pi x i a b   -> Pi x i (go a) (goBind b)
+    App t u ni   -> case goSp t of
+                      Left t  -> quote (valsLen vs) (vApp t (eval vs u) ni)
+                      Right t -> App t (go u) ni
+    Lam x i a t  -> Lam x i (go a) (goBind t)
+    Let x a t u  -> Let x (go a) (go t) (goBind u)
+    Tel          -> Tel
+    TEmpty       -> TEmpty
+    TCons x t u  -> TCons x (go t) (goBind u)
+    Rec a        -> Rec (go a)
+    Tempty       -> Tempty
+    Tcons t u    -> Tcons (go t) (go u)
+    Proj1 t      -> Proj1 (go t)
+    Proj2 t      -> Proj2 (go t)
+    PiTel x a b  -> PiTel x (go a) (goBind b)
+    AppTel a t u -> case goSp t of
+                      Left t  -> quote (valsLen vs) (vAppTel (eval vs a) t (eval vs u))
+                      Right t -> AppTel (go a) t (go u)
+    LamTel x a b -> LamTel x (go a) (goBind b)
+    Skip t       -> Skip (goBind t)

### 2020-02-24 paper.tex
stuff

@@
 for full dependent type theories and compares favorably to prior solutions in
-terms of power, generality and conceptual simplicity. We build atop Norell's
+terms of power, generality and simplicity. We build atop Norell's
 bidirectional elaboration algorithm for Agda, and we note that the key issue is
@@
     Agda-like surface language to a small Martin-L\"of type theory extended with
-    implicit function types, telescopes and \emph{strictly curried} function
-    types with telescope domain. We use these extensions to accummulate
-    information about implicit insertions.
-  \item Our algorithm is a conservative extension of Norell's bidirectional
-    elaborator for Agda \cite[Chapter~3]{norell07thesis}; it accepts strictly more
-    programs, and does not require new constructs in the surface language.
-  \item Our target language serves as a general platform for elaborating
-    implicit functions. The concrete elaborator presented in this paper is a
-    relatively simple one, and there is plenty of room to develop more advanced
-    elaboration and unification. However, our simple algorithm is already
-    comparable or superior to previous solutions for impredicative inference.
+    implicit function types, telescopes and \emph{strictly curried function
+      types} with telescope domain. We use these extensions to accummulate
+    information about implicit insertions. Our algorithm is a conservative
+    extension of Norell's bidirectional elaborator for Agda
+    \cite[Chapter~3]{norell07thesis}.
+  \item The presented elaborator is generally more powerful than previous
+    solutions for impredicative inference. In particular, it goes beyond prior
+    art by supporting full dependent type theory and by being global,
+    i.e.\ inference may consider the whole program and not just particular n-ary
+    applications.
+  \item The target theory of elaboration serves as a general platform for
+    elaborating implicit function types: our concrete elaborator is a relatively
+    simple one, and there is room to further develop it.
   \item We provide an executable implementation of the elaborator described in
     this paper.
+  \item Our solution is also \emph{simple}: we implemented elaboration,
+    evaluation and unification in 650 lines of Haskell, of which about 200 lines
+    implement the novel enhancements on the top of Norell's basic elaborator.
 \end{itemize}
@@
 
-\begin{notation}
-Both in the surface and core syntax, we use Agda-like notation:
+\emph{Notations.} We use Agda-like syntactic sugar both in the surface syntax and
+in core syntax.
 \begin{itemize}
@@
 \end{itemize}
-\end{notation}
-
-\begin{notation}\label{not:spines}
+\begin{definition}[Spines]\label{def:spines}
   We use a spine notation for neutral terms. A spine is a list of
@@
   particular, if $t$ is a metavariable, the neutral term is \emph{meta-headed}.
-\end{notation}
-
+\end{definition}
 \begin{example}
@@
 \end{example}
-%% In the core syntax, all abstractions and applications must be notated. Thanks to
-%% elaboration, in the surface syntax we may omit some of these.
 
@@
 \end{mathpar}
+
+%% \end{mathpar}
+%% \begin{mathpar}
 
@@
 
-We provide a implementation of the elaborator described in Section
+We provide an implementation of the elaborator from Section
 \ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
@@
 or optionally the type or the normal form of the result.
+
+It is implemented in 834 lines of Haskell. Of this, 660 lines constitute the
+core syntax, evaluation, unification and elaboration. Of this 660 lines, 445
+lines implement the basic elaborator of Section
+\ref{sec:bidirectional_elaboration} and 215 lines implement the extended
+elaborator. We give here implementation-specific details and describe
+differences to the presentation in this paper.
+
+Elaboration is implemented in the style of Coquand's algorithm
+\cite{coquand1996algorithm}, where elaboration is interleaved with
+normalization-by-evaluation. Hence, we do not perform any substitution
+operations on core syntax, instead we evaluate core terms into the semantic
+domain, and perform unification, strengthening and occurrence checking on
+semantic values. There is also a quoting (or readback) operation which yields
+normalized core terms from values, and which is used most prominently when we
+have to generate solutions for metavariables.
+
+Metacontexts are a key point in the implementation. We avoid the tedious (and
+inefficient) threading of composed metasubstitutions, instead we have a mutable
+reference which stores the current metasubstitution. We have a ``forcing''
+operation which computes a semantic value to a head normal form, with respect to
+the current metasubstitution. Hence, instead of constantly updating every term
+and type by metasubstitution, we only force them whenever we need to match on
+the shape of types, for example when we are inserting implicit arguments based
+on types. In this paper we still stick with the ``threaded metasubstitution''
+presentation because it is more conventional, and also more straightforward to
+formalize.
+
+We use normalization-by-evaluation in Abel's style (see
+e.g.\ \cite[Chapter~3]{abel2013normalization}), where semantic values use de
+Bruijn levels, and the core syntax uses de Bruijn indices. This is practically
+very favorable, because the evaluator never has to perform weakening on values.
+The implementation of curried functions presents a bit of a complication,
+because the computation rules are type-directed, so we have to annotate
+applications and abstractions with telescopes (just as in our notation for the
+core syntax).
+
+We provide a fairly optimized implementation for constraints: we keep track of
+``blocking'' metas for each constraint, which are relevant to checking the
+constancy condition, and upon solving a meta we only review constraints which
+were blocked on the meta.
+
+\section{Evaluation}
+\label{sec:impl_and_eval}
+
+%% \begin{itemize}
+%% \item Elaboration is implemented in the style of Coquand's algorithm \cite{coquand1996algorithm},
+%%       where elaboration is interleaved with normalization-by-evaluation.
+%% \end{itemize}
 

### 2020-02-24 implicit-fun-elab/Elaboration.hs
RELEASE
this version has top-level lambda printing
and no postulates

@@ -344,8 +344,8 @@ unify cxt l r = go l r where
     (VLamTel x a t, VLamTel x' a' t')        -> goBind x (VRec a) t t'
     (VLamTel x a t, t')                      -> goBind x (VRec a) t (vAppTel a t')
     (t, VLamTel x' a' t')                    -> goBind x' (VRec a') (vAppTel a' t) t'
-    (VNe (HMeta m) sp, VNe (HMeta m') sp')   -> flexFlex m sp m' sp'
     (VNe h sp, VNe h' sp') | h == h'         -> goSp (forceSp sp) (forceSp sp')
+    (VNe (HMeta m) sp, VNe (HMeta m') sp')   -> flexFlex m sp m' sp'
     (VNe (HMeta m) sp, t')                   -> solveMeta cxt m sp t'
     (t, VNe (HMeta m') sp')                  -> solveMeta cxt m' sp' t
 

### 2020-02-25 paper.tex
stuff
stuff, nicer top lambda printing

@@
   \item Our solution is also \emph{simple}: we implemented elaboration,
-    evaluation and unification in 650 lines of Haskell, of which about 200 lines
+    evaluation and unification in about 670 lines of Haskell, of which 210 lines
     implement the novel enhancements on the top of Norell's basic elaborator.
@@
 
-It is implemented in 834 lines of Haskell. Of this, 660 lines constitute the
-core syntax, evaluation, unification and elaboration. Of this 660 lines, 445
+\todo{update loc}
+
+It is implemented in about 1200 lines of Haskell. Of this, 673 lines constitute
+the core syntax, evaluation, unification and elaboration. Of this 673 lines, 458
 lines implement the basic elaborator of Section

### 2020-02-25 implicit-fun-elab/Elaboration.hs

@@ -443,6 +443,21 @@ check cxt topT ~topA = case (topT, force topA) of
     unifyWhile cxt va topA
     pure t
 
+-- | We specialcase top-level lambdas (serving as postulates) for better
+--   printing: we don't print them in meta spines.
+inferTopLams :: Cxt -> Raw -> IO (Tm, VTy)
+inferTopLams cxt = \case
+  RLam x ann i t -> do
+    a <- case ann of
+      Just ann -> check cxt ann VU
+      Nothing  -> freshMeta cxt VU
+    let ~va = eval (cxt^.vals) a
+    (t, liftVal cxt -> b) <- inferTopLams (bind ('*':x) NOSource va cxt) t
+    pure (Lam x i a t, VPi x i va b)
+  RSrcPos p t ->
+    addSrcPos p $ inferTopLams cxt t
+  t ->
+    infer cxt True t
 
 infer :: Cxt -> MetaInsertion -> Raw -> IO (Tm, VTy)
 infer cxt ins = \case
@@ -453,7 +468,7 @@ infer cxt ins = \case
 
   RVar x -> insert cxt ins $ do
     let go :: [Name] -> [NameOrigin] -> Types -> Int -> IO (Tm, VTy)
-        go (y:xs) (NOSource:os) (TSnoc _  a) i | x == y = pure (Var i, a)
+        go (y:xs) (NOSource:os) (TSnoc _  a) i | x == y || ('*':x) == y = pure (Var i, a)
         go (_:xs) (_       :os) (TSnoc as _) i = go xs os as (i + 1)
         go []     []            TNil         _ = report (cxt^.names) (NameNotInScope x)
         go _ _ _ _ = error "impossible"

### 2020-02-25 implicit-fun-elab/Main.hs

@@ -43,7 +43,7 @@ mainWith getOpt getTm = do
       elab = do
         reset
         (t, src) <- getTm
-        (t, a) <- infer emptyCxt True t `catch` displayError src
+        (t, a) <- inferTopLams emptyCxt t `catch` displayError src
         t <- pure $ zonk VNil t
         let ~nt = quote 0 $ eval VNil t
         let ~na = quote 0 a
@@ -82,13 +82,14 @@ pruneTest2 = main' "elab" $ unlines [
  ]
 
 test0 = main' "elab" $ unlines [
-  "λ (Bool : U)(true : Bool).",
+  "λ (A : U)(a : A).",
 
-  "let x : _ = true in x"
+  "let x = a in x"
   ]
 
 
 test = main' "elab" $ unlines [
+  "λ (FOO : U)(foobarr : FOO).",
   "let the : (A : U) → A → A = \\A x.x in",
   "let IdTy = {A} → A → A in",
   "let id : IdTy = λ x. x in",

### 2020-02-25 implicit-fun-elab/Pretty.hs

@@ -1,116 +1,126 @@
 {-# options_ghc -Wno-orphans #-}
 
-module Pretty where
+module Pretty (showTm, showTopTm) where
 
 import Types
 
-prettyTm :: Int -> [Name] -> Tm -> ShowS
-prettyTm prec = topLams (prec /= 0) "λ" "" where
+-- special case printing of top lambdas, since they are usually used
+-- to postulate stuff
+topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
+topLams p pre post ns (Lam (fresh ns -> x) i a t) =
+  showParen p (
+    (pre++)
+  . icit i bracket parens (
+         ((if null x then "_" else x)++) . (" : "++) . go False ns a)
+  . topLams False "\n " ".\n\n" (('*':x):ns) t)
+topLams _ pre post ns t = (post++) . go False ns t
 
-  -- special case printing of top lambdas, since they are usually used
-  -- to postulate stuff
-  topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
-  topLams p pre post ns (Lam (fresh ns -> x) i a t) =
-    showParen p (
-      (pre++)
-    . icit i bracket parens (
-           ((if null x then "_" else x)++) . (" : "++) . go False ns a)
-    . topLams False "\n" ".\n\n" (x:ns) t)
-  topLams _ pre post ns t = (post++) . go False ns t
+fresh :: [Name] -> Name -> Name
+fresh _ "_" = "_"
+fresh ns n | elem n ns = fresh ns (n++"'")
+           | otherwise = n
 
-  fresh :: [Name] -> Name -> Name
-  fresh _ "_" = "_"
-  fresh ns n | elem n ns = fresh ns (n++"'")
-             | otherwise = n
+goVar :: [Name] -> Ix -> ShowS
+goVar ns x = case ns !! x of
+  '*':n -> (n++)
+  n     -> (n++)
 
-  goVar :: [Name] -> Ix -> ShowS
-  goVar ns topX = go ns topX where
-    -- go []     _ = error "impossible"
-    go []     _ = (show topX++)
-    go (n:ns) 0 = (n++)
-    go (n:ns) x = go ns (x - 1)
+goArg :: [Name] -> Tm -> Icit -> ShowS
+goArg ns t i = icit i (bracket (go False ns t)) (go True ns t)
 
[…]

### 2020-02-25 paper.tex
stuff

@@
 \usepackage{amssymb}
+
 %% \usepackage{cleveref}
@@
   are uncertain about whether we should be uncertain about implicit
-  insertions. So far, this higher-order insertion problem seems to seems to be
+  insertions. So far, this higher-order insertion problem seems to be
   irrelevant in practice, in the prototype implementation.}. We prepend the
@@
 \begin{definition}[Contextualization]
-We extend our spine notation to applications of curried functions. For example,
-we may have a spine $\ol{t} \equiv (\{x : \ol{A}\}\,\{y : \ol{B}\})$. We
+We extend spine notation to applications of curried functions. For example, we
+may have a spine $\ol{t} \equiv (\{x : \ol{A}\}\,\{y : \ol{B}\})$. We
 accordingly revise Definition \ref{def:contextualization} for
@@
 allow elaboration to insert $\lambda$-s for strictly curried functions. We present
-some examples of the algorithm at action.
+some examples of the algorithm in action.
 
@@
 
-\todo{update loc}
-
 It is implemented in about 1200 lines of Haskell. Of this, 673 lines constitute
@@
 reference which stores the current metasubstitution. We have a ``forcing''
-operation which computes a semantic value to a head normal form, with respect to
+operation which computes a semantic value to a head normal form with respect to
 the current metasubstitution. Hence, instead of constantly updating every term
-and type by metasubstitution, we only force them whenever we need to match on
-the shape of types, for example when we are inserting implicit arguments based
-on types. In this paper we still stick with the ``threaded metasubstitution''
-presentation because it is more conventional, and also more straightforward to
-formalize.
+and type by performing metasubstitution, we only force them whenever we need to
+match on the shape of types, for example when we are inserting implicit
+arguments based on types. In this paper we still stick with the ``threaded
+metasubstitution'' presentation because it is more conventional, and also more
+straightforward to formalize.
 
@@
 
-%% \begin{itemize}
-%% \item Elaboration is implemented in the style of Coquand's algorithm \cite{coquand1996algorithm},
-%%       where elaboration is interleaved with normalization-by-evaluation.
-%% \end{itemize}
+\newcommand{\yes}{\text{Yes}}
+\newcommand{\yesst}{\text{Yes*}}
+\newcommand{\no}{\text{No}}
+
+\begin{figure}
+\begin{tabular}{|lll|}
+\hline
+A & \multicolumn{2}{l|}{\textsc{polymorphic instantiation}}\\
+\hline
+A1 & $\lambda\,x\,y.\,y$ & \yesst\\
+A2 & $choose\,id$ & \yesst \\
+A3 & $choose\,nil\,ids$ & \yes \\
+A4 & $\lambda\,(x : \{A : \U\}\to A \to A).\,x\,x$ & \yesst\\
+A5 & $id\,auto$ & \yes\\
+A6 & $id\,auto'$ & \yesst\\
+A7 & $choose\,id\,auto$ & \yes\\
+A8 & $choose\,id\,auto'$ & \no\\
+A9 & $\lambda\,(f : \{A : \U\}\to(A\to A)\to List\,A \to A).\,f\,(choose\,id)\,ids$ & \yesst\\
+A10 & $poly\,id$ & \yes\\
+A11 & $poly\,(\lambda\,x.\,x)$ & \yes\\
+\hline
+B & \multicolumn{2}{l|}{\textsc{inference of polymorphic arguments}}\\
+\hline
+B1 & $\lambda\,f.\,pair\,(f\,zero)\,(f\,true)$ & \no\\
+B2 & $\lambda\,xs.\,poly\,(head\,xs)$ & \no\\
+\hline
+C & \multicolumn{2}{l|}{\textsc{functions on polymorphic lists}}\\
+\hline
+C1  & $length\,ids$ & \yes \\
+C2  & $tail\,ids$ & \yes \\
+C3  & $head\,ids$ & \yesst \\
+C4  & $single\,id$ & \yesst \\
+C5  & $cons\,id\,ids$ & \yes \\
+C6  & $cons\,(\lambda\,x.\,x)\,ids$ & \yes \\
+C7  & $append\,(single\,inc)\,(single\,id)$ & \yes \\
+C8  & $\lambda\,(g:\{A:\U\}\to List\,A \to List\,A \to A).\,g\,(single\,id)\,ids$ & \yesst \\
+C9  & $map\,poly\,(single\,id)$ & \yes \\
+C10 & $map\,head\,(single\,ids)$ & \yes \\
+\hline
+D & \multicolumn{2}{l|}{\textsc{application functions}}\\
+\hline
+D1 & $app\,poly\,id$        & \yes \\
+D2 & $revapp\,id\,poly$     & \yes \\
+D3 & $runST\,argST$         & \yes \\
+D4 & $app\,runST\,argST$    & \yes \\
+D5 & $revapp\,argST\,runST$ & \yes \\
+\hline
+\end{tabular}
+
+\caption{Elaboration benchmark from \cite{serrano2018guarded}.}
+\end{figure}
+
+%% λ (g : {A} → List A → List A → A). g (single id) ids
 

### 2020-02-25 paper.tex
stuff

@@
 
+%% \begin{alignat*}{3}
+%% & \scriptstyle{0}\qquad\qquad && \einfert{\slet\,x : \_ = true\,\sin\,x}{\true}
+%%       \hspace{21em}\\
+%% & \scriptstyle{1} && \quad \echeckt{\_}{\U}\\
+%% & \scriptstyle{2} && \quad = \alpha_0\\
+%% & \scriptstyle{3} && \quad \echeckt{true}{\alpha_0}\\
+%% & \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel\\
+%% & \scriptstyle{5} && \qquad = \alpha_1\\
+%% & \scriptstyle{6} && \qquad \einfert{true}{\true}\\
+%% & \scriptstyle{7} && \qquad = true : Bool\\
+%% & \scriptstyle{8} && \qquad \mathsf{newConstancy}_{\emptycon,\,x : \Rec\,\alpha_1}\,Bool\\
+%% & \scriptstyle{9} && \qquad\quad \unify\,\alpha_1\,\epsilon\\
+%% & \scriptstyle{10} && \qquad \unify\,\alpha_0\,Bool\\
+%% & \scriptstyle{11} && \quad = true\\
+%% & \scriptstyle{12} && \quad \einfert{x}{\true}\\
+%% & \scriptstyle{13} && \quad = true : Bool\\
+%% & \scriptstyle{14} && = (\slet\,x : Bool = true\,\sin\,x) : Bool
+%% \end{alignat*}
+
 \begin{example}
@@
       \hspace{21em}\\
-& \scriptstyle{1} && \quad \echeckt{\_}{\U}\\
-& \scriptstyle{2} && \quad = \alpha_0\\
+& \scriptstyle{1} && \quad \echeckt{\_}{\U} = \alpha_0\\
 & \scriptstyle{3} && \quad \echeckt{true}{\alpha_0}\\
-& \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel\\
-& \scriptstyle{5} && \qquad = \alpha_1\\
-& \scriptstyle{6} && \qquad \einfert{true}{\true}\\
-& \scriptstyle{7} && \qquad = true : Bool\\
+& \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel = \alpha_1\\
+& \scriptstyle{6} && \qquad \einfert{true}{\true} = true : Bool\\
 & \scriptstyle{8} && \qquad \mathsf{newConstancy}_{\emptycon,\,x : \Rec\,\alpha_1}\,Bool\\
@@
 & \scriptstyle{11} && \quad = true\\
-& \scriptstyle{12} && \quad \einfert{x}{\true}\\
-& \scriptstyle{13} && \quad = true : Bool\\
+& \scriptstyle{12} && \quad \einfert{x}{\true} = true : Bool\\
 & \scriptstyle{14} && = (\slet\,x : Bool = true\,\sin\,x) : Bool
@@
 \newcommand{\no}{\text{No}}
+
+{\small
 
@@
 \hline
+E & \multicolumn{2}{l|}{\textsc{$\eta$-expansion}}\\
+\hline
+& \multicolumn{2}{l|}{assuming $k : \{A : \U\}\to A \to List\,A \to A$,} \\
+& \multicolumn{2}{l|}{$h : Int \to \{A:\U\}\to A \to A$}\\
+& \multicolumn{2}{l|}{and $lst : List\,(\{A:\U\} \to Int \to A \to A)$} \\
+E1 & $k\,h\,lst$ & \no \\
+E2 & $k\,(\lambda\,x.\,h\,x)\,lst$ & \yesst \\
+E3 & $\lambda\,(r : (\{A :\U\}\to A \to \{B : \U\}\to B \to B) \to Int).\,r\,(\lambda\,x\,y.\,y)$ & \yes \\
+\hline
+\multicolumn{3}{l}{\text{{\footnotesize
+``Yes*'' means that our system can infer a type for the
+expression, but the lack of}}}\\
+\multicolumn{3}{l}{\text{{\footnotesize
+let-generalization causes unsolved metas in the type.}}}
 \end{tabular}
@@
 
-%% λ (g : {A} → List A → List A → A). g (single id) ids
+}
+
+%% (h : Int → {A} → A → A)(k : {A} → A → List A → A)(lst : List ({A} → Int → A → A))
 

### 2020-02-26 paper.tex
stuff

@@
 \newcommand{\ol}[1]{\overline{#1}}
+
+\newcommand{\yes}{\text{Yes}}
+\newcommand{\yesst}{\text{Yes*}}
+\newcommand{\no}{\text{No}}
 
@@
 
-Our telescopes and records are derivable from natural numbers, the unit type and
+Telescopes and records are derivable from natural numbers, the unit type and
 $\Sigma$-types. We use Agda-like pattern matching notation in the following.
@@
 
-\section{Evaluation}
+\section{Related Works and Evaluation}
 \label{sec:impl_and_eval}
 
-\newcommand{\yes}{\text{Yes}}
-\newcommand{\yesst}{\text{Yes*}}
-\newcommand{\no}{\text{No}}
+In this section we examine how the elaborator fares in practice, its
+limitations, and how it compares to related works. We abbreviate our elaborator
+as FCIF when comparing it to others.
+
+\subsubsection{Related works}
+MLF \cite{le2014mlf} is an extension of System F with polymorphic subtyping
+bounds, which supports strong inference for first class polymorphism. HML
+\cite{leijen2009flexible}) is a simplified variant of MLF. Since these systems
+diverge markedly from (fragments of) Martin-Löf type theory and rely critically
+on subtyping, it is unclear how they could be extended to dependent type
+theories. HMF \cite{leijen2008hmf} is a relatively simple system with
+first-class polymorphism; however, it is also weaker than others in terms of
+inferable annotations, and its more powerful variant (extended with n-ary
+application handling) jumps in complexity.
+
+There have been several attempts in the context of GHC. Boxy types
+\cite{vytiniotis2006boxy} and FPH \cite{vytiniotis2008fph} were two early
+iterations which suffered from complex specification, complex implementation or
+fragility. More recent works are guarded impredicativity (GI)
+\cite{serrano2018guarded} and quick look impredicativity (QL)
+\cite{serrano2020a}. They both try to examine n-ary applications to find
+metavariable instantiations which arise from occurrences guarded by rigid type
+constructors. GI's use of \emph{generalization constraints} anticipates our use
+of telescopes, but overall GI is also burdened by inessential complexity. QL
+streamlines GI by eschewing constraints in favor of an eager preprocessing pass
+on neutral expressions which finds polymorphic instantiations. QL also takes
+advantage of bidirectional type propagation.  Although QL seems to be
+practically the most favorable so far, it is far from being elegant; it does
+multiple checking passes, and the restriction of quick looking to neutral
+applications is rather ad-hoc.
 
 {\small
+
+\begin{figure}
+\begin{alignat*}{4}
+  & IdTy   && :\equiv \{A : \U\}\to A \to A   \hspace{10em}         && inc    && : Int \to Int                                        \\
+  & single && : \{A : \U\} \to A \to List\,A                       && auto   && : IdTy \to IdTy                                      \\
+  & id     && : IdTy                                               && auto'  && : \{B : \U\}\to IdTy \to B \to B                     \\
+  & ids    && : List\,IdTy                                         && choose && : \{A : \U\} \to A \to A \to A                       \\
+  & nil    && : \{A : \U\}\to List\,A                              && app    && : \{A\,B : \U\} \to (A \to B) \to A \to B            \\
+  & cons   && : \{A : \U\} \to A \to List\,A \to List\,A           && revapp && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
+  & head   && : \{A : \U\} List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
+  & tail   && : \{A : \U\} List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
+  & map    && : \{A B : \U\} \to (A \to B) \to List\,A \to List\,B && poly   && : IdTy \to Pair\,Int\,Bool
+\end{alignat*}
+\caption{Types used in Figure \ref{fig:benchmark}.}
+\label{fig:benchmark_types}
+\end{figure}
 
@@
 A1 & $\lambda\,x\,y.\,y$ & \yesst\\
+   & \text{we infer a type which may be solved to $\{A\,B : \U\}\to A \to B \to A$} &\\
+   & \text{but not to $\{A : \U\}\to A \to \{B : \U\} \to B \to A$} & \\
 A2 & $choose\,id$ & \yesst \\
@@
 \multicolumn{3}{l}{\text{{\footnotesize
-let-generalization causes unsolved metas in the type.}}}
+let-generalization yields unsolved metas in the type.}}}
 \end{tabular}
@@
 \caption{Elaboration benchmark from \cite{serrano2018guarded}.}
+\label{fig:benchmark}
 \end{figure}
-
 }
 
-%% (h : Int → {A} → A → A)(k : {A} → A → List A → A)(lst : List ({A} → Int → A → A))
+\subsubsection{Evaluation}
+We borrow a very useful collection of inference benchmarks from the GI
+\cite{serrano2018guarded} and QL \cite{serrano2020a} papers: see Figure
+\ref{fig:benchmark_types} and Figure \ref{fig:benchmark}. We also include a
+source file in the implementation which reproduces these results.  In ibid.\ a
+comparison is presented between multiple systems, but here we only include
+FCIF. The relative performance of FCIC here is easy to remember: it handles
+exactly the same cases as QL (which is slightly more than what GI covers). We
+mark some cases as ``Yes*'', where FCIF successfully infers a type, but since
+FCIC does no let-generalization, the inferred types are not fully constrained
+without extra contextual information.
+
+In the $A1$ case, FCIF inserts only a single curried $\lambda$ on the outside,
+which is why the inferred type is not unifiable with $\{A : \U\}\to A \to \{B :
+\U\} \to B \to A$. However, this could be easily remedied by adding an extra
+curried $\lambda$ insertion in the definition of
+$\einfer{\lambda\,x.\,t}{ins}{\Theta}{\Gamma}$.
+
+In $A8$, the failure is intentional: the types of $id$ and $auto'$ are not
+unifiable because of the mismatched order of implicit and explicit arguments. We
+do not float out or reorder implicit arguments in any way, because we want to
+support arbitrary mixing of implicit/explicit arguments in the surface language,
+and such reordering would be problematic anyway in the presence of dependent
+types. The same situation arises in $E1$, where we cannot unify $\{A : \U\}\to
+Int \to A \to A$ with $Int \to \{A : \U\}\to A \to A$.
+
+While in this benchmark FCIF matches QL, in other ways it is significantly
+stronger. Some general observations on the relative strength of FCIF:
+\begin{itemize}
+  \item
+        Most obviously, FCIF supports dependent type theory and prior solutions
+        do not. It is also doubtful whether prior solutions scale to dependent
+        types. In particular, QL's preprocessing step is not applicable, since in
+        a dependently typed spine we cannot do anything with later arguments until
+        previous arguments are actually elaborated, or else it has to be restricted
+        to non-dependent spines.
+  \item
+        FCIF is global and does not consider neutral spines specially in any
+        way. For an example for global inference, $\slet\,x : \_ =
+        single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF but does not
+        work in GI, QL and HMF. It also works in MLF and HML, which are also
+        global in this sense. These work by immediately giving a principal type
+        to $single\,id$ which includes a subtyping bound. In contrast, FCIF
+        makes no promises about principal typing --- which is not feasible with
+        dependent types --- but it can still infer a type for $id$ which can be
+        later constrained to $IdTy$.
+\end{itemize}
+
+\emph{Limitations.} What FCIF definitely cannot do, is inferring an implicit
+function type for a function argument. MLF is the only system which can do this,
+in cases where the polymorphic argument is used only once. Consider the
+following:
+\[
+  \slet\,f : IdTy \to Bool = \lambda\,\_.\,true\,\sin\,\lambda\,x.\,f\,x
+  \]
+FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
+infer a type for $x$, because $A$ is not in the scope of $x$'s type. In the next
+section we sketch an extension of FCIF which can handle this, and can also
+handle $B2$ from Figure \ref{fig:benchmark}.
+
+\section{Future Work}
+
+\section{Conclusion}
+
+%% -- let f : IdTy → Bool = λ x. true in
+%% -- λ x. f x
 

### 2020-02-27 paper.tex
stuff

@@
 The above issue in Agda is a specific instance of the challenges described in
-the mentioned works. So far, none of the above solutions have landed in
+the mentioned works. Currently none of the above solutions are supported in
 production compilers, for reasons of complexity, fragility and interaction with
@@
 solution which is relatively simple, and which is likely to land in an official
-GHC release. However, none of these solutions support dependent types, and they
-also have other limitations which we would like to address.
+GHC release. However, none of these works support dependent types, which is a
+key point in our work.
 
@@
     implicit function types, telescopes and \emph{strictly curried function
-      types} with telescope domain. We use these extensions to accummulate
-    information about implicit insertions. Our algorithm is a conservative
-    extension of Norell's bidirectional elaborator for Agda
-    \cite[Chapter~3]{norell07thesis}.
-  \item The presented elaborator is generally more powerful than previous
-    solutions for impredicative inference. In particular, it goes beyond prior
-    art by supporting full dependent type theory and by being global,
-    i.e.\ inference may consider the whole program and not just particular n-ary
-    applications.
+      types} with telescope domain. We use these extensions to accumulate
+    information about implicit insertions. Our algorithm is based on Norell's
+    bidirectional elaborator for Agda \cite[Chapter~3]{norell07thesis}.
+  \item
+    In the System F fragment, the presented elaborator is comparable or superior
+    to previous solutions for impredicative inference. However, it also supports
+    full dependent type theory. Our inference is also global, i.e.\ it can
+    consider the whole program and not just particular n-ary applications.
+  \item We provide an executable implementation of the elaborator described in
+    this paper.
+  \item Our solution is simple: we implemented elaboration,
+    evaluation and unification in about 670 lines of Haskell, of which 210 lines
+    implement the novel enhancements on the top of Norell's basic elaborator.
   \item The target theory of elaboration serves as a general platform for
@@
     simple one, and there is room to further develop it.
-  \item We provide an executable implementation of the elaborator described in
-    this paper.
-  \item Our solution is also \emph{simple}: we implemented elaboration,
-    evaluation and unification in about 670 lines of Haskell, of which 210 lines
-    implement the novel enhancements on the top of Norell's basic elaborator.
 \end{itemize}
@@
 We do not present any \emph{completeness} result for elaboration in this
-paper. For an example of what this would entail, in \cite{dunfield2013complete},
+paper. For an example of what this would entail, in \cite{dunfield2013complete}
 completeness means that whenever there is a way to fill in missing details in
@@
 that it is not tractable in general to figure out which arguments are inferable,
-by looking at function types; often we have to run elaboration to see what
+by looking at function types, and often we have to run elaboration to see what
 works.
@@
 We can still say something about the behavior of our elaborator. For this, we
-consider a translation from core terms to surface terms, the obvious forgetful
+consider a translation from core terms to surface terms, the evident forgetful
 translation, which maps core terms to surface counterparts. Now, this is an
@@
 application insertions requires extending unification; we discuss this in
-Section \ref{TODO}. First, we modify closing types and contextualization to take
-advantage of telescopes.
+Section \ref{sec:appinsert}. First, we modify closing types and
+contextualization to take advantage of telescopes.
 
@@
 a surface expression from standard input, and outputs the result of elaboration,
-or optionally the type or the normal form of the result.
-
-It is implemented in about 1200 lines of Haskell. Of this, 673 lines constitute
-the core syntax, evaluation, unification and elaboration. Of this 673 lines, 458
-lines implement the basic elaborator of Section
-\ref{sec:bidirectional_elaboration} and 215 lines implement the extended
-elaborator. We give here implementation-specific details and describe
-differences to the presentation in this paper.
+or optionally the type or the normal form of the result. It is implemented in
+about 1200 lines of Haskell. Of this, 673 lines constitute the core syntax,
+evaluation, unification and elaboration. Of this 673 lines, 458 lines implement
+the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215
+lines implement the extended elaborator.
 
@@
 and type by performing metasubstitution, we only force them whenever we need to
-match on the shape of types, for example when we are inserting implicit
+pattern match on the shape of types, for example when we are inserting implicit
 arguments based on types. In this paper we still stick with the ``threaded
@@
 applications and abstractions with telescopes (just as in our notation for the
-core syntax).
-
-We provide a fairly optimized implementation for constraints: we keep track of
-``blocking'' metas for each constraint, which are relevant to checking the
-constancy condition, and upon solving a meta we only review constraints which
-were blocked on the meta.
+core syntax). We implement constraints in a fairly optimized way: we keep track
+of relevant ``blocking'' metas for each constraint and upon solving a meta we
+only review constraints which were blocked on the meta.
 
@@
 
-\subsubsection{Related works}
+\subsection{Related Works}
 MLF \cite{le2014mlf} is an extension of System F with polymorphic subtyping
@@
 advantage of bidirectional type propagation.  Although QL seems to be
-practically the most favorable so far, it is far from being elegant; it does
-multiple checking passes, and the restriction of quick looking to neutral
-applications is rather ad-hoc.
+practically the most favorable so far, it is far from being elegant: it rechecks
+expressions multiple times, and the restriction of preprocessing to (nested)
+neutral applications is rather ad-hoc.
 
@@
 
-\subsubsection{Evaluation}
+\subsection{Evaluation}
 We borrow a very useful collection of inference benchmarks from the GI
@@
 comparison is presented between multiple systems, but here we only include
-FCIF. The relative performance of FCIC here is easy to remember: it handles
+FCIF. The relative performance of FCIF here is easy to remember: it handles
 exactly the same cases as QL (which is slightly more than what GI covers). We
 mark some cases as ``Yes*'', where FCIF successfully infers a type, but since
-FCIC does no let-generalization, the inferred types are not fully constrained
+FCIF does no let-generalization, the inferred types are not fully constrained
 without extra contextual information.
@@
 
-While in this benchmark FCIF matches QL, in other ways it is significantly
-stronger. Some general observations on the relative strength of FCIF:
-\begin{itemize}
-  \item
-        Most obviously, FCIF supports dependent type theory and prior solutions
-        do not. It is also doubtful whether prior solutions scale to dependent
-        types. In particular, QL's preprocessing step is not applicable, since in
-        a dependently typed spine we cannot do anything with later arguments until
-        previous arguments are actually elaborated, or else it has to be restricted
-        to non-dependent spines.
-  \item
-        FCIF is global and does not consider neutral spines specially in any
-        way. For an example for global inference, $\slet\,x : \_ =
-        single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF but does not
-        work in GI, QL and HMF. It also works in MLF and HML, which are also
-        global in this sense. These work by immediately giving a principal type
-        to $single\,id$ which includes a subtyping bound. In contrast, FCIF
-        makes no promises about principal typing --- which is not feasible with
-        dependent types --- but it can still infer a type for $id$ which can be
-        later constrained to $IdTy$.
-\end{itemize}
-
-\emph{Limitations.} What FCIF definitely cannot do, is inferring an implicit
-function type for a function argument. MLF is the only system which can do this,
-in cases where the polymorphic argument is used only once. Consider the
-following:
+Some general comments on the comparison of FCIF to prior works. First, FCIF
+supports dependent types and prior solutions do not. It is also unclear whether
+prior solutions can scale to dependent types. In particular, QL's preprocessing
+step is not applicable, since in a dependently typed spine we cannot do anything
+with later arguments until previous arguments are actually elaborated, or else
+it has to be restricted to non-dependent spines.
+
+Secondly, FCIF supports global inference and does not handle neutral spines
+specially. For an example for global inference, $\slet\,x : \_ =
+single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF, MLF and HMF but does
+not work in GI, QL and HMF. MLF and HML work here by immediately giving a
+principal type to $single\,id$ which involves generalization with a subtyping
+bound. In contrast, FCIF makes no promises about principal typing --- which is
+not feasible with dependent types --- and does no generalization, but it can
+still infer a type for $id$ which can be later constrained to $IdTy$.
+
+\subsection{Inferring Polymorphic Types for Function Arguments}
+\label{sec:appinsert}
+
+It is apparent from the $B1$ and $B2$ cases on Figure \ref{fig:benchmark} that FCIF
+cannot infer implicit function types for function arguments.  MLF is the only
+system which can do this, in cases where the polymorphic argument is used only
+once. Consider the following:
 \[
@@
 FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
-infer a type for $x$, because $A$ is not in the scope of $x$'s type. In the next
-section we sketch an extension of FCIF which can handle this, and can also
-handle $B2$ from Figure \ref{fig:benchmark}.
-
-\section{Future Work}
-
-\section{Conclusion}
-
-%% -- let f : IdTy → Bool = λ x. true in
-%% -- λ x. f x
+infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
+an extension of FCIF with \emph{implicit application insertion}, which could
+possibly handle this example. We note though that this is not yet fleshed out
+and is a subject of future work.
+
+\section{Future Work and Conclusion}
 

### 2020-02-28 paper.tex
stuff

@@
 inferable annotations, and its more powerful variant (extended with n-ary
-application handling) jumps in complexity.
+application handling) is more complex.
 
@@
 supports dependent types and prior solutions do not. It is also unclear whether
-prior solutions can scale to dependent types. In particular, QL's preprocessing
-step is not applicable, since in a dependently typed spine we cannot do anything
-with later arguments until previous arguments are actually elaborated, or else
-it has to be restricted to non-dependent spines.
+prior solutions can scale to dependent types. MLF and HML rely on subtyping, and
+solutions which work one neutral spine at a time (HMF, GI, QL) also face issues
+with dependently typed spines, since it is not possible to do anything with
+later arguments in such spines until previous arguments are elaborated.
 

### 2020-02-29 paper.tex
stuff

@@
 \newcommand{\echeck}[4]{\llbracket#1\rrbracket\!\Downarrow\,_{#2|#3}\,#4}
-\newcommand{\einfer}[4]{\llbracket#1\rrbracket\!\Uparrow\,_{#3|#4}\,#2}
+\newcommand{\einfer}[3]{\llbracket#1\rrbracket\!\Uparrow\,_{#2|#3}}
 \newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
@@
 \newcommand{\false}{\mathsf{false}}
-\newcommand{\einsert}[3]{\mathsf{insert}\,#1\,#2\,#3}
-\newcommand{\einsertpar}[5]{\mathsf{insert}_{#1|#2}\,#3\,#4\,#5}
+\newcommand{\einsert}{\mathsf{insert}}
 \newcommand{\Set}{\mathsf{Set}}
@@
 
-\begin{definition}[Contextualization]\label{def:contextualization} For each $\Theta|\Gamma \vdash t : \Gamma\To A$,
-we define the spine $\overline{\mathsf{vars}_{\Gamma}}$ such that that
-$\Theta|\Gamma\vdash t\,\overline{\mathsf{vars}_{\Gamma}} : A$. Informally, this
-is $t$ applied to all bound variables in $\Gamma$.
+\begin{definition}[Contextualization]\label{def:contextualization}
+For each $\Theta|\Gamma \vdash t : \Gamma\To A$, we define the spine
+$\overline{\mathsf{vars}_{\Gamma}}$ such that that $\Theta|\Gamma\vdash
+t\,\overline{\mathsf{vars}_{\Gamma}} : A$, which is $t$ applied to all bound
+variables in $\Gamma$.
   \begin{alignat*}{3}
@@
 \end{definition}
-
-\subsection{Implicit Argument Insertion}
-
-We define a helper function which inserts implicit applications around a core
-term. For example, if we have a defined name $id$ with type $\{A : \U\}\to A \to
-A$ in a surface program, we usually want to expand $id$ to $id\,\{\alpha\}$,
-where $\alpha$ is a fresh metavariable.
-\begin{mathpar}
-\inferrule*{ins \in \{\true,\,\false\} \\ \Theta|\Gamma\vdash t : A}
-           {\einsert{ins}{t}{A} \in \{(\Theta',\,\theta\,t',\,A')\,|\,
-             (\theta : \Theta' \To \Theta)\,\land\,(\Theta'|\Gamma[\theta]\vdash t' : A')\}}
-\end{mathpar}
-We have an additional $ins \in \{\true,\,\false\}$ parameter, which simply
-toggles whether any insertion is to be performed. We will use this in the
-definition of elaboration. Insertion is defined by recursion on $ins$ and the
-$A$ type, as follows. Here, we notate $\Theta$ and $\Gamma$ as subscripts but
-later we will leave them implicit. The clauses of the following definition are
-matched top-down.
-\begin{alignat*}{3}
-  &\einsertpar{\Theta}{\Gamma}{\false}{t}{A} && :\equiv\,\,&& (\Theta,\,\id,\,t,\,A)\\
-  &\einsertpar{\Theta}{\Gamma}{\true\,}{t}{(\{x : A\}\to B)} && :\equiv&&
-       \slet\,(\Theta',\,\theta,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
-  & && &&\sin\,\einsertpar{\Theta'}{\Gamma[\theta']}{\true}{((t[\theta])\,\{u\})}{(B[\theta][x\mapsto u])}\\
-  &\einsertpar{\Theta}{\Gamma}{\true\,}{t}{A} && :\equiv&& (\Theta,\,\id,\,t,\,A)
-\end{alignat*}
 
@@
 
-In this section we define the elaboration algorithm. First, we explain the used notations.
-
+Elaboration consists of two (partial) functions, checking and inferring, which
+are defined by mutual induction on surface syntax. We also have implicit
+argument insertion as a helper function, which is defined by recursion on core
+types, and which is used as a post-processing step after inference. First, about
+the used notations:
 \begin{itemize}
@@
     whether a type is a function type. This assumes an evaluation/normalization
-    procedure on core terms; but note that we already assume this in implicit
-    argument insertion and unification.
+    procedure on core terms; but note that we already assume this in
+    unification.
   \item We abbreviate $\theta_1 \circ \theta_2$ as $\theta_{12}$, $\theta_1
@@
 \end{itemize}
-
-Elaboration consists of two (partial) functions, checking and inferring, which
-are defined by mutual induction on surface syntax. They are specified as
-follows. We also have the additional $ins \in \{\true,\,\false\}$ argument for
-inference, which toggles whether we perform implicit argument insertion on the
-output. We explain the cases of the definition in order.
-
+We present the specifications and definitions below, then we describe them in order.
+\begin{mathpar}
+\inferrule*[lab=insert]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
+             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta_0 \To \Theta)\,\land\,
+               (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
+           {\einsert(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
+             (\theta : \Theta_0 \To \Theta)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
+\end{mathpar}
+\begin{alignat*}{3}
+  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
+       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
+  & && &&\sin\,\einsert\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
+  &\einsert\,(\Theta_1,\,\theta_1,\,t,\,A) && :\equiv&& \ereturn\,(\Theta_1,\,\theta_1,\,t,\,A)\\
+  &\einsert\,\fail && :\equiv && \fail
+\end{alignat*}
+%% \pagebreak
+\begingroup
+\allowdisplaybreaks
 \begin{mathpar}
@@
 \inferrule*[lab=infer]
-           {ins \in \{\true,\,\false\} \\
-                \text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
-           {\einfer{t}{ins}{\Theta}{\Gamma} \in
+           {\text{$t$ is a surface expression} \\ \Theta|\Gamma \vdash}
+           {\einfer{t}{\Theta}{\Gamma}\,\in
              \{(\Theta',\,\theta,\,t',\,A)\,|\,(\theta : \Theta' \To \Theta)\,\land\,
@@
 \end{mathpar}
-
-\begingroup
-\allowdisplaybreaks
 \begin{alignat*}{3}
@@
   &\echeck{t}{\Theta_0}{\Gamma}{A} :\equiv \edo \\
-  &\quad(\Theta_1,\,\theta_1,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
+  &\quad(\Theta_1,\,\theta_1,\,t',\,B) \leftarrow \einsert\,(\einfer{t}{\Theta_0}{\Gamma})\\
   &\quad(\Theta_2,\,\theta_2) \leftarrow \unify\,(A[\theta_1])\,B\\
   &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,t'[\theta_2])\\\\
-  &\einfer{x}{ins}{\Theta}{\Gamma} :\equiv \edo\\
+  &\einfer{x}{\Theta}{\Gamma} :\equiv \edo\\
   &\quad\eif\,(\Gamma = (\Gamma_0,\,x : A,\,\Gamma_1))\,\lor\,(\Gamma = (\Gamma_0,\,x : A = t,\,\Gamma_1)) \\
-  &\qquad   \ethen\, \ereturn\,(\einsert{ins}{x}{A})\\
+  &\qquad   \ethen\, \ereturn\,(\Theta,\,\id,\,x,\,A)\\
   &\qquad   \eelse\, \fail\\
-  &\einfer{\U}{ins}{\Theta}{\Gamma} :\equiv \edo\\
+  &\einfer{\U}{\Theta}{\Gamma} :\equiv \edo\\
   &\quad\ereturn\,(\Theta,\,\id,\,\U,\,\U)\\
-  &\einfer{(x : A)\to B}{ins}{\Theta_0}{\Gamma} :\equiv \edo\\
+  &\einfer{(x : A)\to B}{\Theta_0}{\Gamma} :\equiv \edo\\
   &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_1}{\Gamma}{\U}\\
@@
                  \,((x : A'[\theta_2])\to B'),\,\U)\\
-  &\einfer{\{x : A\}\to B}{ins}{\Theta_0}{\Gamma} :\equiv \edo\\
+  &\einfer{\{x : A\}\to B}{\Theta_0}{\Gamma} :\equiv \edo\\
   &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_1}{\Gamma}{\U}\\
@@
                  \,(\{x : A'[\theta_2]\}\to B'),\,\U)\\
-  &\einfer{\lambda\,x.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
-  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
+  &\einfer{\lambda\,x.\,t}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad \slet\, (\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A})\\
   &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,\lambda\,x.\,t',\,(x : A[\theta_2])\to B)\\
-  &\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  &\quad(\Theta_1,\,\theta_1,\,A)\leftarrow \freshMeta{\Theta_0}{\Gamma}{\U}\\
-  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:A}\\
-  &\quad\ereturn\,(\einsert{ins}{(\lambda\,\{x\}.\,t')}{(\{x : A[\theta_2]\}\to B)})\\
-  &\einfer{t\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\true}{\Theta_0}{\Gamma}\\
-  &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
-  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
+  &\einfer{\lambda\,\{x\}.\,t}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\U}\\
+  &\quad(\Theta_2,\,\theta_2,\,t',\,B)\leftarrow \einsert\,(\einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:A})\\
+  &\quad\ereturn\,(\Theta_2,\,\theta_{12},\,\lambda\,\{x\}.\,t',\,\{x : A[\theta_2]\}\to B)\\
+  &\einfer{t\,u}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einsert\,(\einfer{t}{\Theta_0}{\Gamma})\\
+  &\quad \slet\,(\Theta_2,\,\theta_2,A_0) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
+  &\quad \slet\,(\Theta_3,\,\theta_3,A_1) = \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
   &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_{23}])\,((x : A_0[\theta_3])\to A_1)\\
   &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_{1234}]}{(A_0[\theta_{34}])}\\
-  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,u')}\,{(A_1[\theta_{45}][x\mapsto u'])})\\
-  &\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
-  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\false}{\Theta_0}{\Gamma}\\
-  &\quad (\Theta_2,\,\theta_2,A_0) \leftarrow \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
-  &\quad (\Theta_3,\,\theta_3,A_1) \leftarrow \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
+  &\quad \ereturn\,(\Theta_5,\,\theta_{12345},\,(t'[\theta_{2345}])\,u',\,A_1[\theta_{45}][x\mapsto u'])\\
+  &\einfer{t\,\{u\}}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad (\Theta_1,\,\theta_1,\,t',\,A) \leftarrow \einfer{t}{\Theta_0}{\Gamma}\\
+  &\quad \slet\,(\Theta_2,\,\theta_2,A_0) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{\U}\\
+  &\quad \slet\,(\Theta_3,\,\theta_3,A_1) = \freshMeta{\Theta_2}{\Gamma[\theta_{12}],\,x:A_0}{\U}\\
   &\quad (\Theta_4,\,\theta_4) \leftarrow \unify\,(A[\theta_{23}])\,(\{x : A_0[\theta_3]\}\to A_1)\\
   &\quad (\Theta_5,\,\theta_5,\,u') \leftarrow \echeck{u}{\Theta_4}{\Gamma[\theta_{1234}]}{(A_0[\theta_{34}])}\\
-  &\quad \ereturn\,(\mathsf{insert}\,{ins}\,{((t'[\theta_{2345}])\,\{u'\})}\,{(A_1[\theta_{45}][x\mapsto u'])})\\
-  &\einfer{\slet\,x:A=t\,\sin\,u}{ins}{\Theta_0}{\Gamma} :\equiv \edo \\
+  &\quad \ereturn\,(\Theta_5,\,\theta_{12345},\,(t'[\theta_{2345}])\,\{u'\},\,A_1[\theta_{45}][x\mapsto u'])\\
+  &\einfer{\slet\,x:A=t\,\sin\,u}{\Theta_0}{\Gamma} :\equiv \edo \\
   &\quad(\Theta_1,\,\theta_1,\,A') \leftarrow \echeck{A}{\Theta_0}{\Gamma}{\U}\\
@@
   &\quad(\Theta_3,\,\theta_3,\,u',\,B) \leftarrow
-                \einfer{u}{ins}{\Theta_2}{\Gamma[\theta_{12}]}\\
+                \einfer{u}{\Theta_2}{\Gamma[\theta_{12}]}\\
   &\quad\ereturn\,(\Theta_3,\,\theta_{123},\,(\slet\,x:A'[\theta_{23}]=t'[\theta_3]\,\sin\,u'),\,B)\\
-  &\einfer{\_}{ins}{\Theta}{\Gamma} :\equiv \edo \\
+  &\einfer{\_}{\Theta}{\Gamma} :\equiv \edo \\
   &\quad \slet\,(\Theta',\,\theta,\,A) = \freshMeta{\Theta}{\Gamma}{\U}\\
@@
 \endgroup
+\subsubsection{Implicit argument insertion}
+This inserts implicit applications around a core term. For example, if we have a
+defined name $id$ with type $\{A : \U\}\to A \to A$ in a surface program, we
+usually want to expand $id$ to $id\,\{\alpha\}$, where $\alpha$ is a fresh
+metavariable. We define $\einsert$ to take as input the output of
+$\einferblank$, so that it is more convenient to use as an optional
+post-processing step after inference.
 
@@
 or optionally the type or the normal form of the result. It is implemented in
-about 1200 lines of Haskell. Of this, 673 lines constitute the core syntax,
-evaluation, unification and elaboration. Of this 673 lines, 458 lines implement
+1061 lines of Haskell. Of this, 673 lines constitute the core syntax,
+evaluation, unification and elaboration. Of these 673 lines, 458 lines implement
 the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215

### 2020-02-29 implicit-fun-elab/Elaboration.hs

@@ -385,9 +385,8 @@ unify cxt l r = go l r where
 -- Elaboration
 --------------------------------------------------------------------------------
 
-insert :: Cxt -> MetaInsertion -> IO (Tm, VTy) -> IO (Tm, VTy)
-insert cxt False act = act
-insert cxt True  act = do
+insert :: Cxt -> IO (Tm, VTy) -> IO (Tm, VTy)
+insert cxt act = do
   (t, va) <- act
   let go t va = case force va of
         VPi x Impl a b -> do
@@ -422,7 +421,8 @@ check cxt topT ~topA = case (topT, force topA) of
     x <- ("Γ"++) . show <$> readIORef nextMId
     dom <- freshMeta cxt VTel
     let vdom = eval (cxt^.vals) dom
-    (t, liftVal cxt -> a) <- infer (bind x NOInserted (VRec vdom) cxt) True t
+    let cxt' = bind x NOInserted (VRec vdom) cxt
+    (t, liftVal cxt -> a) <- insert cxt' $ infer cxt' t
     newConstancy cxt vdom a
     unifyWhile cxt topA (VPiTel x vdom a)
     pure $ LamTel x dom t
@@ -439,11 +439,11 @@ check cxt topT ~topA = case (topT, force topA) of
     freshMeta cxt topA
 
   (t, topA) -> do
-    (t, va) <- infer cxt True t
+    (t, va) <- insert cxt $ infer cxt t
     unifyWhile cxt va topA
     pure t
 
--- | We specialcase top-level lambdas (serving as postulates) for better
+-- | We special case top-level lambdas (serving as postulates) for better
 --   printing: we don't print them in meta spines.
 inferTopLams :: Cxt -> Raw -> IO (Tm, VTy)
 inferTopLams cxt = \case
@@ -456,17 +456,16 @@ inferTopLams cxt = \case
     pure (Lam x i a t, VPi x i va b)
   RSrcPos p t ->
     addSrcPos p $ inferTopLams cxt t
-  t ->
-    infer cxt True t
 
-infer :: Cxt -> MetaInsertion -> Raw -> IO (Tm, VTy)
-infer cxt ins = \case
-  RSrcPos p t ->
-    addSrcPos p $ infer cxt ins t
+  t -> insert cxt $ infer cxt t
+
+infer :: Cxt -> Raw -> IO (Tm, VTy)
+infer cxt = \case
+  RSrcPos p t -> addSrcPos p $ infer cxt t
 
[…]

### 2020-02-29 implicit-fun-elab/Main.hs

@@ -69,6 +69,7 @@ main = mainWith getArgs parseStdin
 main' :: String -> String -> IO ()
 main' mode src = mainWith (pure [mode]) ((,src) <$> parseString src)
 
+
 pruneTest = main' "elab" $ unlines [
   "λ (Bool : U)(not : Bool → Bool → Bool)(true : Bool).",
   "λ f (x : Bool). not (f x)"
@@ -84,7 +85,8 @@ pruneTest2 = main' "elab" $ unlines [
 test0 = main' "elab" $ unlines [
   "λ (A : U)(a : A).",
 
-  "let x = a in x"
+  "let id : {A} → A → A = λ x. x in",
+  "_"
   ]
 
 

### 2020-02-29 implicit-fun-elab/Pretty.hs

@@ -68,7 +68,7 @@ isMetaSp (AppTel _ t _) = isMetaSp t
 isMetaSp _              = False
 
 goMetaSp :: [Name] -> Tm -> ShowS
-goMetaSp ns (App t (Var x) i) | '_':_ <- ns !! x =
+goMetaSp ns (App t (Var x) i) | '*':_ <- ns !! x =
   goMetaSp ns t
 goMetaSp ns (App t u i)    =
   goMetaSp ns t . (' ':) . goArg ns u i

### 2020-02-29 paper.tex
pretty nice fmt now

@@
 \newcommand{\echeckt}[2]{\llbracket#1\rrbracket\!\Downarrow\,#2}
-\newcommand{\einfert}[2]{\llbracket#1\rrbracket\!\Uparrow\,#2}
+\newcommand{\einfert}[1]{\llbracket#1\rrbracket\!\Uparrow}
 \newcommand{\edo}{\boldsymbol{\mathsf{do}}}
@@
 ``impredicative instantiation'' or ``impredicative inference''. We propose a new
-framework for elaborating first-class implicit functions, which is applicable
-for full dependent type theories and compares favorably to prior solutions in
+solution for elaborating first-class implicit functions, which is applicable
+to full dependent type theories and compares favorably to prior solutions in
 terms of power, generality and simplicity. We build atop Norell's
@@
 details, translating terse surface-level languages to explicit core
-languages. Modern compilers such as Agda have gotten quite adept at this
-task. However, in practice, programmers still have to tell the compiler when and
-where to try filling in details on its own.
+languages. Modern systems such as Agda have gotten quite adept at this
+task. However, in practice, programmers still have to tell the compiler where to
+try filling in details on its own.
 
@@
 $\kw{@}id\,bool\,true$, and in GHC we can enable the language extension
-$\mathsf{TypeApplications}$ and write $id\,\kw{@}Bool\,True$.
+$\mathsf{TypeApplications}$ \cite{eisenberg2016visible} and write
+$id\,\kw{@}Bool\,True$.
 
@@
 From now on, we will only consider well-formed core syntax, and unless otherwise
-mentioned, constructions on core syntax which respect definitional equality.
+mentioned, presented constructions on the core syntax respect definitional
+equality.
 
@@
 metacontext to the old one, by ``dropping'' the new entry. Lastly,
-$\alpha\,\overline{\mathsf{vars}_{\Gamma}}$ is the new meta
+$\alpha\,\overline{\mathsf{vars}_{\Gamma}}$ is the fresh meta
 applied to all bound variables.
@@
 \end{alignat*}
-%% \pagebreak
+\pagebreak
 \begingroup
@@
 For $\echeck{\slet\,x:A=t\,\sin\,u}{\Theta_0}{\Gamma}{B}$, we simply let
-checking fall through. The definition here is a bit noisy, because we need to
-thread metasubstitutions through, and we always have to ``update'' core terms
-and contexts with the current metasubstitution.
-
-For $\echeck{\_}{\Theta}{\Gamma}{A}$, we return a fresh metavariable with the
-expected type. In any other $\echeck{t}{\Theta_0}{\Gamma}{A}$ case, we have a
-\emph{change of direction}: we infer a type for $t$, then unify the expected and
+checking fall through. For $\echeck{\_}{\Theta}{\Gamma}{A}$, we return a fresh
+metavariable with the expected type. In any other
+$\echeck{t}{\Theta_0}{\Gamma}{A}$ case, we have a \emph{change of direction}: we
+infer a type for $t$ (with implicit insertions) then unify the expected and
 inferred types.
@@
 
-For $\einfer{x}{ins}{\Theta}{\Gamma}$, we look up the type of $x$ in $\Gamma$,
-and insert implicit applications if needed. In the case of $\U$, we always
-succeed and infer $\U$ as type. In the cases for function types, we check that
-the domains and codomains have type $\U$.
-
-For $\lambda$-s, we create a fresh meta for the domain type (since our surface
-$\lambda$-s are not annotated), and infer types for the bodies. In the case of
-$\einfer{\lambda\,\{x\}.\,t}{ins}{\Theta_0}{\Gamma}$, we additionally perform
-$\einsert$ on the output.
-
-The $\einfer{t\,u}{ins}{\Theta_0}{\Gamma}$ and
-$\einfer{t\,\{u\}}{ins}{\Theta_0}{\Gamma}$ cases are again interesting. Here, we
-first infer a type for the term which is being applied, then refine the type to
-a function type, and lastly check the argument with the domain type. Note the
-difference between the explicit and implicit case. In the former case, we use
-$\einfer{t}{\true}{\Theta_0}{\Gamma}$, which inserts implicit applications. In
-the latter case we do no insertion.  This, together with the inference
-definition for variables, ensures that implicit applications in the surface
-syntax behave similarly as in Agda. For example, given $id : \{A : \U\}\to A\to
-A$ in scope, we elaborate $id\,\U$ as follows:
+For $\einfer{x}{\Theta}{\Gamma}$, we look up the type of $x$ in $\Gamma$. In the
+case of $\U$, we always succeed and infer $\U$ as type. In the cases for
+function types, we check that the domains and codomains have type $\U$.  For
+$\lambda$-s, we create a fresh meta for the domain type (since our surface
+$\lambda$-s are not annotated), and infer types for the bodies.
+
+The $\einfer{t\,u}{\Theta_0}{\Gamma}$ and $\einfer{t\,\{u\}}{\Theta_0}{\Gamma}$
+cases are again interesting. Here, we first infer a type for $t$, then refine
+the type to a function type, and lastly check the argument with the domain
+type. Note the difference between the explicit and implicit case. In the former
+case we use $\einsert\,(\einfer{t}{\Theta_0}{\Gamma})$, which inserts implicit
+applications. In the latter case we do no insertion. This ensures that implicit
+applications in the surface syntax behave similarly as in Agda. For example,
+given $id : \{A : \U\}\to A\to A$ in scope, we elaborate $id\,\U$ as follows:
 \begin{enumerate}
@@
     we check that $\U$ has type $\U$.
-  \item This changes direction again, and we infer $\U$ type for $\U$, and
-    successfully unify $\U$ with $\U$.
+  \item We change direction and infer $\U$ as type for $\U$, and successfully
+    unify $\U$ with $\U$.
 \end{enumerate}
-Note that it would be more efficient (and also allow more user-friendly error
+It would be more efficient (and also allow more user-friendly error
 messages) to not immediately refine the inferred type of $t$ to a function type,
@@
 \subsubsection{Starting and finishing elaboration} Given a surface term $t$,
-we initiate elaboration by computing $\einfer{t}{\true}{\emptycon}{\emptycon}$. If this
+we initiate elaboration by computing $\einfer{t}{\emptycon}{\emptycon}$. If this
 succeeds, we get a $(\Theta,\,\theta,t')$ result. Elaboration is successful overall
@@
 
+%% \begin{alignat*}{3}
+%%   & List && : \U \to \U \\
+%%   & nil  && : \{A : \U\} \to List\,A\\
+%%   & cons && : \{A : \U\} \to A \to List\,A
+%% \end{alignat*}
+
 We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
 assume the following:
-\begin{alignat*}{3}
-  & List && : \U \to \U \\
-  & nil  && : \{A : \U\} \to List\,A\\
-  & cons && : \{A : \U\} \to A \to List\,A
-\end{alignat*}
+\[
+List : \U \to \U \qquad nil : \{A : \U\} \to List\,A \qquad
+cons : \{A : \U\} \to A \to List\,A
+\]
 In the following, we present a trace of checking $cons\,(\lambda\,x.\,x)\,nil$
@@
      \hspace{14em}\\
-  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil}{\true} \\
-  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}{\true}\\
-  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}{\true}\\
+  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil} \\
+  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}\\
+  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}\\
   & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
@@
 \item $\echeck{t}{\Theta}{\Gamma}{A}$ can insert a $\lambda$ only if $A$ is an implicit function type.
-\item $\einsert{\true}{A}$ inserts an application only if $A$ is an implicit function type.
+\item $\einsert\,(\Theta,\,\theta,\,t,\,A)$ inserts an application only if $A$ is an implicit function type.
 \end{enumerate}
@@
 
-At least for $\lambda$-insertion, there is a potential solution: just
-\emph{postpone} checking a term until the shape of the checking type is known
-for sure. This was included as part of a proposed solution for smarter
-$\lambda$-insertions in \cite{johansson2015eliminating}. This means that
-checking with a meta-headed type returns a ``guarded constant''
-\cite[Chapter~3]{norell07thesis}, an opaque stand-in which only computes to an
-actual core term when the checking type becomes known. In practice, this
-solution has a painful drawback: \emph{we get no information at all from checked
-  terms before the guard is unblocked}. For an example for unexpected behavior
-with this solution, let us assume $Bool : \U$ and $true : Bool$, and try to
-infer type for the following surface term:
-\begin{alignat*}{2}
-  \slet\,x : \_ = true\,\sin\, x
-\end{alignat*}
-We first insert a fresh meta $\alpha$ for the hole, and then check $true$ with
-$\alpha$. We postpone this checking, returning a guarded constant, and then
-infer a type for $x$, which is $\alpha$. Hence, this small example yields an
-unsolved meta and a guarded constant in the output.
+There is a potential naive solution: just \emph{postpone} checking a term until
+the shape of the checking type is known for sure. This was included as part of a
+proposed solution for smarter $\lambda$-insertions in
+\cite{johansson2015eliminating}. This means that checking with a meta-headed
+type returns a ``guarded constant'' \cite[Chapter~3]{norell07thesis}, an opaque
+stand-in which only computes to an actual core term when the checking type
+becomes known. In practice, this solution has a painful drawback: \emph{we get
+  no information at all from checked terms before the guard is unblocked}.
+
+For an example for unexpected behavior with this solution, let us assume $Bool :
+\U$ and $true : Bool$, and try to infer type for the surface term $\slet\,x : \_
+= true\,\sin\, x$. We first insert a fresh meta $\alpha$ for the hole, and then
+check $true$ with $\alpha$. We postpone this checking, returning a guarded
+constant, and then infer a type for $x$, which is $\alpha$. Hence, this small
+example yields an unsolved meta and a guarded constant in the output.
 
 Now, this particular example can be repaired by special-casing the elaboration
-of a $\slet$-definition without an explicit type annotation. However, the
-current author's experience from playing with an implementation of this
-solution, is that we are missing too much information by postponing, and this
-cascades in an unfortunate way: postponing yields more unsolved metas, which
-cause more postponing.
+of a $\slet$-definition without an explicit type annotation. However, our
+experience from playing with an implementation of this solution, is that we are
+missing too much information by postponing, and this cascades in an unfortunate
+way: postponing yields more unsolved metas, which cause more postponing.
 
@@
 algebras as follows:
-\begin{alignat*}{3}
-  & \slet\,NatAlgSig : \Tel = (N : \U) \TCons (zero : N) \TCons (suc : N \to N) \TCons \epsilon\,\sin\,...
-\end{alignat*}
+\[
+  \slet\,NatAlgSig : \Tel = (N : \U) \TCons (zero : N) \TCons (suc : N \to N) \TCons \epsilon\,\sin\,...
+\]
 We interpret an $A : \Tel$ as a record type as $\Rec\,A$, which behaves as the
@@
 \item $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
-\item $\einsert{\true}{A}$ does not insert an implicit application if $A$ is meta-headed.
+\item $\einsert\,(\Theta,\,\theta,\,t,\,A)$ does not insert an implicit application when $A$ is meta-headed.
 \end{enumerate}
@@
     metasubstitution can possibly remove this occurrence. In this case the
-    constraint holds vacuously, so we can use the rule
-    \LabTirName{metasub/constancy} to return a $\theta'$ substitution which also
-    solves the constraint.
+    constraint holds vacuously, so we can use the \LabTirName{metasub/constancy}
+    rule to return a $\theta'$ substitution which also solves the constraint.
   \item If we have $x \notin \FreeVars(B[\theta])$, we recursively unify
     $A[\theta]$ with $\epsilon$. If that succeeds, we get a $\theta'$ which
-    unifies $A$ and $\epsilon$ and thus forces the constraint to hold, so we
+    unifies $A[\theta]$ and $\epsilon$ and thus forces the constraint to hold, so we
     can again use \LabTirName{metasub/constancy} to solve the constraint.
@@
 sound to solve the constraint when the occurrence is not rigid. However, this
-way we could lose potential non-$\epsilon$ solutions of $A$.
+way we could lose potential non-$\epsilon$ solutions of $A[\theta]$.
 
@@
 & \unify_{\Theta_0|\Gamma}(\{x : \ol{A}\}\to B)\,(\{x_0 : A_0\}\to B') :\equiv \edo\\
-& \quad (\Theta_1,\,\theta_1,\,A_1) \leftarrow \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
+& \quad \slet\,(\Theta_1,\,\theta_1,\,A_1) = \freshMeta{\Theta_0}{\Gamma,\,x_0 : A_0}{\Tel}\\
 & \quad (\Theta_2,\,\theta_2) \leftarrow
   \unify_{\Theta_1|\Gamma[\theta_1]}\,(A[\theta_1])\,((x : A_0[\theta_1])\TCons A_1)\\
-& \quad (\Theta_3,\,\theta_3) \leftarrow
+& \quad \slet\,(\Theta_3,\,\theta_3) =
     \newConstancy{\Theta_2}{\Gamma[\theta_{12}],\,x_0 : A_0[\theta_{12}]}{x_1}{(A_1[\theta_2])}{(B[\theta_{12}][x\mapsto (x_0 :: x_1)])}\\
@@
 & \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{t})} :\equiv \edo\\
-& \quad (\Theta_1,\,\theta_1,\,A) \leftarrow \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
+& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
 & \quad (\Theta_2,\,\theta_2,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
-& \quad (\Theta_3,\,\theta_3) \leftarrow \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
+& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
 & \quad (\Theta_4,\,\theta_4) \leftarrow
@@
       \hspace{14em}\\
-  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil}{\true} \\
-  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)}{\true}\\
-  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}{\true}\\
+  & \scriptstyle{1 }  && \quad \einfert{cons\,(\lambda\,x.\,x)\,nil} \\
+  & \scriptstyle{2 }  && \qquad \einfert{cons\,(\lambda\,x.\,x)} \\
+  & \scriptstyle{3 }  && \qquad\quad \einfert{cons}\\
   & \scriptstyle{4 }  && \qquad\quad = cons\,\{\alpha_0\}\,:\,\alpha_0 \to List\,\alpha_0 \to List\,\alpha_0\\
@@
 \begin{alignat*}{3}
-& \scriptstyle{0}\qquad\qquad && \einfert{\slet\,x : \_ = true\,\sin\,x}{\true}
+& \scriptstyle{0}\qquad\qquad && \einfert{\slet\,x : \_ = true\,\sin\,x}
       \hspace{21em}\\
@@
 & \scriptstyle{4} && \qquad \mathsf{freshMeta}\,\Tel = \alpha_1\\
-& \scriptstyle{6} && \qquad \einfert{true}{\true} = true : Bool\\
+& \scriptstyle{6} && \qquad \einfert{true}\,= true : Bool\\
 & \scriptstyle{8} && \qquad \mathsf{newConstancy}_{\emptycon,\,x : \Rec\,\alpha_1}\,Bool\\
@@
 create a fresh telescope meta $\alpha_1$, then infer $Bool$ for $true$. On line
-$\scriptstyle{8}$ we create a constancy constraint, and immediately try to solve
-it, in accordance with the algorithmic implementation in Section
-\ref{sec:algorithmic}. Since $Bool$ does not depend on the domain, we solve the
-constraint and thereby solve $\alpha_1$ as $\epsilon$. Hence, we simply return
-$true$ on line $\scriptstyle{11}$, since $\lambda\,\{x : \ol{\epsilon}\}.\,true$
-computes to that, and elaboration succeeds with the expected output.
+$\scriptstyle{8}$ we create a constancy constraint and immediately try to solve
+it, as described in Section \ref{sec:algorithmic}. Since $Bool$ does not depend
+on the domain, we solve the constraint and thereby solve $\alpha_1$ as
+$\epsilon$. Hence, we simply return $true$ on line $\scriptstyle{11}$, since
+$\lambda\,\{x : \ol{\epsilon}\}.\,true$ computes to that, and elaboration
+succeeds with the expected output.
 \end{example}
@@
 In this section we examine how the elaborator fares in practice, its
-limitations, and how it compares to related works. We abbreviate our elaborator
-as FCIF when comparing it to others.
+limitations, and how it compares to related works. We refer to our elaborator as
+FCIF when comparing it to others.
 
 \subsection{Related Works}
-MLF \cite{le2014mlf} is an extension of System F with polymorphic subtyping
-bounds, which supports strong inference for first class polymorphism. HML
-\cite{leijen2009flexible}) is a simplified variant of MLF. Since these systems
-diverge markedly from (fragments of) Martin-Löf type theory and rely critically
-on subtyping, it is unclear how they could be extended to dependent type
-theories. HMF \cite{leijen2008hmf} is a relatively simple system with
+MLF \cite{le2014mlf} extends System F with polymorphic subtyping
+bounds, and supports strong inference for first class polymorphism. HML
+\cite{leijen2009flexible}) is a simplified variant of MLF. These systems rely
+critically on subtyping and thus diverge markedly from (fragments of) Martin-Löf
+type theory. HMF \cite{leijen2008hmf} is a relatively simple system with
 first-class polymorphism; however, it is also weaker than others in terms of
@@
 \cite{serrano2018guarded} and quick look impredicativity (QL)
-\cite{serrano2020a}. They both try to examine n-ary applications to find
+\cite{serrano2020a}. They both work by examining n-ary applications to find
 metavariable instantiations which arise from occurrences guarded by rigid type
 constructors. GI's use of \emph{generalization constraints} anticipates our use
-of telescopes, but overall GI is also burdened by inessential complexity. QL
+of telescopes, but overall GI is also burdened by a great deal of complexity. QL
 streamlines GI by eschewing constraints in favor of an eager preprocessing pass
@@
 curried $\lambda$ insertion in the definition of
-$\einfer{\lambda\,x.\,t}{ins}{\Theta}{\Gamma}$.
+$\einfer{\lambda\,x.\,t}{\Theta}{\Gamma}$.
 
@@
 
-Secondly, FCIF supports global inference and does not handle neutral spines
-specially. For an example for global inference, $\slet\,x : \_ =
-single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF, MLF and HMF but does
-not work in GI, QL and HMF. MLF and HML work here by immediately giving a
-principal type to $single\,id$ which involves generalization with a subtyping
-bound. In contrast, FCIF makes no promises about principal typing --- which is
-not feasible with dependent types --- and does no generalization, but it can
-still infer a type for $id$ which can be later constrained to $IdTy$.
+Secondly, FCIF supports global inference. For an example for global inference,
+$\slet\,x : \_ = single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF, MLF
+and HMF but does not work in GI, QL and HMF. MLF and HML work here by
+immediately giving a principal type to $single\,id$ which involves
+generalization with a subtyping bound. In contrast, FCIF makes no promises about
+principal typing --- which is not feasible with dependent types --- and does no
+generalization, but it can still infer a type for $id$ which can be later
+constrained to $IdTy$.
 

### 2020-02-29 paper.tex
stuff

@@
 \begin{alignat*}{3}
-& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{t})} :\equiv \edo\\
+& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
 & \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
@@
 & \quad (\Theta_4,\,\theta_4) \leftarrow
-            \unify\, ((\alpha\,\ol{t})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
+            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
 & \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
@@
 infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
-an extension of FCIF with \emph{implicit application insertion}, which could
-possibly handle this example. We note though that this is not yet fleshed out
-and is a subject of future work.
-
-\section{Future Work and Conclusion}
+an extension of FCIF with \emph{curried application insertion}, which could
+possibly handle this example. We extend $\einsert$ with the following case. We
+omit metacontexts and metasubstitutions for brevity.
+\begin{alignat*}{3}
+& \einsert\,(t,\,\alpha\,\ol{u}) :\equiv \edo \\
+& \quad A \leftarrow \mathsf{freshMeta}_{\Gamma}\,\Tel\\
+& \quad B \leftarrow \mathsf{freshMeta}_{\Gamma,\,x: \Rec\,A}\,\U\\
+& \quad \unify\,(\alpha\,\ol{u})\,(\{x : \ol{A}\} \to B)\\
+& \quad \mathsf{newConstancy}_{\Gamma, x : \Rec\,A}\,B\\
+& \quad u \leftarrow \mathsf{freshMeta}_{\Gamma}\,(\Rec\,A)\\
+& \quad \ereturn\,(t\,\{u : \ol{A}\},\,B[x \mapsto u])
+\end{alignat*}
+In short, we insert a new curried application whenever we $\einsert$ with
+meta-headed type. Now, elaboration returns
+$f\,(\lambda\,\{A\}.\,x\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})$, and right before
+we try to unify expected and inferred types, we have $x : \{y :
+\ol{\alpha_0}\}\to \alpha_1\,y$, and we have to compute
+$\unify\,(\alpha_1\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})\,(A \to A)$. This
+problem does not fall in any usual pattern fragment, and does not have an
+obvious most general solution. However, it is reasonable to make the following
+assumption: metas like $\alpha_2$ are only solved with \emph{order-preserving
+  embeddings}, so for $\alpha_2$ the four potential solutions are
+$(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$, $(\lambda\,x\,A.\,(x ::
+[]))$ and $(\lambda\,x\,A.\,(x :: A :: []))$. The reason for only allowing such
+solutions is that these cases can possibly yield solvable pattern unification
+problems, while solutions containing arbitrary terms or non-linear variable
+occurrences cannot. Out of these, $(\lambda\,x\,A.\,(x :: A :: []))$ and
+$(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would require a cyclic
+solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled out because it
+would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$ problem. Hence, there
+is a unique order-preserving embedding solution in this case, which does yield
+the expected elaboration output. However, note that this idea is far from being
+fully fleshed out, and it could be subject of future research.
+
+\section{Conclusions and Future Work}
 

### 2020-02-29 paper.tex
start preparing for submission

@@
 \inferrule*[lab=insert]{\Theta_0|\Gamma\vdash \\ (\Theta,\,\theta,\,t,\,A) \in
-             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta_0 \To \Theta)\,\land\,
+             \{(\Theta,\,\theta,\,t,\,A)\,|\,(\theta : \Theta \To \Theta_0)\,\land\,
                (\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
            {\einsert(\Theta,\,\theta,\,t,\,A) \in \{(\Theta,\,\theta,\,t,\,A)\,|\,
-             (\theta : \Theta_0 \To \Theta)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
+             (\theta : \Theta \To \Theta_0)\,\land\,(\Theta|\Gamma[\theta]\vdash t : A)\}\cup\{\fail\}}
 \end{mathpar}
@@
 solutions which work one neutral spine at a time (HMF, GI, QL) also face issues
-with dependently typed spines, since it is not possible to do anything with
-later arguments in such spines until previous arguments are elaborated.
+with dependently typed spines. With such spines, we cannot simply process later
+arguments when previous ones are not yet elaborated, as return types depend on
+argument values, and skipping over arguments may clog up type computation in an
+unacceptable way.
 
@@
 an extension of FCIF with \emph{curried application insertion}, which could
-possibly handle this example. We extend $\einsert$ with the following case. We
+possibly handle this example. We extend insertion with the following case. We
 omit metacontexts and metasubstitutions for brevity.
@@
 obvious most general solution. However, it is reasonable to make the following
-assumption: metas like $\alpha_2$ are only solved with \emph{order-preserving
-  embeddings}, so for $\alpha_2$ the four potential solutions are
-$(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$, $(\lambda\,x\,A.\,(x ::
-[]))$ and $(\lambda\,x\,A.\,(x :: A :: []))$. The reason for only allowing such
-solutions is that these cases can possibly yield solvable pattern unification
-problems, while solutions containing arbitrary terms or non-linear variable
-occurrences cannot. Out of these, $(\lambda\,x\,A.\,(x :: A :: []))$ and
-$(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would require a cyclic
-solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled out because it
-would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$ problem. Hence, there
-is a unique order-preserving embedding solution in this case, which does yield
-the expected elaboration output. However, note that this idea is far from being
-fully fleshed out, and it could be subject of future research.
-
-\section{Conclusions and Future Work}
+assumption: metas which return in unknown record types, like $\alpha_2$, are
+only solved with \emph{order-preserving embeddings}, so for $\alpha_2$ the four
+potential solutions are $(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$,
+$(\lambda\,x\,A.\,(x :: []))$ and $(\lambda\,x\,A.\,(x :: A :: []))$. The reason
+for only allowing such solutions is that these can possibly yield solvable
+pattern unification problems, while solutions containing arbitrary terms or
+non-linear variable occurrences cannot. Out of these, $(\lambda\,x\,A.\,(x :: A
+:: []))$ and $(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would
+require a cyclic solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled
+out because it would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$
+problem. Thus, $(\lambda\,x\,A.\,(A :: []))$ is the unique order-preserving
+embedding solution in this case, which does yield the expected elaboration
+output. However, this idea is far from being fully fleshed out, and it
+could be subject of future research.
+
+\section{Conclusion and Future Work}
+
+In type theory, it is a common endeavor to search for theories and features
+which confer the greatest amount of expressive power for the least amount of
+formal and conceptual complexity. In relation to elaboration and inference
+algorithms, we would similarly like to use tools and concepts with high
+power-to-weight ratio. Sometimes core theories need to be extended to allow more
+powerful elaboration. Certainly, type inference would be cumbersome without the
+notion of metavariables. Likewise, contextual metavariables are essential
+for elaboration in the presence of type dependencies.
+
+In this paper we propose the concept of strictly curried functions, which
+supports elaboration of first-class implicit functions. It seems that trying to
+solve this problem by fiddling with postponing and heuristics, without extending
+the core theory, does not really cut it. In a dependently typed spine $t\,u\,v$, when
+elaborating $v$ we already want to have an elaborated version of $u$ with
+computational behavior. While contextual metavariables yield a nice modal type
+theory which computes in the presence of \emph{unknown terms}, our curried
+functions extend this to a system which also computes in the presence of
+unknown \emph{implicit insertions}. Since implicit insertions are in a way also
+just unknown terms, we can reuse most of the infrastructure of contextual/crisp
+type theory, and only add modest extensions.
+
+Often it is the case that constructions on dependent type theory are forced to
+be more principled and structured than constructions on less powerful theories,
+because the many interactions and intricacies leave less room for whims. In our
+case, we believe that focusing on dependent type theory helped us hone in on the
+essential parts of the problem, and it is likely that restrictions of our
+solution would be also favorably simple and powerful in simpler settings such as
+System F.
+
+In future work, we would like to
+\begin{itemize}
+\item Investigate extending elaboration with curried application insertions, along the lines
+      of Section \ref{sec:appinsert}.
+\item Formalize elaboration and unification from an algebraic/categorical perspective, in particular
+      give algebraic definitions for the core theories.
+\item Investigate whether features of our elaborator could be implemented in a
+      reasonably practical way in production systems such as Agda. Also, investigate
+      simplifying our algorithm to non-dependent (e.g.\ System F) settings.
+\item Investigate completeness and properties related to inferability.
+\end{itemize}
 

### 2020-02-29 Notes.hs

@@ -1,12 +1,0 @@
-{-# LANGUAGE KindSignatures #-}
-{-# LANGUAGE RankNTypes #-}
-{-# LANGUAGE TypeApplications #-}
-
-import Prelude hiding (id)
-
-id :: forall (a :: *). a -> a
-id x = x
-
-majom = 100022
-
-foo = id @Bool True

### 2020-02-29 fcif/Elaboration.hs
stuff

@@ -1,5 +1,4 @@
 
-
 module Elaboration where
 
 import Control.Exception
@@ -14,8 +13,6 @@ import Types
 import Evaluation
 import ElabState
 import Errors
-
-import Debug.Trace
 
 -- Context operations
 --------------------------------------------------------------------------------

### 2020-02-29 fcif/Main.hs

@@ -7,7 +7,6 @@ import System.Exit
 import System.Environment
 
 import Types
-import Pretty
 import Evaluation
 import Elaboration
 import Errors
@@ -68,290 +67,3 @@ main = mainWith getArgs parseStdin
 -- | Run main with inputs as function arguments.
 main' :: String -> String -> IO ()
 main' mode src = mainWith (pure [mode]) ((,src) <$> parseString src)
-
-
-pruneTest = main' "elab" $ unlines [
-  "λ (Bool : U)(not : Bool → Bool → Bool)(true : Bool).",
-  "λ f (x : Bool). not (f x)"
-  ]
-
-pruneTest2 = main' "elab" $ unlines [
- "λ (ℕ : U)(zero : ℕ)(suc : ℕ → ℕ)(add : ℕ → ℕ → ℕ).",
- "let test1 = (P : _) → (n : ℕ) → P n → P n in",
- "let test2 = λ f x . add (suc (f x)) x in",
- "U"
- ]
-
-test0 = main' "elab" $ unlines [
-  "λ (A : U)(a : A).",
-
-  "let id : {A} → A → A = λ x. x in",
-  "_"
-  ]
-
-
-test = main' "elab" $ unlines [
-  "λ (FOO : U)(foobarr : FOO).",
-  "let the : (A : U) → A → A = \\A x.x in",
-  "let IdTy = {A} → A → A in",
-  "let id : IdTy = λ x. x in",
-  "let choose : {A} → A → A → A = λ x y. x in",
-
-  "let Nat : U",
-  "    = (N : U) -> (N -> N) -> N -> N in",
-  "let zero : Nat = λ _ s z. z in",
-  "let inc : Nat → Nat = λ n N s z. s (n _ s z) in",
-
-  "let mul : Nat -> Nat -> Nat",
-  "    = \\a b N s z. a _ (b N s) z in",
-  "let ten : Nat",
-  "    = \\N s z. s (s (s (s (s (s (s (s (s (s z))))))))) in",
-  "let hundred = mul ten ten in",
-
-  "let List : U -> U",
-  "    = \\A. (L : U) -> (A -> L -> L) -> L -> L in",
-  "let nil : {A : U} -> List A",
-  "    = \\cons L nil. nil in",
-  "let cons : {A : U} -> A -> List A -> List A",
-  "    = \\{A} x xs L cons nil. cons x (xs _ cons nil) in",
-
-  "let append : {A} → List A → List A → List A",
[…]

### 2020-02-29 fcif/Types.hs

@@ -4,9 +4,7 @@ module Types (
   module Text.Megaparsec
   ) where
 
-import Control.Exception
 import Text.Megaparsec (SourcePos(..), unPos, initialPos)
-import Text.Printf
 import Lens.Micro.Platform
 
 import qualified Data.IntMap.Strict as IM

### 2020-03-01 paper.tex
stuff

@@
 simple implementation which only requires minimal changes to unification, and
-which is already remarkably powerful. It seems that handling implicit
+which is already remarkably powerful. It seems that enhancing implicit
 application insertions requires extending unification; we discuss this in
@@
 potential solutions are $(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$,
-$(\lambda\,x\,A.\,(x :: []))$ and $(\lambda\,x\,A.\,(x :: A :: []))$. The reason
+$(\lambda\,x\,A.\,(x :: []))$ and $(\lambda\,x\,A.\,(x :: A :: []))$\footnote{Technically, we could also allow permutations such as $\lambda\,x\,A.\,(A :: x :: [])$, but these are superfluous since nothing in our system depends on the ordering of permutable function arguments.}.
+The reason
 for only allowing such solutions is that these can possibly yield solvable

### 2020-03-01 paper.tex
fix
cosmetics

@@
   \item Our solution is simple: we implemented elaboration,
-    evaluation and unification in about 670 lines of Haskell, of which 210 lines
+    evaluation and unification in about 670 lines of Haskell, of which 215 lines
     implement the novel enhancements on the top of Norell's basic elaborator.
@@
 or optionally the type or the normal form of the result. It is implemented in
-1061 lines of Haskell. Of this, 673 lines constitute the core syntax,
-evaluation, unification and elaboration. Of these 673 lines, 458 lines implement
+1056 lines of Haskell. Of this, 669 lines constitute the core syntax,
+evaluation, unification and elaboration. Of these 669 lines, 454 lines implement
 the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215
@@
 we try to unify expected and inferred types, we have $x : \{y :
-\ol{\alpha_0}\}\to \alpha_1\,y$, and we have to compute
+\ol{\alpha_0}\}\to \alpha_1\,\{y : \ol{\alpha_0}\}$, and we have to compute
 $\unify\,(\alpha_1\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})\,(A \to A)$. This

### 2020-03-01 fcif/ElabState.hs

@@ -1,4 +1,8 @@
 
+-- | Top-level mutable state involved in elaboration. We use actual mutable
+--   top-level references simply because it's convenient and our simple
+--   program does not call for anything more sophisticated.
+
 module ElabState where
 
 import Data.IORef

### 2020-03-01 fcif/Elaboration.hs

@@ -131,6 +131,7 @@ newConstancy cxt dom cod =
 -- | Checks that a spine consists only of distinct bound vars.
 --   Returns a partial variable renaming on success, alongside the size
 --   of the spine, and the list of variables in the spine.
+--   May throw SpineError.
 checkSp :: Spine -> IO (Renaming, Lvl, [Lvl])
 checkSp = (over _3 reverse <$>) . go . forceSp where
   go :: Spine -> IO (Renaming, Lvl, [Lvl])
@@ -176,8 +177,9 @@ closingTm = go 0 where
       LamTel x (quote d a) $ go (d + 1) (b (VVar d), len-1, drop 1 xs) rhs
     _            -> error "impossible"
 
--- | Strengthen a value, returns quoted normal result. This performs scope
+-- | Strengthens a value, returns a quoted normal result. This performs scope
 --   checking, meta occurs checking and (recursive) pruning at the same time.
+--   May throw StrengtheningError.
 strengthen :: Str -> Val -> IO Tm
 strengthen str = go where
 
@@ -269,6 +271,7 @@ strengthen str = go where
     SProj1 sp      -> Proj1 <$> goSp h sp
     SProj2 sp      -> Proj2 <$> goSp h sp
 
+-- | May throw UnifyError.
 solveMeta :: Cxt -> MId -> Spine -> Val -> IO ()
 solveMeta cxt m sp rhs = do
 
@@ -307,13 +310,14 @@ freshMeta cxt (quote (cxt^.len) -> a) = do
   let sp = fst $ vars (cxt^.types)
   pure (quote (cxt^.len) (VNe (HMeta m) sp))
 
-
+-- | Wrap the inner UnifyError arising from unification in an UnifyErrorWhile.
 unifyWhile :: Cxt -> Val -> Val -> IO ()
 unifyWhile cxt l r =
   unify cxt l r
   `catch`
   (report (cxt^.names) . UnifyErrorWhile (quote (cxt^.len) l) (quote (cxt^.len) r))
 
+-- | May throw UnifyError.
 unify :: Cxt -> Val -> Val -> IO ()
 unify cxt l r = go l r where
 
@@ -440,7 +444,7 @@ check cxt topT ~topA = case (topT, force topA) of
     unifyWhile cxt va topA
     pure t
 
--- | We special case top-level lambdas (serving as postulates) for better
+-- | We specialcase top-level lambdas (serving as postulates) for better
[…]

### 2020-03-01 fcif/Main.hs

@@ -16,11 +16,11 @@ import Zonk
 
 
 helpMsg = unlines [
-  "usage: holes [--help|nf|type]",
+  "usage: fcif [--help|elab|nf|type]",
   "  --help : display this message",
-  "  elab   : read & elaborate expression from stdin",
+  "  elab   : read & elaborate expression from stdin, print elaboration output",
   "  nf     : read & elaborate expression from stdin, print its normal form",
-  "  type   : read & elaborate expression from stdin, print its type"]
+  "  type   : read & elaborate expression from stdin, print its (normal) type"]
 
 displayError :: String -> Err -> IO a
 displayError file (Err ns err (Just (SPos (SourcePos path (unPos -> linum) (unPos -> colnum))))) = do

### 2020-03-01 fcif/Pretty.hs

@@ -4,15 +4,17 @@ module Pretty (showTm, showTopTm) where
 
 import Types
 
--- special case printing of top lambdas, since they are usually used
--- to postulate stuff
+-- | We specialcase printing of top lambdas, since they are usually used
+--   to postulate stuff. We use '*' in a somewhat hacky way to mark
+--   names bound in top lambdas, so that later we can avoid printing
+--   them in meta spines.
 topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
 topLams p pre post ns (Lam (fresh ns -> x) i a t) =
   showParen p (
     (pre++)
   . icit i bracket parens (
          ((if null x then "_" else x)++) . (" : "++) . go False ns a)
-  . topLams False "\n " ".\n\n" (('*':x):ns) t)
+  . topLams False "\n " ".\n\n" (('*':x):ns) t) -- note the '*'
 topLams _ pre post ns t = (post++) . go False ns t
 
 fresh :: [Name] -> Name -> Name

### 2020-03-01 fcif/Types.hs

@@ -28,6 +28,7 @@ icit :: Icit -> a -> a -> a
 icit Impl i e = i
 icit Expl i e = e
 
+-- | Surface syntax.
 data Raw
   = RVar Name                        -- ^ x
   | RLam Name (Maybe Raw) Icit Raw   -- ^ λx.t  or λ{x}.t with optional type annotation
@@ -61,7 +62,7 @@ data MetaEntry
   | Constancy Cxt VTy VTy BlockedBy
 
 
--- | A partial mapping from levels to levels. Undefined domain reresents
+-- | A partial mapping from levels to levels. Undefined domain represents
 --   out-of-scope or "illegal" variables.
 type Renaming = IM.IntMap Lvl
 
@@ -71,7 +72,7 @@ data Str = Str {
   _strDom :: Lvl,        -- ^ size of renaming domain
   _strCod :: Lvl,        -- ^ size of renaming codomain
   _strRen :: Renaming,   -- ^ partial renaming
-  _strOcc :: Maybe MId   -- ^ meta for occurs strengthening
+  _strOcc :: Maybe MId   -- ^ meta for occurs checking
   }
 
 -- | Lift over a bound variable.
@@ -82,9 +83,12 @@ liftStr (Str c d r occ) = Str (c + 1) (d + 1) (IM.insert d c r) occ
 skipStr :: Str -> Str
 skipStr (Str c d r occ) = Str c (d + 1) r occ
 
-
+-- | Value environment. `VSkip` skips over a bound variable.
 data Vals  = VNil | VDef Vals ~Val | VSkip Vals
+
+-- | Type environment.
 data Types = TNil | TDef Types ~VTy | TBound Types ~VTy
+
 type Ix    = Int
 type Lvl   = Int
 type Ty    = Tm
@@ -100,11 +104,13 @@ pattern TSnoc as a <- ((\case TBound as a -> Just (as, a)
 lvlName :: [Name] -> Lvl -> Name
 lvlName ns x = ns !! (length ns - x - 1)
 
+-- | We need to distinguish invented names from source names, as
+--   we don't want the former to shadow the latter during name lookup
+--   in elaboration.
 data NameOrigin =
     NOSource        -- ^ Names which come from surface syntax.
   | NOInserted      -- ^ Names of binders inserted by elaboration.
 
-type MetaInsertion = Bool
 
 -- | Context for elaboration and unification.
 data Cxt = Cxt {

### 2020-03-01 paper.tex
minor fix

@@
 recursion on the $A$ telescope, but the $\beta\eta$ rules are specified
-generically for an arbitrary (possibly neutral)
-telescope. \LabTirName{curried-$\beta$} is still provable as a propositional
+generically for arbitrary (possibly neutral)
+telescopes. \LabTirName{curried-$\beta$} is still provable as a propositional
 equality, and assuming function extensionality \LabTirName{curried-$\eta$} is
@@
 Hence, we can derive a somewhat weaker version of curried functions, with
-propositional $\beta$ and $\eta$. From this, we still get consistency almost for
-free. That is because whenever we have a model of the base theory with $\Nat$,
-$\top$ and $\Sigma$, such that the model also validates equality reflection,
-then every propositionally provable equation is also validated as a definitional
-equation. And fortunately for us, standard models which prove consistency
-usually validate equality reflection, e.g.\ set-theoretical models or standard
-models in extensional type theory.
-
-In contrast, showing canonicity, normalization and decidability of conversion
-would require some extra work. We leave this to future work, but we expect that
-it is straightforward to extend previous proofs to cover strict $\beta$ and
-$\eta$ for curried functions.
+propositional $\beta$ and $\eta$. From this, we still get consistency and
+canonicity very cheaply. This is because models proving consistency or
+canonicity for the base theory with $\Nat$, $\top$ and $\Sigma$ usually support
+equality reflection. For consistency, standard set-theoretical models and models
+in extensional type theory have this property, for canonicity, glued models
+e.g.\ as in \cite{kaposi2019gluing, sterling2019algebraic} also have this property.
+
+In contrast, showing normalization and decidability of conversion would require
+some extra work. We leave this to future work, but we expect that it is not
+difficult to extend previous proofs to cover strict $\beta$ and $\eta$ for
+curried functions.
 

### 2020-03-01 paper.tex
fix

@@
 \subsection{Surface syntax}
-Figure \ref{fig:surface} shows the the possible constructs in the surface
+Figure \ref{fig:surface} shows the possible constructs in the surface
 language. We only have terms, as we have Russell-style universe in the core, and
@@
 
-The syntactic constructs are the almost the same in the surface language as in
+The syntactic constructs are almost the same in the surface language as in
 the core syntax. The difference is that $\_$ holes only appear in surface
@@
 typing rule for $\slet$-definitions, and we have that any defined variable is
-equal to its definition. We mainly support this as a convenience feature in the
-surface language.
+equal to its definition. We support $\slet$ as a convenience feature which
+is important for defining more involved example programs\footnote{In dependent type theories, the $\slet$ rule is not
+  derivable from function application, unlike in simple type theories.}.
 
@@
 The core syntax is quite expressive as a programming language, thanks to
-$\slet$-definitions\footnote{In dependent type theories, the $\slet$ rule is not
-  derivable from function application, unlike in simple type theories.} and the
-type-in-type rule which allows Church-encodings of a large class of inductive
+$\slet$-definitions and the
+type-in-type rule which allows Church-encodings for a large class of inductive
 types. For example, the following term computes a list of types by mapping:
@@
 \begin{definition}[Contextualization]\label{def:contextualization}
-For each $\Theta|\Gamma \vdash t : \Gamma\To A$, we define the spine
+For each $\Theta|\emptycon \vdash t : \Gamma\To A$, we define the spine
 $\overline{\mathsf{vars}_{\Gamma}}$ such that that $\Theta|\Gamma\vdash
@@
 assuming we have a constancy constraint for $A$ and $B$ in context, if $B$ does
-not depend on the $x : \Rec\,A$ domain variable, then $A$ is equal to the the
-empty telescope $\epsilon$.
+not depend on the $x : \Rec\,A$ domain variable, then $A$ is equal to the empty
+telescope $\epsilon$.
 
@@
 $t\,\{u : \ol{A}\}$, unification is structural, and other cases remain the same
-as in the the basic elaborator of Section \ref{sec:bidirectional_elaboration}.
+as in the basic elaborator of Section \ref{sec:bidirectional_elaboration}.
 

### 2020-03-02 paper.tex
fix type of cons
fixes
typo fix

@@
   &\einsert\,(\Theta_1,\,\theta_1,\,t,\,\{x : A\}\to B) && :\equiv\hspace{0.5em}&&
-       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta}{\Gamma}{A}\\
+       \slet\,(\Theta_2,\,\theta_2,\,u) = \freshMeta{\Theta_1}{\Gamma[\theta_1]}{A}\\
   & && &&\sin\,\einsert\,(\Theta_2,\,\theta_{12},\,(t[\theta_2])\,\{u\},\,B[\theta_2][x\mapsto u])\\
@@
 \[
-NatInd : \{P : Nat \to \U\}\to \,P\,\zero \to (\{n : Nat\}\to P\,n \to
+NatInd : (P : Nat \to \U)\to \,P\,\zero \to (\{n : Nat\}\to P\,n \to
          P\,(\suc\,n)) \to (n : Nat) \to P\,n
@@
 
-%% \begin{alignat*}{3}
-%%   & List && : \U \to \U \\
-%%   & nil  && : \{A : \U\} \to List\,A\\
-%%   & cons && : \{A : \U\} \to A \to List\,A
-%% \end{alignat*}
-
 We revisit now the $polyList$ example from Section \ref{sec:introduction}. We
@@
 List : \U \to \U \qquad nil : \{A : \U\} \to List\,A \qquad
-cons : \{A : \U\} \to A \to List\,A
+cons : \{A : \U\} \to A \to List\,A \to List\,A
 \]
@@
 We extend spine notation to applications of curried functions. For example, we
-may have a spine $\ol{t} \equiv (\{x : \ol{A}\}\,\{y : \ol{B}\})$. We
+may have a spine $\ol{t} \equiv (\{x : \ol{A}\},\,\{y : \ol{B}\})$. We
 accordingly revise Definition \ref{def:contextualization} for
@@
 $\slet\,x : \_ = single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF, MLF
-and HMF but does not work in GI, QL and HMF. MLF and HML work here by
+and HML but does not work in GI, QL and HMF. MLF and HML work here by
 immediately giving a principal type to $single\,id$ which involves

### 2020-03-03 paper.tex
fix

@@
   & cons   && : \{A : \U\} \to A \to List\,A \to List\,A           && revapp && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
-  & head   && : \{A : \U\} List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
-  & tail   && : \{A : \U\} List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
+  & head   && : \{A : \U\} \to List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
+  & tail   && : \{A : \U\} \to List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
   & map    && : \{A B : \U\} \to (A \to B) \to List\,A \to List\,B && poly   && : IdTy \to Pair\,Int\,Bool

### 2020-03-24 paper.tex
typo
fix

--- 2020-03-02 paper.tex
@@
 & \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
-& \quad (\Theta_2,\,\theta_2,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
+& \quad (\Theta_2,\,\theta_2,\,t',\,B) \leftarrow \einfer{t}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
 & \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
@@
   & cons   && : \{A : \U\} \to A \to List\,A \to List\,A           && revapp && : \{A\,B : \U\} \to A \to (A \to B) \to B            \\
-  & head   && : \{A : \U\} List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
-  & tail   && : \{A : \U\} List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
-  & map    && : \{A B : \U\} \to (A \to B) \to List\,A \to List\,B && poly   && : IdTy \to Pair\,Int\,Bool
+  & head   && : \{A : \U\} \to List\,A \to A                           && runST  && : \{A : \U\} \to (\{S : \U\} \to ST\,S\,A) \to A     \\
+  & tail   && : \{A : \U\} \to List\,A \to List\,A                     && argST  && : \{S : \U\}\to ST\,S\,Int                           \\
+  & map    && : \{A\,B : \U\} \to (A \to B) \to List\,A \to List\,B && poly   && : IdTy \to Pair\,Int\,Bool
 \end{alignat*}

### 2020-03-24 fcif/Elaboration.hs
progress

@@ -21,18 +21,18 @@ emptyCxt :: Cxt
 emptyCxt = Cxt VNil TNil [] [] 0
 
 -- | Add a bound variable.
-bind :: Name -> NameOrigin -> VTy -> Cxt -> Cxt
-bind x o ~a (Cxt vs tys ns no d) =
-  Cxt (VSkip vs) (TBound tys a) (x:ns) (o:no) (d + 1)
+bind :: Name -> NameOrigin -> VTy -> VUniv -> Cxt -> Cxt
+bind x o ~a ~un (Cxt vs tys ns no d) =
+  Cxt (VSkip vs) (TBound tys a un) (x:ns) (o:no) (d + 1)
 
 -- | Add a bound variable which comes from surface syntax.
-bindSrc :: Name -> VTy -> Cxt -> Cxt
+bindSrc :: Name -> VTy -> VUniv -> Cxt -> Cxt
 bindSrc x = bind x NOSource
 
 -- | Define a new variable.
-define :: Name -> VTy -> Val -> Cxt -> Cxt
-define x ~a ~t (Cxt vs tys ns no d) =
-  Cxt (VDef vs t) (TDef tys a) (x:ns) (NOSource:no) (d + 1)
+define :: Name -> VTy -> VUniv -> Val -> Cxt -> Cxt
+define x ~a ~un ~t (Cxt vs tys ns no d) =
+  Cxt (VDef vs t) (TDef tys a un) (x:ns) (NOSource:no) (d + 1)
 
 -- | Lift ("skolemize") a value in an extended context to a function in a
 --   non-extended context.
@@ -68,10 +68,7 @@ occurs d topX = occurs' d mempty where
 
     goSp ms sp = case forceSp sp of
       SNil           -> mempty
-      SApp sp u i    -> goSp ms sp <> go u
-      SAppTel a sp u -> go a <> goSp ms sp <> go u
-      SProj1 sp      -> goSp ms sp
-      SProj2 sp      -> goSp ms sp
+      SApp sp u un i -> goSp ms sp <> go u <> go un
 
     goBind t =
       occurs' (d + 1) ms (t (VVar d))
@@ -80,50 +77,10 @@ occurs d topX = occurs' d mempty where
       VNe (HVar x) sp | x == topX -> occurrence ms <> goSp ms sp
       VNe (HVar x) sp             -> goSp ms sp
       VNe (HMeta m) sp            -> goSp (IS.insert m ms) sp
-      VPi _ i a b   -> go a <> goBind b
[…]

### 2020-03-24 fcif/Errors.hs

@@ -30,6 +30,7 @@ data ElabError
   = UnifyErrorWhile Tm Tm UnifyError
   | NameNotInScope Name
   | ExpectedFunction Tm
+  | ExpectedType Tm Tm
   | IcitMismatch Icit Icit
 
 data Err = Err {
@@ -93,6 +94,11 @@ showError ns = \case
     "Name not in scope: " ++ x
   ExpectedFunction ty ->
     "Expected a function type, instead inferred:\n\n  " ++ showTm ns ty
+  ExpectedType a un -> printf (
+    "Expected type Set or Prop for expression:\n\n" ++
+    "  %s\n\n" ++
+    "inferred type\n\n" ++
+    "  %s") (showTm ns a) (showTm ns un)
   IcitMismatch i i' -> printf (
     "Function icitness mismatch: expected %s, got %s.")
     (show i) (show i')

### 2020-03-24 fcif/Evaluation.hs

@@ -4,18 +4,6 @@ module Evaluation where
 import Types
 import ElabState
 
-vProj1 :: Val -> Val
-vProj1 (VTcons t u)    = t
-vProj1 (VNe h sp)      = VNe h (SProj1 sp)
-vProj1 (VLamTel x a t) = vProj1 (vLamTel id x a t)
-vProj1 _               = error "impossible"
-
-vProj2 :: Val -> Val
-vProj2 (VTcons t u)    = u
-vProj2 (VNe h sp)      = VNe h (SProj2 sp)
-vProj2 (VLamTel x a t) = vProj2 (vLamTel id x a t)
-vProj2 _               = error "impossible"
-
 vVar :: Ix -> Vals -> Val
 vVar 0 (VDef vs v) = v
 vVar 0 (VSkip vs)  = VVar (valsLen vs)
@@ -27,7 +15,6 @@ vMeta :: MId -> Val
 vMeta m = case runLookupMeta m of
   Unsolved{} -> VMeta m
   Solved v   -> v
-  _          -> error "impossible"
 
 -- | Force the outermost constructor in a value, do not force the spine
 --   of a neutral value.
@@ -36,10 +23,7 @@ force = \case
   v@(VNe (HMeta m) sp) -> case runLookupMeta m of
     Unsolved{} -> v
     Solved v   -> force (vAppSp v sp)
-    _          -> error "impossible"
-  VPiTel x a b  -> vPiTel force x a b
-  VLamTel x a t -> vLamTel force x a t
-  v             -> v
+  v            -> v
 
 -- | Force a spine, computing telescope applications where possible.
 forceSp :: Spine -> Spine
@@ -51,69 +35,28 @@ forceSp sp =
     VNe _ sp -> sp
     _        -> error "impossible"
 
-vApp :: Val -> Val -> Icit -> Val
-vApp (VLam _ _ _ t)  ~u i = t u
-vApp (VNe h sp)      ~u i = VNe h (SApp sp u i)
-vApp (VLamTel x a t) ~u i = vApp (vLamTel id x a t) u i
-vApp _                _ _ = error "impossible"
-
-vPiTel :: (Val -> Val) -> Name -> VTy -> (Val -> Val) -> Val
-vPiTel k x a b = case force a of
-  VTEmpty        -> k (b VTempty)
-  VTCons x1 a as -> let x1 = x ++ "1"
-                        x2 = x ++ "2"
-                    in VPi x1 Impl a $ \ ~x1 ->
[…]

### 2020-03-24 fcif/Main.hs

@@ -1,4 +1,3 @@
-
 module Main where
 
 import Text.Printf

### 2020-03-24 fcif/Parser.hs

@@ -35,7 +35,7 @@ pBind    = pIdent <|> symbol "_"
 
 keyword :: String -> Bool
 keyword x =
-  x == "let" || x == "in" || x == "λ" || x == "U" || x == "assume"
+  x == "let" || x == "in" || x == "λ" || x == "Set" || x == "Prop"
 
 pIdent :: Parser Name
 pIdent = try $ do
@@ -46,7 +46,7 @@ pIdent = try $ do
 pAtom :: Parser Raw
 pAtom  =
       withPos (    (RVar <$> pIdent)
-               <|> (RU <$ char 'U')
+               <|> (RSet <$ symbol "Set")
                <|> (RHole <$ char '_'))
   <|> parens pTm
 

### 2020-03-24 fcif/Pretty.hs

@@ -9,7 +9,7 @@ import Types
 --   names bound in top lambdas, so that later we can avoid printing
 --   them in meta spines.
 topLams :: Bool -> String -> String -> [Name] -> Tm -> ShowS
-topLams p pre post ns (Lam (fresh ns -> x) i a t) =
+topLams p pre post ns (Lam (fresh ns -> x) i a _ t) =
   showParen p (
     (pre++)
   . icit i bracket parens (
@@ -37,9 +37,7 @@ bracket s = ('{':).s.('}':)
 parens  s = ('(':).s.(')':)
 
 goLam :: [Name] -> Tm -> ShowS
-goLam ns (Lam (fresh ns -> x) i a t)  = (' ':) . goLamBind x i . goLam (x:ns) t
-goLam ns (LamTel(fresh ns -> x) a t) =
-  (' ':) . bracket ((x++) . (" : "++) . go False ns a) . goLam (x:ns) t
+goLam ns (Lam (fresh ns -> x) i a _ t) = (' ':) . goLamBind x i . goLam (x:ns) t
 goLam ns t = (". "++) . go False ns t
 
 goPiBind :: [Name] -> Name -> Icit -> Tm -> ShowS
@@ -47,76 +45,47 @@ goPiBind ns x i a =
   icit i bracket (showParen True) ((x++) . (" : "++) . go False ns a)
 
 goPi :: [Name] -> Bool -> Tm -> ShowS
-goPi ns p (Pi (fresh ns -> x) i a b)
+goPi ns p (Pi (fresh ns -> x) i a _ b)
   | x /= "_" = goPiBind ns x i a . goPi (x:ns) True b
   | otherwise =
      (if p then (" → "++) else id) .
-     go (case a of App{} -> False; AppTel{} -> False; _ -> True) ns a .
-     (" → "++) . go False (x:ns) b
-
-goPi ns p (PiTel (fresh ns -> x) a b)
-  | x /= "_" = goPiBind ns x Impl a . goPi (x:ns) True b
-  | otherwise =
-     (if p then (" → "++) else id) .
-     go (case a of App{} -> False; AppTel{} -> False; _ -> True) ns a .
+     go (case a of App{} -> False; _ -> True) ns a .
      (" → "++) . go False (x:ns) b
 
 goPi ns p t = (if p then (" → "++) else id) . go False ns t
 
 isMetaSp :: Tm -> Bool
 isMetaSp Meta{}         = True
[…]

### 2020-03-24 fcif/Types.hs

@@ -34,7 +34,8 @@ data Raw
   | RLam Name (Maybe Raw) Icit Raw   -- ^ λx.t  or λ{x}.t with optional type annotation
                                      --   on x
   | RApp Raw Raw Icit                -- ^ t u  or  t {u}
-  | RU                               -- ^ U
+  | RSet                             -- ^ Set
+  | RProp                            -- ^ Prop
   | RPi Name Icit Raw Raw            -- ^ (x : A) → B  or  {x : A} → B
   | RLet Name Raw Raw Raw            -- ^ let x : A = t in u
   | RHole                            -- ^ _
@@ -56,11 +57,6 @@ type BlockedBy = IS.IntSet
 data MetaEntry
   = Unsolved Blocking ~VTy
   | Solved Val
-
-  -- | Constancy (Γ, x : Rec A) B   + a list of blocking metas.
-  --   When B becomes constant, A is solved to ε
-  | Constancy Cxt VTy VTy BlockedBy
-
 
 -- | A partial mapping from levels to levels. Undefined domain represents
 --   out-of-scope or "illegal" variables.
@@ -87,19 +83,21 @@ skipStr (Str c d r occ) = Str c (d + 1) r occ
 data Vals  = VNil | VDef Vals ~Val | VSkip Vals
 
 -- | Type environment.
-data Types = TNil | TDef Types ~VTy | TBound Types ~VTy
+data Types = TNil | TDef Types ~VTy ~VUniv | TBound Types ~VTy ~VUniv
 
 type Ix    = Int
 type Lvl   = Int
 type Ty    = Tm
 type VTy   = Val
 type MCxt  = IM.IntMap MetaEntry
+type Univ  = Ty
+type VUniv = VTy
 
 -- | Extending `Types` with any type.
-pattern TSnoc :: Types -> VTy -> Types
-pattern TSnoc as a <- ((\case TBound as a -> Just (as, a)
-                              TDef as a   -> Just (as, a)
-                              TNil        -> Nothing) -> Just (as, a))
+pattern TSnoc :: Types -> VTy -> VUniv -> Types
[…]

### 2020-03-24 fcif/Zonk.hs

@@ -12,43 +12,27 @@ zonk vs t = go t where
 
   goSp :: Tm -> Either Val Tm
   goSp = \case
-    Meta m       -> case runLookupMeta m of
-                      Solved v -> Left v
-                      _        -> Right (Meta m)
-    App t u ni   -> case goSp t of
-                      Left t  -> Left (vApp t (eval vs u) ni)
-                      Right t -> Right $ App t (go u) ni
-    AppTel a t u -> case goSp t of
-                      Left t  -> Left (vAppTel (eval vs a) t (eval vs u))
-                      Right t -> Right $ AppTel (go a) t (go u)
-    t            -> Right (zonk vs t)
+    Meta m        -> case runLookupMeta m of
+                       Solved v -> Left v
+                       _        -> Right (Meta m)
+    App t u un ni -> case goSp t of
+                       Left t  -> Left (vApp t (eval vs u) (eval vs un) ni)
+                       Right t -> Right $ App t (go u) (go un) ni
+    t             -> Right (zonk vs t)
 
   goBind = zonk (VSkip vs)
 
   go = \case
-    Var x        -> Var x
-    Meta m       -> case runLookupMeta m of
-                      Solved v   -> quote (valsLen vs) v
-                      Unsolved{} -> Meta m
-                      _          -> error "impossible"
-    U            -> U
-    Pi x i a b   -> Pi x i (go a) (goBind b)
-    App t u ni   -> case goSp t of
-                      Left t  -> quote (valsLen vs) (vApp t (eval vs u) ni)
-                      Right t -> App t (go u) ni
-    Lam x i a t  -> Lam x i (go a) (goBind t)
-    Let x a t u  -> Let x (go a) (go t) (goBind u)
-    Tel          -> Tel
-    TEmpty       -> TEmpty
-    TCons x t u  -> TCons x (go t) (goBind u)
-    Rec a        -> Rec (go a)
-    Tempty       -> Tempty
-    Tcons t u    -> Tcons (go t) (go u)
-    Proj1 t      -> Proj1 (go t)
-    Proj2 t      -> Proj2 (go t)
-    PiTel x a b  -> PiTel x (go a) (goBind b)
-    AppTel a t u -> case goSp t of
-                      Left t  -> quote (valsLen vs) (vAppTel (eval vs a) t (eval vs u))
-                      Right t -> AppTel (go a) t (go u)
-    LamTel x a b -> LamTel x (go a) (goBind b)
[…]
